### 2024-02-14 paper-data-platforms.qmd
first commit

---
title: "Conceptualizing open health data platforms for low- and middle income countries"
format:
  typst:
    columns: 1
# acm-pdf:
  #   keep-tex: true  
  # acm-html: default
author:
  - name: Daniel Kapitan
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
      - name: Eindhoven University of Technology
        city: Eindhoven
        country: the Netherlands
    orcid: 0000-0001-8979-9194
    email: daniel@kapitan.net
  - name: Femke Heddema
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Julie Fleischer
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Chris Ihure
    affiliations:
      - name: PharmAccess Kenya
        city: Nairobi
        country: Kenya
  - name: Antragama Abbas
    affiliations:
      - name: Delft University of Technology
        city: Delft
        country: the Netherlands
  - name: Steven Wanyee
    affiliations:
      - name: IntelliSOFT
        city: Nairobi
        country: Kenya
  - name: XXX
    affiliations:
      - name: ONA
        city: Nairobi
        country: Kenya
  - name: John Grimes
    affiliations:
      - name: Australian e-Health Research Centre
        city: Brisbane
        country: Australia
  - name: Mark van der Graaf
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Mark de Reuver
    affiliations:
      - name: Delft University of Technology
        city: Delft
        country: the Netherlands

abstract: |
  This document is only a demo explaining how to use the template.
keywords: [template, demo]
reference-section-title: References
bibliography: pharmaccess.bib  
---

## Introduction {#sec-intro}

### Digital platforms vs. data platforms for healthcare

It is a widely held belief that digital technologies have an important role to play in strengthening health systems in low- and middle income countries (LMICs), as exemplified by the WHO global strategy on digital health [@who2021global]. The adoption rate of mobile phones in LMICs has been an important driver in implementing digital health solutions [@mccool2022mobile]. Yet, there are many shortcomings and challenges, including the current fragmentation of digital platforms and the lack of clear-cut pathways of scaling up digital health programmes, such that they can support sustainable and equitable change of national health systems in LMICs [@mccool2022mobile;@who2019recommendations;@neumark2021digital].

A commonly used perspective to scrutinize digital health is to consider it as a digital platform [@dereuver2018digital]. Digital platforms have disrupted many sectors but have just started to make inroads into highly regulated industries such as healthcare [@ozalp2022digital]. In this light, the challenges faced by LMICs in establishing national digital health platforms have a lot in common with those faced by high income countries. From a technological perspective, interoperability issues, weak integrations, siloed data repositories and overall lack of openness are often reported as key impediments [@malm-nicolais2023exploring;@mehl2023fullstac]. From a societal perspective, issues pertaining to the winner-takes-all nature of digital platforms are hotly debated as many jurisdictions make work to ensure these platforms indeed serve the common good of achieving universal health coverage [@sharon2018when].

Case studies on digital platforms in healthcare point to an emerging pattern where the focus shifts from the digital platform with its defining software and hardware components, to the data as the primary object of interest in and of itself [@ozalp2022digital;@alaimo2022organizations]. This observation ties into the proposed research agenda by de Reuver et al. to consider data platforms as a phenomenon distinct from digital platforms [@dereuver2022openness]. Generally, data platforms inherit the characteristics of digital platforms. For example, the economic perspective on digital platforms stresses their multi-sidedness with developers and consumers, while data platforms are used by data owners, data consumers and third party solution providers. At the same time, data platforms differ as their main offerings revolve around data. From a market perspective, data platforms have more moderate network effects and are more susceptible to fragmentation and heterogeneity.

Particularly relevant in the context of health data platforms (HDPs), is the conceptualization of openness. The objective of openness, the ways to realize openness and the risks of opening up can be considered as elements within the ongoing debate on data spaces [@otto2022designing] and data solidarity [@kickbush2021governing;@prainsack2022data;@prainsack2023beyond] in the healthcare domain, which is a current issue for high income countries and LMICs alike. Openness is particularly relevant if we are to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@otto2022designing].

### From health information exchanges to health data platforms

This paper is motivated by the conflation of a number of developments relevant to the design and implementation of solidarity-based HDPs in LMICs. First, the OpenHIE framework [@openhie] has been adopted by many sub-Saharan African countries [@mamuye2022health] as the architectural blueprint for implementing nation-wide health information exchanges (HIE), including Nigeria [@ohienigeria], Kenya [@khisif] and Tanzania [@tzhea2020]. These countries have, as a matter of course, extended the framework to include “analytics services” as an additional domain. The rationale for this addition is to facilitate primary and secondary re-use of health data for academic research, real-world evidence studies etc. which can be framed within the context of ongoing efforts towards Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair]. In doing so, however, we have implicitly moved from conceptualizing digital health platforms (the original OpenHIE specification) to health data platforms. This is problematic because the notion of openness, which is assumed to be essential in establishing solidarity-based approaches to data sharing, is inherently different for a data platform compared to a digital platform.

Conceptually, the OpenHIE framework constitutes a framework for an open digital platform. Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods[@digitalpublicgoods]. If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for data sharing and re-use. Distinguishing four types of data sharing (@tbl-types-data-sharing), the purpose of this paper is to investigate how new standards and technologies that can establish openness of health data platforms can be integrated into the OpenHIE architecture framework. The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.

|   | __Type of data sharing__ | __New standards and technology enablers to create openness__ |
|-|--|-------|
| __1__ | Data at the most granular (patient) level, which is persisted and used to provide a longitudinal record. | __Bulk FHIR API__ which has by now been incorporated in all major FHIR implementations the FHIR standard can be readily used to support type 1 data sharing to patient-level data across a patient population [@mandl2020push;@jones2021landscape]. |
| __2__ | Aggregated data, for example statistics for policy evaluation and benchmarking |__SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows. Builds on FHIRPath [@fhirpath] expressions in a logical structure to specify things like column names and unnested items. Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].|
| __3__ | Data analytics modules, that provide access to work and access the data. | __Federated learning (FL) and privacy- enhancing technologies (PETs)__ [@rieke2020future;@scheibner2021revolutionizing;@jordan2022selecting]: new paradigms that address the problem of data governance and privacy by training algorithms collaboratively without exchanging the data itself. Models can be trained on combined datasets and made available as open source artifacts for decision support. Data analysts can use FL and PETs to work with the data in a collaborative, decentralized fashion. |
| __4__ | Trained models that have been derived from the data and can be used stand-alone for decision support. | ^^ |

: Types of data sharing and in relation to new standards and technology enablers to create openness. {#tbl-types-data-sharing}

## Methods

In the paper, we present a design that extends the OpenHIE specification to include the four types of data sharing mentioned above. Using the full-STAC approach [@mehl2023fullstac] we combine open standards, open technologies and open architectures into a coherent modular HDP platform that can be configured and re-used across a variety of use-cases. Subsequently, we employ the Framework for Evaluation of Design Science (FEDS) research [@venable2016feds] to evaluate the design. The purpose of this evaluation is formative: we aim to to help improve the outcomes of the design and the artifacts presented. We do so by taking a naturalistic approach, that is, the expected performance of the design is considered with its real environment where we draw on our experience in two use-cases, namely the Momcare programme and the ONA Canopy data platform.

As part of our design research, we have taken a narrative approach in surveying existing scientific studies on health data platforms, focusing on the seminal reports and subsequently searching forward citations. In addition, we have searched the open source repositories (most notably GitHub) and the online communities (OpenHIE community, FHIR community) to search for relevant open standards, technologies and architectures. This paper should not be considered as a proper systematic review.

## Abbreviations

|    |    |
|:---|:---|
| FAIR |Findable, Accessible, Interoperable and Reusable |
| FHIR |Fast Healthcare Interoperability Resources |
| HDP	|	Health data platform, explicitly differentiated from health digital platform |
| HIE | Health Information Exchange |
| FL | Federated learning |
| LMIC | Low- and middle income countries |
| PET | Privacy-enhancing technologies |

### 2024-02-14 paper-data-platforms.tex
first commit

\documentclass[manuscript,screen,review]{acmart}

\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother

%%
%% This is file `sample-manuscript.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `manuscript')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-manuscript.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}

\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}

%% PANDOC PREAMBLE BEGINS

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\definecolor{mypink}{RGB}{219, 48, 122}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
%% PANDOC PREAMBLE ENDS

\setlength{\parindent}{10pt}
\setlength{\parskip}{0pt}

\hypersetup{
  pdftitle={Conceptualizing open health data platforms for low- and middle income countries},
  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Antragama Abbas; Steven Wanyee; XXX; John Grimes; Mark van der Graaf; Mark de Reuver},
  pdfkeywords={template, demo},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={red},
  pdfcreator={LaTeX via pandoc, via quarto}}

%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{}
\copyrightyear{}
\acmYear{}
\acmDOI{}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[]{}{}{}
\acmPrice{}
\acmISBN{}

%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Conceptualizing open health data platforms for low- and middle
income countries}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.

  \author{Daniel Kapitan}
  \orcid{0000-0001-8979-9194}
            \affiliation{%
                  \institution{PharmAccess Foundation}
                                  \city{Amsterdam}
                                  \country{the Netherlands}
                      }
          \affiliation{%
                  \institution{Eindhoven University of Technology}
                                  \city{Eindhoven}
                                  \country{the Netherlands}
                      }
        \author{Femke Heddema}

            \affiliation{%
                  \institution{PharmAccess Foundation}
                                  \city{Amsterdam}
                                  \country{the Netherlands}
                      }
        \author{Julie Fleischer}

            \affiliation{%
                  \institution{PharmAccess Foundation}
                                  \city{Amsterdam}
                                  \country{the Netherlands}
                      }
        \author{Chris Ihure}

            \affiliation{%
                  \institution{PharmAccess Kenya}
                                  \city{Nairobi}
                                  \country{Kenya}
                      }
        \author{Antragama Abbas}

            \affiliation{%
                  \institution{Delft University of Technology}
                                  \city{Delft}
                                  \country{the Netherlands}
                      }
        \author{Steven Wanyee}

            \affiliation{%
                  \institution{IntelliSOFT}
                                  \city{Nairobi}
                                  \country{Kenya}
                      }
        \author{XXX}

            \affiliation{%
                  \institution{ONA}
                                  \city{Nairobi}
                                  \country{Kenya}
                      }
        \author{John Grimes}

            \affiliation{%
                  \institution{Australian e-Health Research Centre}
                                  \city{Brisbane}
                                  \country{Australia}
                      }
        \author{Mark van der Graaf}

            \affiliation{%
                  \institution{PharmAccess Foundation}
                                  \city{Amsterdam}
                                  \country{the Netherlands}
                      }
        \author{Mark de Reuver}

            \affiliation{%
                  \institution{Delft University of Technology}
                                  \city{Delft}
                                  \country{the Netherlands}
                      }

%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato et al.}
%%  
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
This document is only a demo explaining how to use the template.    
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\setlength{\parskip}{-0.1pt}

\subsection{Introduction}\label{sec-intro}

This is a dummy example only for the purpose to use this repo as a
template starter for creating new formats. For this
\texttt{article-format-template} we call our dummy article \texttt{aft}.

This quarto extension format supports PDF and HTML outputs.
\texttt{quarto-journals} is aiming at porting existing {\LaTeX} template
from journals to be used with quarto. PDF format is what require the
most work to fit the journals guideline, but Quarto offer a nice
rendering for HTML output too. This demo format template only use basic
HTML format without any customization for now.

\subsection{About Quarto Extensions formats And Quarto Journals
Article}\label{about-quarto-extensions-formats-and-quarto-journals-article}

First, please get familiar with the following resources:

\begin{itemize}
\tightlist
\item
  \href{https://quarto.org/docs/extensions/formats.html}{Creating
  Formats} in Quarto as part of the
  \href{https://quarto.org/docs/extensions/}{Extensions} mechanism.
\item
  \href{https://quarto.org/docs/journals/}{Journals Articles} for
  Quarto.
\end{itemize}

\subsection{Structure of this
repository}\label{structure-of-this-repository}

Everything for the extensions is in \texttt{\_extensions}. See Quarto
doc for details.

\begin{itemize}
\item
  In \texttt{partials}, you'll find the \texttt{.tex} partials that can
  be used and should be removed or tweaked,s
\item
  Your extension can make shortcodes and lua filters available. This
  document shows the effect of the one provided in the \texttt{aft}
  format.
\item
  \texttt{aft} format sets some defaults which are different from
  \texttt{pdf} or \texttt{html}, link setting links to URL in read
  inside PDF output.
\end{itemize}

Source repository for this template format can found on
\href{https://github.com/quarto-journals/article-format-template}{Github}

\subsubsection{\texorpdfstring{\texttt{\_extensions\textbackslash{}aft}}{\_extensions\textbackslash aft}}\label{extensionsaft}

In this folder you'll find everything that defines the extensions which
could be installed using \texttt{quarto\ install\ extension} or be part
of the template when using \texttt{quarto\ use\ template}

\begin{description}
\item[Format Metadata]
This is in \texttt{\_extension.yml} is where all the metadata about the
format are defined so that Quarto knows what to use. Adapt this file for
you own template.
\item[Partials]
In \texttt{partials}, there are the \texttt{.tex} files that will be
used as Pandoc's template. We provide here all the partials supported by
Quarto and custom one for this format. Quarto allows to provide partials
to ease the process of tweaking the default latex Pandoc's template and
keeping it up to date.\\
This template repo contains all the relevant partials that you can use
with Quarto \emph{as example}. We only tweaked \texttt{title.tex} to
show the usage of a custom partials called \texttt{\_custom.tex}.\\
\textbf{Only keep the partials that you need to tweak for the format you
are creating}

If you need to completely change the default template (i.g customizing
partials is not enough), then you need to provide your own template to
Pandoc based on
\href{https://github.com/quarto-dev/quarto-cli/blob/main/src/resources/formats/pdf/pandoc/template.tex}{\texttt{template.tex}}
and also using partials or not. This can be provided using the
\texttt{template} YAML key in \texttt{\_extension.yml} for Quarto to use
it.

This is considered advanced configuration as it will be harder to
maintain than only using partials but could be required for some
specific format. Be aware that this may lead to loose some Pandoc or
Quarto features tied to default \texttt{template.tex} content if you
remove some specific parts.
\item[Lua Filters]
Most of the time, custom formats will need Lua filters to provide
specific features like cross format supports or provides custom
shortcodes through the Quarto extension mechanism. Those filters will be
available to the user and could be used in the custom formats (according
to \texttt{\_extensions} metadata). We have provided two examples:

\begin{itemize}
\tightlist
\item
  \texttt{color-text.lua}, a Lua filter used to add color to inline text
  for PDF and HTML outputs using the same Markdown syntax
\item
  \texttt{shorcodes.lua}, a Lua filter which follow
  \href{https://quarto.org/docs/authoring/shortcodes.html\#custom-shortcodes}{Quarto
  custom shortcodes} guidelines to provide a \texttt{} shortcode to
  nicely print LaTeX in PDF and HTML.
\end{itemize}

\textbf{Remove or replace with your own Lua filters}
\item[Format resources]
Resources required by the format needs to be available. We have provided
two examples:

\begin{itemize}
\tightlist
\item
  \texttt{te.bst} is a biblio style file for demo. It has been
  downloaded from
  https://www.economics.utoronto.ca/osborne/latex/TEBST.HTM
  (http://econtheory.org/technical/te.bst) - Licence
  \href{https://www.latex-project.org/lppl/}{LPPL}
\item
  \texttt{aft.cls} is a dummy class file for this example format. It is
  a copy of official \texttt{article.cls}, the one provided in LaTeX
  installation (i.e at \texttt{kpsewhich\ article.cls}) and renamed as
  example (Licence LaTeX Project Public License)
\item
  \texttt{custom.scss} is a style file to have a custom theme for our
  HTML format so that our Lua filter feature \texttt{color-tex.lua}
  works.
\end{itemize}

Those files are referenced within the \texttt{\_extension.yml} to be
used with our example format.

\textbf{Remove and replace with your own resources}
\item[\texttt{.quartoignore}]
Sometimes it is useful to have some files only needed for reference or
for development. They should be available in the source repository but
not downloaded to the user when \texttt{quarto\ use\ template} is used.

\textbf{Use \texttt{.quartoignore} to register such file and folder (one
file or folder per line)}
\item[\texttt{style-guide} folder]
For \texttt{quarto-journals} format, use \texttt{style-guide} folder to
include any documentation and resourced used for format creation, like a
journal style guide or original \texttt{.tex} template. This folder is
already added in \texttt{.quartoignore} in this example repo.

\textbf{Remove, rename or add to this folder}
\item[\texttt{template.qmd}]
This file is the template document that shows how to use the custom
format. It will be downloaded with other resource by
\texttt{quarto\ use\ template}, and even offered to be renamed if the
name \texttt{template.qmd} is used.

This file will usually use the custom format (here \texttt{aft-pdf} and
\texttt{aft-html}) and show how to use the template. When you'll copy
this template, you should be able to render this document to the demo
format.

\textbf{Adapt this file to provide a suitable template for your custom
format}
\item[Other files]
Other files are needed by the template and are usually user provided -
they are not part of the custom format.

Here \texttt{bibliography.bib} is here to demo the usage of the bst file
from the custom format.

\textbf{Remove this file and provide a suitable one for your template}
\end{description}

\newpage{}

\subsection{Checklist: Creating a custom
format}\label{checklist-creating-a-custom-format}

Here is the checklist to help you know what to modify:

\begin{itemize}
\tightlist
\item
  Read the resources mentioned at the top,
\item
  Use this template repo to create a new repository for your format
  (Click on ``Use this template'' to create new github repo)
\item
  Once you are acquainted with the content, remove the resources that
  are there only as example (see above)
\item
  Update README by replacing \texttt{aft} and
  \texttt{Article\ Format\ Template} mentions for your journal format
\item
  Keep only the template partials that you need to tweak, and add custom
  ones if needed
\item
  Add any Lua filters for shortcodes and other that would be useful to
  create the expected output format
\item
  Add any external resource your format will need, and that should be
  part of the extension format that will be downloaded,
\item
  Check \texttt{\_extension.yml} is updated correctly
\item
  Modify the skeleton \texttt{template.qmd} to your format and add any
  required resources to be downloaded to user.
\item
  Check \texttt{.quartoignore} is updated which everything that should
  not be downloaded.
\item
  Publish a demo of you format to github pages of the repo by using
  \texttt{quarto\ publish} command
\end{itemize}

\subsection{Demo of some features found in this demo journal
template}\label{demo-of-some-features-found-in-this-demo-journal-template}

\subsubsection{Shortcode demo}\label{sec-shortcode}

PDF are rendered using {\LaTeX} but it is best if one can use a Markdown
syntax for cross format support.

\texttt{} used in source is a shortcode syntax where the shortcode is
included in the extension folder \texttt{\_extensions}

\subsubsection{Code chunk}\label{sec-chunks}

This format hide chunks by default as option has been set in
\texttt{\_extension.yml} file.

\subsubsection{Text color}\label{sec-summary}

Our format makes applying color on inline text possible using the
\texttt{{[}content{]}\{color=\textless{}name\textgreater{}\}} syntax.
Let's see an example.

Here we are using a special feature of our format which is the coloring
because \textcolor{mypink}{pink is a \textbf{nice} color}.

This is possible thanks to the Lua Filter included in the custom
extension format.

\subsubsection*{Using references}\label{using-references}
\addcontentsline{toc}{subsubsection}{Using references}

I did not read this book \citep{CameronTrivedi2013} but it must be
interesting.

Differences between \texttt{aft-html} and \texttt{aft-pdf}:

\begin{itemize}
\tightlist
\item
  For the HTML format, we are using Pandoc citeproc to include the
  bibliography. Here \texttt{reference-section-title} controls the title
  for the chapter that will be used.
\item
  For the PDF format, \texttt{natbib} is used by default and the
  bibliography is included with a title by the LaTeX template.
\end{itemize}

%% begin pandoc before-bib
%% end pandoc before-bib
%% begin pandoc biblio
%% end pandoc biblio
%% begin pandoc include-after
%% end pandoc include-after
%% begin pandoc after-body
%% end pandoc after-body

\end{document}
\endinput
%%
%% End of file `sample-manuscript.tex'.

### 2024-02-14 paper-data-platforms.qmd
wip

@@
 format:
-  typst:
-    columns: 1
-# acm-pdf:
-  #   keep-tex: true  
-  # acm-html: default
+  # docx
+  # typst:
+  #   columns: 1
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      cite-style: authoryear
 author:
@@
 Particularly relevant in the context of health data platforms (HDPs), is the conceptualization of openness.
-The objective of openness, the ways to realize openness and the risks of opening up can be considered as elements within the ongoing debate on data spaces [@otto2022designing] and data solidarity [@kickbush2021governing;@prainsack2022data;@prainsack2023beyond] in the healthcare domain, which is a current issue for high income countries and LMICs alike.
+The objective of openness, the ways to realize openness and the risks of opening up can be considered as elements within the ongoing debate on data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond] in the healthcare domain, which is a current issue for high income countries and LMICs alike.
 Openness is particularly relevant if we are to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@otto2022designing].
@@
 |   | __Type of data sharing__ | __New standards and technology enablers to create openness__ |
-|-|--|-------|
+|:-:|:-------|:-------------|
 | __1__ | Data at the most granular (patient) level, which is persisted and used to provide a longitudinal record.
@@
 | __3__ | Data analytics modules, that provide access to work and access the data.
-| __Federated learning (FL) and privacy- enhancing technologies (PETs)__ [@rieke2020future;@scheibner2021revolutionizing;@jordan2022selecting]: new paradigms that address the problem of data governance and privacy by training algorithms collaboratively without exchanging the data itself.
+| __Federated learning (FL)__ [@rieke2020future] and __privacy- enhancing technologies (PETs)__ [@scheibner2021revolutionizing;@jordan2022selecting]: new paradigms that address the problem of data governance and privacy by training algorithms collaboratively without exchanging the data itself.
 Models can be trained on combined datasets and made available as open source artifacts for decision support.
@@
 |
-| __4__ | Trained models that have been derived from the data and can be used stand-alone for decision support. | ^^ |
+| __4__ | Trained models that have been derived from the data and can be used stand-alone for decision support. | __ONNX__ [@onnx]: an open format built to represent machine learning models. |
 
@@
 
+### Momcare
+
+MomCare was launched in Kenya [@huisman2022digital;@sanctis2022maintaining] and Tanzania [@shija2021access;@mrema2021application] in 2017 and 2019 respectively, with the objective to create transparency on the journeys of pregnant mothers.
+The programme is built on three pillars: journey tracking, quality support and a mobile wallet (@fig-momcare).
+MomCare distinguishes two user groups: mothers are supported during their pregnancy through reminders and surveys, using SMS as the digital mode of engagement.
+Health workers are equipped with an Android-based application, in which visits, care activities and clinical observations are recorded.
+Reimbursements of the maternal clinic are based on the data captured with SMS and the app, thereby creating a conditional payment scheme, where providers are partially reimbursed up-front for a fixed bundle of activities, supplemented by bonus payments based on a predefined set of care activities.
+
+![Overview of the Momcare programme](images/momcare.png){width=auto #fig-momcare}
+
+### Use-case: ONA Canopy
+
+https://ona.io/home/services/canopy/
+
+### Evaluation of other reported HDP designs
+
+Besides the two case studies, we will discuss how the design relates to other reported open architectures
+
+- Ayaz et al. presented an end-to-end framework that automates the mapping from FHIR Resources to FHIR-compliant tables that are subsequently exposed for consumption [@ayaz2023transforming]
+- Mullie et al.
+presented the CODA project that aims to facilitate distributed computations for multi-institutional data analysis while avoiding the costs and complexity of data pooling [@mullie2023coda]
+- PHAS [@abbasian2023phas]
+- OpenHIM platform with analytics
+
+## Results
+
+### Open standards: using FHIR as the common data model
+
+The recent convergence to FHIR as the de facto standard for information exchange has fuelled the development of OpenHIE.
+FHIR is currently used both for routine healthcare settings40 and clinical research settings [@duda2022hl7;@vorisek2022fast] and is increasingly being used in LMICs as well.
+The FHIR-native OpenSRP platform [@mehl2020open] has been deployed in 14 countries targeting various patient populations, amongst which a reference implementation of the WHO antenatal and neonatal care guidelines for midwives in Lombok, Indonesia [@summitinstitutefordevelopment2023bunda;@kurniawan2019midwife].
+In India, FHIR is used as the underlying technology for the open Health Claims Exchange protocol specification, which has been adopted by the Indian government as the standard for e-claims handling [@hcx].
+Additionally, the guidelines and standards of the African Union explicitly state FHIR is to be used as the messaging standard [@2023african].
+This range of utilizations showcase the standards’ widespread applicability.
+The proceedings of the OpenHIE conference 2023 attest to the fact that FHIR and open source technologies are embraced as critical enablers in implementing health information exchanges in LMICs [@ohie23].
+
+However, despite the increased use of FHIR as a common data model, various studies have investigated its merits and performance vis-a-vis other healthcare standards.
+Comparisons between OpenEHR, ISO 13606, OMOP and FHIR have been made [@ayaz2023transforming;@mullie2023coda;@rinaldi2021openehr;@cremonesi2023need;@sinaci2023data].
+A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR concluded that i) these three standards are functionally and technically compatible, and therefore can be used side by side; and that ii) each of these standards have their strengths and limitations that correlate with their intended use as summarized in the @tbl-comparison.
+
+![Comparison of OpenEHR, ISO 13606 and FHIR standards](images/comparison-ehr-standards.png){width=auto #tbl-comparison}
+
+For an infectious diseases dataset with a limited scope, OpenEHR, OMOP and FHIR have been compared and found all to be equally suitable [@rinaldi2021openehr].
+Comparing OMOP and FHIR, the latter has been found to support more granular mappings required for analytics and was therefore chosen as the standard for the CODA project[@mullie2023coda].
+
+Although FHIR was originally designed only for exchange between systems, we propose to use it as the common data model for the design presented here for the following reasons:
+
+- Industry adoption has significantly increased, as exemplified by FHIR-based offering by major cloud providers such as Google, Azure and AWS.
+Also, Africa CDC has explicitly chosen FHIR as the preferred standard;
+- The concept of FHIR Profiles allow localisation to tailor the standard to a specific use case.
+A profile defines rules, extensions, and constraints for a resource.
+We posit that the possible penalty of this flexibility, namely having to manage different FHIR versions and/or profiles, is less of an issue in the context of LMICs where first priority is to exchange datasets such as the International Patient Summary (IPS) that are less complex compared to the requirements for high income countries;
+- Being based on webstandards, the FHIR standard lends itself best for further separation of concerns as envisioned by the composable data stack.
+This is an important enabler for the downward scalability of the solution;
+- With its inherent, graph-like nature, FHIR can be readily incorporated into the principles of FAIR data sharing, where FHIR-based data repositories can be integrated in an overarching netwerk of FAIR data stations [@sinaci2023data;@pedrera-jimenez2023can].
+
+### Open architecture: extending OpenHIE framework with a lakehouse data platform
+
+Data management and analytics platforms have undergone significant changes since the first generation of data warehouses were introduced.
+Recent studies have shown that the current practice has converged towards the lakehouse as one of the most commonly used solution designs [@armbrust2021lakehouse;@hai2023data;@harby2022data].
+Lakehouses typically have a zonal architecture [@hai2023data] where data is ingested from the source systems in bulk (E), delivered to storage with aligned schemas (L) and transformed into a format ready for analysis (T).
+The discerning characteristic of the lakehouse architecture is its foundation on low-cost and directly-accessible storage that also provides traditional analytical DBMS management and performance features such as ACID transactions, data versioning, auditing, indexing, caching, and query optimization [@armbrust2021lakehouse].
+Lakehouses thus combine the key benefits of data lakes and data warehouses: low-cost storage in an open format accessible by a variety of systems from the former, and powerful management and optimization features from the latter.
+
+With respect to current implementations of lakehouse data platform, we observe a proliferation of tools with as yet limited standards to improve technical interoperability.
+In the analysis of Pedreira et al.
+[@pedreira2023composable] the requirement for specialization in data management systems has evolved faster than our software development practices.
+This situation has created a siloed landscape composed of hundreds of products developed and maintained as monoliths, with limited reuse between systems.
+It has also affected the end users, who are often required to learn the idiosyncrasies of dozens of incompatible SQL and non-SQL API dialects, and settle for systems with incomplete functionality and inconsistent semantics.
+To remedy this, Pedreira et al.
+call to (re-)design and implement modern data platforms in terms of a 'composable data stack’ as a means to decrease development and maintenance cost and pick-up the speed of innovation.
+
+While the lakehouse architecture separating the concerns of compute and storage, the composable data stack takes the separation of concerns is taken one step further.
+A composable data system (@fig-composable-data-stack), not only separates the storage (layer 3) and execution (layer 2), but also separates the user interface (layer 1) from the execution engine by introducing standards for Intermediate Representation (standard A) and Connectivity (standard B).
+The composable data stack can be implemented with current open source technologies (@fig-cds-examples).
+As an example, the Ibis user interface is currently sufficiently mature to offer a standardized dataframe interface to 19 different execution engines.
+
+![Composable data stack](images/composable-data-stack.png){#fig-composable-data-stack}
+
+![Examples of implementations of the composable data stack](images/composable-data-stack-implementation.png){#fig-cds-examples}
+
+### Extended OpenHIE framework
+Our proposal for extending HIE is as follows.
+
+- add diagram, explicitly adding the component
+- fit in structure of OpenHIE specification
+- check which workflows are related to analytics
+
+Following Hai we take a subset of the core functionalities of a data lake. Because the SHR is the key source, don’t need full stack of ingestion of various sources.
+
+- Query-driven data discovery
+
+### Open technologies: available digital public goods
+
+Many components of the OpenHIE specification are now available as a digital public good.
+@tbl-digital-public-goods lists components that are currently available for implementing the OpenHIE framework using open source, digital public goods that are compliant with the FHIR standard, illustrating the maturity of this ecosystem and development community.
+With the launch of the Instant OpenHIE configuration toolkit54, it has become easier to set up, explore and develop HIEs thereby reducing costs and skills required for software developers to deploy an OpenHIE architecture for quicker solution testing and as a starting point for faster production implementation and customisation.
+Several frameworks are available that offer a set of preconfigured components out of the box, such as for example:
+
+- the  “Open Smart Register Platform” (OpenSRP), that focuses on providing a mobile-first platform, including a FHIR native app designed to support the WHO Smart Guidelines
+- the OpenHIM Platform, a reference implementation of the Instant OpenHIE framework, providing an easy way to set up, manage and operate various HIE configurations
+
+![ Overview of current open source implementations of components included in the OpenHIE specification that are FHIR-compatible.
+The category Analytics Services is not a part of the original OpenHIE and is discussed in the paper.
+Point-of-Service systems are excluded for brevity.
+A systematic review of such digital public goods is beyond the scope of this document.](images/digital-public-goods.png){#tbl-digital-public-goods}
+
+### Evaluation: Momcare programme
+
+### Evaluation: ONA Canopy
+
+### Evaluation: reported HDP solutions
+
+## Discussion
+
+### Openness of data platforms
+
+We specifically address the notion of openness of HDPs in LMICs in terms of the design-related questions put forward by de Reuver at al.11:
+
+- Object of openness: what data-related resources should data platforms make available when opening up (e.g.
+data, data products, datadriven insights, analytics modules)?
+Which user groups derive value from accessing data-related resources from data platforms (e.g.
+data providers, data users, intermediaries, developers)?
+- Unit of analysis: what is platform-to-platform openness in the context of data platforms, given the expectation that different HDPs will emerge at various aggregation levels?
+How do we distinguish meta-platforms, forking, and platform interoperability?
+- Risk of openness: What are the novel (negative) implications of opening up data platforms? How can reflexivity in design help providers to resolve the negative implications of openness?
+
+### Comparison with HMIS component
+
+- Workflow requirements: Report aggregate data (link): receiver is HMIS, mADX
+- Functional requirements: https://guides.ohie.org/arch-spec/openhie-component-specifications-1/openhie-health-management-information-system-hmis
+
+Requirements are similar, but implementation differs: Datamodel is non-FHIR, focused on DataValue, which conceptually equates to FHIR Measure
+
+### FHIR and FAIR
+
+- How does FHIR relate to approaches taken by the FAIR community, which tend to take more an approach of using knowledge graphs.
+For example, VODAN Africa [@gebreslassie2023fhir4fair;@purnamajati2022data].
+- FAIR principles vs FHIR graph: is FHIR a FAIR Data Object
+
+### Attribute-based access control
+
+- Cerbos, attribute based on lineage or anonymized tables
+
 ## Abbreviations
@@
 |:---|:---|
+| ELT | Extract, Load and Transform |
 | FAIR |Findable, Accessible, Interoperable and Reusable |

### 2024-02-14 paper-data-platforms.tex
wip

@@
-\documentclass[manuscript,screen,review]{acmart}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%
+\documentclass[
+  authoryear]{elsarticle}
 
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 \makeatother
-
-%%
-%% This is file `sample-manuscript.tex',
-%% generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `manuscript')
-%% 
-%% IMPORTANT NOTICE:
-%% 
-%% For the copyright see the source file.
-%% 
-%% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-manuscript.tex.
-%% 
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%% 
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%%
-%% The first command in your LaTeX source must be the \documentclass command.
-
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-
-%% PANDOC PREAMBLE BEGINS
+\usepackage{xcolor}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{5}
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 
@@
 
-\definecolor{mypink}{RGB}{219, 48, 122}
 \makeatletter
@@
 \makeatother
-%% PANDOC PREAMBLE ENDS
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage[]{natbib}
+\bibliographystyle{elsarticle-harv}
+\usepackage{bookmark}
 
-\setlength{\parindent}{10pt}
-\setlength{\parskip}{0pt}
-
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
 \hypersetup{
@@
   citecolor={Blue},
-  urlcolor={red},
-  pdfcreator={LaTeX via pandoc, via quarto}}
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
 
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{}
-\copyrightyear{}
-\acmYear{}
-\acmDOI{}
-
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[]{}{}{}
-\acmPrice{}
-\acmISBN{}
-
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
-
-%%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
-
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
-%%\citestyle{acmauthoryear}
-
-%% end of the preamble, start of the body of the document source.
+\setlength{\parindent}{6pt}
 \begin{document}
 
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
+\begin{frontmatter}
 \title{Conceptualizing open health data platforms for low- and middle
 income countries}
-
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
-
-  \author{Daniel Kapitan}
-  \orcid{0000-0001-8979-9194}
-            \affiliation{%
-                  \institution{PharmAccess Foundation}
-                                  \city{Amsterdam}
-                                  \country{the Netherlands}
+\author[1,2]{Daniel Kapitan%
+%
 }
-          \affiliation{%
-                  \institution{Eindhoven University of Technology}
-                                  \city{Eindhoven}
-                                  \country{the Netherlands}
-                      }
-        \author{Femke Heddema}
-
-            \affiliation{%
-                  \institution{PharmAccess Foundation}
-                                  \city{Amsterdam}
-                                  \country{the Netherlands}
-                      }
-        \author{Julie Fleischer}
-
-            \affiliation{%
-                  \institution{PharmAccess Foundation}
-                                  \city{Amsterdam}
-                                  \country{the Netherlands}
-                      }
-        \author{Chris Ihure}
-
-            \affiliation{%
-                  \institution{PharmAccess Kenya}
-                                  \city{Nairobi}
-                                  \country{Kenya}
-                      }
-        \author{Antragama Abbas}
-
-            \affiliation{%
-                  \institution{Delft University of Technology}
-                                  \city{Delft}
-                                  \country{the Netherlands}
-                      }
-        \author{Steven Wanyee}
-
-            \affiliation{%
-                  \institution{IntelliSOFT}
-                                  \city{Nairobi}
-                                  \country{Kenya}
-                      }
-        \author{XXX}
-
-            \affiliation{%
-                  \institution{ONA}
-                                  \city{Nairobi}
-                                  \country{Kenya}
-                      }
-        \author{John Grimes}
-
-            \affiliation{%
-                  \institution{Australian e-Health Research Centre}
-                                  \city{Brisbane}
-                                  \country{Australia}
-                      }
-        \author{Mark van der Graaf}
-
-            \affiliation{%
-                  \institution{PharmAccess Foundation}
-                                  \city{Amsterdam}
-                                  \country{the Netherlands}
-                      }
-        \author{Mark de Reuver}
-
-            \affiliation{%
-                  \institution{Delft University of Technology}
-                                  \city{Delft}
-                                  \country{the Netherlands}
+ \ead{daniel@kapitan.net} 
+\author[1]{Femke Heddema%
+%
 }
 
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-%\renewcommand{\shortauthors}{Trovato et al.}
-%%  
-%% The abstract is a short summary of the work to be presented in the
-%% article.
+\author[1]{Julie Fleischer%
+%
+}
+
+\author[3]{Chris Ihure%
+%
+}
+
+\author[4]{Antragama Abbas%
+%
+}
+
+\author[5]{Steven Wanyee%
+%
+}
+
+\author[6]{XXX%
+%
+}
+
+\author[7]{John Grimes%
+%
+}
+
+\author[1]{Mark van der Graaf%
+%
+}
+
+\author[4]{Mark de Reuver%
+%
+}
+
+\affiliation[1]{organization={PharmAccess
+Foundation},city={Amsterdam},country={the
+Netherlands},countrysep={,},postcodesep={}}
+\affiliation[2]{organization={Eindhoven University of
+Technology},city={Eindhoven},country={the
+Netherlands},countrysep={,},postcodesep={}}
+\affiliation[3]{organization={PharmAccess
+Kenya},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
+\affiliation[4]{organization={Delft University of
+Technology},city={Delft},country={the
+Netherlands},countrysep={,},postcodesep={}}
+\affiliation[5]{organization={IntelliSOFT},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
+\affiliation[6]{organization={ONA},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
+\affiliation[7]{organization={Australian e-Health Research
+Centre},city={Brisbane},country={Australia},countrysep={,},postcodesep={}}
+
+\cortext[cor1]{Corresponding author}
+
 \begin{abstract}
@@
 
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
+\begin{keyword}
+    template \sep 
+    demo
+\end{keyword}
+\end{frontmatter}
 
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
+\section{Introduction}\label{sec-intro}
 
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
-\maketitle
+\subsection{Digital platforms vs.~data platforms for
+healthcare}\label{digital-platforms-vs.-data-platforms-for-healthcare}
 
-\setlength{\parskip}{-0.1pt}
+It is a widely held belief that digital technologies have an important
+role to play in strengthening health systems in low- and middle income
+countries (LMICs), as exemplified by the WHO global strategy on digital
+health \citep{who2021global}. The adoption rate of mobile phones in
+LMICs has been an important driver in implementing digital health
+solutions \citep{mccool2022mobile}. Yet, there are many shortcomings and
+challenges, including the current fragmentation of digital platforms and
+the lack of clear-cut pathways of scaling up digital health programmes,
+such that they can support sustainable and equitable change of national
+health systems in LMICs
+\citep{mccool2022mobile, who2019recommendations, neumark2021digital}.
 
-\subsection{Introduction}\label{sec-intro}
+A commonly used perspective to scrutinize digital health is to consider
+it as a digital platform \citep{dereuver2018digital}. Digital platforms
+have disrupted many sectors but have just started to make inroads into
+highly regulated industries such as healthcare \citep{ozalp2022digital}.
+In this light, the challenges faced by LMICs in establishing national
+digital health platforms have a lot in common with those faced by high
+income countries. From a technological perspective, interoperability
+issues, weak integrations, siloed data repositories and overall lack of
+openness are often reported as key impediments
+\citep{malm-nicolais2023exploring, mehl2023fullstac}. From a societal
+perspective, issues pertaining to the winner-takes-all nature of digital
+platforms are hotly debated as many jurisdictions make work to ensure
+these platforms indeed serve the common good of achieving universal
+health coverage \citep{sharon2018when}.
 
-This is a dummy example only for the purpose to use this repo as a
-template starter for creating new formats. For this
-\texttt{article-format-template} we call our dummy article \texttt{aft}.
+Case studies on digital platforms in healthcare point to an emerging
+pattern where the focus shifts from the digital platform with its
+defining software and hardware components, to the data as the primary
+object of interest in and of itself
+\citep{ozalp2022digital, alaimo2022organizations}. This observation ties
+into the proposed research agenda by de Reuver et al.~to consider data
+platforms as a phenomenon distinct from digital platforms
+\citep{dereuver2022openness}. Generally, data platforms inherit the
+characteristics of digital platforms. For example, the economic
+perspective on digital platforms stresses their multi-sidedness with
+developers and consumers, while data platforms are used by data owners,
+data consumers and third party solution providers. At the same time,
+data platforms differ as their main offerings revolve around data. From
+a market perspective, data platforms have more moderate network effects
+and are more susceptible to fragmentation and heterogeneity.
 
-This quarto extension format supports PDF and HTML outputs.
-\texttt{quarto-journals} is aiming at porting existing {\LaTeX} template
-from journals to be used with quarto. PDF format is what require the
-most work to fit the journals guideline, but Quarto offer a nice
-rendering for HTML output too. This demo format template only use basic
-HTML format without any customization for now.
+Particularly relevant in the context of health data platforms (HDPs), is
+the conceptualization of openness. The objective of openness, the ways
+to realize openness and the risks of opening up can be considered as
+elements within the ongoing debate on data spaces
+\citep{otto2022designing} and data solidarity
+\citep{kickbusch2021lancet, prainsack2022data, prainsack2023beyond} in
+the healthcare domain, which is a current issue for high income
+countries and LMICs alike. Openness is particularly relevant if we are
+to realize a solidarity-based approach to health data sharing that i)
+gives people a greater control over their data as active decision
+makers; ii) ensures that the value of data is harnessed for public good;
+and iii) moves society towards equity and justice by counteracting
+dynamics of data extraction \citep{otto2022designing}.
 
-\subsection{About Quarto Extensions formats And Quarto Journals
-Article}\label{about-quarto-extensions-formats-and-quarto-journals-article}
+\subsection{From health information exchanges to health data
+platforms}\label{from-health-information-exchanges-to-health-data-platforms}
 
-First, please get familiar with the following resources:
+This paper is motivated by the conflation of a number of developments
+relevant to the design and implementation of solidarity-based HDPs in
+LMICs. First, the OpenHIE framework \citep{openhie} has been adopted by
+many sub-Saharan African countries \citep{mamuye2022health} as the
+architectural blueprint for implementing nation-wide health information
+exchanges (HIE), including Nigeria \citep{ohienigeria}, Kenya
+\citep{khisif} and Tanzania \citep{tzhea2020}. These countries have, as
+a matter of course, extended the framework to include ``analytics
+services'' as an additional domain. The rationale for this addition is
+to facilitate primary and secondary re-use of health data for academic
+research, real-world evidence studies etc. which can be framed within
+the context of ongoing efforts towards Findable, Accessible,
+Interoperable and Reusable (FAIR) sharing of health data
+\citep{guillot2023fair}. In doing so, however, we have implicitly moved
+from conceptualizing digital health platforms (the original OpenHIE
+specification) to health data platforms. This is problematic because the
+notion of openness, which is assumed to be essential in establishing
+solidarity-based approaches to data sharing, is inherently different for
+a data platform compared to a digital platform.
+
+Conceptually, the OpenHIE framework constitutes a framework for an open
+digital platform. Openness for digital platforms refers to i) the use of
+open boundary resources, that is, specifications for the various
+healthcare specific workflows and information standards such as FHIR;
+and ii) the use of open source components that are available as digital
+public goods\citep{digitalpublicgoods}. If we are to use the OpenHIE
+framework as an open data platform, we need to extend the standards,
+technologies and architecture to include functionality for data sharing
+and re-use. Distinguishing four types of data sharing
+(Table~\ref{tbl-types-data-sharing}), the purpose of this paper is to
+investigate how new standards and technologies that can establish
+openness of health data platforms can be integrated into the OpenHIE
+architecture framework. The lack of detailed specifications and
+consensus of this addition to OpenHIE currently stands in the way of
+development projects that aim to establish HDPs in LMICs.
+
+\begin{longtable}[]{@{}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1200}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3200}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5600}}@{}}
+\caption{Types of data sharing and in relation to new standards and
+technology enablers to create
+openness.}\label{tbl-types-data-sharing}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\centering
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Type of data sharing}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{New standards and technology enablers to create openness}
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\centering
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Type of data sharing}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{New standards and technology enablers to create openness}
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+\textbf{1} & Data at the most granular (patient) level, which is
+persisted and used to provide a longitudinal record. & \textbf{Bulk FHIR
+API} which has by now been incorporated in all major FHIR
+implementations the FHIR standard can be readily used to support type 1
+data sharing to patient-level data across a patient population
+\citep{mandl2020push, jones2021landscape}. \\
+\textbf{2} & Aggregated data, for example statistics for policy
+evaluation and benchmarking & \textbf{SQL-on-FHIR specification}
+\citep{sql-on-fhir}: provides a standardised approach to make FHIR work
+well with familiar and efficient SQL engines that are most commonly used
+in analytical workflows. Builds on FHIRPath \citep{fhirpath} expressions
+in a logical structure to specify things like column names and unnested
+items. Implementations of this approach are available or forthcoming,
+including open source implementations such as Pathling
+\citep{grimes2022pathling} and commercial offerings like Aidbox
+\citep{aidbox}. \\
+\textbf{3} & Data analytics modules, that provide access to work and
+access the data. & \textbf{Federated learning (FL)
+\citep{rieke2020future} and privacy- enhancing technologies (PETs)}
+\citep{scheibner2021revolutionizing, jordan2022selecting}: new paradigms
+that address the problem of data governance and privacy by training
+algorithms collaboratively without exchanging the data itself. Models
+can be trained on combined datasets and made available as open source
+artifacts for decision support. Data analysts can use FL and PETs to
+work with the data in a collaborative, decentralized fashion. \\
+\textbf{4} & Trained models that have been derived from the data and can
+be used stand-alone for decision support. & \textbf{ONNX} \citep{onnx}:
+an open format built to represent machine learning models. \\
+\end{longtable}
+
+\section{Methods}\label{methods}
+
+In the paper, we present a design that extends the OpenHIE specification
+to include the four types of data sharing mentioned above. Using the
+full-STAC approach \citep{mehl2023fullstac} we combine open standards,
+open technologies and open architectures into a coherent modular HDP
+platform that can be configured and re-used across a variety of
+use-cases. Subsequently, we employ the Framework for Evaluation of
+Design Science (FEDS) research \citep{venable2016feds} to evaluate the
+design. The purpose of this evaluation is formative: we aim to to help
+improve the outcomes of the design and the artifacts presented. We do so
+by taking a naturalistic approach, that is, the expected performance of
+the design is considered with its real environment where we draw on our
+experience in two use-cases, namely the Momcare programme and the ONA
+Canopy data platform.
+
+As part of our design research, we have taken a narrative approach in
+surveying existing scientific studies on health data platforms, focusing
+on the seminal reports and subsequently searching forward citations. In
+addition, we have searched the open source repositories (most notably
+GitHub) and the online communities (OpenHIE community, FHIR community)
+to search for relevant open standards, technologies and architectures.
+This paper should not be considered as a proper systematic review.
+
+\subsection{Momcare}\label{momcare}
+
+MomCare was launched in Kenya
+\citep{huisman2022digital, sanctis2022maintaining} and Tanzania
+\citep{shija2021access, mrema2021application} in 2017 and 2019
+respectively, with the objective to create transparency on the journeys
+of pregnant mothers. The programme is built on three pillars: journey
+tracking, quality support and a mobile wallet
+(Figure~\ref{fig-momcare}). MomCare distinguishes two user groups:
+mothers are supported during their pregnancy through reminders and
+surveys, using SMS as the digital mode of engagement. Health workers are
+equipped with an Android-based application, in which visits, care
+activities and clinical observations are recorded. Reimbursements of the
+maternal clinic are based on the data captured with SMS and the app,
+thereby creating a conditional payment scheme, where providers are
+partially reimbursed up-front for a fixed bundle of activities,
+supplemented by bonus payments based on a predefined set of care
+activities.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{images/momcare.png}
+
+}
+
+\caption{\label{fig-momcare}Overview of the Momcare programme}
+
+\end{figure}%
+
+\subsection{Use-case: ONA Canopy}\label{use-case-ona-canopy}
+
+https://ona.io/home/services/canopy/
+
+\subsection{Evaluation of other reported HDP
+designs}\label{evaluation-of-other-reported-hdp-designs}
+
+Besides the two case studies, we will discuss how the design relates to
+other reported open architectures
 
@@
 \item
-  \href{https://quarto.org/docs/extensions/formats.html}{Creating
-  Formats} in Quarto as part of the
-  \href{https://quarto.org/docs/extensions/}{Extensions} mechanism.
+  Ayaz et al.~presented an end-to-end framework that automates the
+  mapping from FHIR Resources to FHIR-compliant tables that are
+  subsequently exposed for consumption \citep{ayaz2023transforming}
 \item
-  \href{https://quarto.org/docs/journals/}{Journals Articles} for
-  Quarto.
+  Mullie et al.~presented the CODA project that aims to facilitate
+  distributed computations for multi-institutional data analysis while
+  avoiding the costs and complexity of data pooling
+  \citep{mullie2023coda}
+\item
+  PHAS \citep{abbasian2023phas}
+\item
+  OpenHIM platform with analytics
 \end{itemize}
 
-\subsection{Structure of this
-repository}\label{structure-of-this-repository}
+\section{Results}\label{results}
 
-Everything for the extensions is in \texttt{\_extensions}. See Quarto
-doc for details.
+\subsection{Open standards: using FHIR as the common data
+model}\label{open-standards-using-fhir-as-the-common-data-model}
 
-\begin{itemize}
-\item
-  In \texttt{partials}, you'll find the \texttt{.tex} partials that can
-  be used and should be removed or tweaked,s
-\item
-  Your extension can make shortcodes and lua filters available. This
-  document shows the effect of the one provided in the \texttt{aft}
-  format.
-\item
-  \texttt{aft} format sets some defaults which are different from
-  \texttt{pdf} or \texttt{html}, link setting links to URL in read
-  inside PDF output.
-\end{itemize}
+The recent convergence to FHIR as the de facto standard for information
+exchange has fuelled the development of OpenHIE. FHIR is currently used
+both for routine healthcare settings40 and clinical research settings
+\citep{duda2022hl7, vorisek2022fast} and is increasingly being used in
+LMICs as well. The FHIR-native OpenSRP platform \citep{mehl2020open} has
+been deployed in 14 countries targeting various patient populations,
+amongst which a reference implementation of the WHO antenatal and
+neonatal care guidelines for midwives in Lombok, Indonesia
+\citep{summitinstitutefordevelopment2023bunda, kurniawan2019midwife}. In
+India, FHIR is used as the underlying technology for the open Health
+Claims Exchange protocol specification, which has been adopted by the
+Indian government as the standard for e-claims handling \citep{hcx}.
+Additionally, the guidelines and standards of the African Union
+explicitly state FHIR is to be used as the messaging standard
+\citep{2023african}. This range of utilizations showcase the standards'
+widespread applicability. The proceedings of the OpenHIE conference 2023
+attest to the fact that FHIR and open source technologies are embraced
+as critical enablers in implementing health information exchanges in
+LMICs \citep{ohie23}.
 
-Source repository for this template format can found on
-\href{https://github.com/quarto-journals/article-format-template}{Github}
+However, despite the increased use of FHIR as a common data model,
+various studies have investigated its merits and performance vis-a-vis
+other healthcare standards. Comparisons between OpenEHR, ISO 13606, OMOP
+and FHIR have been made
+\citep{ayaz2023transforming, mullie2023coda, rinaldi2021openehr, cremonesi2023need, sinaci2023data}.
+A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR
+concluded that i) these three standards are functionally and technically
+compatible, and therefore can be used side by side; and that ii) each of
+these standards have their strengths and limitations that correlate with
+their intended use as summarized in the Table~\ref{tbl-comparison}.
 
-\subsubsection{\texorpdfstring{\texttt{\_extensions\textbackslash{}aft}}{\_extensions\textbackslash aft}}\label{extensionsaft}
+\begin{table}
 
-In this folder you'll find everything that defines the extensions which
-could be installed using \texttt{quarto\ install\ extension} or be part
-of the template when using \texttt{quarto\ use\ template}
+\caption{\label{tbl-comparison}Comparison of OpenEHR, ISO 13606 and FHIR
+standards}
 
-\begin{description}
-\item[Format Metadata]
-This is in \texttt{\_extension.yml} is where all the metadata about the
-format are defined so that Quarto knows what to use. Adapt this file for
-you own template.
-\item[Partials]
-In \texttt{partials}, there are the \texttt{.tex} files that will be
-used as Pandoc's template. We provide here all the partials supported by
-Quarto and custom one for this format. Quarto allows to provide partials
-to ease the process of tweaking the default latex Pandoc's template and
-keeping it up to date.\\
-This template repo contains all the relevant partials that you can use
-with Quarto \emph{as example}. We only tweaked \texttt{title.tex} to
-show the usage of a custom partials called \texttt{\_custom.tex}.\\
-\textbf{Only keep the partials that you need to tweak for the format you
-are creating}
+\centering{
 
-If you need to completely change the default template (i.g customizing
-partials is not enough), then you need to provide your own template to
-Pandoc based on
-\href{https://github.com/quarto-dev/quarto-cli/blob/main/src/resources/formats/pdf/pandoc/template.tex}{\texttt{template.tex}}
-and also using partials or not. This can be provided using the
-\texttt{template} YAML key in \texttt{\_extension.yml} for Quarto to use
-it.
+\includegraphics{images/comparison-ehr-standards.png}
 
-This is considered advanced configuration as it will be harder to
-maintain than only using partials but could be required for some
-specific format. Be aware that this may lead to loose some Pandoc or
-Quarto features tied to default \texttt{template.tex} content if you
-remove some specific parts.
-\item[Lua Filters]
-Most of the time, custom formats will need Lua filters to provide
-specific features like cross format supports or provides custom
-shortcodes through the Quarto extension mechanism. Those filters will be
-available to the user and could be used in the custom formats (according
-to \texttt{\_extensions} metadata). We have provided two examples:
+}
+
+\end{table}%
+
+For an infectious diseases dataset with a limited scope, OpenEHR, OMOP
+and FHIR have been compared and found all to be equally suitable
+\citep{rinaldi2021openehr}. Comparing OMOP and FHIR, the latter has been
+found to support more granular mappings required for analytics and was
+therefore chosen as the standard for the CODA
+project\citep{mullie2023coda}.
+
+Although FHIR was originally designed only for exchange between systems,
+we propose to use it as the common data model for the design presented
+here for the following reasons:
 
@@
 \item
-  \texttt{color-text.lua}, a Lua filter used to add color to inline text
-  for PDF and HTML outputs using the same Markdown syntax
+  Industry adoption has significantly increased, as exemplified by
+  FHIR-based offering by major cloud providers such as Google, Azure and
+  AWS. Also, Africa CDC has explicitly chosen FHIR as the preferred
+  standard;
 \item
-  \texttt{shorcodes.lua}, a Lua filter which follow
-  \href{https://quarto.org/docs/authoring/shortcodes.html\#custom-shortcodes}{Quarto
-  custom shortcodes} guidelines to provide a \texttt{} shortcode to
-  nicely print LaTeX in PDF and HTML.
+  The concept of FHIR Profiles allow localisation to tailor the standard
+  to a specific use case. A profile defines rules, extensions, and
+  constraints for a resource. We posit that the possible penalty of this
+  flexibility, namely having to manage different FHIR versions and/or
+  profiles, is less of an issue in the context of LMICs where first
+  priority is to exchange datasets such as the International Patient
+  Summary (IPS) that are less complex compared to the requirements for
+  high income countries;
+\item
+  Being based on webstandards, the FHIR standard lends itself best for
+  further separation of concerns as envisioned by the composable data
+  stack. This is an important enabler for the downward scalability of
+  the solution;
+\item
+  With its inherent, graph-like nature, FHIR can be readily incorporated
+  into the principles of FAIR data sharing, where FHIR-based data
+  repositories can be integrated in an overarching netwerk of FAIR data
+  stations \citep{sinaci2023data, pedrera-jimenez2023can}.
 \end{itemize}
 
-\textbf{Remove or replace with your own Lua filters}
-\item[Format resources]
-Resources required by the format needs to be available. We have provided
-two examples:
+\subsection{Open architecture: extending OpenHIE framework with a
+lakehouse data
+platform}\label{open-architecture-extending-openhie-framework-with-a-lakehouse-data-platform}
+
+Data management and analytics platforms have undergone significant
+changes since the first generation of data warehouses were introduced.
+Recent studies have shown that the current practice has converged
+towards the lakehouse as one of the most commonly used solution designs
+\citep{armbrust2021lakehouse, hai2023data, harby2022data}. Lakehouses
+typically have a zonal architecture \citep{hai2023data} where data is
+ingested from the source systems in bulk (E), delivered to storage with
+aligned schemas (L) and transformed into a format ready for analysis
+(T). The discerning characteristic of the lakehouse architecture is its
+foundation on low-cost and directly-accessible storage that also
+provides traditional analytical DBMS management and performance features
+such as ACID transactions, data versioning, auditing, indexing, caching,
+and query optimization \citep{armbrust2021lakehouse}. Lakehouses thus
+combine the key benefits of data lakes and data warehouses: low-cost
+storage in an open format accessible by a variety of systems from the
+former, and powerful management and optimization features from the
+latter.
+
+With respect to current implementations of lakehouse data platform, we
+observe a proliferation of tools with as yet limited standards to
+improve technical interoperability. In the analysis of Pedreira et al.
+\citep{pedreira2023composable} the requirement for specialization in
+data management systems has evolved faster than our software development
+practices. This situation has created a siloed landscape composed of
+hundreds of products developed and maintained as monoliths, with limited
+reuse between systems. It has also affected the end users, who are often
+required to learn the idiosyncrasies of dozens of incompatible SQL and
+non-SQL API dialects, and settle for systems with incomplete
+functionality and inconsistent semantics. To remedy this, Pedreira et
+al.~call to (re-)design and implement modern data platforms in terms of
+a `composable data stack' as a means to decrease development and
+maintenance cost and pick-up the speed of innovation.
+
+While the lakehouse architecture separating the concerns of compute and
+storage, the composable data stack takes the separation of concerns is
+taken one step further. A composable data system
+(Figure~\ref{fig-composable-data-stack}), not only separates the storage
+(layer 3) and execution (layer 2), but also separates the user interface
+(layer 1) from the execution engine by introducing standards for
+Intermediate Representation (standard A) and Connectivity (standard B).
+The composable data stack can be implemented with current open source
+technologies (Figure~\ref{fig-cds-examples}). As an example, the Ibis
+user interface is currently sufficiently mature to offer a standardized
+dataframe interface to 19 different execution engines.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{images/composable-data-stack.png}
+
+}
+
+\caption{\label{fig-composable-data-stack}Composable data stack}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{images/composable-data-stack-implementation.png}
+
+}
+
+\caption{\label{fig-cds-examples}Examples of implementations of the
+composable data stack}
+
+\end{figure}%
+
+\subsection{Extended OpenHIE
+framework}\label{extended-openhie-framework}
+
+Our proposal for extending HIE is as follows.
 
@@
 \item
-  \texttt{te.bst} is a biblio style file for demo. It has been
-  downloaded from
-  https://www.economics.utoronto.ca/osborne/latex/TEBST.HTM
-  (http://econtheory.org/technical/te.bst) - Licence
-  \href{https://www.latex-project.org/lppl/}{LPPL}
+  add diagram, explicitly adding the component
 \item
-  \texttt{aft.cls} is a dummy class file for this example format. It is
-  a copy of official \texttt{article.cls}, the one provided in LaTeX
-  installation (i.e at \texttt{kpsewhich\ article.cls}) and renamed as
-  example (Licence LaTeX Project Public License)
+  fit in structure of OpenHIE specification
 \item
-  \texttt{custom.scss} is a style file to have a custom theme for our
-  HTML format so that our Lua filter feature \texttt{color-tex.lua}
-  works.
+  check which workflows are related to analytics
 \end{itemize}
 
-Those files are referenced within the \texttt{\_extension.yml} to be
-used with our example format.
-
-\textbf{Remove and replace with your own resources}
-\item[\texttt{.quartoignore}]
-Sometimes it is useful to have some files only needed for reference or
-for development. They should be available in the source repository but
-not downloaded to the user when \texttt{quarto\ use\ template} is used.
-
-\textbf{Use \texttt{.quartoignore} to register such file and folder (one
-file or folder per line)}
-\item[\texttt{style-guide} folder]
-For \texttt{quarto-journals} format, use \texttt{style-guide} folder to
-include any documentation and resourced used for format creation, like a
-journal style guide or original \texttt{.tex} template. This folder is
-already added in \texttt{.quartoignore} in this example repo.
-
-\textbf{Remove, rename or add to this folder}
-\item[\texttt{template.qmd}]
-This file is the template document that shows how to use the custom
-format. It will be downloaded with other resource by
-\texttt{quarto\ use\ template}, and even offered to be renamed if the
-name \texttt{template.qmd} is used.
-
-This file will usually use the custom format (here \texttt{aft-pdf} and
-\texttt{aft-html}) and show how to use the template. When you'll copy
-this template, you should be able to render this document to the demo
-format.
-
-\textbf{Adapt this file to provide a suitable template for your custom
-format}
-\item[Other files]
-Other files are needed by the template and are usually user provided -
-they are not part of the custom format.
-
-Here \texttt{bibliography.bib} is here to demo the usage of the bst file
-from the custom format.
-
-\textbf{Remove this file and provide a suitable one for your template}
-\end{description}
-
-\newpage{}
-
-\subsection{Checklist: Creating a custom
-format}\label{checklist-creating-a-custom-format}
-
-Here is the checklist to help you know what to modify:
+Following Hai we take a subset of the core functionalities of a data
+lake. Because the SHR is the key source, don't need full stack of
+ingestion of various sources.
 
@@
 \item
-  Read the resources mentioned at the top,
-\item
-  Use this template repo to create a new repository for your format
-  (Click on ``Use this template'' to create new github repo)
-\item
-  Once you are acquainted with the content, remove the resources that
-  are there only as example (see above)
-\item
-  Update README by replacing \texttt{aft} and
-  \texttt{Article\ Format\ Template} mentions for your journal format
-\item
-  Keep only the template partials that you need to tweak, and add custom
-  ones if needed
-\item
-  Add any Lua filters for shortcodes and other that would be useful to
-  create the expected output format
-\item
-  Add any external resource your format will need, and that should be
-  part of the extension format that will be downloaded,
-\item
-  Check \texttt{\_extension.yml} is updated correctly
-\item
-  Modify the skeleton \texttt{template.qmd} to your format and add any
-  required resources to be downloaded to user.
-\item
-  Check \texttt{.quartoignore} is updated which everything that should
-  not be downloaded.
-\item
-  Publish a demo of you format to github pages of the repo by using
-  \texttt{quarto\ publish} command
+  Query-driven data discovery
 \end{itemize}
 
-\subsection{Demo of some features found in this demo journal
-template}\label{demo-of-some-features-found-in-this-demo-journal-template}
+\subsection{Open technologies: available digital public
+goods}\label{open-technologies-available-digital-public-goods}
 
-\subsubsection{Shortcode demo}\label{sec-shortcode}
-
-PDF are rendered using {\LaTeX} but it is best if one can use a Markdown
-syntax for cross format support.
-
-\texttt{} used in source is a shortcode syntax where the shortcode is
-included in the extension folder \texttt{\_extensions}
-
-\subsubsection{Code chunk}\label{sec-chunks}
-
-This format hide chunks by default as option has been set in
-\texttt{\_extension.yml} file.
-
-\subsubsection{Text color}\label{sec-summary}
-
-Our format makes applying color on inline text possible using the
-\texttt{{[}content{]}\{color=\textless{}name\textgreater{}\}} syntax.
-Let's see an example.
-
-Here we are using a special feature of our format which is the coloring
-because \textcolor{mypink}{pink is a \textbf{nice} color}.
-
-This is possible thanks to the Lua Filter included in the custom
-extension format.
-
-\subsubsection*{Using references}\label{using-references}
-\addcontentsline{toc}{subsubsection}{Using references}
-
-I did not read this book \citep{CameronTrivedi2013} but it must be
-interesting.
-
-Differences between \texttt{aft-html} and \texttt{aft-pdf}:
+Many components of the OpenHIE specification are now available as a
+digital public good. Table~\ref{tbl-digital-public-goods} lists
+components that are currently available for implementing the OpenHIE
+framework using open source, digital public goods that are compliant
+with the FHIR standard, illustrating the maturity of this ecosystem and
+development community. With the launch of the Instant OpenHIE
+configuration toolkit54, it has become easier to set up, explore and
+develop HIEs thereby reducing costs and skills required for software
+developers to deploy an OpenHIE architecture for quicker solution
+testing and as a starting point for faster production implementation and
+customisation. Several frameworks are available that offer a set of
+preconfigured components out of the box, such as for example:
 
@@
 \item
-  For the HTML format, we are using Pandoc citeproc to include the
-  bibliography. Here \texttt{reference-section-title} controls the title
-  for the chapter that will be used.
+  the ``Open Smart Register Platform'' (OpenSRP), that focuses on
+  providing a mobile-first platform, including a FHIR native app
+  designed to support the WHO Smart Guidelines
 \item
-  For the PDF format, \texttt{natbib} is used by default and the
-  bibliography is included with a title by the LaTeX template.
+  the OpenHIM Platform, a reference implementation of the Instant
+  OpenHIE framework, providing an easy way to set up, manage and operate
+  various HIE configurations
 \end{itemize}
 
-%% begin pandoc before-bib
-%% end pandoc before-bib
-%% begin pandoc biblio
-%% end pandoc biblio
-%% begin pandoc include-after
-%% end pandoc include-after
-%% begin pandoc after-body
-%% end pandoc after-body
+\begin{table}
+
+\caption{\label{tbl-digital-public-goods}Overview of current open source
+implementations of components included in the OpenHIE specification that
+are FHIR-compatible. The category Analytics Services is not a part of
+the original OpenHIE and is discussed in the paper. Point-of-Service
+systems are excluded for brevity. A systematic review of such digital
+public goods is beyond the scope of this document.}
+
+\centering{
+
+\includegraphics{images/digital-public-goods.png}
+
+}
+
+\end{table}%
+
+\subsection{Evaluation: Momcare
+programme}\label{evaluation-momcare-programme}
+
+\subsection{Evaluation: ONA Canopy}\label{evaluation-ona-canopy}
+
+\subsection{Evaluation: reported HDP
+solutions}\label{evaluation-reported-hdp-solutions}
+
+\section{Discussion}\label{discussion}
+
+\subsection{Openness of data
+platforms}\label{openness-of-data-platforms}
+
+We specifically address the notion of openness of HDPs in LMICs in terms
+of the design-related questions put forward by de Reuver at al.11:
+
+\begin{itemize}
+\tightlist
+\item
+  Object of openness: what data-related resources should data platforms
+  make available when opening up (e.g.~data, data products, datadriven
+  insights, analytics modules)? Which user groups derive value from
+  accessing data-related resources from data platforms (e.g.~data
+  providers, data users, intermediaries, developers)?
+\item
+  Unit of analysis: what is platform-to-platform openness in the context
+  of data platforms, given the expectation that different HDPs will
+  emerge at various aggregation levels? How do we distinguish
+  meta-platforms, forking, and platform interoperability?
+\item
+  Risk of openness: What are the novel (negative) implications of
+  opening up data platforms? How can reflexivity in design help
+  providers to resolve the negative implications of openness?
+\end{itemize}
+
+\subsection{Comparison with HMIS
+component}\label{comparison-with-hmis-component}
+
+\begin{itemize}
+\tightlist
+\item
+  Workflow requirements: Report aggregate data (link): receiver is HMIS,
+  mADX
+\item
+  Functional requirements:
+  https://guides.ohie.org/arch-spec/openhie-component-specifications-1/openhie-health-management-information-system-hmis
+\end{itemize}
+
+Requirements are similar, but implementation differs: Datamodel is
+non-FHIR, focused on DataValue, which conceptually equates to FHIR
+Measure
+
+\subsection{FHIR and FAIR}\label{fhir-and-fair}
+
+\begin{itemize}
+\tightlist
+\item
+  How does FHIR relate to approaches taken by the FAIR community, which
+  tend to take more an approach of using knowledge graphs. For example,
+  VODAN Africa \citep{gebreslassie2023fhir4fair, purnamajati2022data}.
+\item
+  FAIR principles vs FHIR graph: is FHIR a FAIR Data Object
+\end{itemize}
+
+\subsection{Attribute-based access
+control}\label{attribute-based-access-control}
+
+\begin{itemize}
+\tightlist
+\item
+  Cerbos, attribute based on lineage or anonymized tables
+\end{itemize}
+
+\section{Abbreviations}\label{abbreviations}
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}@{}}
+\toprule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+ELT & Extract, Load and Transform \\
+FAIR & Findable, Accessible, Interoperable and Reusable \\
+FHIR & Fast Healthcare Interoperability Resources \\
+HDP & Health data platform, explicitly differentiated from health
+digital platform \\
+HIE & Health Information Exchange \\
+FL & Federated learning \\
+LMIC & Low- and middle income countries \\
+PET & Privacy-enhancing technologies \\
+\end{longtable}
+
+  \bibliography{pharmaccess.bib}
 
 \end{document}
-\endinput
-%%
-%% End of file `sample-manuscript.tex'.

### 2024-02-14 paper-data-platforms.qmd
wip

@@
     journal:
+      name: JIMR
       cite-style: authoryear

### 2024-02-16 paper-data-platforms.qmd
wip

@@
 The objective of openness, the ways to realize openness and the risks of opening up can be considered as elements within the ongoing debate on data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond] in the healthcare domain, which is a current issue for high income countries and LMICs alike.
-Openness is particularly relevant if we are to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@otto2022designing].
+Openness is particularly relevant if we are to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].
 

### 2024-02-16 paper-data-platforms.tex
wip

@@
 \makeatother
+\journal{JIMR}
 \ifLuaTeX
@@
 and iii) moves society towards equity and justice by counteracting
-dynamics of data extraction \citep{otto2022designing}.
+dynamics of data extraction \citep{prainsack2022data}.
 
@@
 \textbf{3} & Data analytics modules, that provide access to work and
-access the data. & \textbf{Federated learning (FL)
-\citep{rieke2020future} and privacy- enhancing technologies (PETs)}
-\citep{scheibner2021revolutionizing, jordan2022selecting}: new paradigms
-that address the problem of data governance and privacy by training
-algorithms collaboratively without exchanging the data itself. Models
-can be trained on combined datasets and made available as open source
-artifacts for decision support. Data analysts can use FL and PETs to
-work with the data in a collaborative, decentralized fashion. \\
+access the data. & \textbf{Federated learning (FL)}
+\citep{rieke2020future} and \textbf{privacy- enhancing technologies
+(PETs)} \citep{scheibner2021revolutionizing, jordan2022selecting}: new
+paradigms that address the problem of data governance and privacy by
+training algorithms collaboratively without exchanging the data itself.
+Models can be trained on combined datasets and made available as open
+source artifacts for decision support. Data analysts can use FL and PETs
+to work with the data in a collaborative, decentralized fashion. \\
 \textbf{4} & Trained models that have been derived from the data and can

### 2024-02-16 paper-data-platforms.qmd
version 0.3

@@
 The recent convergence to FHIR as the de facto standard for information exchange has fuelled the development of OpenHIE.
-FHIR is currently used both for routine healthcare settings40 and clinical research settings [@duda2022hl7;@vorisek2022fast] and is increasingly being used in LMICs as well.
+FHIR is currently used both for routine healthcare settings[ @ayaz2021fast] and clinical research settings [@duda2022hl7;@vorisek2022fast] and is increasingly being used in LMICs as well.
 The FHIR-native OpenSRP platform [@mehl2020open] has been deployed in 14 countries targeting various patient populations, amongst which a reference implementation of the WHO antenatal and neonatal care guidelines for midwives in Lombok, Indonesia [@summitinstitutefordevelopment2023bunda;@kurniawan2019midwife].
@@
 For an infectious diseases dataset with a limited scope, OpenEHR, OMOP and FHIR have been compared and found all to be equally suitable [@rinaldi2021openehr].
-Comparing OMOP and FHIR, the latter has been found to support more granular mappings required for analytics and was therefore chosen as the standard for the CODA project[@mullie2023coda].
+Comparing OMOP and FHIR, the latter has been found to support more granular mappings required for analytics and was therefore chosen as the standard for the CODA project [@mullie2023coda].
 
@@
 
-With respect to current implementations of lakehouse data platform, we observe a proliferation of tools with as yet limited standards to improve technical interoperability.
+With respect to current implementations of lakehouse data platforms, we observe a proliferation of tools with as yet limited standards to improve technical interoperability.
 In the analysis of Pedreira et al.
@@
 
-While the lakehouse architecture separating the concerns of compute and storage, the composable data stack takes the separation of concerns is taken one step further.
+While the lakehouse architecture separates the concerns of compute and storage, the composable data stack takes the separation of concerns is taken one step further.
 A composable data system (@fig-composable-data-stack), not only separates the storage (layer 3) and execution (layer 2), but also separates the user interface (layer 1) from the execution engine by introducing standards for Intermediate Representation (standard A) and Connectivity (standard B).
@@
 
-Many components of the OpenHIE specification are now available as a digital public good.
+Many components of the OpenHIE specification are now available as a digital public goods.
 @tbl-digital-public-goods lists components that are currently available for implementing the OpenHIE framework using open source, digital public goods that are compliant with the FHIR standard, illustrating the maturity of this ecosystem and development community.
-With the launch of the Instant OpenHIE configuration toolkit54, it has become easier to set up, explore and develop HIEs thereby reducing costs and skills required for software developers to deploy an OpenHIE architecture for quicker solution testing and as a starting point for faster production implementation and customisation.
+With the launch of the Instant OpenHIE configuration toolkit[@InstantOpenHIEv2], it has become easier to set up, explore and develop HIEs thereby reducing costs and skills required for software developers to deploy an OpenHIE architecture for quicker solution testing and as a starting point for faster production implementation and customisation.
 Several frameworks are available that offer a set of preconfigured components out of the box, such as for example:

### 2024-02-16 paper-data-platforms.tex
version 0.3

@@
 exchange has fuelled the development of OpenHIE. FHIR is currently used
-both for routine healthcare settings40 and clinical research settings
-\citep{duda2022hl7, vorisek2022fast} and is increasingly being used in
-LMICs as well. The FHIR-native OpenSRP platform \citep{mehl2020open} has
-been deployed in 14 countries targeting various patient populations,
-amongst which a reference implementation of the WHO antenatal and
-neonatal care guidelines for midwives in Lombok, Indonesia
+both for routine healthcare settings\citep{ayaz2021fast} and clinical
+research settings \citep{duda2022hl7, vorisek2022fast} and is
+increasingly being used in LMICs as well. The FHIR-native OpenSRP
+platform \citep{mehl2020open} has been deployed in 14 countries
+targeting various patient populations, amongst which a reference
+implementation of the WHO antenatal and neonatal care guidelines for
+midwives in Lombok, Indonesia
 \citep{summitinstitutefordevelopment2023bunda, kurniawan2019midwife}. In
@@
 found to support more granular mappings required for analytics and was
-therefore chosen as the standard for the CODA
-project\citep{mullie2023coda}.
+therefore chosen as the standard for the CODA project
+\citep{mullie2023coda}.
 
@@
 
-With respect to current implementations of lakehouse data platform, we
+With respect to current implementations of lakehouse data platforms, we
 observe a proliferation of tools with as yet limited standards to
@@
 
-While the lakehouse architecture separating the concerns of compute and
+While the lakehouse architecture separates the concerns of compute and
 storage, the composable data stack takes the separation of concerns is
@@
 Many components of the OpenHIE specification are now available as a
-digital public good. Table~\ref{tbl-digital-public-goods} lists
+digital public goods. Table~\ref{tbl-digital-public-goods} lists
 components that are currently available for implementing the OpenHIE
@@
 development community. With the launch of the Instant OpenHIE
-configuration toolkit54, it has become easier to set up, explore and
-develop HIEs thereby reducing costs and skills required for software
-developers to deploy an OpenHIE architecture for quicker solution
-testing and as a starting point for faster production implementation and
-customisation. Several frameworks are available that offer a set of
-preconfigured components out of the box, such as for example:
+configuration toolkit\citep{InstantOpenHIEv2}, it has become easier to
+set up, explore and develop HIEs thereby reducing costs and skills
+required for software developers to deploy an OpenHIE architecture for
+quicker solution testing and as a starting point for faster production
+implementation and customisation. Several frameworks are available that
+offer a set of preconfigured components out of the box, such as for
+example:
 

### 2024-03-09 paper-data-platforms.qmd
Add d2 hie diagram

@@
 abstract: |
-  This document is only a demo explaining how to use the template.
-keywords: [template, demo]
+  TO DO: add abstract.
+keywords: [Analytics-on-FHIR, SQL-on-FHIR, HIE, data platforms, LMICs, digital health]
 reference-section-title: References
@@
 The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.
+In addition, we explicitly address the requirement of downward scalability as we aim to implement health data platforms in resource constrained setting of LMICs.
 
@@
 Using the full-STAC approach [@mehl2023fullstac] we combine open standards, open technologies and open architectures into a coherent modular HDP platform that can be configured and re-used across a variety of use-cases.
-Subsequently, we employ the Framework for Evaluation of Design Science (FEDS) research [@venable2016feds] to evaluate the design.
-The purpose of this evaluation is formative: we aim to to help improve the outcomes of the design and the artifacts presented.
-We do so by taking a naturalistic approach, that is, the expected performance of the design is considered with its real environment where we draw on our experience in two use-cases, namely the Momcare programme and the ONA Canopy data platform.
-
+Subsequently, we employ a formative, naturalistic evaluation to assess the technical risk and efficacy of the design [@venable2016feds].
+Given that it is prohibitively expensive to evaluate with real users and real systems in the real setting, we aim to minimize technological risks and maximize the efficacy of the design by considering three real-world examples of health data platoform in LMICs, namely i) the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)); ii) the ONA Canopy platform [https://ona.io/home/services/canopy/](https://ona.io/home/services/canopy/); and iii) the MomCare platform ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org).
 As part of our design research, we have taken a narrative approach in surveying existing scientific studies on health data platforms, focusing on the seminal reports and subsequently searching forward citations.
@@
 
-### Momcare
-
-MomCare was launched in Kenya [@huisman2022digital;@sanctis2022maintaining] and Tanzania [@shija2021access;@mrema2021application] in 2017 and 2019 respectively, with the objective to create transparency on the journeys of pregnant mothers.
-The programme is built on three pillars: journey tracking, quality support and a mobile wallet (@fig-momcare).
-MomCare distinguishes two user groups: mothers are supported during their pregnancy through reminders and surveys, using SMS as the digital mode of engagement.
-Health workers are equipped with an Android-based application, in which visits, care activities and clinical observations are recorded.
-Reimbursements of the maternal clinic are based on the data captured with SMS and the app, thereby creating a conditional payment scheme, where providers are partially reimbursed up-front for a fixed bundle of activities, supplemented by bonus payments based on a predefined set of care activities.
-
-![Overview of the Momcare programme](images/momcare.png){width=auto #fig-momcare}
-
-### Use-case: ONA Canopy
-
-https://ona.io/home/services/canopy/
-
-### Evaluation of other reported HDP designs
-
-Besides the two case studies, we will discuss how the design relates to other reported open architectures
-
-- Ayaz et al. presented an end-to-end framework that automates the mapping from FHIR Resources to FHIR-compliant tables that are subsequently exposed for consumption [@ayaz2023transforming]
-- Mullie et al.
-presented the CODA project that aims to facilitate distributed computations for multi-institutional data analysis while avoiding the costs and complexity of data pooling [@mullie2023coda]
-- PHAS [@abbasian2023phas]
-- OpenHIM platform with analytics
-
-## Results
+## Design
 
@@
 FHIR is currently used both for routine healthcare settings[ @ayaz2021fast] and clinical research settings [@duda2022hl7;@vorisek2022fast] and is increasingly being used in LMICs as well.
+The guidelines and standards of the African Union explicitly state FHIR is to be used as the messaging standard [@2023african].
 The FHIR-native OpenSRP platform [@mehl2020open] has been deployed in 14 countries targeting various patient populations, amongst which a reference implementation of the WHO antenatal and neonatal care guidelines for midwives in Lombok, Indonesia [@summitinstitutefordevelopment2023bunda;@kurniawan2019midwife].
 In India, FHIR is used as the underlying technology for the open Health Claims Exchange protocol specification, which has been adopted by the Indian government as the standard for e-claims handling [@hcx].
-Additionally, the guidelines and standards of the African Union explicitly state FHIR is to be used as the messaging standard [@2023african].
 This range of utilizations showcase the standards’ widespread applicability.
@@
 Also, Africa CDC has explicitly chosen FHIR as the preferred standard;
+- The widespread availibility of the Bulk FHIR API [@mandl2020push;@jones2021landscape] enables bulk, fileb-based nbatchwise processing for analytics using the lakehouse architecture as detailed in the next section;
 - The concept of FHIR Profiles allow localisation to tailor the standard to a specific use case.
@@
 - With its inherent, graph-like nature, FHIR can be readily incorporated into the principles of FAIR data sharing, where FHIR-based data repositories can be integrated in an overarching netwerk of FAIR data stations [@sinaci2023data;@pedrera-jimenez2023can].
+
+Possible risks pertaining to the use of FHIR as the common data model, most notably the possible incompatibilities and/or high costs of maintenance in supporting different versions, will be addressed in the Discussion.
 
@@
 
-### Extended OpenHIE framework
+The premise of separating the user interface from the execution engine is directly related to the key objective of the SQL-on-FHIR project ([https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/](https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/)), namely to make large-scale analysis of FHIR data accessible to a larger audience, portable between systems and to make FHIR data work well with the best available analytic tools, regardless of the technology stack.
+However, as FHIR is represented as a graph of resources, with detailed semantics defined for references between resources, data types, terminology, extensions, and many other aspects of the specification; to use FHIR effectively analysts require a thorough understanding of the specification.
+Most analytic and machine learning use cases require the preparation of FHIR data using transformations and tabular projections from its original form.
+The task of authoring these transformations and projections is not trivial and there is currently no standard mechanisms to support reuse.
+
+The solution of the SQL-on-FHIR project is to provide a specification for defining tabular, use case-specific views of FHIR data.
+The view definition and the execution of the view are separated, in such a way that the definition is portable across systems while the execution engine (called runners) are system-specific tools or libraries that apply view definitions to the underlying data layer, optionally making use of annotations to optimize performance.
+
 Our proposal for extending HIE is as follows.
@@
 Following Hai we take a subset of the core functionalities of a data lake. Because the SHR is the key source, don’t need full stack of ingestion of various sources.
+
+![](./images/hie.png)
 
@@
 
-### Evaluation: Momcare programme
+## Evaluation
 
-### Evaluation: ONA Canopy
+### OpenHIM platform
 
-### Evaluation: reported HDP solutions
+- View generation in different packages, not portable
+  - Logstash for bulk
+  - Kafka for streaming
+
+-
+
+### ONA Canopy
+
+### Momcare programme
+
+MomCare was launched in Kenya [@huisman2022digital;@sanctis2022maintaining] and Tanzania [@shija2021access;@mrema2021application] in 2017 and 2019 respectively, with the objective to create transparency on the journeys of pregnant mothers.
+The programme is built on three pillars: journey tracking, quality support and a mobile wallet (@fig-momcare).
+MomCare distinguishes two user groups: mothers are supported during their pregnancy through reminders and surveys, using SMS as the digital mode of engagement.
+Health workers are equipped with an Android-based application, in which visits, care activities and clinical observations are recorded.
+Reimbursements of the maternal clinic are based on the data captured with SMS and the app, thereby creating a conditional payment scheme, where providers are partially reimbursed up-front for a fixed bundle of activities, supplemented by bonus payments based on a predefined set of care activities.
+
+![Overview of the Momcare programme](images/momcare.png){width=auto #fig-momcare}
 
@@
 
+- TO DO: if you have generated flattened SQL tables, how are you going to manage security?
 - Cerbos, attribute based on lineage or anonymized tables

### 2024-03-09 paper-data-platforms.tex
Add d2 hie diagram

@@
   pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Antragama Abbas; Steven Wanyee; XXX; John Grimes; Mark van der Graaf; Mark de Reuver},
-  pdfkeywords={template, demo},
+  pdfkeywords={Analytics-on-FHIR, SQL-on-FHIR, HIE, data
+platforms, LMICs, digital health},
   colorlinks=true,
@@
 \begin{abstract}
-This document is only a demo explaining how to use the template.
+TO DO: add abstract.
 \end{abstract}
@@
 \begin{keyword}
-    template \sep 
-    demo
+    Analytics-on-FHIR \sep SQL-on-FHIR \sep HIE \sep data
+platforms \sep LMICs \sep 
+    digital health
 \end{keyword}
@@
 consensus of this addition to OpenHIE currently stands in the way of
-development projects that aim to establish HDPs in LMICs.
+development projects that aim to establish HDPs in LMICs. In addition,
+we explicitly address the requirement of downward scalability as we aim
+to implement health data platforms in resource constrained setting of
+LMICs.
 
@@
 platform that can be configured and re-used across a variety of
-use-cases. Subsequently, we employ the Framework for Evaluation of
-Design Science (FEDS) research \citep{venable2016feds} to evaluate the
-design. The purpose of this evaluation is formative: we aim to to help
-improve the outcomes of the design and the artifacts presented. We do so
-by taking a naturalistic approach, that is, the expected performance of
-the design is considered with its real environment where we draw on our
-experience in two use-cases, namely the Momcare programme and the ONA
-Canopy data platform.
-
-As part of our design research, we have taken a narrative approach in
+use-cases. Subsequently, we employ a formative, naturalistic evaluation
+to assess the technical risk and efficacy of the design
+\citep{venable2016feds}. Given that it is prohibitively expensive to
+evaluate with real users and real systems in the real setting, we aim to
+minimize technological risks and maximize the efficacy of the design by
+considering three real-world examples of health data platoform in LMICs,
+namely i) the OpenHIM platform
+(\url{https://jembi.gitbook.io/openhim-platform/}); ii) the ONA Canopy
+platform \url{https://ona.io/home/services/canopy/}; and iii) the
+MomCare platform (\url{https://health-data-commons.pharmaccess.org}. As
+part of our design research, we have taken a narrative approach in
 surveying existing scientific studies on health data platforms, focusing
@@
 
-\subsection{Momcare}\label{momcare}
-
-MomCare was launched in Kenya
-\citep{huisman2022digital, sanctis2022maintaining} and Tanzania
-\citep{shija2021access, mrema2021application} in 2017 and 2019
-respectively, with the objective to create transparency on the journeys
-of pregnant mothers. The programme is built on three pillars: journey
-tracking, quality support and a mobile wallet
-(Figure~\ref{fig-momcare}). MomCare distinguishes two user groups:
-mothers are supported during their pregnancy through reminders and
-surveys, using SMS as the digital mode of engagement. Health workers are
-equipped with an Android-based application, in which visits, care
-activities and clinical observations are recorded. Reimbursements of the
-maternal clinic are based on the data captured with SMS and the app,
-thereby creating a conditional payment scheme, where providers are
-partially reimbursed up-front for a fixed bundle of activities,
-supplemented by bonus payments based on a predefined set of care
-activities.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{images/momcare.png}
-
-}
-
-\caption{\label{fig-momcare}Overview of the Momcare programme}
-
-\end{figure}%
-
-\subsection{Use-case: ONA Canopy}\label{use-case-ona-canopy}
-
-https://ona.io/home/services/canopy/
-
-\subsection{Evaluation of other reported HDP
-designs}\label{evaluation-of-other-reported-hdp-designs}
-
-Besides the two case studies, we will discuss how the design relates to
-other reported open architectures
-
-\begin{itemize}
-\tightlist
-\item
-  Ayaz et al.~presented an end-to-end framework that automates the
-  mapping from FHIR Resources to FHIR-compliant tables that are
-  subsequently exposed for consumption \citep{ayaz2023transforming}
-\item
-  Mullie et al.~presented the CODA project that aims to facilitate
-  distributed computations for multi-institutional data analysis while
-  avoiding the costs and complexity of data pooling
-  \citep{mullie2023coda}
-\item
-  PHAS \citep{abbasian2023phas}
-\item
-  OpenHIM platform with analytics
-\end{itemize}
-
-\section{Results}\label{results}
+\section{Design}\label{design}
 
@@
 research settings \citep{duda2022hl7, vorisek2022fast} and is
-increasingly being used in LMICs as well. The FHIR-native OpenSRP
-platform \citep{mehl2020open} has been deployed in 14 countries
-targeting various patient populations, amongst which a reference
-implementation of the WHO antenatal and neonatal care guidelines for
-midwives in Lombok, Indonesia
+increasingly being used in LMICs as well. The guidelines and standards
+of the African Union explicitly state FHIR is to be used as the
+messaging standard \citep{2023african}. The FHIR-native OpenSRP platform
+\citep{mehl2020open} has been deployed in 14 countries targeting various
+patient populations, amongst which a reference implementation of the WHO
+antenatal and neonatal care guidelines for midwives in Lombok, Indonesia
 \citep{summitinstitutefordevelopment2023bunda, kurniawan2019midwife}. In
@@
 Indian government as the standard for e-claims handling \citep{hcx}.
-Additionally, the guidelines and standards of the African Union
-explicitly state FHIR is to be used as the messaging standard
-\citep{2023african}. This range of utilizations showcase the standards'
-widespread applicability. The proceedings of the OpenHIE conference 2023
-attest to the fact that FHIR and open source technologies are embraced
-as critical enablers in implementing health information exchanges in
-LMICs \citep{ohie23}.
+This range of utilizations showcase the standards' widespread
+applicability. The proceedings of the OpenHIE conference 2023 attest to
+the fact that FHIR and open source technologies are embraced as critical
+enablers in implementing health information exchanges in LMICs
+\citep{ohie23}.
 
@@
 \item
+  The widespread availibility of the Bulk FHIR API
+  \citep{mandl2020push, jones2021landscape} enables bulk, fileb-based
+  nbatchwise processing for analytics using the lakehouse architecture
+  as detailed in the next section;
+\item
   The concept of FHIR Profiles allow localisation to tailor the standard
@@
 \end{itemize}
+
+Possible risks pertaining to the use of FHIR as the common data model,
+most notably the possible incompatibilities and/or high costs of
+maintenance in supporting different versions, will be addressed in the
+Discussion.
 
@@
 
-\subsection{Extended OpenHIE
-framework}\label{extended-openhie-framework}
+The premise of separating the user interface from the execution engine
+is directly related to the key objective of the SQL-on-FHIR project
+(\url{https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/}), namely to make
+large-scale analysis of FHIR data accessible to a larger audience,
+portable between systems and to make FHIR data work well with the best
+available analytic tools, regardless of the technology stack. However,
+as FHIR is represented as a graph of resources, with detailed semantics
+defined for references between resources, data types, terminology,
+extensions, and many other aspects of the specification; to use FHIR
+effectively analysts require a thorough understanding of the
+specification. Most analytic and machine learning use cases require the
+preparation of FHIR data using transformations and tabular projections
+from its original form. The task of authoring these transformations and
+projections is not trivial and there is currently no standard mechanisms
+to support reuse.
+
+The solution of the SQL-on-FHIR project is to provide a specification
+for defining tabular, use case-specific views of FHIR data. The view
+definition and the execution of the view are separated, in such a way
+that the definition is portable across systems while the execution
+engine (called runners) are system-specific tools or libraries that
+apply view definitions to the underlying data layer, optionally making
+use of annotations to optimize performance.
 
@@
 ingestion of various sources.
+
+\includegraphics{./images/hie.png}
 
@@
 
-\subsection{Evaluation: Momcare
-programme}\label{evaluation-momcare-programme}
+\section{Evaluation}\label{evaluation}
 
-\subsection{Evaluation: ONA Canopy}\label{evaluation-ona-canopy}
+\subsection{OpenHIM platform}\label{openhim-platform}
 
-\subsection{Evaluation: reported HDP
-solutions}\label{evaluation-reported-hdp-solutions}
+\begin{itemize}
+\tightlist
+\item
+  View generation in different packages, not portable
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Logstash for bulk
+  \item
+    Kafka for streaming
+  \end{itemize}
+\item
+\end{itemize}
+
+\subsection{ONA Canopy}\label{ona-canopy}
+
+\subsection{Momcare programme}\label{momcare-programme}
+
+MomCare was launched in Kenya
+\citep{huisman2022digital, sanctis2022maintaining} and Tanzania
+\citep{shija2021access, mrema2021application} in 2017 and 2019
+respectively, with the objective to create transparency on the journeys
+of pregnant mothers. The programme is built on three pillars: journey
+tracking, quality support and a mobile wallet
+(Figure~\ref{fig-momcare}). MomCare distinguishes two user groups:
+mothers are supported during their pregnancy through reminders and
+surveys, using SMS as the digital mode of engagement. Health workers are
+equipped with an Android-based application, in which visits, care
+activities and clinical observations are recorded. Reimbursements of the
+maternal clinic are based on the data captured with SMS and the app,
+thereby creating a conditional payment scheme, where providers are
+partially reimbursed up-front for a fixed bundle of activities,
+supplemented by bonus payments based on a predefined set of care
+activities.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{images/momcare.png}
+
+}
+
+\caption{\label{fig-momcare}Overview of the Momcare programme}
+
+\end{figure}%
 
@@
 \item
+  TO DO: if you have generated flattened SQL tables, how are you going
+  to manage security?
+\item
   Cerbos, attribute based on lineage or anonymized tables

### 2024-03-25 paper-data-platforms.qmd
WIP: description design

@@
 format:
-  # docx
-  # typst:
-  #   columns: 1
   elsevier-pdf:
@@
       cite-style: authoryear
+  elsevier-html:
+    toc: true
+    toc-depth: 2
+
 author:
@@
 From a technological perspective, interoperability issues, weak integrations, siloed data repositories and overall lack of openness are often reported as key impediments [@malm-nicolais2023exploring;@mehl2023fullstac].
-From a societal perspective, issues pertaining to the winner-takes-all nature of digital platforms are hotly debated as many jurisdictions make work to ensure these platforms indeed serve the common good of achieving universal health coverage [@sharon2018when].
+From a societal perspective, issues pertaining to the winner-takes-all nature of digital platforms are hotly debated as many jurisdictions make work to ensure these new digital health platforms indeed serve the common good of achieving universal health coverage [@sharon2018when].
 
@@
 Particularly relevant in the context of health data platforms (HDPs), is the conceptualization of openness.
-The objective of openness, the ways to realize openness and the risks of opening up can be considered as elements within the ongoing debate on data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond] in the healthcare domain, which is a current issue for high income countries and LMICs alike.
+The shift in perspective from digital platforms to data platforms coincides with the paradox of open [@keller2021paradox].
+Originally, openness of digital platforms focused on open source, open standards and copyrights, which by has been superseded by “...
+conflicts about privacy, economic value extraction, the emergence of artificial intelligence, and the destabilizing effects of dominant platforms on (democratic) societies.
+Instead of access to information, the control of personal data has emerged in the age of platforms as the critical contention.” [@keller2021paradox].
+These conflicts are particularly salient in the healthcare domain, as is evidenced by polemic surrounding data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond].
 Openness is particularly relevant if we are to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].
@@
 This paper is motivated by the conflation of a number of developments relevant to the design and implementation of solidarity-based HDPs in LMICs.
-First, the OpenHIE framework [@openhie] has been adopted by many sub-Saharan African countries [@mamuye2022health] as the architectural blueprint for implementing nation-wide health information exchanges (HIE), including Nigeria [@ohienigeria], Kenya [@khisif] and Tanzania [@tzhea2020].
+First, the OpenHIE framework [@openhie] has been adopted by many sub-Saharan African countries [@mamuye2022health] as the architectural blueprint for implementing nation-wide health information exchanges (HIE), including Nigeria [@dalhatu2023paper], Kenya [@thaiya2021adoption] and Tanzania [@nsaghurwe2021one].
 These countries have, as a matter of course, extended the framework to include “analytics services” as an additional domain.
-The rationale for this addition is to facilitate primary and secondary re-use of health data for academic research, real-world evidence studies etc.
+The rationale for this addition is to facilitate secondary reuse of health data for academic research, real-world evidence studies etc.
 which can be framed within the context of ongoing efforts towards Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair].
@@
 Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods[@digitalpublicgoods].
-If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for data sharing and re-use.
+If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for data sharing and reuse.
 Distinguishing four types of data sharing (@tbl-types-data-sharing), the purpose of this paper is to investigate how new standards and technologies that can establish openness of health data platforms can be integrated into the OpenHIE architecture framework.
@@
 
-|   | __Type of data sharing__ | __New standards and technology enablers to create openness__ |
+|   | __Type of data sharing__ | __Technology enablers to create openness of health data platforms__ |
 |:-:|:-------|:-------------|
 | __1__ | Data at the most granular (patient) level, which is persisted and used to provide a longitudinal record.
-| __Bulk FHIR API__ which has by now been incorporated in all major FHIR implementations the FHIR standard can be readily used to support type 1 data sharing to patient-level data across a patient population [@mandl2020push;@jones2021landscape].
+| __Bulk FHIR API__: has by now been incorporated in all major FHIR implementations the FHIR standard can be readily used to support data sharing to patient-level data across a patient population [@mandl2020push;@jones2021landscape].
 |
 | __2__ | Aggregated data, for example statistics for policy evaluation and benchmarking |__SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows.
-Builds on FHIRPath [@fhirpath] expressions in a logical structure to specify things like column names and unnested items.
+Builds on FHIRPath [@hl72020fhirpath] expressions in a logical structure to specify things like column names and unnested items.
 Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].|
@@
 In the paper, we present a design that extends the OpenHIE specification to include the four types of data sharing mentioned above.
-Using the full-STAC approach [@mehl2023fullstac] we combine open standards, open technologies and open architectures into a coherent modular HDP platform that can be configured and re-used across a variety of use-cases.
+Using the full-STAC approach [@mehl2023fullstac] we combine open standards, open technologies and open architectures into a coherent modular HDP platform that can be configured and reused across a variety of use-cases.
 Subsequently, we employ a formative, naturalistic evaluation to assess the technical risk and efficacy of the design [@venable2016feds].
-Given that it is prohibitively expensive to evaluate with real users and real systems in the real setting, we aim to minimize technological risks and maximize the efficacy of the design by considering three real-world examples of health data platoform in LMICs, namely i) the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)); ii) the ONA Canopy platform [https://ona.io/home/services/canopy/](https://ona.io/home/services/canopy/); and iii) the MomCare platform ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org).
+Given that it is prohibitively expensive to evaluate with real users and real systems in the real setting, we aim to minimize technological risks and maximize the efficacy of the design by considering three real-world examples of health data platoform in LMICs, namely i) the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)); ii) the ONA Canopy platform [https://ona.io/home/services/canopy/](https://ona.io/home/services/canopy/); and iii) the work conducted at PharmAccess Foundation as part of the MomCare programme ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org)).
 As part of our design research, we have taken a narrative approach in surveying existing scientific studies on health data platforms, focusing on the seminal reports and subsequently searching forward citations.
@@
 The recent convergence to FHIR as the de facto standard for information exchange has fuelled the development of OpenHIE.
-FHIR is currently used both for routine healthcare settings[ @ayaz2021fast] and clinical research settings [@duda2022hl7;@vorisek2022fast] and is increasingly being used in LMICs as well.
+FHIR is currently used both for routine healthcare settings [@ayaz2021fast] and clinical research settings [@duda2022hl7;@vorisek2022fast] and is increasingly being used in LMICs as well.
 The guidelines and standards of the African Union explicitly state FHIR is to be used as the messaging standard [@2023african].
@@
 This range of utilizations showcase the standards’ widespread applicability.
-The proceedings of the OpenHIE conference 2023 attest to the fact that FHIR and open source technologies are embraced as critical enablers in implementing health information exchanges in LMICs [@ohie23].
+The proceedings of the OpenHIE conference 2023 attest to the fact that FHIR and open source technologies are embraced as critical enablers in implementing health information exchanges in LMICs [@ohie2023unconference].
 
-However, despite the increased use of FHIR as a common data model, various studies have investigated its merits and performance vis-a-vis other healthcare standards.
+Various studies have investigated the merits of FHIR and its performance vis-a-vis other healthcare standards.
 Comparisons between OpenEHR, ISO 13606, OMOP and FHIR have been made [@ayaz2023transforming;@mullie2023coda;@rinaldi2021openehr;@cremonesi2023need;@sinaci2023data].
@@
 
-### Open architecture: extending OpenHIE framework with a lakehouse data platform
+### Open architecture: extending OpenHIE with a composable data stack
+
+#### Evolving the data lakehouse to a composable data stack
 
@@
 Recent studies have shown that the current practice has converged towards the lakehouse as one of the most commonly used solution designs [@armbrust2021lakehouse;@hai2023data;@harby2022data].
-Lakehouses typically have a zonal architecture [@hai2023data] where data is ingested from the source systems in bulk (E), delivered to storage with aligned schemas (L) and transformed into a format ready for analysis (T).
+Lakehouses typically have a zonal architecture that follow the Extract-Load-Transform pattern (ELT) [@hai2023data] where data is ingested from the source systems in bulk (E), delivered to storage with aligned schemas (L) and transformed into a format ready for analysis (T).
 The discerning characteristic of the lakehouse architecture is its foundation on low-cost and directly-accessible storage that also provides traditional analytical DBMS management and performance features such as ACID transactions, data versioning, auditing, indexing, caching, and query optimization [@armbrust2021lakehouse].
@@
 While the lakehouse architecture separates the concerns of compute and storage, the composable data stack takes the separation of concerns is taken one step further.
-A composable data system (@fig-composable-data-stack), not only separates the storage (layer 3) and execution (layer 2), but also separates the user interface (layer 1) from the execution engine by introducing standards for Intermediate Representation (standard A) and Connectivity (standard B).
+A composable data system (@fig-composable-data-stack), not only separates the storage (layer 3) and execution (layer 2), but also separates the user interface (layer 1) from the execution engine by introducing standards including Substrait for Intermediate Representation (standard A) and Apache Arrow (standard B and C).
 The composable data stack can be implemented with current open source technologies (@fig-cds-examples).
@@
 
-![Examples of implementations of the composable data stack](images/composable-data-stack-implementation.png){#fig-cds-examples}
+![Examples of implementations of the composable data stack](images/composable-data-stack-implementation-2.png){#fig-cds-examples}
+
+#### SQL-on-FHIR v2 as an intermediate representation for FHIR data in tabular format
 
 The premise of separating the user interface from the execution engine is directly related to the key objective of the SQL-on-FHIR project ([https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/](https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/)), namely to make large-scale analysis of FHIR data accessible to a larger audience, portable between systems and to make FHIR data work well with the best available analytic tools, regardless of the technology stack.
-However, as FHIR is represented as a graph of resources, with detailed semantics defined for references between resources, data types, terminology, extensions, and many other aspects of the specification; to use FHIR effectively analysts require a thorough understanding of the specification.
+However, as FHIR is represented as a graph of resources, with detailed semantics defined for references between resources, data types, terminology, extensions, and many other aspects of the specification.
+To use FHIR effectively analysts require a thorough understanding of the specification.
 Most analytic and machine learning use cases require the preparation of FHIR data using transformations and tabular projections from its original form.
@@
 
-Our proposal for extending HIE is as follows.
+#### Extending OpenHIE with a FHIR-based composable data stack
 
-- add diagram, explicitly adding the component
+We propose to extend the OpenHIE architecture with a "Data and Analytics Services" domain with 5 services by synthesizing the architecture of Hai et al.
+[@hai2023data] and the composable data stack [@pedreira2023composable] (@fig-ohie, @tbl-data-and-analysis-services).
+
+![Proposed extension of the OpenHIE architecture that includes "Data and Analytics Services" as an additional service domain.](./images/hie.png){#fig-ohie}
+
++-------------------+---------------------------------------------------------------------------------------+
+| Service           | Functional requirements                                                               |
++===================+=======================================================================================+
+| Ingestion         | - Metadata extraction                                                                 |
+|                   | - Metadata modeling                                                                   |
++-------------------+---------------------------------------------------------------------------------------+
+| Storage           | - File-based blob storage                                                             |
++-------------------+---------------------------------------------------------------------------------------+
+| Query &           | - SQL-on-FHIR Runner                                                                  |
+| Processing        | - Execution engine on tabular data as defined in composable data stack                |  
++-------------------+---------------------------------------------------------------------------------------+
+| Maintenance       | - SQL-on-FHIR View defintions                                                         |
+|                   | - Maintenance-related functions as defined by Hai et al.                              |
++-------------------+---------------------------------------------------------------------------------------+
+| Exploration       | - Interactive computing environment [@granger2021jupyter]                             |
+|                   | - Capability to participate as a node in a federative learning network, including MPC |
++-------------------+---------------------------------------------------------------------------------------+
+
+: Definition of Data and Analysis Services {#tbl-data-and-analysis-services}
+
+##### Ingestion
+
+Default workflow is extraction of data from SHR using Bulk FHIR API.
+Data contains metadata (incl.
+FHIR versions) and fully qualified semantics, for example, coding systems.
+Despite this, metadata extraction and metadata modeling is still required to meet the FAIR requirements.
+Issues that need to be solved by these services:
+
+- To prepare for future updates of FHIR versions
+- Implement late-binding principle of having increasingly more specific FHIR profiles as bulk FHIR data propagates through lakehouse
+
+##### Storage
+
+- from ndjson to parquet
+- possibly used delta lake for time versioning
+- separation of storage from compute not only for benefits of lower TCO, but also be ready for federated learning and MPC in future
+
+##### Query & Processing
+
 - fit in structure of OpenHIE specification
 - check which workflows are related to analytics
+- Hai calls this 'Maintenance'
 
-Following Hai we take a subset of the core functionalities of a data lake. Because the SHR is the key source, don’t need full stack of ingestion of various sources.
+##### Maintenance
 
-![](./images/hie.png)
+- SQL-on-FHIR Views provide new standard to support mADX aggregate reporting
+- Maintenance-related functions remain the same
+- NB: orchestration falls under data provenance
 
-- Query-driven data discovery
+##### Exploration
 
@@
 
+### Federated learning and multiparty computation
+
+- data stations??!
+
 ## Abbreviations

### 2024-03-25 index.tex
Remove project, revert to single page html

--- 2024-03-09 paper-data-platforms.tex
@@
 platforms are hotly debated as many jurisdictions make work to ensure
-these platforms indeed serve the common good of achieving universal
-health coverage \citep{sharon2018when}.
+these new digital health platforms indeed serve the common good of
+achieving universal health coverage \citep{sharon2018when}.
 
@@
 Particularly relevant in the context of health data platforms (HDPs), is
-the conceptualization of openness. The objective of openness, the ways
-to realize openness and the risks of opening up can be considered as
-elements within the ongoing debate on data spaces
-\citep{otto2022designing} and data solidarity
-\citep{kickbusch2021lancet, prainsack2022data, prainsack2023beyond} in
-the healthcare domain, which is a current issue for high income
-countries and LMICs alike. Openness is particularly relevant if we are
-to realize a solidarity-based approach to health data sharing that i)
-gives people a greater control over their data as active decision
-makers; ii) ensures that the value of data is harnessed for public good;
-and iii) moves society towards equity and justice by counteracting
-dynamics of data extraction \citep{prainsack2022data}.
+the conceptualization of openness. The shift in perspective from digital
+platforms to data platforms coincides with the paradox of open
+\citep{keller2021paradox}. Originally, openness of digital platforms
+focused on open source, open standards and copyrights, which by has been
+superseded by ``\ldots{} conflicts about privacy, economic value
+extraction, the emergence of artificial intelligence, and the
+destabilizing effects of dominant platforms on (democratic) societies.
+Instead of access to information, the control of personal data has
+emerged in the age of platforms as the critical contention.''
+\citep{keller2021paradox}. These conflicts are particularly salient in
+the healthcare domain, as is evidenced by polemic surrounding data
+spaces \citep{otto2022designing} and data solidarity
+\citep{kickbusch2021lancet, prainsack2022data, prainsack2023beyond}.
+Openness is particularly relevant if we are to realize a
+solidarity-based approach to health data sharing that i) gives people a
+greater control over their data as active decision makers; ii) ensures
+that the value of data is harnessed for public good; and iii) moves
+society towards equity and justice by counteracting dynamics of data
+extraction \citep{prainsack2022data}.
 
@@
 architectural blueprint for implementing nation-wide health information
-exchanges (HIE), including Nigeria \citep{ohienigeria}, Kenya
-\citep{khisif} and Tanzania \citep{tzhea2020}. These countries have, as
-a matter of course, extended the framework to include ``analytics
-services'' as an additional domain. The rationale for this addition is
-to facilitate primary and secondary re-use of health data for academic
+exchanges (HIE), including Nigeria \citep{dalhatu2023paper}, Kenya
+\citep{thaiya2021adoption} and Tanzania \citep{nsaghurwe2021one}. These
+countries have, as a matter of course, extended the framework to include
+``analytics services'' as an additional domain. The rationale for this
+addition is to facilitate secondary reuse of health data for academic
 research, real-world evidence studies etc. which can be framed within
@@
 technologies and architecture to include functionality for data sharing
-and re-use. Distinguishing four types of data sharing
+and reuse. Distinguishing four types of data sharing
 (Table~\ref{tbl-types-data-sharing}), the purpose of this paper is to
@@
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-\textbf{New standards and technology enablers to create openness}
+\textbf{Technology enablers to create openness of health data platforms}
 \end{minipage} \\
@@
 persisted and used to provide a longitudinal record. & \textbf{Bulk FHIR
-API} which has by now been incorporated in all major FHIR
-implementations the FHIR standard can be readily used to support type 1
-data sharing to patient-level data across a patient population
+API}: has by now been incorporated in all major FHIR implementations the
+FHIR standard can be readily used to support data sharing to
+patient-level data across a patient population
 \citep{mandl2020push, jones2021landscape}. \\
@@
 well with familiar and efficient SQL engines that are most commonly used
-in analytical workflows. Builds on FHIRPath \citep{fhirpath} expressions
-in a logical structure to specify things like column names and unnested
-items. Implementations of this approach are available or forthcoming,
-including open source implementations such as Pathling
+in analytical workflows. Builds on FHIRPath \citep{hl72020fhirpath}
+expressions in a logical structure to specify things like column names
+and unnested items. Implementations of this approach are available or
+forthcoming, including open source implementations such as Pathling
 \citep{grimes2022pathling} and commercial offerings like Aidbox
@@
 open technologies and open architectures into a coherent modular HDP
-platform that can be configured and re-used across a variety of
+platform that can be configured and reused across a variety of
 use-cases. Subsequently, we employ a formative, naturalistic evaluation
@@
 (\url{https://jembi.gitbook.io/openhim-platform/}); ii) the ONA Canopy
-platform \url{https://ona.io/home/services/canopy/}; and iii) the
-MomCare platform (\url{https://health-data-commons.pharmaccess.org}. As
-part of our design research, we have taken a narrative approach in
-surveying existing scientific studies on health data platforms, focusing
-on the seminal reports and subsequently searching forward citations. In
+platform \url{https://ona.io/home/services/canopy/}; and iii) the work
+conducted at PharmAccess Foundation as part of the MomCare programme
+(\url{https://health-data-commons.pharmaccess.org}). As part of our
+design research, we have taken a narrative approach in surveying
+existing scientific studies on health data platforms, focusing on the
+seminal reports and subsequently searching forward citations. In
 addition, we have searched the open source repositories (most notably
@@
 exchange has fuelled the development of OpenHIE. FHIR is currently used
-both for routine healthcare settings\citep{ayaz2021fast} and clinical
+both for routine healthcare settings \citep{ayaz2021fast} and clinical
 research settings \citep{duda2022hl7, vorisek2022fast} and is
@@
 enablers in implementing health information exchanges in LMICs
-\citep{ohie23}.
+\citep{ohie2023unconference}.
 
-However, despite the increased use of FHIR as a common data model,
-various studies have investigated its merits and performance vis-a-vis
-other healthcare standards. Comparisons between OpenEHR, ISO 13606, OMOP
-and FHIR have been made
+Various studies have investigated the merits of FHIR and its performance
+vis-a-vis other healthcare standards. Comparisons between OpenEHR, ISO
+13606, OMOP and FHIR have been made
 \citep{ayaz2023transforming, mullie2023coda, rinaldi2021openehr, cremonesi2023need, sinaci2023data}.
@@
 
-\subsection{Open architecture: extending OpenHIE framework with a
-lakehouse data
-platform}\label{open-architecture-extending-openhie-framework-with-a-lakehouse-data-platform}
+\subsection{Open architecture: extending OpenHIE with a composable data
+stack}\label{open-architecture-extending-openhie-with-a-composable-data-stack}
+
+\subsubsection{Evolving the data lakehouse to a composable data
+stack}\label{evolving-the-data-lakehouse-to-a-composable-data-stack}
 
@@
 \citep{armbrust2021lakehouse, hai2023data, harby2022data}. Lakehouses
-typically have a zonal architecture \citep{hai2023data} where data is
+typically have a zonal architecture that follow the
+Extract-Load-Transform pattern (ELT) \citep{hai2023data} where data is
 ingested from the source systems in bulk (E), delivered to storage with
@@
 (layer 3) and execution (layer 2), but also separates the user interface
-(layer 1) from the execution engine by introducing standards for
-Intermediate Representation (standard A) and Connectivity (standard B).
-The composable data stack can be implemented with current open source
-technologies (Figure~\ref{fig-cds-examples}). As an example, the Ibis
-user interface is currently sufficiently mature to offer a standardized
-dataframe interface to 19 different execution engines.
+(layer 1) from the execution engine by introducing standards including
+Substrait for Intermediate Representation (standard A) and Apache Arrow
+(standard B and C). The composable data stack can be implemented with
+current open source technologies (Figure~\ref{fig-cds-examples}). As an
+example, the Ibis user interface is currently sufficiently mature to
+offer a standardized dataframe interface to 19 different execution
+engines.
 
@@
 
-\includegraphics{images/composable-data-stack-implementation.png}
+\includegraphics{images/composable-data-stack-implementation-2.png}
 
@@
 \end{figure}%
+
+\subsubsection{SQL-on-FHIR v2 as an intermediate representation for FHIR
+data in tabular
+format}\label{sql-on-fhir-v2-as-an-intermediate-representation-for-fhir-data-in-tabular-format}
 
@@
 defined for references between resources, data types, terminology,
-extensions, and many other aspects of the specification; to use FHIR
+extensions, and many other aspects of the specification. To use FHIR
 effectively analysts require a thorough understanding of the
@@
 
-Our proposal for extending HIE is as follows.
+\subsubsection{Extending OpenHIE with a FHIR-based composable data
+stack}\label{extending-openhie-with-a-fhir-based-composable-data-stack}
+
+We propose to extend the OpenHIE architecture with a ``Data and
+Analytics Services'' domain with 5 services by synthesizing the
+architecture of Hai et al. \citep{hai2023data} and the composable data
+stack \citep{pedreira2023composable} (Figure~\ref{fig-ohie},
+Table~\ref{tbl-data-and-analysis-services}).
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{./images/hie.png}
+
+}
+
+\caption{\label{fig-ohie}Proposed extension of the OpenHIE architecture
+that includes ``Data and Analytics Services'' as an additional service
+domain.}
+
+\end{figure}%
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1852}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8148}}@{}}
+\caption{Definition of Data and Analysis
+Services}\label{tbl-data-and-analysis-services}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Service
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Functional requirements
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Service
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Functional requirements
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Ingestion & \begin{minipage}[t]{\linewidth}\raggedright
+\begin{itemize}
+\tightlist
+\item
+  Metadata extraction
+\item
+  Metadata modeling
+\end{itemize}
+\end{minipage} \\
+Storage & \begin{minipage}[t]{\linewidth}\raggedright
+\begin{itemize}
+\tightlist
+\item
+  File-based blob storage
+\end{itemize}
+\end{minipage} \\
+Query \& Processing & \begin{minipage}[t]{\linewidth}\raggedright
+\begin{itemize}
+\tightlist
+\item
+  SQL-on-FHIR Runner
+\item
+  Execution engine on tabular data as defined in composable data stack
+\end{itemize}
+\end{minipage} \\
+Maintenance & \begin{minipage}[t]{\linewidth}\raggedright
+\begin{itemize}
+\tightlist
+\item
+  SQL-on-FHIR View defintions
+\item
+  Maintenance-related functions as defined by Hai et al.
+\end{itemize}
+\end{minipage} \\
+Exploration & \begin{minipage}[t]{\linewidth}\raggedright
+\begin{itemize}
+\tightlist
+\item
+  Interactive computing environment \citep{granger2021jupyter}
+\item
+  Capability to participate as a node in a federative learning network,
+  including MPC
+\end{itemize}
+\end{minipage} \\
+\end{longtable}
+
+\paragraph{Ingestion}\label{ingestion}
+
+Default workflow is extraction of data from SHR using Bulk FHIR API.
+Data contains metadata (incl.~FHIR versions) and fully qualified
+semantics, for example, coding systems. Despite this, metadata
+extraction and metadata modeling is still required to meet the FAIR
+requirements. Issues that need to be solved by these services:
 
@@
 \item
-  add diagram, explicitly adding the component
+  To prepare for future updates of FHIR versions
+\item
+  Implement late-binding principle of having increasingly more specific
+  FHIR profiles as bulk FHIR data propagates through lakehouse
+\end{itemize}
+
+\paragraph{Storage}\label{storage}
+
+\begin{itemize}
+\tightlist
+\item
+  from ndjson to parquet
+\item
+  possibly used delta lake for time versioning
+\item
+  separation of storage from compute not only for benefits of lower TCO,
+  but also be ready for federated learning and MPC in future
+\end{itemize}
+
+\paragraph{Query \& Processing}\label{query-processing}
+
+\begin{itemize}
+\tightlist
 \item
@@
   check which workflows are related to analytics
+\item
+  Hai calls this `Maintenance'
 \end{itemize}
 
-Following Hai we take a subset of the core functionalities of a data
-lake. Because the SHR is the key source, don't need full stack of
-ingestion of various sources.
-
-\includegraphics{./images/hie.png}
+\paragraph{Maintenance}\label{maintenance}
 
@@
 \item
-  Query-driven data discovery
+  SQL-on-FHIR Views provide new standard to support mADX aggregate
+  reporting
+\item
+  Maintenance-related functions remain the same
+\item
+  NB: orchestration falls under data provenance
 \end{itemize}
+
+\paragraph{Exploration}\label{exploration}
 
@@
 
+\subsection{Federated learning and multiparty
+computation}\label{federated-learning-and-multiparty-computation}
+
+\begin{itemize}
+\tightlist
+\item
+  data stations??!
+\end{itemize}
+
 \section{Abbreviations}\label{abbreviations}

### 2024-03-26 index.qmd
fix typos

--- 2024-03-25 paper-data-platforms.qmd
@@
 
-### Digital platforms vs. data platforms for healthcare
+### The paradox of open for digital vs. data platforms in healthcare
 
@@
 Also, Africa CDC has explicitly chosen FHIR as the preferred standard;
-- The widespread availibility of the Bulk FHIR API [@mandl2020push;@jones2021landscape] enables bulk, fileb-based nbatchwise processing for analytics using the lakehouse architecture as detailed in the next section;
+- The widespread availibility of the Bulk FHIR API [@mandl2020push;@jones2021landscape] enables bulk, file-based batchwise processing for analytics using the lakehouse architecture as detailed in the next section;
 - The concept of FHIR Profiles allow localisation to tailor the standard to a specific use case.

### 2024-03-26 index.tex
fix typos

@@
 
-\subsection{Digital platforms vs.~data platforms for
-healthcare}\label{digital-platforms-vs.-data-platforms-for-healthcare}
+\subsection{The paradox of open for digital vs.~data platforms in
+healthcare}\label{the-paradox-of-open-for-digital-vs.-data-platforms-in-healthcare}
 
@@
   The widespread availibility of the Bulk FHIR API
-  \citep{mandl2020push, jones2021landscape} enables bulk, fileb-based
-  nbatchwise processing for analytics using the lakehouse architecture
-  as detailed in the next section;
+  \citep{mandl2020push, jones2021landscape} enables bulk, file-based
+  batchwise processing for analytics using the lakehouse architecture as
+  detailed in the next section;
 \item

### 2024-03-26 index.qmd
wip

@@
     toc-depth: 2
+    fig-cap-location: top
 
@@
 
-This paper is motivated by the conflation of a number of developments relevant to the design and implementation of solidarity-based HDPs in LMICs.
+This paper is motivated by the conflation of a number of developments relevant to the design and implementation of open, solidarity-based HDPs in LMICs.
 First, the OpenHIE framework [@openhie] has been adopted by many sub-Saharan African countries [@mamuye2022health] as the architectural blueprint for implementing nation-wide health information exchanges (HIE), including Nigeria [@dalhatu2023paper], Kenya [@thaiya2021adoption] and Tanzania [@nsaghurwe2021one].
@@
 Recent studies have shown that the current practice has converged towards the lakehouse as one of the most commonly used solution designs [@armbrust2021lakehouse;@hai2023data;@harby2022data].
-Lakehouses typically have a zonal architecture that follow the Extract-Load-Transform pattern (ELT) [@hai2023data] where data is ingested from the source systems in bulk (E), delivered to storage with aligned schemas (L) and transformed into a format ready for analysis (T).
+Lakehouses typically have a zonal architecture that follow the Extract-Load-Transform pattern (ELT) where data is ingested from the source systems in bulk (E), delivered to storage with aligned schemas (L) and transformed into a format ready for analysis (T) [@hai2023data].
 The discerning characteristic of the lakehouse architecture is its foundation on low-cost and directly-accessible storage that also provides traditional analytical DBMS management and performance features such as ACID transactions, data versioning, auditing, indexing, caching, and query optimization [@armbrust2021lakehouse].
@@
 While the lakehouse architecture separates the concerns of compute and storage, the composable data stack takes the separation of concerns is taken one step further.
-A composable data system (@fig-composable-data-stack), not only separates the storage (layer 3) and execution (layer 2), but also separates the user interface (layer 1) from the execution engine by introducing standards including Substrait for Intermediate Representation (standard A) and Apache Arrow (standard B and C).
+A composable data system (@fig-composable-data-stack), not only separates the storage (layer 3) and execution (layer 2), but also separates the user interface (layer 1) from the execution engine by introducing standards including Substrait for Intermediate Representation (standard A, IR) and Apache Arrow for data connectivity and data memory layout (standards B and C, respectively).
 The composable data stack can be implemented with current open source technologies (@fig-cds-examples).
@@
 
-![Composable data stack](images/composable-data-stack.png){#fig-composable-data-stack}
+![The composable data stack. Image source: [https://voltrondata.com/codex](https://voltrondata.com/codex).](images/composable-data-stack.png){#fig-composable-data-stack}
 
-![Examples of implementations of the composable data stack](images/composable-data-stack-implementation-2.png){#fig-cds-examples}
+![Examples of implementations of the composable data stack.
+Image source: [https://voltrondata.com/codex](https://voltrondata.com/codex).](images/composable-data-stack-implementation-2.png){#fig-cds-examples}
 
@@
 MomCare was launched in Kenya [@huisman2022digital;@sanctis2022maintaining] and Tanzania [@shija2021access;@mrema2021application] in 2017 and 2019 respectively, with the objective to create transparency on the journeys of pregnant mothers.
-The programme is built on three pillars: journey tracking, quality support and a mobile wallet (@fig-momcare).
+The programme is built on three pillars: journey tracking, quality support and a mobile wallet.
 MomCare distinguishes two user groups: mothers are supported during their pregnancy through reminders and surveys, using SMS as the digital mode of engagement.
@@
 
-![Overview of the Momcare programme](images/momcare.png){width=auto #fig-momcare}
+In its original form, the MomCare programme uses predominantly closed digital platforms.
+In Kenya, M-TIBA is the primary digital platform, on top of which a relatively lightweight custom app has been built as the engagement layer for the health workers [@huisman2022digital].
+M-TIBA provides data access through its data warehouse platform for the MomCare programme, however, this is not a standardized, general purpose API.
+In the case of Tanzania, a stand-alone custom app is used which does not provide an interface of any kind for interacting with the platform [@mrema2021application].
+Given these constraints, to date the MomCare programme uses a custom-built data warehouse environment as its main data platform, on which data extractions, transformations and analysis are performed to generate the operational reports.
+Feedback reports for the health workers, in the form of operational dashboards, are made accessible through the app.
+Similar reports are provided to the back-office for the periodic reimbursement to the clinics.
+
+In the periode 2022-2023, a new iteration of MomCare ...One of the practical objectives of the DataCare project was to assess the complexity and effort required to implement the data transformation scripts, and reproduce the existing analytic reports.
+The data was transformed into 10 FHIR v4 resources as listed in table 1 with the number of records per resource type.
+The conceptual data model of the existing MomCare app could readily be transformed into the FHIR standard using SQL and validated with a Python library [@islam2023fhir].
+
+The largest challenge during the transformation process pertained to the absence of unique business identifiers for patients and healthcare organizations.
+For patients, either the mobile phone number or the healthcare insurance number was taken, depending on availability.
+A combination of name, address and latitude/longitude coordinates were used to uniquely identify organizations and locations, as Tanzania does not have a system in place for this purpose.
+
+The transformed and validated data is uploaded into the FHIR server on a daily basis using an automated cloud function.
+Analysis of bulk data was done by directly reading the standard newline delimited JSON into the Python pandas data analysis library.
+Cross checking the output with queries on the original data confirmed that the whole data pipeline produced consistent results.
+For example, the report of the antenatal coverage metric (number of pregnancies with four or more visits) could be reproduced per patient journey and aggregated (per year, per organization etc.) as required for the MomCare reports.
 
@@
 | FHIR |Fast Healthcare Interoperability Resources |
+| FL | Federated learning |
 | HDP	|	Health data platform, explicitly differentiated from health digital platform |
 | HIE | Health Information Exchange |
-| FL | Federated learning |
+| IR | Intermediate Representation |
 | LMIC | Low- and middle income countries |

### 2024-03-26 index.tex
wip

@@
 This paper is motivated by the conflation of a number of developments
-relevant to the design and implementation of solidarity-based HDPs in
-LMICs. First, the OpenHIE framework \citep{openhie} has been adopted by
-many sub-Saharan African countries \citep{mamuye2022health} as the
+relevant to the design and implementation of open, solidarity-based HDPs
+in LMICs. First, the OpenHIE framework \citep{openhie} has been adopted
+by many sub-Saharan African countries \citep{mamuye2022health} as the
 architectural blueprint for implementing nation-wide health information
@@
 typically have a zonal architecture that follow the
-Extract-Load-Transform pattern (ELT) \citep{hai2023data} where data is
-ingested from the source systems in bulk (E), delivered to storage with
-aligned schemas (L) and transformed into a format ready for analysis
-(T). The discerning characteristic of the lakehouse architecture is its
-foundation on low-cost and directly-accessible storage that also
-provides traditional analytical DBMS management and performance features
-such as ACID transactions, data versioning, auditing, indexing, caching,
-and query optimization \citep{armbrust2021lakehouse}. Lakehouses thus
-combine the key benefits of data lakes and data warehouses: low-cost
-storage in an open format accessible by a variety of systems from the
-former, and powerful management and optimization features from the
-latter.
+Extract-Load-Transform pattern (ELT) where data is ingested from the
+source systems in bulk (E), delivered to storage with aligned schemas
+(L) and transformed into a format ready for analysis (T)
+\citep{hai2023data}. The discerning characteristic of the lakehouse
+architecture is its foundation on low-cost and directly-accessible
+storage that also provides traditional analytical DBMS management and
+performance features such as ACID transactions, data versioning,
+auditing, indexing, caching, and query optimization
+\citep{armbrust2021lakehouse}. Lakehouses thus combine the key benefits
+of data lakes and data warehouses: low-cost storage in an open format
+accessible by a variety of systems from the former, and powerful
+management and optimization features from the latter.
 
@@
 (layer 1) from the execution engine by introducing standards including
-Substrait for Intermediate Representation (standard A) and Apache Arrow
-(standard B and C). The composable data stack can be implemented with
-current open source technologies (Figure~\ref{fig-cds-examples}). As an
-example, the Ibis user interface is currently sufficiently mature to
-offer a standardized dataframe interface to 19 different execution
-engines.
+Substrait for Intermediate Representation (standard A, IR) and Apache
+Arrow for data connectivity and data memory layout (standards B and C,
+respectively). The composable data stack can be implemented with current
+open source technologies (Figure~\ref{fig-cds-examples}). As an example,
+the Ibis user interface is currently sufficiently mature to offer a
+standardized dataframe interface to 19 different execution engines.
 
@@
 
-\caption{\label{fig-composable-data-stack}Composable data stack}
+\caption{\label{fig-composable-data-stack}The composable data stack.
+Image source: \url{https://voltrondata.com/codex}.}
 
@@
 \caption{\label{fig-cds-examples}Examples of implementations of the
-composable data stack}
+composable data stack. Image source:
+\url{https://voltrondata.com/codex}.}
 
@@
 of pregnant mothers. The programme is built on three pillars: journey
-tracking, quality support and a mobile wallet
-(Figure~\ref{fig-momcare}). MomCare distinguishes two user groups:
-mothers are supported during their pregnancy through reminders and
-surveys, using SMS as the digital mode of engagement. Health workers are
-equipped with an Android-based application, in which visits, care
-activities and clinical observations are recorded. Reimbursements of the
-maternal clinic are based on the data captured with SMS and the app,
-thereby creating a conditional payment scheme, where providers are
-partially reimbursed up-front for a fixed bundle of activities,
-supplemented by bonus payments based on a predefined set of care
-activities.
+tracking, quality support and a mobile wallet. MomCare distinguishes two
+user groups: mothers are supported during their pregnancy through
+reminders and surveys, using SMS as the digital mode of engagement.
+Health workers are equipped with an Android-based application, in which
+visits, care activities and clinical observations are recorded.
+Reimbursements of the maternal clinic are based on the data captured
+with SMS and the app, thereby creating a conditional payment scheme,
+where providers are partially reimbursed up-front for a fixed bundle of
+activities, supplemented by bonus payments based on a predefined set of
+care activities.
 
-\begin{figure}
+In its original form, the MomCare programme uses predominantly closed
+digital platforms. In Kenya, M-TIBA is the primary digital platform, on
+top of which a relatively lightweight custom app has been built as the
+engagement layer for the health workers \citep{huisman2022digital}.
+M-TIBA provides data access through its data warehouse platform for the
+MomCare programme, however, this is not a standardized, general purpose
+API. In the case of Tanzania, a stand-alone custom app is used which
+does not provide an interface of any kind for interacting with the
+platform \citep{mrema2021application}. Given these constraints, to date
+the MomCare programme uses a custom-built data warehouse environment as
+its main data platform, on which data extractions, transformations and
+analysis are performed to generate the operational reports. Feedback
+reports for the health workers, in the form of operational dashboards,
+are made accessible through the app. Similar reports are provided to the
+back-office for the periodic reimbursement to the clinics.
 
-\centering{
+In the periode 2022-2023, a new iteration of MomCare \ldots One of the
+practical objectives of the DataCare project was to assess the
+complexity and effort required to implement the data transformation
+scripts, and reproduce the existing analytic reports. The data was
+transformed into 10 FHIR v4 resources as listed in table 1 with the
+number of records per resource type. The conceptual data model of the
+existing MomCare app could readily be transformed into the FHIR standard
+using SQL and validated with a Python library \citep{islam2023fhir}.
 
-\includegraphics{images/momcare.png}
+The largest challenge during the transformation process pertained to the
+absence of unique business identifiers for patients and healthcare
+organizations. For patients, either the mobile phone number or the
+healthcare insurance number was taken, depending on availability. A
+combination of name, address and latitude/longitude coordinates were
+used to uniquely identify organizations and locations, as Tanzania does
+not have a system in place for this purpose.
 
-}
-
-\caption{\label{fig-momcare}Overview of the Momcare programme}
-
-\end{figure}%
+The transformed and validated data is uploaded into the FHIR server on a
+daily basis using an automated cloud function. Analysis of bulk data was
+done by directly reading the standard newline delimited JSON into the
+Python pandas data analysis library. Cross checking the output with
+queries on the original data confirmed that the whole data pipeline
+produced consistent results. For example, the report of the antenatal
+coverage metric (number of pregnancies with four or more visits) could
+be reproduced per patient journey and aggregated (per year, per
+organization etc.) as required for the MomCare reports.
 
@@
 FHIR & Fast Healthcare Interoperability Resources \\
+FL & Federated learning \\
 HDP & Health data platform, explicitly differentiated from health
@@
 HIE & Health Information Exchange \\
-FL & Federated learning \\
+IR & Intermediate Representation \\
 LMIC & Low- and middle income countries \\

### 2024-03-27 index.qmd
Add new diagram and layers

@@
 Instead of access to information, the control of personal data has emerged in the age of platforms as the critical contention.” [@keller2021paradox].
-These conflicts are particularly salient in the healthcare domain, as is evidenced by polemic surrounding data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond].
+These conflicts are particularly salient in the healthcare domain, as is evidenced by the polemic surrounding data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond].
 Openness is particularly relevant if we are to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].
@@
 If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for data sharing and reuse.
-Distinguishing four types of data sharing (@tbl-types-data-sharing), the purpose of this paper is to investigate how new standards and technologies that can establish openness of health data platforms can be integrated into the OpenHIE architecture framework.
+Distinguishing four types of data sharing (@tbl-types-data-sharing), the purpose of this paper is to investigate how new standards and technologies that can establish openness of HDPs can be integrated into the OpenHIE architecture framework.
 The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.
@@
 Subsequently, we employ a formative, naturalistic evaluation to assess the technical risk and efficacy of the design [@venable2016feds].
-Given that it is prohibitively expensive to evaluate with real users and real systems in the real setting, we aim to minimize technological risks and maximize the efficacy of the design by considering three real-world examples of health data platoform in LMICs, namely i) the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)); ii) the ONA Canopy platform [https://ona.io/home/services/canopy/](https://ona.io/home/services/canopy/); and iii) the work conducted at PharmAccess Foundation as part of the MomCare programme ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org)).
+Given that it is prohibitively expensive to evaluate the design a real-world setting, we aim to minimize technological risks and maximize the efficacy of the design by considering three examples of health data platforms in LMICs, namely i) the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)); ii) the ONA Canopy platform [https://ona.io/home/services/canopy/](https://ona.io/home/services/canopy/); and iii) the work conducted at PharmAccess Foundation as part of the MomCare programme ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org)).
 As part of our design research, we have taken a narrative approach in surveying existing scientific studies on health data platforms, focusing on the seminal reports and subsequently searching forward citations.
@@
 This paper should not be considered as a proper systematic review.
+
+The main contributions of this paper is that:
+- It provides a framework for the components of the Data & Analysis Services that builds on current best practices from the data engineering community into the OpenHIE framework
+- Provides details on type 1 and 2 data sharing, comparing the different design options
 
@@
 The premise of separating the user interface from the execution engine is directly related to the key objective of the SQL-on-FHIR project ([https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/](https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/)), namely to make large-scale analysis of FHIR data accessible to a larger audience, portable between systems and to make FHIR data work well with the best available analytic tools, regardless of the technology stack.
-However, as FHIR is represented as a graph of resources, with detailed semantics defined for references between resources, data types, terminology, extensions, and many other aspects of the specification.
-To use FHIR effectively analysts require a thorough understanding of the specification.
+However, to use FHIR effectively analysts require a thorough understanding of the specification as FHIR is represented as a graph of resources, with detailed semantics defined for references between resources, data types, terminology, extensions, and many other aspects of the specification.
 Most analytic and machine learning use cases require the preparation of FHIR data using transformations and tabular projections from its original form.
@@
 
-We propose to extend the OpenHIE architecture with a "Data and Analytics Services" domain with 5 services by synthesizing the architecture of Hai et al.
-[@hai2023data] and the composable data stack [@pedreira2023composable] (@fig-ohie, @tbl-data-and-analysis-services).
+We propose to extend the OpenHIE architecture with a "Data and Analytics Services" domain with different service layers by synthesizing the current best practices of the a lakehouse architecture of [@hai2023data;@harby2022data;@harby2024data] and the composable data stack [@pedreira2023composable] (@fig-ohie, @tbl-data-and-analysis-services).
+These 5 services are considered the core of the FHIR data platform, and although it practice often a downstream dashboarding or visualization component is used, that component is not considered part of this architecture as such.
+NB: distinction between exploration and dashboarding is not absolute.
 
-![Proposed extension of the OpenHIE architecture that includes "Data and Analytics Services" as an additional service domain.](./images/hie.png){#fig-ohie}
+![Proposed extension of the OpenHIE architecture that includes "Data and Analytics Services" as an additional service domain.](./images/openhie-extended-architecture.png){#fig-ohie}
 
@@
 +===================+=======================================================================================+
-| Ingestion         | - Metadata extraction                                                                 |
-|                   | - Metadata modeling                                                                   |
+| Ingestion         | - Bulk                                                                                |
+|                   | - Streaming                                                                           |
 +-------------------+---------------------------------------------------------------------------------------+
 | Storage           | - File-based blob storage                                                             |
-+-------------------+---------------------------------------------------------------------------------------+
-| Query &           | - SQL-on-FHIR Runner                                                                  |
-| Processing        | - Execution engine on tabular data as defined in composable data stack                |  
+|                   | - Database optimized for online analytical processing (OLAP)
 +-------------------+---------------------------------------------------------------------------------------+
 | Maintenance       | - SQL-on-FHIR View defintions                                                         |
-|                   | - Maintenance-related functions as defined by Hai et al.                              |
+|                   | - Catalog and other maintenance-related functions as defined by Hai et al.            |
 +-------------------+---------------------------------------------------------------------------------------+
-| Exploration       | - Interactive computing environment [@granger2021jupyter]                             |
-|                   | - Capability to participate as a node in a federative learning network, including MPC |
+| Processing        | - SQL-on-FHIR Runner                                                                  |
+| & API             | - Execution engine on tabular data as defined in composable data stack                |
+|                   | - Capability to participate as a node in federated learning / MPC network             |
+|                   | - Read-only access to storage                                                         |  
++-------------------+---------------------------------------------------------------------------------------+
+| Data              | - SQL interactive development environment (IDE)                                       |
+| Consumer          | - Interactive notebook computing environment [@granger2021jupyter]                    |
+|                   | - BI reporting, dashboarding and data visualization                                   |  
 +-------------------+---------------------------------------------------------------------------------------+
@@
 
+- File-based:
   - from ndjson to parquet
@@
   - separation of storage from compute not only for benefits of lower TCO, but also be ready for federated learning and MPC in future
+
+- OLAP DBMS
+  - Often columnar, like Clickhouse and BigQuery
+  -
 
@@
 
-##### Exploration
+##### Data consumers
+
+Strictly speaking not part of the architecture,
 
@@
 
+Allows for the Data Warehouse data to be visualised via a BI tool (Apache Superset)
+
 ## Evaluation
@@
 
+- One of the most used IOL layers in LMICs. Leading development of InstantHIE library to configure OpenHIE infrastructures
+
+[https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing](https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing)
+
+- Accept FHIR bundles submitted securely through an IOL (OpenHIM)
+- Stores Clinical FHIR data to a FHIR store (HAPI FHIR)
+- Stores Patient Demographic data to an MPI (JeMPI)
+- Pushes FHIR resources to Kafka for the reporting pipeline (and other systems) to use
+- Pulls FHIR data out of Kafka and maps it to flattened tables in the Data Warehouse (Clickhouse)
+
 - View generation in different packages, not portable
   - Logstash for bulk
-  - Kafka for streaming
+  - Kafka for streaming, uses JSON for mappings which looks similar to SQL-on-FHIR
 
@@
 
-### FHIR and FAIR
+### The need to a semantic layer?
 
+- FHIR and FAIR
   - How does FHIR relate to approaches taken by the FAIR community, which tend to take more an approach of using knowledge graphs.
@@
   - FAIR principles vs FHIR graph: is FHIR a FAIR Data Object
+
+- Since we use FHIR, we don't need a semantic layer because that is already provided
+-
 
@@
 - Cerbos, attribute based on lineage or anonymized tables
+- Catalogs solve this: Tabular.io, Google BigLake. What is open source option?
 

### 2024-03-27 index.tex
Add new diagram and layers

@@
 \citep{keller2021paradox}. These conflicts are particularly salient in
-the healthcare domain, as is evidenced by polemic surrounding data
+the healthcare domain, as is evidenced by the polemic surrounding data
 spaces \citep{otto2022designing} and data solidarity
@@
 investigate how new standards and technologies that can establish
-openness of health data platforms can be integrated into the OpenHIE
-architecture framework. The lack of detailed specifications and
-consensus of this addition to OpenHIE currently stands in the way of
-development projects that aim to establish HDPs in LMICs. In addition,
-we explicitly address the requirement of downward scalability as we aim
-to implement health data platforms in resource constrained setting of
-LMICs.
+openness of HDPs can be integrated into the OpenHIE architecture
+framework. The lack of detailed specifications and consensus of this
+addition to OpenHIE currently stands in the way of development projects
+that aim to establish HDPs in LMICs. In addition, we explicitly address
+the requirement of downward scalability as we aim to implement health
+data platforms in resource constrained setting of LMICs.
 
@@
 \citep{venable2016feds}. Given that it is prohibitively expensive to
-evaluate with real users and real systems in the real setting, we aim to
-minimize technological risks and maximize the efficacy of the design by
-considering three real-world examples of health data platoform in LMICs,
-namely i) the OpenHIM platform
-(\url{https://jembi.gitbook.io/openhim-platform/}); ii) the ONA Canopy
-platform \url{https://ona.io/home/services/canopy/}; and iii) the work
-conducted at PharmAccess Foundation as part of the MomCare programme
-(\url{https://health-data-commons.pharmaccess.org}). As part of our
-design research, we have taken a narrative approach in surveying
-existing scientific studies on health data platforms, focusing on the
-seminal reports and subsequently searching forward citations. In
+evaluate the design a real-world setting, we aim to minimize
+technological risks and maximize the efficacy of the design by
+considering three examples of health data platforms in LMICs, namely i)
+the OpenHIM platform (\url{https://jembi.gitbook.io/openhim-platform/});
+ii) the ONA Canopy platform \url{https://ona.io/home/services/canopy/};
+and iii) the work conducted at PharmAccess Foundation as part of the
+MomCare programme (\url{https://health-data-commons.pharmaccess.org}).
+As part of our design research, we have taken a narrative approach in
+surveying existing scientific studies on health data platforms, focusing
+on the seminal reports and subsequently searching forward citations. In
 addition, we have searched the open source repositories (most notably
@@
 This paper should not be considered as a proper systematic review.
+
+The main contributions of this paper is that: - It provides a framework
+for the components of the Data \& Analysis Services that builds on
+current best practices from the data engineering community into the
+OpenHIE framework - Provides details on type 1 and 2 data sharing,
+comparing the different design options
 
@@
 available analytic tools, regardless of the technology stack. However,
-as FHIR is represented as a graph of resources, with detailed semantics
-defined for references between resources, data types, terminology,
-extensions, and many other aspects of the specification. To use FHIR
-effectively analysts require a thorough understanding of the
-specification. Most analytic and machine learning use cases require the
-preparation of FHIR data using transformations and tabular projections
-from its original form. The task of authoring these transformations and
+to use FHIR effectively analysts require a thorough understanding of the
+specification as FHIR is represented as a graph of resources, with
+detailed semantics defined for references between resources, data types,
+terminology, extensions, and many other aspects of the specification.
+Most analytic and machine learning use cases require the preparation of
+FHIR data using transformations and tabular projections from its
+original form. The task of authoring these transformations and
 projections is not trivial and there is currently no standard mechanisms
@@
 We propose to extend the OpenHIE architecture with a ``Data and
-Analytics Services'' domain with 5 services by synthesizing the
-architecture of Hai et al. \citep{hai2023data} and the composable data
-stack \citep{pedreira2023composable} (Figure~\ref{fig-ohie},
-Table~\ref{tbl-data-and-analysis-services}).
+Analytics Services'' domain with different service layers by
+synthesizing the current best practices of the a lakehouse architecture
+of \citep{hai2023data, harby2022data, harby2024data} and the composable
+data stack \citep{pedreira2023composable} (Figure~\ref{fig-ohie},
+Table~\ref{tbl-data-and-analysis-services}). These 5 services are
+considered the core of the FHIR data platform, and although it practice
+often a downstream dashboarding or visualization component is used, that
+component is not considered part of this architecture as such. NB:
+distinction between exploration and dashboarding is not absolute.
 
@@
 
-\includegraphics{./images/hie.png}
+\includegraphics{./images/openhie-extended-architecture.png}
 
@@
 \item
-  Metadata extraction
-\item
-  Metadata modeling
+  Bulk
+\item
+  Streaming
 \end{itemize}
@@
   File-based blob storage
-\end{itemize}
-\end{minipage} \\
-Query \& Processing & \begin{minipage}[t]{\linewidth}\raggedright
-\begin{itemize}
-\tightlist
-\item
-  SQL-on-FHIR Runner
-\item
-  Execution engine on tabular data as defined in composable data stack
+\item
+  Database optimized for online analytical processing (OLAP)
 \end{itemize}
@@
 \item
-  Maintenance-related functions as defined by Hai et al.
+  Catalog and other maintenance-related functions as defined by Hai et
+  al.
 \end{itemize}
 \end{minipage} \\
-Exploration & \begin{minipage}[t]{\linewidth}\raggedright
-\begin{itemize}
-\tightlist
-\item
-  Interactive computing environment \citep{granger2021jupyter}
-\item
-  Capability to participate as a node in a federative learning network,
-  including MPC
+Processing \& API & \begin{minipage}[t]{\linewidth}\raggedright
+\begin{itemize}
+\tightlist
+\item
+  SQL-on-FHIR Runner
+\item
+  Execution engine on tabular data as defined in composable data stack
+\item
+  Capability to participate as a node in federated learning / MPC
+  network
+\item
+  Read-only access to storage
+\end{itemize}
+\end{minipage} \\
+Data Consumer & \begin{minipage}[t]{\linewidth}\raggedright
+\begin{itemize}
+\tightlist
+\item
+  SQL interactive development environment (IDE)
+\item
+  Interactive notebook computing environment \citep{granger2021jupyter}
+\item
+  BI reporting, dashboarding and data visualization
 \end{itemize}
@@
 \item
+  File-based:
+
+  \begin{itemize}
+  \tightlist
+  \item
     from ndjson to parquet
@@
   \item
-  separation of storage from compute not only for benefits of lower TCO,
-  but also be ready for federated learning and MPC in future
+    separation of storage from compute not only for benefits of lower
+    TCO, but also be ready for federated learning and MPC in future
+  \end{itemize}
+\item
+  OLAP DBMS
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Often columnar, like Clickhouse and BigQuery
+  \item
+  \end{itemize}
 \end{itemize}
@@
 
-\paragraph{Exploration}\label{exploration}
+\paragraph{Data consumers}\label{data-consumers}
+
+Strictly speaking not part of the architecture,
 
@@
 
+Allows for the Data Warehouse data to be visualised via a BI tool
+(Apache Superset)
+
 \section{Evaluation}\label{evaluation}
@@
 \item
+  One of the most used IOL layers in LMICs. Leading development of
+  InstantHIE library to configure OpenHIE infrastructures
+\end{itemize}
+
+\url{https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing}
+
+\begin{itemize}
+\item
+  Accept FHIR bundles submitted securely through an IOL (OpenHIM)
+\item
+  Stores Clinical FHIR data to a FHIR store (HAPI FHIR)
+\item
+  Stores Patient Demographic data to an MPI (JeMPI)
+\item
+  Pushes FHIR resources to Kafka for the reporting pipeline (and other
+  systems) to use
+\item
+  Pulls FHIR data out of Kafka and maps it to flattened tables in the
+  Data Warehouse (Clickhouse)
+\item
   View generation in different packages, not portable
@@
   \item
-    Kafka for streaming
+    Kafka for streaming, uses JSON for mappings which looks similar to
+    SQL-on-FHIR
   \end{itemize}
@@
 
-\subsection{FHIR and FAIR}\label{fhir-and-fair}
-
-\begin{itemize}
-\tightlist
-\item
-  How does FHIR relate to approaches taken by the FAIR community, which
-  tend to take more an approach of using knowledge graphs. For example,
-  VODAN Africa \citep{gebreslassie2023fhir4fair, purnamajati2022data}.
+\subsection{The need to a semantic
+layer?}\label{the-need-to-a-semantic-layer}
+
+\begin{itemize}
+\tightlist
+\item
+  FHIR and FAIR
+
+  \begin{itemize}
+  \tightlist
+  \item
+    How does FHIR relate to approaches taken by the FAIR community,
+    which tend to take more an approach of using knowledge graphs. For
+    example, VODAN Africa
+    \citep{gebreslassie2023fhir4fair, purnamajati2022data}.
   \item
     FAIR principles vs FHIR graph: is FHIR a FAIR Data Object
+  \end{itemize}
+\item
+  Since we use FHIR, we don't need a semantic layer because that is
+  already provided
+\item
 \end{itemize}
@@
   Cerbos, attribute based on lineage or anonymized tables
+\item
+  Catalogs solve this: Tabular.io, Google BigLake. What is open source
+  option?
 \end{itemize}

### 2024-04-02 index.qmd
Add OpenHIM section

@@
 
-The main contributions of this paper is that:
-- It provides a framework for the components of the Data & Analysis Services that builds on current best practices from the data engineering community into the OpenHIE framework
-- Provides details on type 1 and 2 data sharing, comparing the different design options
+The main contributions of this paper are i) description of a framework for the components of the Data & Analysis Services that builds on current best practices from the data engineering community into the OpenHIE framework; and ii) evaluation of different implementations and design options for type 1 and 2 data sharing within an extended OpenHIE architecture.
+As such, it aims to inform future developments and implementation of open digital health platforms in LMICs.
 
@@
 Various studies have investigated the merits of FHIR and its performance vis-a-vis other healthcare standards.
-Comparisons between OpenEHR, ISO 13606, OMOP and FHIR have been made [@ayaz2023transforming;@mullie2023coda;@rinaldi2021openehr;@cremonesi2023need;@sinaci2023data].
+Comparisons between OpenEHR, ISO 13606, OMOP and FHIR have been made [@ayaz2023transforming;@mullie2023coda;@rinaldi2021openehr;@cremonesi2023need;@sinaci2023data;@tsafnat2024converge].
 A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR concluded that i) these three standards are functionally and technically compatible, and therefore can be used side by side; and that ii) each of these standards have their strengths and limitations that correlate with their intended use as summarized in the @tbl-comparison.
@@
 We propose to extend the OpenHIE architecture with a "Data and Analytics Services" domain with different service layers by synthesizing the current best practices of the a lakehouse architecture of [@hai2023data;@harby2022data;@harby2024data] and the composable data stack [@pedreira2023composable] (@fig-ohie, @tbl-data-and-analysis-services).
-These 5 services are considered the core of the FHIR data platform, and although it practice often a downstream dashboarding or visualization component is used, that component is not considered part of this architecture as such.
-NB: distinction between exploration and dashboarding is not absolute.
+These 5 services are considered the core of the FHIR data platform, and although in practice often a downstream dashboarding or visualization component is used, that component is not the main focus of our analysis.
+Rather, we aim to elucidate and conceptualize the core "FHIR Data Lakehouse".
 
@@
 
-- SQL-on-FHIR Views provide new standard to support mADX aggregate reporting
+- SQL-on-FHIR Views provide new standard to support mADX aggregate reporting !! We need to stress this, because this is an existing OpenHIE workflow
 - Maintenance-related functions remain the same
@@
 
-Strictly speaking not part of the architecture,
+- Many tools, often focus on creating information dashboards and visualizations
+- Compatibility with processing & API: which query languages and interfaces are supported. Some dialect of SQL, some dialect of NoSQL, dataframe API, all of the above?
 
@@
 
-![ Overview of current open source implementations of components included in the OpenHIE specification that are FHIR-compatible.
+![Overview of current open source implementations of components included in the OpenHIE specification that are FHIR-compatible.
 The category Analytics Services is not a part of the original OpenHIE and is discussed in the paper.
@@
 A systematic review of such digital public goods is beyond the scope of this document.](images/digital-public-goods.png){#tbl-digital-public-goods}
-
-Allows for the Data Warehouse data to be visualised via a BI tool (Apache Superset)
 
@@
 
-- One of the most used IOL layers in LMICs. Leading development of InstantHIE library to configure OpenHIE infrastructures
+To evaluate the extended OpenHIE architecture described above, we first consider the OpenHIM Platform.
+The Open Health Information Mediator (OpenHIM, [http://openhim.org/](OpenHIMhttp://openhim.org/))) component is the reference implementation of the Interoperability Layer (IOL) as defined in the OpenHIE specification.
+The most current version (8.4.2 at the time of writing) provides all the core functions including central point of access for the services of the HIE; routing functions; central logging for auditing and debugging purposes; and orchestration/mediation mechanisms to co-ordinate requests.
+This reference implementation of the IOL has recently been extended as a reference implementation of the so-called 'Instant OpenHIE' concept, that aims to address the primary needs of (i) allowing implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) having a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@digitalsquare2020instantopenhie].
+The Instant HIE project allows OpenHIE components to be packages up, deployed, operated and scaled via a simple Command Line Interface (CLI, version 2 released in 2023 [https://github.com/openhie/instant-v2](https://github.com/openhie/instant-v2)).
+The OpenHIM Platform ([https://jembi.gitbook.io/openhim-platform](https://jembi.gitbook.io/openhim-platform)) provides a set of bundled packages (called recipes) that can be deployed using the Instant OpenHIE deployment tool.
+In the following we will evaluate the recipe for "a central data repository with a data warehouse" that provides "A FHIR-based Shared Health record linked to a Master Patient Index (MPI) for linking and mathing patient demographics and a default reporting pipeline to transform and visualise FHIR data" ([https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing](https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing)).
 
-[https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing](https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing)
+![Overview of the default data stack of the OpenHIM Platform.
+The default stack (top, darkblue) consists of Kafka, Clickhouse and Superset.
+An alternative solution based on the ELK stack is also supported (bottom, lightblue), consisting of Elasticsearch, Logstash and Kibana.](images/openhim-platform.png){#fig-openhim-platform}
 
-- Accept FHIR bundles submitted securely through an IOL (OpenHIM)
-- Stores Clinical FHIR data to a FHIR store (HAPI FHIR)
-- Stores Patient Demographic data to an MPI (JeMPI)
-- Pushes FHIR resources to Kafka for the reporting pipeline (and other systems) to use
-- Pulls FHIR data out of Kafka and maps it to flattened tables in the Data Warehouse (Clickhouse)
+@fig-openhim-platform shows a schematic overview of two data stacks that are supported in the OpenHIM platform.
+The Shared Health Record (SHR, implemented with HAPI FHIR server) and the Client Registry (CR, implemented with JeMPI server) are the sources that store clinical FHIR data and patient demographic data, respectively.
+The default data stack is based on streaming ingestion using Kafka into a Clickhouse database.
+As part of the ingestion, incoming FHIR bundles that contain multiple FHIR resources are unbundled in separate topics.
+Subsequently, each FHIR resource topic is flatted using mappers Unbundling and flattening of FHIR resources is done in Kafka using FHIRPath.
+Superset is used as the tool for consuming the data to create dashboard visualizations.
 
-- View generation in different packages, not portable
-  - Logstash for bulk
-  - Kafka for streaming, uses JSON for mappings which looks similar to SQL-on-FHIR
+The OpenHIM platform also support data analytics based on the ELK stack, where data is ingested in bulk using Logstash, stored in Elasticsearch and made available for consumption in Kibana.
+Also here, the incoming FHIR bundles are unbundled in Logstash into separate FHIR resources.
+However, given that Elasticsearch is a document-based search engine, the FHIR resources are stored as-is with no flattening.
+Exploring and analysing the data requires writing queries in Elasticsearch Query Language (ES|QL), either through the query interface of Elasticsearch or using Kibana.
 
--
+Evaluating these two data stacks, we see the following:
+
+- Pattern of flattening FHIR resources with FHIRPath expressions is very close to the idea of SQL-on-FHIR.
+Although it doesn't adhere to this new standard in the strict sense, the philosophy of generating tabular views is the same
+- When using the ELK stack, flattening is done at the end. Implementations of FHIRPath support Elasticsearch as an execution engine, also here 
+- Main limitations: both Clickhouse en Elasticsearch don't follow decomposition of storage, compute and UI. Therefore, downward scaleability is limited.
 
@@
 |:---|:---|
+| CLI | Command-line Interface |
+| CR | Client Registry |
+| ELK | Elasticsearch, Logstach and Kibana stack |
 | ELT | Extract, Load and Transform |
@@
 | LMIC | Low- and middle income countries |
+| MPC | Multiparty Computation |
 | PET | Privacy-enhancing technologies |
+| SHR | Shared Health Record |

### 2024-04-02 index.tex
Add OpenHIM section

@@
 
-The main contributions of this paper is that: - It provides a framework
+The main contributions of this paper are i) description of a framework
 for the components of the Data \& Analysis Services that builds on
 current best practices from the data engineering community into the
-OpenHIE framework - Provides details on type 1 and 2 data sharing,
-comparing the different design options
+OpenHIE framework; and ii) evaluation of different implementations and
+design options for type 1 and 2 data sharing within an extended OpenHIE
+architecture. As such, it aims to inform future developments and
+implementation of open digital health platforms in LMICs.
 
@@
 13606, OMOP and FHIR have been made
-\citep{ayaz2023transforming, mullie2023coda, rinaldi2021openehr, cremonesi2023need, sinaci2023data}.
+\citep{ayaz2023transforming, mullie2023coda, rinaldi2021openehr, cremonesi2023need, sinaci2023data, tsafnat2024converge}.
 A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR
@@
 Table~\ref{tbl-data-and-analysis-services}). These 5 services are
-considered the core of the FHIR data platform, and although it practice
+considered the core of the FHIR data platform, and although in practice
 often a downstream dashboarding or visualization component is used, that
-component is not considered part of this architecture as such. NB:
-distinction between exploration and dashboarding is not absolute.
+component is not the main focus of our analysis. Rather, we aim to
+elucidate and conceptualize the core ``FHIR Data Lakehouse''.
 
@@
   SQL-on-FHIR Views provide new standard to support mADX aggregate
-  reporting
+  reporting !! We need to stress this, because this is an existing
+  OpenHIE workflow
 \item
@@
 
-Strictly speaking not part of the architecture,
+\begin{itemize}
+\tightlist
+\item
+  Many tools, often focus on creating information dashboards and
+  visualizations
+\item
+  Compatibility with processing \& API: which query languages and
+  interfaces are supported. Some dialect of SQL, some dialect of NoSQL,
+  dataframe API, all of the above?
+\end{itemize}
 
@@
 
-Allows for the Data Warehouse data to be visualised via a BI tool
-(Apache Superset)
-
 \section{Evaluation}\label{evaluation}
@@
 
-\begin{itemize}
-\tightlist
-\item
-  One of the most used IOL layers in LMICs. Leading development of
-  InstantHIE library to configure OpenHIE infrastructures
-\end{itemize}
-
-\url{https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing}
-
-\begin{itemize}
-\item
-  Accept FHIR bundles submitted securely through an IOL (OpenHIM)
-\item
-  Stores Clinical FHIR data to a FHIR store (HAPI FHIR)
-\item
-  Stores Patient Demographic data to an MPI (JeMPI)
-\item
-  Pushes FHIR resources to Kafka for the reporting pipeline (and other
-  systems) to use
-\item
-  Pulls FHIR data out of Kafka and maps it to flattened tables in the
-  Data Warehouse (Clickhouse)
-\item
-  View generation in different packages, not portable
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Logstash for bulk
-  \item
-    Kafka for streaming, uses JSON for mappings which looks similar to
-    SQL-on-FHIR
-  \end{itemize}
-\item
+To evaluate the extended OpenHIE architecture described above, we first
+consider the OpenHIM Platform. The Open Health Information Mediator
+(OpenHIM, \href{OpenHIMhttp://openhim.org/}{http://openhim.org/}))
+component is the reference implementation of the Interoperability Layer
+(IOL) as defined in the OpenHIE specification. The most current version
+(8.4.2 at the time of writing) provides all the core functions including
+central point of access for the services of the HIE; routing functions;
+central logging for auditing and debugging purposes; and
+orchestration/mediation mechanisms to co-ordinate requests. This
+reference implementation of the IOL has recently been extended as a
+reference implementation of the so-called `Instant OpenHIE' concept,
+that aims to address the primary needs of (i) allowing implementers to
+engage with a preconfigured health information exchange solution and
+running tools (based on the architecture) and test their applicability
+and functionality with a real health context problem; and (ii) having a
+packaged reference version of the OpenHIE architecture that is comprised
+of a set of reference technologies and other appropriate tools that form
+the building blocks of the health information exchange that can be
+configured and extended to support particular use cases
+\citep{digitalsquare2020instantopenhie}. The Instant HIE project allows
+OpenHIE components to be packages up, deployed, operated and scaled via
+a simple Command Line Interface (CLI, version 2 released in 2023
+\url{https://github.com/openhie/instant-v2}). The OpenHIM Platform
+(\url{https://jembi.gitbook.io/openhim-platform}) provides a set of
+bundled packages (called recipes) that can be deployed using the Instant
+OpenHIE deployment tool. In the following we will evaluate the recipe
+for ``a central data repository with a data warehouse'' that provides
+``A FHIR-based Shared Health record linked to a Master Patient Index
+(MPI) for linking and mathing patient demographics and a default
+reporting pipeline to transform and visualise FHIR data''
+(\url{https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing}).
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{images/openhim-platform.png}
+
+}
+
+\caption{\label{fig-openhim-platform}Overview of the default data stack
+of the OpenHIM Platform. The default stack (top, darkblue) consists of
+Kafka, Clickhouse and Superset. An alternative solution based on the ELK
+stack is also supported (bottom, lightblue), consisting of
+Elasticsearch, Logstash and Kibana.}
+
+\end{figure}%
+
+Figure~\ref{fig-openhim-platform} shows a schematic overview of two data
+stacks that are supported in the OpenHIM platform. The Shared Health
+Record (SHR, implemented with HAPI FHIR server) and the Client Registry
+(CR, implemented with JeMPI server) are the sources that store clinical
+FHIR data and patient demographic data, respectively. The default data
+stack is based on streaming ingestion using Kafka into a Clickhouse
+database. As part of the ingestion, incoming FHIR bundles that contain
+multiple FHIR resources are unbundled in separate topics. Subsequently,
+each FHIR resource topic is flatted using mappers Unbundling and
+flattening of FHIR resources is done in Kafka using FHIRPath. Superset
+is used as the tool for consuming the data to create dashboard
+visualizations.
+
+The OpenHIM platform also support data analytics based on the ELK stack,
+where data is ingested in bulk using Logstash, stored in Elasticsearch
+and made available for consumption in Kibana. Also here, the incoming
+FHIR bundles are unbundled in Logstash into separate FHIR resources.
+However, given that Elasticsearch is a document-based search engine, the
+FHIR resources are stored as-is with no flattening. Exploring and
+analysing the data requires writing queries in Elasticsearch Query
+Language (ES\textbar QL), either through the query interface of
+Elasticsearch or using Kibana.
+
+Evaluating these two data stacks, we see the following:
+
+\begin{itemize}
+\tightlist
+\item
+  Pattern of flattening FHIR resources with FHIRPath expressions is very
+  close to the idea of SQL-on-FHIR. Although it doesn't adhere to this
+  new standard in the strict sense, the philosophy of generating tabular
+  views is the same
+\item
+  When using the ELK stack, flattening is done at the end.
+  Implementations of FHIRPath support Elasticsearch as an execution
+  engine, also here
+\item
+  Main limitations: both Clickhouse en Elasticsearch don't follow
+  decomposition of storage, compute and UI. Therefore, downward
+  scaleability is limited.
 \end{itemize}
@@
 \endlastfoot
+CLI & Command-line Interface \\
+CR & Client Registry \\
+ELK & Elasticsearch, Logstach and Kibana stack \\
 ELT & Extract, Load and Transform \\
@@
 LMIC & Low- and middle income countries \\
+MPC & Multiparty Computation \\
 PET & Privacy-enhancing technologies \\
+SHR & Shared Health Record \\
 \end{longtable}

### 2024-04-02 index.qmd
Update section on MomCare

@@
 
-### Open technologies: available digital public goods
+### Open technologies: deploying Instant OpenHIE with digital public goods
 
-Many components of the OpenHIE specification are now available as a digital public goods.
-@tbl-digital-public-goods lists components that are currently available for implementing the OpenHIE framework using open source, digital public goods that are compliant with the FHIR standard, illustrating the maturity of this ecosystem and development community.
-With the launch of the Instant OpenHIE configuration toolkit[@InstantOpenHIEv2], it has become easier to set up, explore and develop HIEs thereby reducing costs and skills required for software developers to deploy an OpenHIE architecture for quicker solution testing and as a starting point for faster production implementation and customisation.
-Several frameworks are available that offer a set of preconfigured components out of the box, such as for example:
-
-- the  “Open Smart Register Platform” (OpenSRP), that focuses on providing a mobile-first platform, including a FHIR native app designed to support the WHO Smart Guidelines
-- the OpenHIM Platform, a reference implementation of the Instant OpenHIE framework, providing an easy way to set up, manage and operate various HIE configurations
+Today, many components of the OpenHIE specification are now available as a digital public goods, as listed in @tbl-digital-public-goods.
+To further ease the development, configuration and deployment of health information exchanges, the concept of 'Instant OpenHIE' has been championed to (i) allow implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) have a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@InstantOpenHIEv2].
+Besides the core functional components of the OpenHIE architecture, the Instant OpenHIE toolkit allows packaging and integration of generic components such as Identity and Access Managment (IAM) and a reverse proxy gateway.
+In the following, we will evaluatie three of such configurations, with the aim to conceptualize and evaluate the proposed Data and Analytics Services domain of of the OpenHIE architecture.
 
@@
 The most current version (8.4.2 at the time of writing) provides all the core functions including central point of access for the services of the HIE; routing functions; central logging for auditing and debugging purposes; and orchestration/mediation mechanisms to co-ordinate requests.
-This reference implementation of the IOL has recently been extended as a reference implementation of the so-called 'Instant OpenHIE' concept, that aims to address the primary needs of (i) allowing implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) having a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@digitalsquare2020instantopenhie].
-The Instant HIE project allows OpenHIE components to be packages up, deployed, operated and scaled via a simple Command Line Interface (CLI, version 2 released in 2023 [https://github.com/openhie/instant-v2](https://github.com/openhie/instant-v2)).
-The OpenHIM Platform ([https://jembi.gitbook.io/openhim-platform](https://jembi.gitbook.io/openhim-platform)) provides a set of bundled packages (called recipes) that can be deployed using the Instant OpenHIE deployment tool.
+By extension, the OpenHIM Platform ([https://jembi.gitbook.io/openhim-platform](https://jembi.gitbook.io/openhim-platform)) is a reference implementation of a set of Instant OpenHIE configurations, refered to as 'recipes' in the documentation.
 In the following we will evaluate the recipe for "a central data repository with a data warehouse" that provides "A FHIR-based Shared Health record linked to a Master Patient Index (MPI) for linking and mathing patient demographics and a default reporting pipeline to transform and visualise FHIR data" ([https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing](https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing)).
@@
 ![Overview of the default data stack of the OpenHIM Platform.
-The default stack (top, darkblue) consists of Kafka, Clickhouse and Superset.
-An alternative solution based on the ELK stack is also supported (bottom, lightblue), consisting of Elasticsearch, Logstash and Kibana.](images/openhim-platform.png){#fig-openhim-platform}
+The default stack (top, red) consists of Kafka, Clickhouse and Superset.
+An alternative solution based on the ELK stack is also supported (bottom, orange), consisting of Elasticsearch, Logstash and Kibana.](images/openhim-platform.png){#fig-openhim-platform}
 
@@
 The default data stack is based on streaming ingestion using Kafka into a Clickhouse database.
-As part of the ingestion, incoming FHIR bundles that contain multiple FHIR resources are unbundled in separate topics.
-Subsequently, each FHIR resource topic is flatted using mappers Unbundling and flattening of FHIR resources is done in Kafka using FHIRPath.
+As part of the ingestion, incoming FHIR bundles that contain multiple FHIR resources are unbundled in separate topics using a generic Kafka utility component.
+Subsequently, each FHIR resource topic is flatted with Kafka mappers that use FHIRPath.
 Superset is used as the tool for consuming the data to create dashboard visualizations.
 
-The OpenHIM platform also support data analytics based on the ELK stack, where data is ingested in bulk using Logstash, stored in Elasticsearch and made available for consumption in Kibana.
+The OpenHIM platform also support data and analytics based on the ELK stack, where data is ingested in bulk using Logstash, stored in Elasticsearch and made available for consumption in Kibana.
 Also here, the incoming FHIR bundles are unbundled in Logstash into separate FHIR resources.
@@
 
+TO DO: add section describing ONA Canopy.
+
 ### Momcare programme
 
-MomCare was launched in Kenya [@huisman2022digital;@sanctis2022maintaining] and Tanzania [@shija2021access;@mrema2021application] in 2017 and 2019 respectively, with the objective to create transparency on the journeys of pregnant mothers.
-The programme is built on three pillars: journey tracking, quality support and a mobile wallet.
+MomCare was launched in Kenya [@huisman2022digital;@sanctis2022maintaining] and Tanzania [@shija2021access;@mrema2021application] in 2017 and 2019 respectively, with the objective to improve health outcomes for maternal and antenatal care.
 MomCare distinguishes two user groups: mothers are supported during their pregnancy through reminders and surveys, using SMS as the digital mode of engagement.
@@
 
-In its original form, the MomCare programme uses predominantly closed digital platforms.
+In its original form, the MomCare programme used closed digital platforms.
 In Kenya, M-TIBA is the primary digital platform, on top of which a relatively lightweight custom app has been built as the engagement layer for the health workers [@huisman2022digital].
@@
 In the case of Tanzania, a stand-alone custom app is used which does not provide an interface of any kind for interacting with the platform [@mrema2021application].
-Given these constraints, to date the MomCare programme uses a custom-built data warehouse environment as its main data platform, on which data extractions, transformations and analysis are performed to generate the operational reports.
+Given these constraints, the first iteration of the MomCare programme used a custom-built data warehouse environment as its main data platform, on which data extractions, transformations and analysis are performed to generate the operational reports.
 Feedback reports for the health workers, in the form of operational dashboards, are made accessible through the app.
@@
 
-In the periode 2022-2023, a new iteration of MomCare ...One of the practical objectives of the DataCare project was to assess the complexity and effort required to implement the data transformation scripts, and reproduce the existing analytic reports.
-The data was transformed into 10 FHIR v4 resources as listed in table 1 with the number of records per resource type.
-The conceptual data model of the existing MomCare app could readily be transformed into the FHIR standard using SQL and validated with a Python library [@islam2023fhir].
-
+Clearly, a more open and scaleable platform was required if MomCare was to be implemented in more regions.
+This need led to a redesign of the underlying technical infrastructure of the MomCare project.
+The objectives of this work were in fact to demonstrate a solution design that could support the first three types of data sharing listed in @tbl-types-data-sharing.
+First, to investigate the viability of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed to assess the complexity and effort required to implement the facade pattern to integrate the legacy system into the FHIR data standard.
+Using the longitudinal dataset from approximately 28 thousand patient records, FHIR transformations script were developed and deployed using the mediator function of the IOL.
+The data was transformed into 10 FHIR v4 resources and the conceptual data model of the existing MomCare app could readily be transformed into the FHIR standard using SQL and validated with a Python library [@islam2023fhir].
 The largest challenge during the transformation process pertained to the absence of unique business identifiers for patients and healthcare organizations.
@@
 
+The second objective was to reproduce existing analytic reports, using the bulk FHIR data format as input.
+Here, the focus was to standardize the logic required for producing metrics and reports.
 The transformed and validated data is uploaded into the FHIR server on a daily basis using an automated cloud function.
@@
 For example, the report of the antenatal coverage metric (number of pregnancies with four or more visits) could be reproduced per patient journey and aggregated (per year, per organization etc.) as required for the MomCare reports.
+
+TO DO: explain logic of patient-timeline table.
+Write standard transformation to go from FHIR resources to this standard table.
+On top of that the actual metrics and reporting.
+Explain serverless: we wanted to get rid of resource-heavy data visualization tools.
+This led to the idea of serverless: using duckdb-wasm and pipelines of cloud functions.
+
+The third objective was to run a technical feasibility test for federated analytics. Using the MPC platform of Roseman Labs, we managed to do aggregations in the blind ...
+TO DO: explain that we managed to reproduce the reports we generated in the clear, but then in the blind. Note, however, that in the remainder we will focus on first two types of data sharing.
+
+Based on these experiments, we arrived at the following design for the data & analytics services
+
+- Use 'serverless' file-based storage: bulk copy of data as-is in parquet
+  - Tension: how to manage change data capture
+  - Tension: how to manage access rights
+- Use SQL-on-FHIR-v2 to create tabular views.
+  - Example: patient timeline
+  - TO DO: rewrite patient timeline queries with SQL-on-FHIR-v2 and run it with Pathling
+- Use semantic modeling layer to define metrics
+  - There are many options: dbt, cube.dev
+  - Fulfills same function as ADX/mADX IHE profile in OpenHIE specification
+  - Tension: going from patient-timeline to reported metrics still isn't standardized.
+This is where Ibis/Substrait comes in.
+Substrait as IR for cross-language serialization for relational algebra.
+Can be executed on different backends.
+Write once, run on different engines.
+- Distribute and publish reports on resource-constrained devices 
+  - duckdb
+  - sveltekit
+
+TO DO: Add diagram
 
@@
 - Risk of openness: What are the novel (negative) implications of opening up data platforms? How can reflexivity in design help providers to resolve the negative implications of openness?
+
+- Answers/insights to above:
+  - Openness of standardized view on FHIR data and cross-language serialization of relational algebra makes it possible to fully standardize the workflow from start to finish
+  - Platform-to-platform: MPC
+  - Risk of openness: difficult to answer ...
 

### 2024-04-02 index.tex
Update render

@@
 
-\subsection{Open technologies: available digital public
-goods}\label{open-technologies-available-digital-public-goods}
-
-Many components of the OpenHIE specification are now available as a
-digital public goods. Table~\ref{tbl-digital-public-goods} lists
-components that are currently available for implementing the OpenHIE
-framework using open source, digital public goods that are compliant
-with the FHIR standard, illustrating the maturity of this ecosystem and
-development community. With the launch of the Instant OpenHIE
-configuration toolkit\citep{InstantOpenHIEv2}, it has become easier to
-set up, explore and develop HIEs thereby reducing costs and skills
-required for software developers to deploy an OpenHIE architecture for
-quicker solution testing and as a starting point for faster production
-implementation and customisation. Several frameworks are available that
-offer a set of preconfigured components out of the box, such as for
-example:
-
-\begin{itemize}
-\tightlist
-\item
-  the ``Open Smart Register Platform'' (OpenSRP), that focuses on
-  providing a mobile-first platform, including a FHIR native app
-  designed to support the WHO Smart Guidelines
-\item
-  the OpenHIM Platform, a reference implementation of the Instant
-  OpenHIE framework, providing an easy way to set up, manage and operate
-  various HIE configurations
-\end{itemize}
+\subsection{Open technologies: deploying Instant OpenHIE with digital
+public
+goods}\label{open-technologies-deploying-instant-openhie-with-digital-public-goods}
+
+Today, many components of the OpenHIE specification are now available as
+a digital public goods, as listed in
+Table~\ref{tbl-digital-public-goods}. To further ease the development,
+configuration and deployment of health information exchanges, the
+concept of `Instant OpenHIE' has been championed to (i) allow
+implementers to engage with a preconfigured health information exchange
+solution and running tools (based on the architecture) and test their
+applicability and functionality with a real health context problem; and
+(ii) have a packaged reference version of the OpenHIE architecture that
+is comprised of a set of reference technologies and other appropriate
+tools that form the building blocks of the health information exchange
+that can be configured and extended to support particular use cases
+\citep{InstantOpenHIEv2}. Besides the core functional components of the
+OpenHIE architecture, the Instant OpenHIE toolkit allows packaging and
+integration of generic components such as Identity and Access Managment
+(IAM) and a reverse proxy gateway. In the following, we will evaluatie
+three of such configurations, with the aim to conceptualize and evaluate
+the proposed Data and Analytics Services domain of of the OpenHIE
+architecture.
 
@@
 central logging for auditing and debugging purposes; and
-orchestration/mediation mechanisms to co-ordinate requests. This
-reference implementation of the IOL has recently been extended as a
-reference implementation of the so-called `Instant OpenHIE' concept,
-that aims to address the primary needs of (i) allowing implementers to
-engage with a preconfigured health information exchange solution and
-running tools (based on the architecture) and test their applicability
-and functionality with a real health context problem; and (ii) having a
-packaged reference version of the OpenHIE architecture that is comprised
-of a set of reference technologies and other appropriate tools that form
-the building blocks of the health information exchange that can be
-configured and extended to support particular use cases
-\citep{digitalsquare2020instantopenhie}. The Instant HIE project allows
-OpenHIE components to be packages up, deployed, operated and scaled via
-a simple Command Line Interface (CLI, version 2 released in 2023
-\url{https://github.com/openhie/instant-v2}). The OpenHIM Platform
-(\url{https://jembi.gitbook.io/openhim-platform}) provides a set of
-bundled packages (called recipes) that can be deployed using the Instant
-OpenHIE deployment tool. In the following we will evaluate the recipe
-for ``a central data repository with a data warehouse'' that provides
-``A FHIR-based Shared Health record linked to a Master Patient Index
-(MPI) for linking and mathing patient demographics and a default
+orchestration/mediation mechanisms to co-ordinate requests. By
+extension, the OpenHIM Platform
+(\url{https://jembi.gitbook.io/openhim-platform}) is a reference
+implementation of a set of Instant OpenHIE configurations, refered to as
+`recipes' in the documentation. In the following we will evaluate the
+recipe for ``a central data repository with a data warehouse'' that
+provides ``A FHIR-based Shared Health record linked to a Master Patient
+Index (MPI) for linking and mathing patient demographics and a default
 reporting pipeline to transform and visualise FHIR data''
@@
 \caption{\label{fig-openhim-platform}Overview of the default data stack
-of the OpenHIM Platform. The default stack (top, darkblue) consists of
-Kafka, Clickhouse and Superset. An alternative solution based on the ELK
-stack is also supported (bottom, lightblue), consisting of
-Elasticsearch, Logstash and Kibana.}
+of the OpenHIM Platform. The default stack (top, red) consists of Kafka,
+Clickhouse and Superset. An alternative solution based on the ELK stack
+is also supported (bottom, orange), consisting of Elasticsearch,
+Logstash and Kibana.}
 
@@
 database. As part of the ingestion, incoming FHIR bundles that contain
-multiple FHIR resources are unbundled in separate topics. Subsequently,
-each FHIR resource topic is flatted using mappers Unbundling and
-flattening of FHIR resources is done in Kafka using FHIRPath. Superset
-is used as the tool for consuming the data to create dashboard
-visualizations.
-
-The OpenHIM platform also support data analytics based on the ELK stack,
-where data is ingested in bulk using Logstash, stored in Elasticsearch
-and made available for consumption in Kibana. Also here, the incoming
-FHIR bundles are unbundled in Logstash into separate FHIR resources.
-However, given that Elasticsearch is a document-based search engine, the
-FHIR resources are stored as-is with no flattening. Exploring and
-analysing the data requires writing queries in Elasticsearch Query
-Language (ES\textbar QL), either through the query interface of
-Elasticsearch or using Kibana.
+multiple FHIR resources are unbundled in separate topics using a generic
+Kafka utility component. Subsequently, each FHIR resource topic is
+flatted with Kafka mappers that use FHIRPath. Superset is used as the
+tool for consuming the data to create dashboard visualizations.
+
+The OpenHIM platform also support data and analytics based on the ELK
+stack, where data is ingested in bulk using Logstash, stored in
+Elasticsearch and made available for consumption in Kibana. Also here,
+the incoming FHIR bundles are unbundled in Logstash into separate FHIR
+resources. However, given that Elasticsearch is a document-based search
+engine, the FHIR resources are stored as-is with no flattening.
+Exploring and analysing the data requires writing queries in
+Elasticsearch Query Language (ES\textbar QL), either through the query
+interface of Elasticsearch or using Kibana.
 
@@
 
+TO DO: add section describing ONA Canopy.
+
 \subsection{Momcare programme}\label{momcare-programme}
@@
 \citep{shija2021access, mrema2021application} in 2017 and 2019
-respectively, with the objective to create transparency on the journeys
-of pregnant mothers. The programme is built on three pillars: journey
-tracking, quality support and a mobile wallet. MomCare distinguishes two
-user groups: mothers are supported during their pregnancy through
-reminders and surveys, using SMS as the digital mode of engagement.
-Health workers are equipped with an Android-based application, in which
-visits, care activities and clinical observations are recorded.
-Reimbursements of the maternal clinic are based on the data captured
-with SMS and the app, thereby creating a conditional payment scheme,
-where providers are partially reimbursed up-front for a fixed bundle of
-activities, supplemented by bonus payments based on a predefined set of
-care activities.
-
-In its original form, the MomCare programme uses predominantly closed
-digital platforms. In Kenya, M-TIBA is the primary digital platform, on
-top of which a relatively lightweight custom app has been built as the
+respectively, with the objective to improve health outcomes for maternal
+and antenatal care. MomCare distinguishes two user groups: mothers are
+supported during their pregnancy through reminders and surveys, using
+SMS as the digital mode of engagement. Health workers are equipped with
+an Android-based application, in which visits, care activities and
+clinical observations are recorded. Reimbursements of the maternal
+clinic are based on the data captured with SMS and the app, thereby
+creating a conditional payment scheme, where providers are partially
+reimbursed up-front for a fixed bundle of activities, supplemented by
+bonus payments based on a predefined set of care activities.
+
+In its original form, the MomCare programme used closed digital
+platforms. In Kenya, M-TIBA is the primary digital platform, on top of
+which a relatively lightweight custom app has been built as the
 engagement layer for the health workers \citep{huisman2022digital}.
@@
 does not provide an interface of any kind for interacting with the
-platform \citep{mrema2021application}. Given these constraints, to date
-the MomCare programme uses a custom-built data warehouse environment as
-its main data platform, on which data extractions, transformations and
-analysis are performed to generate the operational reports. Feedback
-reports for the health workers, in the form of operational dashboards,
-are made accessible through the app. Similar reports are provided to the
-back-office for the periodic reimbursement to the clinics.
-
-In the periode 2022-2023, a new iteration of MomCare \ldots One of the
-practical objectives of the DataCare project was to assess the
-complexity and effort required to implement the data transformation
-scripts, and reproduce the existing analytic reports. The data was
-transformed into 10 FHIR v4 resources as listed in table 1 with the
-number of records per resource type. The conceptual data model of the
-existing MomCare app could readily be transformed into the FHIR standard
-using SQL and validated with a Python library \citep{islam2023fhir}.
-
-The largest challenge during the transformation process pertained to the
-absence of unique business identifiers for patients and healthcare
-organizations. For patients, either the mobile phone number or the
-healthcare insurance number was taken, depending on availability. A
-combination of name, address and latitude/longitude coordinates were
-used to uniquely identify organizations and locations, as Tanzania does
-not have a system in place for this purpose.
-
-The transformed and validated data is uploaded into the FHIR server on a
-daily basis using an automated cloud function. Analysis of bulk data was
-done by directly reading the standard newline delimited JSON into the
-Python pandas data analysis library. Cross checking the output with
-queries on the original data confirmed that the whole data pipeline
-produced consistent results. For example, the report of the antenatal
-coverage metric (number of pregnancies with four or more visits) could
-be reproduced per patient journey and aggregated (per year, per
+platform \citep{mrema2021application}. Given these constraints, the
+first iteration of the MomCare programme used a custom-built data
+warehouse environment as its main data platform, on which data
+extractions, transformations and analysis are performed to generate the
+operational reports. Feedback reports for the health workers, in the
+form of operational dashboards, are made accessible through the app.
+Similar reports are provided to the back-office for the periodic
+reimbursement to the clinics.
+
+Clearly, a more open and scaleable platform was required if MomCare was
+to be implemented in more regions. This need led to a redesign of the
+underlying technical infrastructure of the MomCare project. The
+objectives of this work were in fact to demonstrate a solution design
+that could support the first three types of data sharing listed in
+Table~\ref{tbl-types-data-sharing}. First, to investigate the viability
+of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed
+to assess the complexity and effort required to implement the facade
+pattern to integrate the legacy system into the FHIR data standard.
+Using the longitudinal dataset from approximately 28 thousand patient
+records, FHIR transformations script were developed and deployed using
+the mediator function of the IOL. The data was transformed into 10 FHIR
+v4 resources and the conceptual data model of the existing MomCare app
+could readily be transformed into the FHIR standard using SQL and
+validated with a Python library \citep{islam2023fhir}. The largest
+challenge during the transformation process pertained to the absence of
+unique business identifiers for patients and healthcare organizations.
+For patients, either the mobile phone number or the healthcare insurance
+number was taken, depending on availability. A combination of name,
+address and latitude/longitude coordinates were used to uniquely
+identify organizations and locations, as Tanzania does not have a system
+in place for this purpose.
+
+The second objective was to reproduce existing analytic reports, using
+the bulk FHIR data format as input. Here, the focus was to standardize
+the logic required for producing metrics and reports. The transformed
+and validated data is uploaded into the FHIR server on a daily basis
+using an automated cloud function. Analysis of bulk data was done by
+directly reading the standard newline delimited JSON into the Python
+pandas data analysis library. Cross checking the output with queries on
+the original data confirmed that the whole data pipeline produced
+consistent results. For example, the report of the antenatal coverage
+metric (number of pregnancies with four or more visits) could be
+reproduced per patient journey and aggregated (per year, per
 organization etc.) as required for the MomCare reports.
+
+TO DO: explain logic of patient-timeline table. Write standard
+transformation to go from FHIR resources to this standard table. On top
+of that the actual metrics and reporting. Explain serverless: we wanted
+to get rid of resource-heavy data visualization tools. This led to the
+idea of serverless: using duckdb-wasm and pipelines of cloud functions.
+
+The third objective was to run a technical feasibility test for
+federated analytics. Using the MPC platform of Roseman Labs, we managed
+to do aggregations in the blind \ldots{} TO DO: explain that we managed
+to reproduce the reports we generated in the clear, but then in the
+blind. Note, however, that in the remainder we will focus on first two
+types of data sharing.
+
+Based on these experiments, we arrived at the following design for the
+data \& analytics services
+
+\begin{itemize}
+\tightlist
+\item
+  Use `serverless' file-based storage: bulk copy of data as-is in
+  parquet
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Tension: how to manage change data capture
+  \item
+    Tension: how to manage access rights
+  \end{itemize}
+\item
+  Use SQL-on-FHIR-v2 to create tabular views.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Example: patient timeline
+  \item
+    TO DO: rewrite patient timeline queries with SQL-on-FHIR-v2 and run
+    it with Pathling
+  \end{itemize}
+\item
+  Use semantic modeling layer to define metrics
+
+  \begin{itemize}
+  \tightlist
+  \item
+    There are many options: dbt, cube.dev
+  \item
+    Fulfills same function as ADX/mADX IHE profile in OpenHIE
+    specification
+  \item
+    Tension: going from patient-timeline to reported metrics still isn't
+    standardized. This is where Ibis/Substrait comes in. Substrait as IR
+    for cross-language serialization for relational algebra. Can be
+    executed on different backends. Write once, run on different
+    engines.
+  \end{itemize}
+\item
+  Distribute and publish reports on resource-constrained devices
+
+  \begin{itemize}
+  \tightlist
+  \item
+    duckdb
+  \item
+    sveltekit
+  \end{itemize}
+\end{itemize}
+
+TO DO: Add diagram
 
@@
 \begin{itemize}
-\tightlist
 \item
@@
   providers to resolve the negative implications of openness?
+\item
+  Answers/insights to above:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Openness of standardized view on FHIR data and cross-language
+    serialization of relational algebra makes it possible to fully
+    standardize the workflow from start to finish
+  \item
+    Platform-to-platform: MPC
+  \item
+    Risk of openness: difficult to answer \ldots{}
+  \end{itemize}
 \end{itemize}

### 2024-04-02 index.qmd
Fix openhim image

@@
 - Since we use FHIR, we don't need a semantic layer because that is already provided
--
+- We do need different semantic layer, namely with metrics. Explain different types of semantics.
+  - The metrics layer same function as CQL. Discuss CQL vs generic metrics layer.
 

### 2024-04-15 index.qmd
Add evaluation dimensions in Methods section

@@
 format:
+  elsevier-html:
+    toc: true
+    toc-depth: 2
+    fig-cap-location: top
   elsevier-pdf:
@@
       cite-style: authoryear
-  elsevier-html:
-    toc: true
-    toc-depth: 2
-    fig-cap-location: top
-
 author:
@@
 Conceptually, the OpenHIE framework constitutes a framework for an open digital platform.
-Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods[@digitalpublicgoods].
+Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods [@digitalpublicgoods].
 If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for data sharing and reuse.
-Distinguishing four types of data sharing (@tbl-types-data-sharing), the purpose of this paper is to investigate how new standards and technologies that can establish openness of HDPs can be integrated into the OpenHIE architecture framework.
 The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.
-In addition, we explicitly address the requirement of downward scalability as we aim to implement health data platforms in resource constrained setting of LMICs.
+The purpose of this paper, therefore, is to specificy how new standards and technologies can be integrated into the OpenHIE architecture such that an open HDP can be realized that supports the four types of data sharing as defined by de Reuver at al.
+(@tbl-types-data-sharing).
 
@@
 
-: Types of data sharing and in relation to new standards and technology enablers to create openness. {#tbl-types-data-sharing}
+: Types of data sharing and in relation to new standards and technology enablers to create openness. Taken from de Reuver et al. (2022). {#tbl-types-data-sharing}
 
@@
 
-In the paper, we present a design that extends the OpenHIE specification to include the four types of data sharing mentioned above.
+In this paper, we present a design that extends the OpenHIE specification to include the four types of data sharing mentioned above.
 Using the full-STAC approach [@mehl2023fullstac] we combine open standards, open technologies and open architectures into a coherent modular HDP platform that can be configured and reused across a variety of use-cases.
-Subsequently, we employ a formative, naturalistic evaluation to assess the technical risk and efficacy of the design [@venable2016feds].
-Given that it is prohibitively expensive to evaluate the design a real-world setting, we aim to minimize technological risks and maximize the efficacy of the design by considering three examples of health data platforms in LMICs, namely i) the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)); ii) the ONA Canopy platform [https://ona.io/home/services/canopy/](https://ona.io/home/services/canopy/); and iii) the work conducted at PharmAccess Foundation as part of the MomCare programme ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org)).
+We employ a formative, naturalistic evaluation to assess the technical risk and efficacy of the design [@venable2016feds].
+Given that it is prohibitively expensive to evaluate the design a real-world setting, we aim to minimize technological risks and maximize the efficacy of the design by considering three examples of health data platforms in LMICs, namely:
+
+ 1. the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/))
+ 2. the ONA Canopy platform [https://ona.io/home/services/canopy/](https://ona.io/home/services/canopy/)
+ 3. the work conducted at PharmAccess Foundation as part of the MomCare programme ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org)).
+
+These three real-world implementations are evaluated along the following dimensions:
+
+- What is the level of openness of the implementation, specifically in terms of the four types of data sharing and platform-to-platform opennenss?
+- How are the core maintenance functions of the HDP implemented, where we particularly focus on functions related to metadata management, data lineage and access control mechanisms?
+- Is the solution suitable for downward scalability, where we focus on the computer and storage requirements of the solution and assess the minimal requirements for on-site resources (computers, edge devices)
+
 As part of our design research, we have taken a narrative approach in surveying existing scientific studies on health data platforms, focusing on the seminal reports and subsequently searching forward citations.
@@
 
-The main contributions of this paper are i) description of a framework for the components of the Data & Analysis Services that builds on current best practices from the data engineering community into the OpenHIE framework; and ii) evaluation of different implementations and design options for type 1 and 2 data sharing within an extended OpenHIE architecture.
+The main contributions of this paper are i) description of a framework for the components of the Data & Analysis Services that builds on current best practices from the data engineering community into the OpenHIE framework; ii) evaluation of different implementations and design options for type 1 and 2 data sharing within an extended OpenHIE architecture; and iii) provide open content for the MomCare implementation to facilitate adaptation and deployment.
 As such, it aims to inform future developments and implementation of open digital health platforms in LMICs.

### 2024-04-15 index.tex
Fix openhim image
Add evaluation dimensions in Methods section

@@
 platform that can be configured and reused across a variety of
-use-cases. Subsequently, we employ a formative, naturalistic evaluation
-to assess the technical risk and efficacy of the design
-\citep{venable2016feds}. Given that it is prohibitively expensive to
-evaluate the design a real-world setting, we aim to minimize
-technological risks and maximize the efficacy of the design by
-considering three examples of health data platforms in LMICs, namely i)
-the OpenHIM platform (\url{https://jembi.gitbook.io/openhim-platform/});
-ii) the ONA Canopy platform \url{https://ona.io/home/services/canopy/};
-and iii) the work conducted at PharmAccess Foundation as part of the
-MomCare programme (\url{https://health-data-commons.pharmaccess.org}).
+use-cases. We employ a formative, naturalistic evaluation to assess the
+technical risk and efficacy of the design \citep{venable2016feds}. Given
+that it is prohibitively expensive to evaluate the design a real-world
+setting, we aim to minimize technological risks and maximize the
+efficacy of the design by considering three examples of health data
+platforms in LMICs, namely:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  the OpenHIM platform
+  (\url{https://jembi.gitbook.io/openhim-platform/})
+\item
+  the ONA Canopy platform \url{https://ona.io/home/services/canopy/}
+\item
+  the work conducted at PharmAccess Foundation as part of the MomCare
+  programme (\url{https://health-data-commons.pharmaccess.org}).
+\end{enumerate}
+
 As part of our design research, we have taken a narrative approach in
@@
 \item
+  We do need different semantic layer, namely with metrics. Explain
+  different types of semantics.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    The metrics layer same function as CQL. Discuss CQL vs generic
+    metrics layer.
+  \end{itemize}
 \end{itemize}

### 2024-04-15 index.tex
wip

@@
 and ii) the use of open source components that are available as digital
-public goods\citep{digitalpublicgoods}. If we are to use the OpenHIE
+public goods \citep{digitalpublicgoods}. If we are to use the OpenHIE
 framework as an open data platform, we need to extend the standards,
 technologies and architecture to include functionality for data sharing
-and reuse. Distinguishing four types of data sharing
-(Table~\ref{tbl-types-data-sharing}), the purpose of this paper is to
-investigate how new standards and technologies that can establish
-openness of HDPs can be integrated into the OpenHIE architecture
-framework. The lack of detailed specifications and consensus of this
+and reuse. The lack of detailed specifications and consensus of this
 addition to OpenHIE currently stands in the way of development projects
-that aim to establish HDPs in LMICs. In addition, we explicitly address
-the requirement of downward scalability as we aim to implement health
-data platforms in resource constrained setting of LMICs.
+that aim to establish HDPs in LMICs. The purpose of this paper,
+therefore, is to specificy how new standards and technologies can be
+integrated into the OpenHIE architecture such that an open HDP can be
+realized that supports the four types of data sharing as defined by de
+Reuver at al.~(Table~\ref{tbl-types-data-sharing}).
 
@@
 \caption{Types of data sharing and in relation to new standards and
-technology enablers to create
-openness.}\label{tbl-types-data-sharing}\tabularnewline
+technology enablers to create openness. Taken from de Reuver et
+al.~(2022).}\label{tbl-types-data-sharing}\tabularnewline
 \toprule\noalign{}
@@
 
-In the paper, we present a design that extends the OpenHIE specification
-to include the four types of data sharing mentioned above. Using the
-full-STAC approach \citep{mehl2023fullstac} we combine open standards,
-open technologies and open architectures into a coherent modular HDP
-platform that can be configured and reused across a variety of
-use-cases. We employ a formative, naturalistic evaluation to assess the
-technical risk and efficacy of the design \citep{venable2016feds}. Given
-that it is prohibitively expensive to evaluate the design a real-world
-setting, we aim to minimize technological risks and maximize the
-efficacy of the design by considering three examples of health data
+In this paper, we present a design that extends the OpenHIE
+specification to include the four types of data sharing mentioned above.
+Using the full-STAC approach \citep{mehl2023fullstac} we combine open
+standards, open technologies and open architectures into a coherent
+modular HDP platform that can be configured and reused across a variety
+of use-cases. We employ a formative, naturalistic evaluation to assess
+the technical risk and efficacy of the design \citep{venable2016feds}.
+Given that it is prohibitively expensive to evaluate the design a
+real-world setting, we aim to minimize technological risks and maximize
+the efficacy of the design by considering three examples of health data
 platforms in LMICs, namely:
@@
 \end{enumerate}
+
+These three real-world implementations are evaluated along the following
+dimensions:
+
+\begin{itemize}
+\tightlist
+\item
+  What is the level of openness of the implementation, specifically in
+  terms of the four types of data sharing and platform-to-platform
+  opennenss?
+\item
+  How are the core maintenance functions of the HDP implemented, where
+  we particularly focus on functions related to metadata management,
+  data lineage and access control mechanisms?
+\item
+  Is the solution suitable for downward scalability, where we focus on
+  the computer and storage requirements of the solution and assess the
+  minimal requirements for on-site resources (computers, edge devices)
+\end{itemize}
 
@@
 current best practices from the data engineering community into the
-OpenHIE framework; and ii) evaluation of different implementations and
+OpenHIE framework; ii) evaluation of different implementations and
 design options for type 1 and 2 data sharing within an extended OpenHIE
-architecture. As such, it aims to inform future developments and
-implementation of open digital health platforms in LMICs.
+architecture; and iii) provide open content for the MomCare
+implementation to facilitate adaptation and deployment. As such, it aims
+to inform future developments and implementation of open digital health
+platforms in LMICs.
 
@@
 
+\subsubsection{Level of opennness}\label{level-of-opennness}
+
+TO DO: evaluate openness of OpenHIM platform
+
+\subsubsection{Core maintenance
+functions}\label{core-maintenance-functions}
+
+TO DO: evaluate maintenance functions
+
+\subsubsection{Downward scaleability}\label{downward-scaleability}
+
+TO DO: evaluate downward scalability
+
 \subsection{ONA Canopy}\label{ona-canopy}
@@
 TO DO: add section describing ONA Canopy.
+
+\subsubsection{Level of opennness}\label{level-of-opennness-1}
+
+TO DO: evaluate openness of OpenHIM platform
+
+\subsubsection{Core maintenance
+functions}\label{core-maintenance-functions-1}
+
+TO DO: evaluate maintenance functions
+
+\subsubsection{Downward scaleability}\label{downward-scaleability-1}
+
+TO DO: evaluate downward scalability
 
@@
 
+\subsubsection{Level of opennness}\label{level-of-opennness-2}
+
+TO DO: evaluate openness of OpenHIM platform
+
+\subsubsection{Core maintenance
+functions}\label{core-maintenance-functions-2}
+
+TO DO: evaluate maintenance functions
+
+\subsubsection{Downward scaleability}\label{downward-scaleability-2}
+
+TO DO: evaluate downward scalability
+
 \section{Discussion}\label{discussion}
@@
 \begin{itemize}
+\tightlist
 \item

### 2024-04-22 index.qmd
wip
Add references

@@
 The recent convergence to FHIR as the de facto standard for information exchange has fuelled the development of OpenHIE.
-FHIR is currently used both for routine healthcare settings [@ayaz2021fast] and clinical research settings [@duda2022hl7;@vorisek2022fast] and is increasingly being used in LMICs as well.
+FHIR is currently used both for routine healthcare settings [@amar2024electronic;@ayaz2021fast] and clinical research settings [@duda2022hl7;@vorisek2022fast] and is increasingly being used in LMICs as well.
 The guidelines and standards of the African Union explicitly state FHIR is to be used as the messaging standard [@2023african].
@@
 
+#### Level of opennness
+
+TO DO: evaluate openness of OpenHIM platform
+
+#### Core maintenance functions
+
+TO DO: evaluate maintenance functions
+
+#### Downward scaleability
+
+TO DO: evaluate downward scalability
+
 ### ONA Canopy
@@
 TO DO: add section describing ONA Canopy.
+
+#### Level of opennness
+
+TO DO: evaluate openness of OpenHIM platform
+
+#### Core maintenance functions
+
+TO DO: evaluate maintenance functions
+
+#### Downward scaleability
+
+TO DO: evaluate downward scalability
 
@@
 
+#### Level of opennness
+
+TO DO: evaluate openness of OpenHIM platform
+
+#### Core maintenance functions
+
+TO DO: evaluate maintenance functions
+
+#### Downward scaleability
+
+TO DO: evaluate downward scalability
+
 ## Discussion
@@
 - Risk of openness: What are the novel (negative) implications of opening up data platforms? How can reflexivity in design help providers to resolve the negative implications of openness?
-
 - Answers/insights to above:

### 2024-04-22 index.tex
Add references

@@
 exchange has fuelled the development of OpenHIE. FHIR is currently used
-both for routine healthcare settings \citep{ayaz2021fast} and clinical
-research settings \citep{duda2022hl7, vorisek2022fast} and is
-increasingly being used in LMICs as well. The guidelines and standards
-of the African Union explicitly state FHIR is to be used as the
-messaging standard \citep{2023african}. The FHIR-native OpenSRP platform
+both for routine healthcare settings
+\citep{amar2024electronic, ayaz2021fast} and clinical research settings
+\citep{duda2022hl7, vorisek2022fast} and is increasingly being used in
+LMICs as well. The guidelines and standards of the African Union
+explicitly state FHIR is to be used as the messaging standard
+\citep{2023african}. The FHIR-native OpenSRP platform
 \citep{mehl2020open} has been deployed in 14 countries targeting various

### 2024-04-26 index.qmd
Add Alessandro as author

@@
         country: Kenya
-  - name: XXX
+  - name: Alessandro Pietrobon 
     affiliations:

### 2024-04-26 index.tex
Add Alessandro as author

@@
   pdftitle={Conceptualizing open health data platforms for low- and middle income countries},
-  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Antragama Abbas; Steven Wanyee; XXX; John Grimes; Mark van der Graaf; Mark de Reuver},
+  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Antragama Abbas; Steven Wanyee; Alessandro Pietrobon; John Grimes; Mark van der Graaf; Mark de Reuver},
   pdfkeywords={Analytics-on-FHIR, SQL-on-FHIR, HIE, data
@@
 
-\author[6]{XXX%
+\author[6]{Alessandro Pietrobon%
 %

### 2024-05-12 index.qmd
Add markdown table digital public goods

--- 2024-04-22 index.qmd
@@
     fig-cap-location: top
+    code-fold: true
   elsevier-pdf:
@@
 
-![Overview of current open source implementations of components included in the OpenHIE specification that are FHIR-compatible.
+| Category  | Component | Digital Public Good |
+|:----------|:----------|:--------------------|
+| **Interoperability layer (IOL)** | IOL | [OpenHIM](http://openhim.org) |
+| **Registry Services** | Client Registry (CR) | [SanteMPI](https://help.santesuite.org/product-overview/santesuite-products/master-patient-index-santempi) |
+| | | [JeMPI](https://jembi.gitbook.io/jempi/) |
+| | | [OpenCR](https://www.openclientregistry.org/) |
+| | Facility Registry (FR) | [Global Open Facility Registry (GOFR)](https://github.com/intrahealth/gofr) |
+| | Health Worker Registry (HWR) | [iHRIS](https://www.ihris.org/ihris-50) |
+| | Terminology Service (TS) | [OCL Terminology Service](https://openconceptlab.org/terminology-service/) |
+| | Product Catalog (PC) | not available |
+| **Business Domain Services** | Shared Health Record (SHR) | [HAPI FHIR](https://hapifhir.io/) |
+| | | [Fhirbase](https://www.health-samurai.io/fhirbase) |
+| | | [FHIR Server for Azure](https://github.com/microsoft/fhir-server) |
+| | | [I-TECH-UW SHR](https://github.com/I-TECH-UW/shared-health-record) |
+| | Health Management Information System (HMIS) | [DHIS2](http://dhis2.org/) |
+| | Finance and Insurance Service (FIS) | [OpenIMIS](http://openimis.org/) |
+| | Logistics Management Information System (LMIS) | [OpenLMIS](http://openlmis.org/) |
+| **Generic** | Identity and Access Management (IAM) | [Keycloak](https://www.keycloak.org/) |
+| | Gateway & proxy | [FHIR Information Gateway](https://github.com/google/fhir-gateway) |
+| | Admin dashboard for SHR | [OpenSRP FHIR Web](https://github.com/onaio/fhir-web) |
+| | Configuration and deployment | [Instant OpenHIE](https://jembi.gitbook.io/instant-v2/) |
+
+: Overview of current open source implementations of components included in the OpenHIE specification that are FHIR-compatible.
 The category Analytics Services is not a part of the original OpenHIE and is discussed in the paper.
 Point-of-Service systems are excluded for brevity.
-A systematic review of such digital public goods is beyond the scope of this document.](images/digital-public-goods.png){#tbl-digital-public-goods}
+A systematic review of such digital public goods is beyond the scope of this document.
+{#tbl-digital-public-goods .striped .hover .responsive}
 

### 2024-05-12 index.tex
Update render

@@
 
-\begin{table}
-
-\caption{\label{tbl-digital-public-goods}Overview of current open source
-implementations of components included in the OpenHIE specification that
-are FHIR-compatible. The category Analytics Services is not a part of
-the original OpenHIE and is discussed in the paper. Point-of-Service
-systems are excluded for brevity. A systematic review of such digital
-public goods is beyond the scope of this document.}
-
-\centering{
-
-\includegraphics{images/digital-public-goods.png}
-
-}
-
-\end{table}%
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2558}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2558}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4884}}@{}}
+\caption{Overview of current open source implementations of components
+included in the OpenHIE specification that are FHIR-compatible. The
+category Analytics Services is not a part of the original OpenHIE and is
+discussed in the paper. Point-of-Service systems are excluded for
+brevity. A systematic review of such digital public goods is beyond the
+scope of this document.}\label{tbl-digital-public-goods}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Category
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Component
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Digital Public Good
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Category
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Component
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Digital Public Good
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+\textbf{Interoperability layer (IOL)} & IOL &
+\href{http://openhim.org}{OpenHIM} \\
+\textbf{Registry Services} & Client Registry (CR) &
+\href{https://help.santesuite.org/product-overview/santesuite-products/master-patient-index-santempi}{SanteMPI} \\
+& & \href{https://jembi.gitbook.io/jempi/}{JeMPI} \\
+& & \href{https://www.openclientregistry.org/}{OpenCR} \\
+& Facility Registry (FR) &
+\href{https://github.com/intrahealth/gofr}{Global Open Facility Registry
+(GOFR)} \\
+& Health Worker Registry (HWR) &
+\href{https://www.ihris.org/ihris-50}{iHRIS} \\
+& Terminology Service (TS) &
+\href{https://openconceptlab.org/terminology-service/}{OCL Terminology
+Service} \\
+& Product Catalog (PC) & not available \\
+\textbf{Business Domain Services} & Shared Health Record (SHR) &
+\href{https://hapifhir.io/}{HAPI FHIR} \\
+& & \href{https://www.health-samurai.io/fhirbase}{Fhirbase} \\
+& & \href{https://github.com/microsoft/fhir-server}{FHIR Server for
+Azure} \\
+& & \href{https://github.com/I-TECH-UW/shared-health-record}{I-TECH-UW
+SHR} \\
+& Health Management Information System (HMIS) &
+\href{http://dhis2.org/}{DHIS2} \\
+& Finance and Insurance Service (FIS) &
+\href{http://openimis.org/}{OpenIMIS} \\
+& Logistics Management Information System (LMIS) &
+\href{http://openlmis.org/}{OpenLMIS} \\
+\textbf{Generic} & Identity and Access Management (IAM) &
+\href{https://www.keycloak.org/}{Keycloak} \\
+& Gateway \& proxy & \href{https://github.com/google/fhir-gateway}{FHIR
+Information Gateway} \\
+& Admin dashboard for SHR &
+\href{https://github.com/onaio/fhir-web}{OpenSRP FHIR Web} \\
+& Configuration and deployment &
+\href{https://jembi.gitbook.io/instant-v2/}{Instant OpenHIE} \\
+\end{longtable}
 

### 2024-05-14 index.qmd
Update render
Add authors

@@
         country: Kenya
+  - name: Ryan Chrichton
+    affiliations:
+      - name: Jembi Health Systems
+        city: Durban
+        country: South Africa
   - name: John Grimes
@@
   - name: Mark van der Graaf
+    affiliations:
+      - name: PharmAccess Foundation
+        city: Amsterdam
+        country: the Netherlands
+  - name: Nicole Spieker
     affiliations:

### 2024-05-14 index.tex
Add authors

@@
   pdftitle={Conceptualizing open health data platforms for low- and middle income countries},
-  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Antragama Abbas; Steven Wanyee; Alessandro Pietrobon; John Grimes; Mark van der Graaf; Mark de Reuver},
+  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Antragama Abbas; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Mark van der Graaf; Nicole Spieker; Mark de Reuver},
   pdfkeywords={Analytics-on-FHIR, SQL-on-FHIR, HIE, data
@@
 
-\author[7]{John Grimes%
+\author[7]{Ryan Chrichton%
 %
@@
 
+\author[8]{John Grimes%
+%
+}
+
 \author[1]{Mark van der Graaf%
+%
+}
+
+\author[1]{Nicole Spieker%
 %
@@
 \affiliation[6]{organization={ONA},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
-\affiliation[7]{organization={Australian e-Health Research
+\affiliation[7]{organization={Jembi Health
+Systems},city={Durban},country={South
+Africa},countrysep={,},postcodesep={}}
+\affiliation[8]{organization={Australian e-Health Research
 Centre},city={Brisbane},country={Australia},countrysep={,},postcodesep={}}

### 2024-05-16 index.qmd
update Digital Public Goods table

@@
 | **Interoperability layer (IOL)** | IOL | [OpenHIM](http://openhim.org) |
+| | | [mHero](https://www.mhero.org/technology) |
+| | | [OpenFN](https://docs.openfn.org/) |
 | **Registry Services** | Client Registry (CR) | [SanteMPI](https://help.santesuite.org/product-overview/santesuite-products/master-patient-index-santempi) |
@@
 | | | [OpenCR](https://www.openclientregistry.org/) |
+| | | [OpenCRVS](https://www.opencrvs.org/) |
 | | Facility Registry (FR) | [Global Open Facility Registry (GOFR)](https://github.com/intrahealth/gofr) |
+| | | [GeoPrism Registry (GPR)](https://geoprismregistry.com/) `|
 | | Health Worker Registry (HWR) | [iHRIS](https://www.ihris.org/ihris-50) |
 | | Terminology Service (TS) | [OCL Terminology Service](https://openconceptlab.org/terminology-service/) |
-| | Product Catalog (PC) | not available |
+| | Product Catalog (PC) | [PCMT](https://productcatalog.io/) |
 | **Business Domain Services** | Shared Health Record (SHR) | [HAPI FHIR](https://hapifhir.io/) |
@@
 | | Logistics Management Information System (LMIS) | [OpenLMIS](http://openlmis.org/) |
+| | | [OpenBoxes](https://openboxes.com/features/) |
 | **Generic** | Identity and Access Management (IAM) | [Keycloak](https://www.keycloak.org/) |
@@
 
-: Overview of current open source implementations of components included in the OpenHIE specification that are FHIR-compatible.
+: Overview of current open source implementations of components that fit in the OpenHIE specification.
 The category Analytics Services is not a part of the original OpenHIE and is discussed in the paper.
-Point-of-Service systems are excluded for brevity.
+Point-of-Service systems are excluded for brevity.List compiled using [Global Goods for Digital Health](https://digitalsquare.org/digital-health-global-goods), [Digital Public Goods Alliance](https://digitalpublicgoods.net/) and search of open source code repositories.
 A systematic review of such digital public goods is beyond the scope of this document.

### 2024-05-16 index.tex
update Digital Public Goods table

@@
 \caption{Overview of current open source implementations of components
-included in the OpenHIE specification that are FHIR-compatible. The
-category Analytics Services is not a part of the original OpenHIE and is
-discussed in the paper. Point-of-Service systems are excluded for
-brevity. A systematic review of such digital public goods is beyond the
-scope of this document.}\label{tbl-digital-public-goods}\tabularnewline
+that fit in the OpenHIE specification. The category Analytics Services
+is not a part of the original OpenHIE and is discussed in the paper.
+Point-of-Service systems are excluded for brevity.List compiled using
+\href{https://digitalsquare.org/digital-health-global-goods}{Global
+Goods for Digital Health},
+\href{https://digitalpublicgoods.net/}{Digital Public Goods Alliance}
+and search of open source code repositories. A systematic review of such
+digital public goods is beyond the scope of this
+document.}\label{tbl-digital-public-goods}\tabularnewline
 \toprule\noalign{}
@@
 \href{http://openhim.org}{OpenHIM} \\
+& & \href{https://www.mhero.org/technology}{mHero} \\
+& & \href{https://docs.openfn.org/}{OpenFN} \\
 \textbf{Registry Services} & Client Registry (CR) &
@@
 & & \href{https://www.openclientregistry.org/}{OpenCR} \\
+& & \href{https://www.opencrvs.org/}{OpenCRVS} \\
 & Facility Registry (FR) &
@@
 (GOFR)} \\
+& & \href{https://geoprismregistry.com/}{GeoPrism Registry (GPR)} ` \\
 & Health Worker Registry (HWR) &
@@
 Service} \\
-& Product Catalog (PC) & not available \\
+& Product Catalog (PC) & \href{https://productcatalog.io/}{PCMT} \\
 \textbf{Business Domain Services} & Shared Health Record (SHR) &
@@
 \href{http://openlmis.org/}{OpenLMIS} \\
+& & \href{https://openboxes.com/features/}{OpenBoxes} \\
 \textbf{Generic} & Identity and Access Management (IAM) &

### 2024-05-27 index.qmd
wip

--- 2024-05-12 index.qmd
@@
         country: Australia
+  - name: Paula van Brakel
+    affiliations:
+      - name: PharmAccess Foundation
+        city: Amsterdam
+        country: the Netherlands
   - name: Mark van der Graaf
+    affiliations:
+      - name: PharmAccess Foundation
+        city: Amsterdam
+        country: the Netherlands
+  - name: Nicole Spieker
     affiliations:
@@
 Case studies on digital platforms in healthcare point to an emerging pattern where the focus shifts from the digital platform with its defining software and hardware components, to the data as the primary object of interest in and of itself [@ozalp2022digital;@alaimo2022organizations].
-This observation ties into the proposed research agenda by de Reuver et al.
+This development roughly correlates with the commonly used distinction between primary and secondary health data sharing [@cascini2024health].
+While people are generally willing to share their health data to receive the best care (primary use), the attitude towards secondary use of health data varies greatly depending on the type and context.
+The shift in perspective from digital platforms supporting primary data sharing toward data platforms supporting secondary data sharing, ties in with the proposed research agenda by de Reuver et al.
 to consider data platforms as a phenomenon distinct from digital platforms [@dereuver2022openness].
 Generally, data platforms inherit the characteristics of digital platforms.
-For example, the economic perspective on digital platforms stresses their multi-sidedness with developers and consumers, while data platforms are used by data owners, data consumers and third party solution providers.
+From an economic perspective, for example, both types exhibit multi-sided markets.
 At the same time, data platforms differ as their main offerings revolve around data.
-From a market perspective, data platforms have more moderate network effects and are more susceptible to fragmentation and heterogeneity.
+From an ecosystem perspective, data platforms have more moderate network effects and are more susceptible to fragmentation and heterogeneity.
 
@@
 Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods [@digitalpublicgoods].
-If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for data sharing and reuse.
+Typically, these open source components are intended to support deployment in small countries (population up to 10 million) or large NGOs out of the box, and should provide a stepping stone for customized deployments in medium-sized countries (population around 40 million).[^1]  If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for data sharing and reuse.
 The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.
@@
 (@tbl-types-data-sharing).
+
+[^1]: Although the OpenHIE specification does not include details on dimensioning, these are typically the requirements that are used within the community.
+See [OpenHIE Community Wiki](https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements).
 
@@
 Lakehouses typically have a zonal architecture that follow the Extract-Load-Transform pattern (ELT) where data is ingested from the source systems in bulk (E), delivered to storage with aligned schemas (L) and transformed into a format ready for analysis (T) [@hai2023data].
-The discerning characteristic of the lakehouse architecture is its foundation on low-cost and directly-accessible storage that also provides traditional analytical DBMS management and performance features such as ACID transactions, data versioning, auditing, indexing, caching, and query optimization [@armbrust2021lakehouse].
+The discerning characteristic of the lakehouse architecture is its foundation on low-cost and directly-accessible storage that also provides traditional database management and performance features such as ACID transactions, data versioning, auditing, indexing, caching, and query optimization [@armbrust2021lakehouse].
 Lakehouses thus combine the key benefits of data lakes and data warehouses: low-cost storage in an open format accessible by a variety of systems from the former, and powerful management and optimization features from the latter.
@@
 
-![The composable data stack. Image source: [https://voltrondata.com/codex](https://voltrondata.com/codex).](images/composable-data-stack.png){#fig-composable-data-stack}
+::: {#fig-composable-data-stack layout-nrow=2 fig-cap="Schematic overview of the composable data stack showing a) the overall architecture and b) examples of implementations.
+Images taken from [https://voltrondata.com/codex](https://voltrondata.com/codex)."}
+![a)](images/composable-data-stack.png)
 
-![Examples of implementations of the composable data stack.
-Image source: [https://voltrondata.com/codex](https://voltrondata.com/codex).](images/composable-data-stack-implementation-2.png){#fig-cds-examples}
+![b)](images/composable-data-stack-implementation-2.png)
+:::
 
@@
 | | | [OpenCR](https://www.openclientregistry.org/) |
+| | | [OpenCRVS](https://www.opencrvs.org/) |
 | | Facility Registry (FR) | [Global Open Facility Registry (GOFR)](https://github.com/intrahealth/gofr) |
+| | | [GeoPrism Registry (GPR)](https://geoprismregistry.com/) |
+| | | [Resource Map](http://instedd.org/technologies/resource-map/) |
+| | | [Healthsite.io](https://healthsites.io/) |
+| | | [GeoPrism Registry](https://geoprismregistry.com/) |
+| | | [DHIS2](https://github.com/dhis2) |
 | | Health Worker Registry (HWR) | [iHRIS](https://www.ihris.org/ihris-50) |
 | | Terminology Service (TS) | [OCL Terminology Service](https://openconceptlab.org/terminology-service/) |
-| | Product Catalog (PC) | not available |
+| | Product Catalog (PC) | [PCMT](https://productcatalog.io/) |
 | **Business Domain Services** | Shared Health Record (SHR) | [HAPI FHIR](https://hapifhir.io/) |
@@
 | | | [I-TECH-UW SHR](https://github.com/I-TECH-UW/shared-health-record) |
-| | Health Management Information System (HMIS) | [DHIS2](http://dhis2.org/) |
+| | Health Management Information System (HMIS)[^2] | [DHIS2](http://dhis2.org/) |
 | | Finance and Insurance Service (FIS) | [OpenIMIS](http://openimis.org/) |
 | | Logistics Management Information System (LMIS) | [OpenLMIS](http://openlmis.org/) |
+| | | [OpenBoxes](https://openboxes.com/features/) |
 | **Generic** | Identity and Access Management (IAM) | [Keycloak](https://www.keycloak.org/) |
@@
 
-: Overview of current open source implementations of components included in the OpenHIE specification that are FHIR-compatible.
-The category Analytics Services is not a part of the original OpenHIE and is discussed in the paper.
-Point-of-Service systems are excluded for brevity.
+: Overview of current open source implementations of components that fit in the OpenHIE specification.
+The category Analytics Services is not a part of the original OpenHIE and is discussed in this paper.
+Point-of-Service systems are excluded for brevity.List compiled using [OpenHIE Reference Technologies](https://wiki.ohie.org/display/DR/OpenHIE+Reference+Technologies), [Global Goods for Digital Health](https://digitalsquare.org/digital-health-global-goods), [Digital Public Goods Alliance](https://digitalpublicgoods.net/) and search of open source code repositories.
 A systematic review of such digital public goods is beyond the scope of this document.
 {#tbl-digital-public-goods .striped .hover .responsive}
+
+[^2]: Note that there is often a confusion on the acronym HMIS.
+Strictly speaking, the OpenHIE specification uses HMIS to refer to a Health Management Information System that is part of the Business Domain Services.
+Sometimes HMIS is used to refer to a Hospital Management Systems in the Point-of-Service domain, synonymous with an Electronic Medical Record (EMR) system.
 
@@
 |:---|:---|
+| ACID | Atomicity, Consistency, Isolation, and Durability |
 | CLI | Command-line Interface |

### 2024-05-27 index.tex
wip

--- 2024-05-14 index.tex
@@
   pdftitle={Conceptualizing open health data platforms for low- and middle income countries},
-  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Antragama Abbas; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Mark van der Graaf; Nicole Spieker; Mark de Reuver},
+  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Antragama Abbas; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Paula van Brakel; Mark van der Graaf; Nicole Spieker; Mark de Reuver},
   pdfkeywords={Analytics-on-FHIR, SQL-on-FHIR, HIE, data
@@
 \author[8]{John Grimes%
+%
+}
+
+\author[1]{Paula van Brakel%
 %
@@
 object of interest in and of itself
-\citep{ozalp2022digital, alaimo2022organizations}. This observation ties
-into the proposed research agenda by de Reuver et al.~to consider data
+\citep{ozalp2022digital, alaimo2022organizations}. This development
+roughly correlates with the commonly used distinction between primary
+and secondary health data sharing \citep{cascini2024health}. While
+people are generally willing to share their health data to receive the
+best care (primary use), the attitude towards secondary use of health
+data varies greatly depending on the type and context. The shift in
+perspective from digital platforms supporting primary data sharing
+toward data platforms supporting secondary data sharing, ties in with
+the proposed research agenda by de Reuver et al.~to consider data
 platforms as a phenomenon distinct from digital platforms
 \citep{dereuver2022openness}. Generally, data platforms inherit the
-characteristics of digital platforms. For example, the economic
-perspective on digital platforms stresses their multi-sidedness with
-developers and consumers, while data platforms are used by data owners,
-data consumers and third party solution providers. At the same time,
-data platforms differ as their main offerings revolve around data. From
-a market perspective, data platforms have more moderate network effects
+characteristics of digital platforms. From an economic perspective, for
+example, both types exhibit multi-sided markets. At the same time, data
+platforms differ as their main offerings revolve around data. From an
+ecosystem perspective, data platforms have more moderate network effects
 and are more susceptible to fragmentation and heterogeneity.
@@
 and ii) the use of open source components that are available as digital
-public goods \citep{digitalpublicgoods}. If we are to use the OpenHIE
-framework as an open data platform, we need to extend the standards,
-technologies and architecture to include functionality for data sharing
-and reuse. The lack of detailed specifications and consensus of this
-addition to OpenHIE currently stands in the way of development projects
-that aim to establish HDPs in LMICs. The purpose of this paper,
-therefore, is to specificy how new standards and technologies can be
-integrated into the OpenHIE architecture such that an open HDP can be
-realized that supports the four types of data sharing as defined by de
-Reuver at al.~(Table~\ref{tbl-types-data-sharing}).
+public goods \citep{digitalpublicgoods}. Typically, these open source
+components are intended to support deployment in small countries
+(population up to 10 million) or large NGOs out of the box, and should
+provide a stepping stone for customized deployments in medium-sized
+countries (population around 40 million).\footnote{Although the OpenHIE
+  specification does not include details on dimensioning, these are
+  typically the requirements that are used within the community. See
+  \href{https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements}{OpenHIE
+  Community Wiki}.} If we are to use the OpenHIE framework as an open
+data platform, we need to extend the standards, technologies and
+architecture to include functionality for data sharing and reuse. The
+lack of detailed specifications and consensus of this addition to
+OpenHIE currently stands in the way of development projects that aim to
+establish HDPs in LMICs. The purpose of this paper, therefore, is to
+specificy how new standards and technologies can be integrated into the
+OpenHIE architecture such that an open HDP can be realized that supports
+the four types of data sharing as defined by de Reuver at
+al.~(Table~\ref{tbl-types-data-sharing}).
 
@@
 architecture is its foundation on low-cost and directly-accessible
-storage that also provides traditional analytical DBMS management and
+storage that also provides traditional database management and
 performance features such as ACID transactions, data versioning,
@@
 respectively). The composable data stack can be implemented with current
-open source technologies (Figure~\ref{fig-cds-examples}). As an example,
+open source technologies (\textbf{?@fig-cds-examples}). As an example,
 the Ibis user interface is currently sufficiently mature to offer a
@@
 
-\centering{
-
-\includegraphics{images/composable-data-stack.png}
-
-}
-
-\caption{\label{fig-composable-data-stack}The composable data stack.
-Image source: \url{https://voltrondata.com/codex}.}
+\begin{minipage}{\linewidth}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{images/composable-data-stack.png}
+
+}
+
+\subcaption{a)}
 
@@
 
-\begin{figure}
-
-\centering{
-
-\includegraphics{images/composable-data-stack-implementation-2.png}
-
-}
-
-\caption{\label{fig-cds-examples}Examples of implementations of the
-composable data stack. Image source:
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{images/composable-data-stack-implementation-2.png}
+
+}
+
+\subcaption{b)}
+
+\end{figure}%
+
+\end{minipage}%
+
+\caption{\label{fig-composable-data-stack}Schematic overview of the
+composable data stack showing a) the overall architecture and b)
+examples of implementations. Images taken from
 \url{https://voltrondata.com/codex}.}
@@
 \caption{Overview of current open source implementations of components
-included in the OpenHIE specification that are FHIR-compatible. The
-category Analytics Services is not a part of the original OpenHIE and is
-discussed in the paper. Point-of-Service systems are excluded for
-brevity. A systematic review of such digital public goods is beyond the
-scope of this document.}\label{tbl-digital-public-goods}\tabularnewline
+that fit in the OpenHIE specification. The category Analytics Services
+is not a part of the original OpenHIE and is discussed in this paper.
+Point-of-Service systems are excluded for brevity.List compiled using
+\href{https://wiki.ohie.org/display/DR/OpenHIE+Reference+Technologies}{OpenHIE
+Reference Technologies},
+\href{https://digitalsquare.org/digital-health-global-goods}{Global
+Goods for Digital Health},
+\href{https://digitalpublicgoods.net/}{Digital Public Goods Alliance}
+and search of open source code repositories. A systematic review of such
+digital public goods is beyond the scope of this
+document.}\label{tbl-digital-public-goods}\tabularnewline
 \toprule\noalign{}
@@
 (GOFR)} \\
+& & \href{https://geoprismregistry.com/}{GeoPrism Registry (GPR)} \\
+& & \href{http://instedd.org/technologies/resource-map/}{Resource
+Map} \\
+& & \href{https://healthsites.io/}{Healthsite.io} \\
+& & \href{https://geoprismregistry.com/}{GeoPrism Registry} \\
+& & \href{https://github.com/dhis2}{DHIS2} \\
 & Health Worker Registry (HWR) &
@@
 SHR} \\
-& Health Management Information System (HMIS) &
-\href{http://dhis2.org/}{DHIS2} \\
+& Health Management Information System (HMIS)\footnote{Note that there
+  is often a confusion on the acronym HMIS. Strictly speaking, the
+  OpenHIE specification uses HMIS to refer to a Health Management
+  Information System that is part of the Business Domain Services.
+  Sometimes HMIS is used to refer to a Hospital Management Systems in
+  the Point-of-Service domain, synonymous with an Electronic Medical
+  Record (EMR) system.} & \href{http://dhis2.org/}{DHIS2} \\
 & Finance and Insurance Service (FIS) &
@@
 \endlastfoot
+ACID & Atomicity, Consistency, Isolation, and Durability \\
 CLI & Command-line Interface \\

### 2024-05-27 index.qmd
wip - start rewrite table 1

@@
         country: Kenya
-  - name: Antragama Abbas
-    affiliations:
-      - name: Delft University of Technology
-        city: Delft
-        country: the Netherlands
   - name: Steven Wanyee
@@
         country: the Netherlands
+
   - name: Mark van der Graaf
@@
         city: Amsterdam
+        country: the Netherlands
+  - name: Antragama Abbas
+    affiliations:
+      - name: Delft University of Technology
+        city: Delft
         country: the Netherlands
@@
 Case studies on digital platforms in healthcare point to an emerging pattern where the focus shifts from the digital platform with its defining software and hardware components, to the data as the primary object of interest in and of itself [@ozalp2022digital;@alaimo2022organizations].
-This development roughly correlates with the commonly used distinction between primary and secondary health data sharing [@cascini2024health].
-While people are generally willing to share their health data to receive the best care (primary use), the attitude towards secondary use of health data varies greatly depending on the type and context.
-The shift in perspective from digital platforms supporting primary data sharing toward data platforms supporting secondary data sharing, ties in with the proposed research agenda by de Reuver et al.
+This observation ties into with the proposed research agenda by de Reuver et al.
 to consider data platforms as a phenomenon distinct from digital platforms [@dereuver2022openness].
@@
 Instead of access to information, the control of personal data has emerged in the age of platforms as the critical contention.” [@keller2021paradox].
-These conflicts are particularly salient in the healthcare domain, as is evidenced by the polemic surrounding data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond].
+These conflicts are particularly salient in the healthcare domain, where people are generally willing to share their health data to receive the best care (primary use), while the attitude towards secondary use of health data varies greatly depending on the type and context [@cascini2024health].
+The shift in perspective from digital platforms supporting primary data sharing toward data platforms supporting secondary data sharing is one of the key issues surrounding the polemic of data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond].
 Openness is particularly relevant if we are to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].
@@
 First, the OpenHIE framework [@openhie] has been adopted by many sub-Saharan African countries [@mamuye2022health] as the architectural blueprint for implementing nation-wide health information exchanges (HIE), including Nigeria [@dalhatu2023paper], Kenya [@thaiya2021adoption] and Tanzania [@nsaghurwe2021one].
-These countries have, as a matter of course, extended the framework to include “analytics services” as an additional domain.
+These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain.
 The rationale for this addition is to facilitate secondary reuse of health data for academic research, real-world evidence studies etc.
 which can be framed within the context of ongoing efforts towards Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair].
-In doing so, however, we have implicitly moved from conceptualizing digital health platforms (the original OpenHIE specification) to health data platforms.
+In doing so, however, we have implicitly moved from conceptualizing digital health platforms for primary data sharing (the original OpenHIE specification) to health data platforms for secondary data sharing.
 This is problematic because the notion of openness, which is assumed to be essential in establishing solidarity-based approaches to data sharing, is inherently different for a data platform compared to a digital platform.
@@
 Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods [@digitalpublicgoods].
-Typically, these open source components are intended to support deployment in small countries (population up to 10 million) or large NGOs out of the box, and should provide a stepping stone for customized deployments in medium-sized countries (population around 40 million).[^1]  If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for data sharing and reuse.
+If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for secondary data sharing and reuse.
 The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.
@@
 
-[^1]: Although the OpenHIE specification does not include details on dimensioning, these are typically the requirements that are used within the community.
-See [OpenHIE Community Wiki](https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements).
-
-|   | __Type of data sharing__ | __Technology enablers to create openness of health data platforms__ |
+|   | __Type of secondary data sharing__ | __Technology enablers to create openness of health data platforms__ |
 |:-:|:-------|:-------------|
@@
  1. the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/))
- 2. the ONA Canopy platform [https://ona.io/home/services/canopy/](https://ona.io/home/services/canopy/)
+ 2. the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/)
  3. the work conducted at PharmAccess Foundation as part of the MomCare programme ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org)).
@@
 A composable data system (@fig-composable-data-stack), not only separates the storage (layer 3) and execution (layer 2), but also separates the user interface (layer 1) from the execution engine by introducing standards including Substrait for Intermediate Representation (standard A, IR) and Apache Arrow for data connectivity and data memory layout (standards B and C, respectively).
-The composable data stack can be implemented with current open source technologies (@fig-cds-examples).
-As an example, the Ibis user interface is currently sufficiently mature to offer a standardized dataframe interface to 19 different execution engines.
+The first generation of open source components are already available.
+For example, the Ibis user interface is currently sufficiently mature to offer a standardized dataframe interface to 19 different execution engines [@ibisproject].
 
@@
 Today, many components of the OpenHIE specification are now available as a digital public goods, as listed in @tbl-digital-public-goods.
-To further ease the development, configuration and deployment of health information exchanges, the concept of 'Instant OpenHIE' has been championed to (i) allow implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) have a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@InstantOpenHIEv2].
+Typically, these open source components are intended to support deployments in small countries (population up to 10 million) or large NGOs out of the box, and should provide a stepping stone for customized deployments in medium-sized countries (population around 40 million).[^1] To further ease the development, configuration and deployment of health information exchanges, the concept of 'Instant OpenHIE' has been championed to (i) allow implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) have a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@InstantOpenHIEv2].
 Besides the core functional components of the OpenHIE architecture, the Instant OpenHIE toolkit allows packaging and integration of generic components such as Identity and Access Managment (IAM) and a reverse proxy gateway.
 In the following, we will evaluatie three of such configurations, with the aim to conceptualize and evaluate the proposed Data and Analytics Services domain of of the OpenHIE architecture.
+
+[^1]: Although the OpenHIE specification does not include details on dimensioning, these are typically the requirements that are used within the community.
+See [OpenHIE Community Wiki](https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements).
 

### 2024-05-27 index.qmd
wip - rewrite research questions

@@
 The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.
-The purpose of this paper, therefore, is to specificy how new standards and technologies can be integrated into the OpenHIE architecture such that an open HDP can be realized that supports the four types of data sharing as defined by de Reuver at al.
-(@tbl-types-data-sharing).
+The purpose of this paper, therefore, is to specificy how new standards and technologies can be integrated into the OpenHIE architecture such that an open HDP can be realized that supports the following different types of data sharing.
 
-|   | __Type of secondary data sharing__ | __Technology enablers to create openness of health data platforms__ |
-|:-:|:-------|:-------------|
-| __1__ | Data at the most granular (patient) level, which is persisted and used to provide a longitudinal record.
-| __Bulk FHIR API__: has by now been incorporated in all major FHIR implementations the FHIR standard can be readily used to support data sharing to patient-level data across a patient population [@mandl2020push;@jones2021landscape].
-|
-| __2__ | Aggregated data, for example statistics for policy evaluation and benchmarking |__SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows.
-Builds on FHIRPath [@hl72020fhirpath] expressions in a logical structure to specify things like column names and unnested items.
-Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].|
-| __3__ | Data analytics modules, that provide access to work and access the data.
-| __Federated learning (FL)__ [@rieke2020future] and __privacy- enhancing technologies (PETs)__ [@scheibner2021revolutionizing;@jordan2022selecting]: new paradigms that address the problem of data governance and privacy by training algorithms collaboratively without exchanging the data itself.
-Models can be trained on combined datasets and made available as open source artifacts for decision support.
-Data analysts can use FL and PETs to work with the data in a collaborative, decentralized fashion.
-|
-| __4__ | Trained models that have been derived from the data and can be used stand-alone for decision support. | __ONNX__ [@onnx]: an open format built to represent machine learning models. |
+#### 1. Sharing of data sets at the most granular (patient) level, persisted as longitudinal records
+The Shared Health Record (SHR) in OpenHIE provides an operational, real-time transactional data source which is intended for primary data sharing.
+The specification explicitly states that the SHR is distinct from a datawarehouse.
+At the same time, the Health Management Information System (HMIS) component is specified to fulfil functional requirements that are typically provided by datawarehouses.
 
-: Types of data sharing and in relation to new standards and technology enablers to create openness. Taken from de Reuver et al. (2022). {#tbl-types-data-sharing}
+- **Q1:** How can proven solution designs and modern technologies from data warehousing and engineering be integrated in the OpenHIE specification and how does this relate to the curren specifications of the SHR and HMIS.
+
+#### 2. Sharing of data products derived from the original data set
+OpenHIE specifies that the Health Management Information System (HMIS) component should support a workflow to validate and save aggregate data based on the emerging [IHE Aggregate Data Exchange (ADX) standard](https://www.ihe.net/uploadedFiles/Documents/QRPH/IHE_QRPH_Suppl_ADX.pdf).
+This is just one specific workflow within a larger data analytics value chain that covers i) collection, ii) standardization, iii) cleaning, iv) storage, v) analysis and vi) distribution [@curry2016big].
+Sharing of benchmarking, and decision-support tools based on various statistical (machine learning) models that have been trained on data sets and/or data product, can be shared for standalone use.
+
+- **Q2:** How can proven solutions designs from the data warehousing and engineering community be included in OpenHIE to support sharing of wider variety of data products.
+
+#### 3. Sharing of secure computational environments to access and work with the data
+
+Current solution design of centralized data storage is increasingly being challenged.
+Recent work points to federated learning (FL) [@rieke2020future] and privacy- enhancing technologies (PETs) [@scheibner2021revolutionizing;@jordan2022selecting] as a better long-term solution to secure and equitable sharing of data.
+Given the resource constrained environments of LMICs, is is particularly challenging to devise a solution that is economically feasible and ...
+[TO DO: add feedback from Mark vdG]
+
+Federated solutions:
+
+- KETOS OMOP-FHIR [@gruendner2019ketos]
+- Personal Health Train on FHIR [@choudhury2020personal]
+- OHDSI analytics [@khalid2021standardized]
+- CODA project [@mullie2023coda]
+- GenoMed4All, in evalution phase for -omis [@cremonesi2023need]
+
+Similar approaches taken in the context of data sharing for pandemic response, for example:
+
+- VODAN Africa, based on electronic data capture with templates [@purnamajati2022data]
+- Global Data Sharing Initiative (GDSI) [@pirmani2023journey]
+
+- **Q3:** How can a solution be designed that is economically viable, can be deployed in a bottom-up fashion, support data governance principles ...
 
@@
 - With its inherent, graph-like nature, FHIR can be readily incorporated into the principles of FAIR data sharing, where FHIR-based data repositories can be integrated in an overarching netwerk of FAIR data stations [@sinaci2023data;@pedrera-jimenez2023can].
+
+TO DO: re-use these elements from table 1 into this section:
+
+- Given that the Bulk FHIR API has by now been incorporated in all major FHIR implementations [@mandl2020push;@jones2021landscape], the FHIR standard can be readily used to bridge the gap between primary (transactional) and secondary (analytics) data sharing.
+- __SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows.
+Builds on FHIRPath [@hl72020fhirpath] expressions in a logical structure to specify things like column names and unnested items.
+Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].
+Modern data platforms take this Within the HL7 ecosystem much progress has been made to support "analytics-on-FHIR" with standards including [FHIRPath](https://hl7.org/fhirpath/) and the new SQL-on-FHIR specification [@sql-on-fhir].
+Given the use of FHIR as the common data model (the rationale of which will be described later), we formulate the second design question:
 

### 2024-05-27 index.tex
wip - rewrite research questions

@@
   pdftitle={Conceptualizing open health data platforms for low- and middle income countries},
-  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Antragama Abbas; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Paula van Brakel; Mark van der Graaf; Nicole Spieker; Mark de Reuver},
+  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Paula van Brakel; Mark van der Graaf; Nicole Spieker; Antragama Abbas; Mark de Reuver},
   pdfkeywords={Analytics-on-FHIR, SQL-on-FHIR, HIE, data
@@
 
-\author[4]{Antragama Abbas%
+\author[4]{Steven Wanyee%
 %
@@
 
-\author[5]{Steven Wanyee%
+\author[5]{Alessandro Pietrobon%
 %
@@
 
-\author[6]{Alessandro Pietrobon%
+\author[6]{Ryan Chrichton%
 %
@@
 
-\author[7]{Ryan Chrichton%
-%
-}
-
-\author[8]{John Grimes%
+\author[7]{John Grimes%
 %
@@
 
-\author[4]{Mark de Reuver%
+\author[8]{Antragama Abbas%
+%
+}
+
+\author[8]{Mark de Reuver%
 %
@@
 Kenya},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
-\affiliation[4]{organization={Delft University of
+\affiliation[4]{organization={IntelliSOFT},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
+\affiliation[5]{organization={ONA},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
+\affiliation[6]{organization={Jembi Health
+Systems},city={Durban},country={South
+Africa},countrysep={,},postcodesep={}}
+\affiliation[7]{organization={Australian e-Health Research
+Centre},city={Brisbane},country={Australia},countrysep={,},postcodesep={}}
+\affiliation[8]{organization={Delft University of
 Technology},city={Delft},country={the
 Netherlands},countrysep={,},postcodesep={}}
-\affiliation[5]{organization={IntelliSOFT},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
-\affiliation[6]{organization={ONA},city={Nairobi},country={Kenya},countrysep={,},postcodesep={}}
-\affiliation[7]{organization={Jembi Health
-Systems},city={Durban},country={South
-Africa},countrysep={,},postcodesep={}}
-\affiliation[8]{organization={Australian e-Health Research
-Centre},city={Brisbane},country={Australia},countrysep={,},postcodesep={}}
 
@@
 object of interest in and of itself
-\citep{ozalp2022digital, alaimo2022organizations}. This development
-roughly correlates with the commonly used distinction between primary
-and secondary health data sharing \citep{cascini2024health}. While
-people are generally willing to share their health data to receive the
-best care (primary use), the attitude towards secondary use of health
-data varies greatly depending on the type and context. The shift in
-perspective from digital platforms supporting primary data sharing
-toward data platforms supporting secondary data sharing, ties in with
-the proposed research agenda by de Reuver et al.~to consider data
-platforms as a phenomenon distinct from digital platforms
+\citep{ozalp2022digital, alaimo2022organizations}. This observation ties
+into with the proposed research agenda by de Reuver et al.~to consider
+data platforms as a phenomenon distinct from digital platforms
 \citep{dereuver2022openness}. Generally, data platforms inherit the
@@
 \citep{keller2021paradox}. These conflicts are particularly salient in
-the healthcare domain, as is evidenced by the polemic surrounding data
-spaces \citep{otto2022designing} and data solidarity
+the healthcare domain, where people are generally willing to share their
+health data to receive the best care (primary use), while the attitude
+towards secondary use of health data varies greatly depending on the
+type and context \citep{cascini2024health}. The shift in perspective
+from digital platforms supporting primary data sharing toward data
+platforms supporting secondary data sharing is one of the key issues
+surrounding the polemic of data spaces \citep{otto2022designing} and
+data solidarity
 \citep{kickbusch2021lancet, prainsack2022data, prainsack2023beyond}.
@@
 countries have, as a matter of course, extended the framework to include
-``analytics services'' as an additional domain. The rationale for this
-addition is to facilitate secondary reuse of health data for academic
-research, real-world evidence studies etc. which can be framed within
-the context of ongoing efforts towards Findable, Accessible,
+``data \& analytics services'' as an additional domain. The rationale
+for this addition is to facilitate secondary reuse of health data for
+academic research, real-world evidence studies etc. which can be framed
+within the context of ongoing efforts towards Findable, Accessible,
 Interoperable and Reusable (FAIR) sharing of health data
 \citep{guillot2023fair}. In doing so, however, we have implicitly moved
-from conceptualizing digital health platforms (the original OpenHIE
-specification) to health data platforms. This is problematic because the
-notion of openness, which is assumed to be essential in establishing
+from conceptualizing digital health platforms for primary data sharing
+(the original OpenHIE specification) to health data platforms for
+secondary data sharing. This is problematic because the notion of
+openness, which is assumed to be essential in establishing
 solidarity-based approaches to data sharing, is inherently different for
@@
 and ii) the use of open source components that are available as digital
-public goods \citep{digitalpublicgoods}. Typically, these open source
-components are intended to support deployment in small countries
-(population up to 10 million) or large NGOs out of the box, and should
-provide a stepping stone for customized deployments in medium-sized
-countries (population around 40 million).\footnote{Although the OpenHIE
-  specification does not include details on dimensioning, these are
-  typically the requirements that are used within the community. See
-  \href{https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements}{OpenHIE
-  Community Wiki}.} If we are to use the OpenHIE framework as an open
-data platform, we need to extend the standards, technologies and
-architecture to include functionality for data sharing and reuse. The
-lack of detailed specifications and consensus of this addition to
-OpenHIE currently stands in the way of development projects that aim to
-establish HDPs in LMICs. The purpose of this paper, therefore, is to
-specificy how new standards and technologies can be integrated into the
-OpenHIE architecture such that an open HDP can be realized that supports
-the four types of data sharing as defined by de Reuver at
-al.~(Table~\ref{tbl-types-data-sharing}).
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1200}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3200}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5600}}@{}}
-\caption{Types of data sharing and in relation to new standards and
-technology enablers to create openness. Taken from de Reuver et
-al.~(2022).}\label{tbl-types-data-sharing}\tabularnewline
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\centering
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Type of data sharing}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Technology enablers to create openness of health data platforms}
-\end{minipage} \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\centering
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Type of data sharing}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Technology enablers to create openness of health data platforms}
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-\textbf{1} & Data at the most granular (patient) level, which is
-persisted and used to provide a longitudinal record. & \textbf{Bulk FHIR
-API}: has by now been incorporated in all major FHIR implementations the
-FHIR standard can be readily used to support data sharing to
-patient-level data across a patient population
-\citep{mandl2020push, jones2021landscape}. \\
-\textbf{2} & Aggregated data, for example statistics for policy
-evaluation and benchmarking & \textbf{SQL-on-FHIR specification}
-\citep{sql-on-fhir}: provides a standardised approach to make FHIR work
-well with familiar and efficient SQL engines that are most commonly used
-in analytical workflows. Builds on FHIRPath \citep{hl72020fhirpath}
-expressions in a logical structure to specify things like column names
-and unnested items. Implementations of this approach are available or
-forthcoming, including open source implementations such as Pathling
-\citep{grimes2022pathling} and commercial offerings like Aidbox
-\citep{aidbox}. \\
-\textbf{3} & Data analytics modules, that provide access to work and
-access the data. & \textbf{Federated learning (FL)}
-\citep{rieke2020future} and \textbf{privacy- enhancing technologies
-(PETs)} \citep{scheibner2021revolutionizing, jordan2022selecting}: new
-paradigms that address the problem of data governance and privacy by
-training algorithms collaboratively without exchanging the data itself.
-Models can be trained on combined datasets and made available as open
-source artifacts for decision support. Data analysts can use FL and PETs
-to work with the data in a collaborative, decentralized fashion. \\
-\textbf{4} & Trained models that have been derived from the data and can
-be used stand-alone for decision support. & \textbf{ONNX} \citep{onnx}:
-an open format built to represent machine learning models. \\
-\end{longtable}
+public goods \citep{digitalpublicgoods}. If we are to use the OpenHIE
+framework as an open data platform, we need to extend the standards,
+technologies and architecture to include functionality for secondary
+data sharing and reuse. The lack of detailed specifications and
+consensus of this addition to OpenHIE currently stands in the way of
+development projects that aim to establish HDPs in LMICs. The purpose of
+this paper, therefore, is to specificy how new standards and
+technologies can be integrated into the OpenHIE architecture such that
+an open HDP can be realized that supports the following different types
+of data sharing.
+
+\subsubsection{1. Sharing of data sets at the most granular (patient)
+level, persisted as longitudinal
+records}\label{sharing-of-data-sets-at-the-most-granular-patient-level-persisted-as-longitudinal-records}
+
+The Shared Health Record (SHR) in OpenHIE provides an operational,
+real-time transactional data source which is intended for primary data
+sharing. The specification explicitly states that the SHR is distinct
+from a datawarehouse. At the same time, the Health Management
+Information System (HMIS) component is specified to fulfil functional
+requirements that are typically provided by datawarehouses.
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Q1:} How can proven solution designs and modern technologies
+  from data warehousing and engineering be integrated in the OpenHIE
+  specification and how does this relate to the curren specifications of
+  the SHR and HMIS.
+\end{itemize}
+
+\subsubsection{2. Sharing of data products derived from the original
+data
+set}\label{sharing-of-data-products-derived-from-the-original-data-set}
+
+OpenHIE specifies that the Health Management Information System (HMIS)
+component should support a workflow to validate and save aggregate data
+based on the emerging
+\href{https://www.ihe.net/uploadedFiles/Documents/QRPH/IHE_QRPH_Suppl_ADX.pdf}{IHE
+Aggregate Data Exchange (ADX) standard}. This is just one specific
+workflow within a larger data analytics value chain that covers i)
+collection, ii) standardization, iii) cleaning, iv) storage, v) analysis
+and vi) distribution \citep{curry2016big}. Sharing of benchmarking, and
+decision-support tools based on various statistical (machine learning)
+models that have been trained on data sets and/or data product, can be
+shared for standalone use.
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Q2:} How can proven solutions designs from the data
+  warehousing and engineering community be included in OpenHIE to
+  support sharing of wider variety of data products.
+\end{itemize}
+
+\subsubsection{3. Sharing of secure computational environments to access
+and work with the
+data}\label{sharing-of-secure-computational-environments-to-access-and-work-with-the-data}
+
+Current solution design of centralized data storage is increasingly
+being challenged. Recent work points to federated learning (FL)
+\citep{rieke2020future} and privacy- enhancing technologies (PETs)
+\citep{scheibner2021revolutionizing, jordan2022selecting} as a better
+long-term solution to secure and equitable sharing of data. Given the
+resource constrained environments of LMICs, is is particularly
+challenging to devise a solution that is economically feasible and
+\ldots{} {[}TO DO: add feedback from Mark vdG{]}
+
+Federated solutions:
+
+\begin{itemize}
+\tightlist
+\item
+  KETOS OMOP-FHIR \citep{gruendner2019ketos}
+\item
+  Personal Health Train on FHIR \citep{choudhury2020personal}
+\item
+  OHDSI analytics \citep{khalid2021standardized}
+\item
+  CODA project \citep{mullie2023coda}
+\item
+  GenoMed4All, in evalution phase for -omis \citep{cremonesi2023need}
+\end{itemize}
+
+Similar approaches taken in the context of data sharing for pandemic
+response, for example:
+
+\begin{itemize}
+\item
+  VODAN Africa, based on electronic data capture with templates
+  \citep{purnamajati2022data}
+\item
+  Global Data Sharing Initiative (GDSI) \citep{pirmani2023journey}
+\item
+  \textbf{Q3:} How can a solution be designed that is economically
+  viable, can be deployed in a bottom-up fashion, support data
+  governance principles \ldots{}
+\end{itemize}
 
@@
 \item
-  the ONA Canopy platform \url{https://ona.io/home/services/canopy/}
+  the OnaData platform
+  \url{https://ona.io/home/products/ona-data/features/}
 \item
@@
 
+TO DO: re-use these elements from table 1 into this section:
+
+\begin{itemize}
+\tightlist
+\item
+  Given that the Bulk FHIR API has by now been incorporated in all major
+  FHIR implementations \citep{mandl2020push, jones2021landscape}, the
+  FHIR standard can be readily used to bridge the gap between primary
+  (transactional) and secondary (analytics) data sharing.
+\item
+  \textbf{SQL-on-FHIR specification} \citep{sql-on-fhir}: provides a
+  standardised approach to make FHIR work well with familiar and
+  efficient SQL engines that are most commonly used in analytical
+  workflows. Builds on FHIRPath \citep{hl72020fhirpath} expressions in a
+  logical structure to specify things like column names and unnested
+  items. Implementations of this approach are available or forthcoming,
+  including open source implementations such as Pathling
+  \citep{grimes2022pathling} and commercial offerings like Aidbox
+  \citep{aidbox}. Modern data platforms take this Within the HL7
+  ecosystem much progress has been made to support ``analytics-on-FHIR''
+  with standards including \href{https://hl7.org/fhirpath/}{FHIRPath}
+  and the new SQL-on-FHIR specification \citep{sql-on-fhir}. Given the
+  use of FHIR as the common data model (the rationale of which will be
+  described later), we formulate the second design question:
+\end{itemize}
+
 Possible risks pertaining to the use of FHIR as the common data model,
@@
 Arrow for data connectivity and data memory layout (standards B and C,
-respectively). The composable data stack can be implemented with current
-open source technologies (\textbf{?@fig-cds-examples}). As an example,
-the Ibis user interface is currently sufficiently mature to offer a
-standardized dataframe interface to 19 different execution engines.
+respectively). The first generation of open source components are
+already available. For example, the Ibis user interface is currently
+sufficiently mature to offer a standardized dataframe interface to 19
+different execution engines \citep{ibisproject}.
 
@@
 a digital public goods, as listed in
-Table~\ref{tbl-digital-public-goods}. To further ease the development,
-configuration and deployment of health information exchanges, the
-concept of `Instant OpenHIE' has been championed to (i) allow
-implementers to engage with a preconfigured health information exchange
-solution and running tools (based on the architecture) and test their
-applicability and functionality with a real health context problem; and
-(ii) have a packaged reference version of the OpenHIE architecture that
-is comprised of a set of reference technologies and other appropriate
-tools that form the building blocks of the health information exchange
-that can be configured and extended to support particular use cases
+Table~\ref{tbl-digital-public-goods}. Typically, these open source
+components are intended to support deployments in small countries
+(population up to 10 million) or large NGOs out of the box, and should
+provide a stepping stone for customized deployments in medium-sized
+countries (population around 40 million).\footnote{Although the OpenHIE
+  specification does not include details on dimensioning, these are
+  typically the requirements that are used within the community. See
+  \href{https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements}{OpenHIE
+  Community Wiki}.} To further ease the development, configuration and
+deployment of health information exchanges, the concept of `Instant
+OpenHIE' has been championed to (i) allow implementers to engage with a
+preconfigured health information exchange solution and running tools
+(based on the architecture) and test their applicability and
+functionality with a real health context problem; and (ii) have a
+packaged reference version of the OpenHIE architecture that is comprised
+of a set of reference technologies and other appropriate tools that form
+the building blocks of the health information exchange that can be
+configured and extended to support particular use cases
 \citep{InstantOpenHIEv2}. Besides the core functional components of the
@@
 that could support the first three types of data sharing listed in
-Table~\ref{tbl-types-data-sharing}. First, to investigate the viability
+\textbf{?@tbl-types-data-sharing}. First, to investigate the viability
 of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed

### 2024-08-12 index.qmd
wip

@@
     toc: true
-    toc-depth: 2
+    toc-depth: 3
     fig-cap-location: top
@@
         country: the Netherlands
-  - name: Antragama Abbas
-    affiliations:
-      - name: Delft University of Technology
-        city: Delft
-        country: the Netherlands
-  - name: Mark de Reuver
-    affiliations:
-      - name: Delft University of Technology
-        city: Delft
-        country: the Netherlands
 
@@
 The adoption rate of mobile phones in LMICs has been an important driver in implementing digital health solutions [@mccool2022mobile].
-Yet, there are many shortcomings and challenges, including the current fragmentation of digital platforms and the lack of clear-cut pathways of scaling up digital health programmes, such that they can support sustainable and equitable change of national health systems in LMICs [@mccool2022mobile;@who2019recommendations;@neumark2021digital].
+Yet, there are many shortcomings and challenges, including the current fragmentation of digital platforms and the lack of clear-cut pathways of scaling up digital health programmes, such that they can support sustainable and equitable change of national health systems in LMICs [@mehl2023fullstac;@mccool2022mobile;@who2019recommendations;@neumark2021digital].
 
@@
 At the same time, data platforms differ as their main offerings revolve around data.
-From an ecosystem perspective, data platforms have more moderate network effects and are more susceptible to fragmentation and heterogeneity.
+From an ecosystem perspective, data platforms have more moderate network effects and are more susceptible to fragmentation and heterogeneity [@dereuver2022openness].
 
-Particularly relevant in the context of health data platforms (HDPs), is the conceptualization of openness.
+Particularly relevant in the context of open health data platforms (OHDPs), is the conceptualization of openness.
 The shift in perspective from digital platforms to data platforms coincides with the paradox of open [@keller2021paradox].
@@
 These conflicts are particularly salient in the healthcare domain, where people are generally willing to share their health data to receive the best care (primary use), while the attitude towards secondary use of health data varies greatly depending on the type and context [@cascini2024health].
-The shift in perspective from digital platforms supporting primary data sharing toward data platforms supporting secondary data sharing is one of the key issues surrounding the polemic of data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond].
+The shift in perspective from digital platforms supporting primary data sharing toward data platforms supporting secondary data sharing is one of the key issues surrounding the polemic of data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond;@purtova2023data].
 Openness is particularly relevant if we are to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].
@@
 
-This paper is motivated by the conflation of a number of developments relevant to the design and implementation of open, solidarity-based HDPs in LMICs.
+This paper is motivated by the conflation of a number of developments relevant to the design and implementation of open, solidarity-based OHDPs in LMICs.
 First, the OpenHIE framework [@openhie] has been adopted by many sub-Saharan African countries [@mamuye2022health] as the architectural blueprint for implementing nation-wide health information exchanges (HIE), including Nigeria [@dalhatu2023paper], Kenya [@thaiya2021adoption] and Tanzania [@nsaghurwe2021one].
@@
 If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for secondary data sharing and reuse.
-The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.
-The purpose of this paper, therefore, is to specificy how new standards and technologies can be integrated into the OpenHIE architecture such that an open HDP can be realized that supports the following different types of data sharing.
+The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish OHDPs in LMICs.
+The purpose of this paper, therefore, is to specificy how new standards and technologies can be integrated into the OpenHIE architecture such that an open OHDP can be realized that supports the following different types of data sharing.
 
@@
 
-In this paper, we present a design that extends the OpenHIE specification to include the four types of data sharing mentioned above.
-Using the full-STAC approach [@mehl2023fullstac] we combine open standards, open technologies and open architectures into a coherent modular HDP platform that can be configured and reused across a variety of use-cases.
+In this paper, we present a design that extends the OpenHIE specification to include the three types of data sharing mentioned above.
+Using the full-STAC approach [@mehl2023fullstac] we combine open standards, open technologies and open architectures into a coherent modular OHDP that can be configured and reused across a variety of use-cases.
 We employ a formative, naturalistic evaluation to assess the technical risk and efficacy of the design [@venable2016feds].
@@
 
-- What is the level of openness of the implementation, specifically in terms of the four types of data sharing and platform-to-platform opennenss?
-- How are the core maintenance functions of the HDP implemented, where we particularly focus on functions related to metadata management, data lineage and access control mechanisms?
+- What is the level of openness of the implementation, specifically in terms of the three types of data sharing and platform-to-platform opennenss?
+- How are the core maintenance functions of the OHDP implemented, where we particularly focus on functions related to metadata management, data lineage and access control mechanisms?
 - Is the solution suitable for downward scalability, where we focus on the computer and storage requirements of the solution and assess the minimal requirements for on-site resources (computers, edge devices)
@@
 
+### Open content
+
+-
+
 ## Evaluation
 
-### OpenHIM platform
+### Jembi OpenHIM platform
 
@@
 
-#### Level of opennness
+### ONA OpenSRP 2
 
-TO DO: evaluate openness of OpenHIM platform
+Continuing our evaluation of the extended OpenHIE architecture, we can see a different flavor in the implementation driven by Ona.
+Ona is a social enterprise that has pioneered the adoption of FHIR data standard via the development of OpenSRP2, a FHIR-based data collection app built using Google’s FHIR SDK and focused on enabling offline-first workflows for community-based care.
+OpenSRP 2 is a complete rewrite of the original OpenSRP application, a global public good maintained by Ona and deployed in XX countries worldwide.
 
-#### Core maintenance functions
+OpenSRP2 applications are currently implemented in the field in three countries (Uganda, Liberia, and Madagascar) in collaboration with local Ministries of Health and with international donors such as UNICEF, supporting a variety of different workflows including antenatal care (ANC), postnatal Care (PNC), immunization, and last-mile logistics.
+Besides the OpenSRP Android application and HAPI-FHIR backend, in each of its projects Ona also implements a companion set of tools that support analytics and various reporting needs.
 
-TO DO: evaluate maintenance functions
+#### Requirements for data sharing
 
-#### Downward scaleability
+Based on years of work in global health, Ona has learned that the data stack implemented to support a national-scale implementation of its FHIR-based application responds to the following requirements.
 
-TO DO: evaluate downward scalability
+| Requirement | Rationale |
+|:-----|:--------------|
+| Ingest data from multiple sources, both FHIR and non-FHIR based.
+| While most health record data can be collected and aggregated in FHIR, Ministries of Health rely on other data sources to govern their operations.
+For example, operationalizing an immunization campaign usually includes tracking against specific targets for locations to be visited on specific days and number of children to immunize per day.
+Such targets are often stored in spreadsheets or other applications where the data is not FHIR.
+|
+| Ingest data in batches.
+| Most data ingestion can happen in batches, since Ona’s applications are deployed in hard to reach areas where connectivity is an issue.
+Data ingestion closer to real-time can be  relevant for disaster-response and other time-sensitive applications, but this is not a priority.|
+| Support national-scale data volumes.
+| A data store that can grow from dozens to thousands of devices and where data can be aggregated up to the national level, matching the scale of implementation of data collection applications in the field.|
+| Pre-compute complex business metrics.
+| Reporting on health systems requires pre-computing complex metrics and often performing cohort analyses to map trends in service provision.
+For example, understanding quality of care for children requires computing metrics such as the percentage of children fully immunized on schedule (i.e.
+children 6-59 months that have received the set of vaccines required by the Ministry of Health, and have received each of those vaccines within the expected age-window).
+For Business Intelligence applications, calculating such a vital metric cannot be performed at run time, to avoid long and expensive queries.|
+| Outbound integrations.
+| While aggregated data and reports should be accessible by other applications such as BI platforms via pulls, there should be an easy integration framework to push data to other applications used by the Ministry of Health for other purposes, such as DHIS2 for health systems management or RapidPro for communications with program beneficiaries.|
+| Open source and easily deployable in-country.| Given the extremely sensitive nature of health data, it is paramount for governments to have the flexibility to deploy the stack in various different environments, both on premise and in private clouds.|
 
-### ONA Canopy
+: Requirements and rationale for open health data platform developed and used by ONA, based on OpenSRP 2 ([https://opensrp.io/](https://opensrp.io/)).
+{#tbl-ona-requirements .striped .hover .responsive}
 
-TO DO: add section describing ONA Canopy.
+The architecture
+Learning from experience in the field and internal research and development, Ona has developed preferences for a specific data stack responding to the aforementioned requirements.
 
-#### Level of opennness
+[[graphic]]
 
-TO DO: evaluate openness of OpenHIM platform
+Core toolings in the stack include: 
+Data ingestion with Airbyte.
+Ona uses Airbyte as the primary data ingestion tool, leveraging the wide array of connectors that come standard with the application as well as a dedicated suite of connectors developed internally by Ona, including HAPI FHIR, RapidPro, Ona Data, Kobo Toolbox and others.
+Data storage with Clickhouse.
+While different health projects have varying requirements, Ona has found success in using Clickhouse as the main analytics data store in its most recent implementations.
+Clickhouse supports the scale required for analytics at a national level, as well as the speed that enables cross-application integrations and more real-time analytics.
+For example, in Madagascar Ona uses its reporting suite to identify facilities with stock in need of maintenance and can trigger the scheduling of a maintenance visit ad hoc.
+Data transformation with dbt.
+Following global best practice, Ona leverages dbt to segregate the data warehouse in different levels (staging, marts, metrics), as well as pre-computing complex indicators for ease of reporting and for transmissions into other systems.
+For example, in Liberia Ona implements OpenSRP at community health worker level, but can aggregate immunization data at facility level in the data warehouse and then push quarterly summary metrics to DHIS2.
+No recommendation on reporting / BI tooling.
+Ona recognizes that business users have their own strong preferences for BI tooling, and some already have licenses for specific software, so the architecture is flexible to provide easy connections to different BI tools.
 
-#### Core maintenance functions
+Evaluation
 
-TO DO: evaluate maintenance functions
+Evaluating the data stack, we see the following:
+Use of generic best-of-breed tooling.
+Ona focused on utilizing Open HIE tools that are widely adopted outside of the global health and development sectors.
+This approach aims to provide assurance on two main fronts, the ability to handle performance at scale and the long term dependability of the tools, rather than relying on smaller projects with uncertain long term funding or unproven implementations.
+Columnar data warehouse for analytics.
+The scale of Ona’s project requires the implementation of a dedicated database for analytics.
+While original data can still be stored as parquet or other file system, being able to ingest it into a relational data store allows to create well defined indicators.
+Using clickhouse as a tool helps and combine the need accuracy with the speed of reporting as new data is ingested.
+Strong emphasis on SQL.
+While Ona has tested and experimented with FHIR-specific tooling, such as the definition of data projections using sql-on-fhir, Ona found that relying on sql for coding business logic remained the faster and most scalable approach.
 
-#### Downward scaleability
+In summary, for Ona building analytics with FHIR data looks similar to building analytics with any other type of data.
+While FHIR provides a clear and standard data model, managing information for most health systems requires custom integration of data between different sources, as well as computing indicators using business logic specific to the needs of the local users.
+Building upon well established best-of-breed tools allows Ona to implement FHIR applications at scale and provide trusted analytics on top.
 
-TO DO: evaluate downward scalability
-
-### Momcare programme
+### PharmAccess demonstrator Momcare programme
 
@@
 This need led to a redesign of the underlying technical infrastructure of the MomCare project.
-The objectives of this work were in fact to demonstrate a solution design that could support the first three types of data sharing listed in @tbl-types-data-sharing.
+The objectives of this work were in fact to demonstrate a solution design that could support the first three types of data sharing.
 First, to investigate the viability of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed to assess the complexity and effort required to implement the facade pattern to integrate the legacy system into the FHIR data standard.
@@
 
-We specifically address the notion of openness of HDPs in LMICs in terms of the design-related questions put forward by de Reuver at al.11:
+We specifically address the notion of openness of OHDPs in LMICs in terms of the design-related questions put forward by de Reuver at al.11:
 
@@
 data providers, data users, intermediaries, developers)?
-- Unit of analysis: what is platform-to-platform openness in the context of data platforms, given the expectation that different HDPs will emerge at various aggregation levels?
+- Unit of analysis: what is platform-to-platform openness in the context of data platforms, given the expectation that different OHDPs will emerge at various aggregation levels?
 How do we distinguish meta-platforms, forking, and platform interoperability?
@@
 | FL | Federated learning |
-| HDP	|	Health data platform, explicitly differentiated from health digital platform |
+| OHDP	|	Health data platform, explicitly differentiated from health digital platform |
 | HIE | Health Information Exchange |

### 2024-08-12 rewrite.qmd
wip

--- 2024-05-27 index.qmd
@@
 
-## Introduction {#sec-intro}
+## Introduction {#sec-introduction}
 
-### The paradox of open for digital vs. data platforms in healthcare
+Digital transformation have the potential to make inclusive health systems in low- and middle income countries (LMICs) a reality by connecting patients and professionals, using data as a currency of trust.
+Adoption of digital technologies in healthcare means more than simply increasing efficiency or cutting costs.
+Instead, digital technologies have the potential to i) increase the availability, accessibility, acceptability, and quality of health services as we know them; ii) change nature and direction of health care, and the possibility of making it more preventive, personalised, and mobile through the use of such technologies; and iii) enfranchise patients and communities (and particularly vulnerable groups including children and young people) and evolve their relationship with health professionals and providers, thus helping shape the health system according to the needs of the patients and communities [@kickbusch2021lancet].
 
-It is a widely held belief that digital technologies have an important role to play in strengthening health systems in low- and middle income countries (LMICs), as exemplified by the WHO global strategy on digital health [@who2021global].
-The adoption rate of mobile phones in LMICs has been an important driver in implementing digital health solutions [@mccool2022mobile].
-Yet, there are many shortcomings and challenges, including the current fragmentation of digital platforms and the lack of clear-cut pathways of scaling up digital health programmes, such that they can support sustainable and equitable change of national health systems in LMICs [@mccool2022mobile;@who2019recommendations;@neumark2021digital].
+Yet, to date the global digital health ecosystem is still project-centric, resulting in data fragmentation and technology lock-in, compromising health care delivery [@mehl2023fullstac].
+There is no dearth of the number of digital health interventions: sub-Saharan Africa has seen over 700 such projects in the past 10 years.
+There is, however, a fundamental lack in coordination, integration, scalability, sustainability, and equitable distribution of investments in digital health.
+Health policymakers and the global health community at large need to urgently institute coordination mechanisms to terminate unending duplication and disjointed vertical implementations and manage solutions for scale [@karamagi2022ehealth].
 
-A commonly used perspective to scrutinize digital health is to consider it as a digital platform [@dereuver2018digital].
+In this context it is essential to have a good understanding of digital platforms and their associated ecosystem as an essential enabling socio-technical infrastructure upon  which digital health interventions can be realized in a more standardized, scaleable and sustainable fashion.
+Digital platforms are a challenging research object because of their distributed nature and intertwinement with institutions, markets and technologies [@dereuver2018digital].
+Digital platform ecosystems vary according to three core building blocks: (1) platform ownership, (2) value-creating mechanisms, and (3) complementor autonomy [@hein2020digital].
 Digital platforms have disrupted many sectors but have just started to make inroads into highly regulated industries such as healthcare [@ozalp2022digital].
+
 In this light, the challenges faced by LMICs in establishing national digital health platforms have a lot in common with those faced by high income countries.
@@
 From a societal perspective, issues pertaining to the winner-takes-all nature of digital platforms are hotly debated as many jurisdictions make work to ensure these new digital health platforms indeed serve the common good of achieving universal health coverage [@sharon2018when].
-
-Case studies on digital platforms in healthcare point to an emerging pattern where the focus shifts from the digital platform with its defining software and hardware components, to the data as the primary object of interest in and of itself [@ozalp2022digital;@alaimo2022organizations].
-This observation ties into with the proposed research agenda by de Reuver et al.
-to consider data platforms as a phenomenon distinct from digital platforms [@dereuver2022openness].
-Generally, data platforms inherit the characteristics of digital platforms.
-From an economic perspective, for example, both types exhibit multi-sided markets.
-At the same time, data platforms differ as their main offerings revolve around data.
-From an ecosystem perspective, data platforms have more moderate network effects and are more susceptible to fragmentation and heterogeneity.
-
-Particularly relevant in the context of health data platforms (HDPs), is the conceptualization of openness.
-The shift in perspective from digital platforms to data platforms coincides with the paradox of open [@keller2021paradox].
-Originally, openness of digital platforms focused on open source, open standards and copyrights, which by has been superseded by “...
-conflicts about privacy, economic value extraction, the emergence of artificial intelligence, and the destabilizing effects of dominant platforms on (democratic) societies.
-Instead of access to information, the control of personal data has emerged in the age of platforms as the critical contention.” [@keller2021paradox].
-These conflicts are particularly salient in the healthcare domain, where people are generally willing to share their health data to receive the best care (primary use), while the attitude towards secondary use of health data varies greatly depending on the type and context [@cascini2024health].
-The shift in perspective from digital platforms supporting primary data sharing toward data platforms supporting secondary data sharing is one of the key issues surrounding the polemic of data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond].
-Openness is particularly relevant if we are to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].
-
-### From health information exchanges to health data platforms
-
-This paper is motivated by the conflation of a number of developments relevant to the design and implementation of open, solidarity-based HDPs in LMICs.
-First, the OpenHIE framework [@openhie] has been adopted by many sub-Saharan African countries [@mamuye2022health] as the architectural blueprint for implementing nation-wide health information exchanges (HIE), including Nigeria [@dalhatu2023paper], Kenya [@thaiya2021adoption] and Tanzania [@nsaghurwe2021one].
-These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain.
-The rationale for this addition is to facilitate secondary reuse of health data for academic research, real-world evidence studies etc.
-which can be framed within the context of ongoing efforts towards Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair].
-In doing so, however, we have implicitly moved from conceptualizing digital health platforms for primary data sharing (the original OpenHIE specification) to health data platforms for secondary data sharing.
-This is problematic because the notion of openness, which is assumed to be essential in establishing solidarity-based approaches to data sharing, is inherently different for a data platform compared to a digital platform.
-
-Conceptually, the OpenHIE framework constitutes a framework for an open digital platform.
-Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods [@digitalpublicgoods].
-If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for secondary data sharing and reuse.
-The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.
-The purpose of this paper, therefore, is to specificy how new standards and technologies can be integrated into the OpenHIE architecture such that an open HDP can be realized that supports the four types of data sharing as defined by de Reuver at al.
-(@tbl-types-data-sharing).
-
-|   | __Type of secondary data sharing__ | __Technology enablers to create openness of health data platforms__ |
-|:-:|:-------|:-------------|
-| __1__ | Data at the most granular (patient) level, which is persisted and used to provide a longitudinal record.
-| __Bulk FHIR API__: has by now been incorporated in all major FHIR implementations the FHIR standard can be readily used to support data sharing to patient-level data across a patient population [@mandl2020push;@jones2021landscape].
-|
-| __2__ | Aggregated data, for example statistics for policy evaluation and benchmarking |__SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows.
-Builds on FHIRPath [@hl72020fhirpath] expressions in a logical structure to specify things like column names and unnested items.
-Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].|
-| __3__ | Data analytics modules, that provide access to work and access the data.
-| __Federated learning (FL)__ [@rieke2020future] and __privacy- enhancing technologies (PETs)__ [@scheibner2021revolutionizing;@jordan2022selecting]: new paradigms that address the problem of data governance and privacy by training algorithms collaboratively without exchanging the data itself.
-Models can be trained on combined datasets and made available as open source artifacts for decision support.
-Data analysts can use FL and PETs to work with the data in a collaborative, decentralized fashion.
-|
-| __4__ | Trained models that have been derived from the data and can be used stand-alone for decision support. | __ONNX__ [@onnx]: an open format built to represent machine learning models. |
-
-: Types of data sharing and in relation to new standards and technology enablers to create openness. Taken from de Reuver et al. (2022). {#tbl-types-data-sharing}
-
-## Methods
-
-In this paper, we present a design that extends the OpenHIE specification to include the four types of data sharing mentioned above.
-Using the full-STAC approach [@mehl2023fullstac] we combine open standards, open technologies and open architectures into a coherent modular HDP platform that can be configured and reused across a variety of use-cases.
-We employ a formative, naturalistic evaluation to assess the technical risk and efficacy of the design [@venable2016feds].
-Given that it is prohibitively expensive to evaluate the design a real-world setting, we aim to minimize technological risks and maximize the efficacy of the design by considering three examples of health data platforms in LMICs, namely:
-
- 1. the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/))
- 2. the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/)
- 3. the work conducted at PharmAccess Foundation as part of the MomCare programme ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org)).
-
-These three real-world implementations are evaluated along the following dimensions:
-
-- What is the level of openness of the implementation, specifically in terms of the four types of data sharing and platform-to-platform opennenss?
-- How are the core maintenance functions of the HDP implemented, where we particularly focus on functions related to metadata management, data lineage and access control mechanisms?
-- Is the solution suitable for downward scalability, where we focus on the computer and storage requirements of the solution and assess the minimal requirements for on-site resources (computers, edge devices)
-
-As part of our design research, we have taken a narrative approach in surveying existing scientific studies on health data platforms, focusing on the seminal reports and subsequently searching forward citations.
-In addition, we have searched the open source repositories (most notably GitHub) and the online communities (OpenHIE community, FHIR community) to search for relevant open standards, technologies and architectures.
-This paper should not be considered as a proper systematic review.
-
-The main contributions of this paper are i) description of a framework for the components of the Data & Analysis Services that builds on current best practices from the data engineering community into the OpenHIE framework; ii) evaluation of different implementations and design options for type 1 and 2 data sharing within an extended OpenHIE architecture; and iii) provide open content for the MomCare implementation to facilitate adaptation and deployment.
-As such, it aims to inform future developments and implementation of open digital health platforms in LMICs.
-
-## Design
-
-### Open standards: using FHIR as the common data model
-
-The recent convergence to FHIR as the de facto standard for information exchange has fuelled the development of OpenHIE.
-FHIR is currently used both for routine healthcare settings [@amar2024electronic;@ayaz2021fast] and clinical research settings [@duda2022hl7;@vorisek2022fast] and is increasingly being used in LMICs as well.
-The guidelines and standards of the African Union explicitly state FHIR is to be used as the messaging standard [@2023african].
-The FHIR-native OpenSRP platform [@mehl2020open] has been deployed in 14 countries targeting various patient populations, amongst which a reference implementation of the WHO antenatal and neonatal care guidelines for midwives in Lombok, Indonesia [@summitinstitutefordevelopment2023bunda;@kurniawan2019midwife].
-In India, FHIR is used as the underlying technology for the open Health Claims Exchange protocol specification, which has been adopted by the Indian government as the standard for e-claims handling [@hcx].
-This range of utilizations showcase the standards’ widespread applicability.
-The proceedings of the OpenHIE conference 2023 attest to the fact that FHIR and open source technologies are embraced as critical enablers in implementing health information exchanges in LMICs [@ohie2023unconference].
-
-Various studies have investigated the merits of FHIR and its performance vis-a-vis other healthcare standards.
-Comparisons between OpenEHR, ISO 13606, OMOP and FHIR have been made [@ayaz2023transforming;@mullie2023coda;@rinaldi2021openehr;@cremonesi2023need;@sinaci2023data;@tsafnat2024converge].
-A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR concluded that i) these three standards are functionally and technically compatible, and therefore can be used side by side; and that ii) each of these standards have their strengths and limitations that correlate with their intended use as summarized in the @tbl-comparison.
-
-![Comparison of OpenEHR, ISO 13606 and FHIR standards](images/comparison-ehr-standards.png){width=auto #tbl-comparison}
-
-For an infectious diseases dataset with a limited scope, OpenEHR, OMOP and FHIR have been compared and found all to be equally suitable [@rinaldi2021openehr].
-Comparing OMOP and FHIR, the latter has been found to support more granular mappings required for analytics and was therefore chosen as the standard for the CODA project [@mullie2023coda].
-
-Although FHIR was originally designed only for exchange between systems, we propose to use it as the common data model for the design presented here for the following reasons:
-
-- Industry adoption has significantly increased, as exemplified by FHIR-based offering by major cloud providers such as Google, Azure and AWS.
-Also, Africa CDC has explicitly chosen FHIR as the preferred standard;
-- The widespread availibility of the Bulk FHIR API [@mandl2020push;@jones2021landscape] enables bulk, file-based batchwise processing for analytics using the lakehouse architecture as detailed in the next section;
-- The concept of FHIR Profiles allow localisation to tailor the standard to a specific use case.
-A profile defines rules, extensions, and constraints for a resource.
-We posit that the possible penalty of this flexibility, namely having to manage different FHIR versions and/or profiles, is less of an issue in the context of LMICs where first priority is to exchange datasets such as the International Patient Summary (IPS) that are less complex compared to the requirements for high income countries;
-- Being based on webstandards, the FHIR standard lends itself best for further separation of concerns as envisioned by the composable data stack.
-This is an important enabler for the downward scalability of the solution;
-- With its inherent, graph-like nature, FHIR can be readily incorporated into the principles of FAIR data sharing, where FHIR-based data repositories can be integrated in an overarching netwerk of FAIR data stations [@sinaci2023data;@pedrera-jimenez2023can].
-
-Possible risks pertaining to the use of FHIR as the common data model, most notably the possible incompatibilities and/or high costs of maintenance in supporting different versions, will be addressed in the Discussion.
-
-### Open architecture: extending OpenHIE with a composable data stack
-
-#### Evolving the data lakehouse to a composable data stack
-
-Data management and analytics platforms have undergone significant changes since the first generation of data warehouses were introduced.
-Recent studies have shown that the current practice has converged towards the lakehouse as one of the most commonly used solution designs [@armbrust2021lakehouse;@hai2023data;@harby2022data].
-Lakehouses typically have a zonal architecture that follow the Extract-Load-Transform pattern (ELT) where data is ingested from the source systems in bulk (E), delivered to storage with aligned schemas (L) and transformed into a format ready for analysis (T) [@hai2023data].
-The discerning characteristic of the lakehouse architecture is its foundation on low-cost and directly-accessible storage that also provides traditional database management and performance features such as ACID transactions, data versioning, auditing, indexing, caching, and query optimization [@armbrust2021lakehouse].
-Lakehouses thus combine the key benefits of data lakes and data warehouses: low-cost storage in an open format accessible by a variety of systems from the former, and powerful management and optimization features from the latter.
-
-With respect to current implementations of lakehouse data platforms, we observe a proliferation of tools with as yet limited standards to improve technical interoperability.
-In the analysis of Pedreira et al.
-[@pedreira2023composable] the requirement for specialization in data management systems has evolved faster than our software development practices.
-This situation has created a siloed landscape composed of hundreds of products developed and maintained as monoliths, with limited reuse between systems.
-It has also affected the end users, who are often required to learn the idiosyncrasies of dozens of incompatible SQL and non-SQL API dialects, and settle for systems with incomplete functionality and inconsistent semantics.
-To remedy this, Pedreira et al.
-call to (re-)design and implement modern data platforms in terms of a 'composable data stack’ as a means to decrease development and maintenance cost and pick-up the speed of innovation.
-
-While the lakehouse architecture separates the concerns of compute and storage, the composable data stack takes the separation of concerns is taken one step further.
-A composable data system (@fig-composable-data-stack), not only separates the storage (layer 3) and execution (layer 2), but also separates the user interface (layer 1) from the execution engine by introducing standards including Substrait for Intermediate Representation (standard A, IR) and Apache Arrow for data connectivity and data memory layout (standards B and C, respectively).
-The first generation of open source components are already available.
-For example, the Ibis user interface is currently sufficiently mature to offer a standardized dataframe interface to 19 different execution engines [@ibisproject].
-
-::: {#fig-composable-data-stack layout-nrow=2 fig-cap="Schematic overview of the composable data stack showing a) the overall architecture and b) examples of implementations.
-Images taken from [https://voltrondata.com/codex](https://voltrondata.com/codex)."}
-![a)](images/composable-data-stack.png)
-
-![b)](images/composable-data-stack-implementation-2.png)
-:::
-
-#### SQL-on-FHIR v2 as an intermediate representation for FHIR data in tabular format
-
-The premise of separating the user interface from the execution engine is directly related to the key objective of the SQL-on-FHIR project ([https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/](https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/)), namely to make large-scale analysis of FHIR data accessible to a larger audience, portable between systems and to make FHIR data work well with the best available analytic tools, regardless of the technology stack.
-However, to use FHIR effectively analysts require a thorough understanding of the specification as FHIR is represented as a graph of resources, with detailed semantics defined for references between resources, data types, terminology, extensions, and many other aspects of the specification.
-Most analytic and machine learning use cases require the preparation of FHIR data using transformations and tabular projections from its original form.
-The task of authoring these transformations and projections is not trivial and there is currently no standard mechanisms to support reuse.
-
-The solution of the SQL-on-FHIR project is to provide a specification for defining tabular, use case-specific views of FHIR data.
-The view definition and the execution of the view are separated, in such a way that the definition is portable across systems while the execution engine (called runners) are system-specific tools or libraries that apply view definitions to the underlying data layer, optionally making use of annotations to optimize performance.
-
-#### Extending OpenHIE with a FHIR-based composable data stack
-
-We propose to extend the OpenHIE architecture with a "Data and Analytics Services" domain with different service layers by synthesizing the current best practices of the a lakehouse architecture of [@hai2023data;@harby2022data;@harby2024data] and the composable data stack [@pedreira2023composable] (@fig-ohie, @tbl-data-and-analysis-services).
-These 5 services are considered the core of the FHIR data platform, and although in practice often a downstream dashboarding or visualization component is used, that component is not the main focus of our analysis.
-Rather, we aim to elucidate and conceptualize the core "FHIR Data Lakehouse".
-
-![Proposed extension of the OpenHIE architecture that includes "Data and Analytics Services" as an additional service domain.](./images/openhie-extended-architecture.png){#fig-ohie}
-
-+-------------------+---------------------------------------------------------------------------------------+
-| Service           | Functional requirements                                                               |
-+===================+=======================================================================================+
-| Ingestion         | - Bulk                                                                                |
-|                   | - Streaming                                                                           |
-+-------------------+---------------------------------------------------------------------------------------+
-| Storage           | - File-based blob storage                                                             |
-|                   | - Database optimized for online analytical processing (OLAP)
-+-------------------+---------------------------------------------------------------------------------------+
-| Maintenance       | - SQL-on-FHIR View defintions                                                         |
-|                   | - Catalog and other maintenance-related functions as defined by Hai et al.            |
-+-------------------+---------------------------------------------------------------------------------------+
-| Processing        | - SQL-on-FHIR Runner                                                                  |
-| & API             | - Execution engine on tabular data as defined in composable data stack                |
-|                   | - Capability to participate as a node in federated learning / MPC network             |
-|                   | - Read-only access to storage                                                         |  
-+-------------------+---------------------------------------------------------------------------------------+
-| Data              | - SQL interactive development environment (IDE)                                       |
-| Consumer          | - Interactive notebook computing environment [@granger2021jupyter]                    |
-|                   | - BI reporting, dashboarding and data visualization                                   |  
-+-------------------+---------------------------------------------------------------------------------------+
-
-: Definition of Data and Analysis Services {#tbl-data-and-analysis-services}
-
-##### Ingestion
-
-Default workflow is extraction of data from SHR using Bulk FHIR API.
-Data contains metadata (incl.
-FHIR versions) and fully qualified semantics, for example, coding systems.
-Despite this, metadata extraction and metadata modeling is still required to meet the FAIR requirements.
-Issues that need to be solved by these services:
-
-- To prepare for future updates of FHIR versions
-- Implement late-binding principle of having increasingly more specific FHIR profiles as bulk FHIR data propagates through lakehouse
-
-##### Storage
-
-- File-based:
-  - from ndjson to parquet
-  - possibly used delta lake for time versioning
-  - separation of storage from compute not only for benefits of lower TCO, but also be ready for federated learning and MPC in future
-
-- OLAP DBMS
-  - Often columnar, like Clickhouse and BigQuery
-  -
-
-##### Query & Processing
-
-- fit in structure of OpenHIE specification
-- check which workflows are related to analytics
-- Hai calls this 'Maintenance'
-
-##### Maintenance
-
-- SQL-on-FHIR Views provide new standard to support mADX aggregate reporting !! We need to stress this, because this is an existing OpenHIE workflow
-- Maintenance-related functions remain the same
-- NB: orchestration falls under data provenance
-
-##### Data consumers
-
-- Many tools, often focus on creating information dashboards and visualizations
-- Compatibility with processing & API: which query languages and interfaces are supported. Some dialect of SQL, some dialect of NoSQL, dataframe API, all of the above?
-
-### Open technologies: deploying Instant OpenHIE with digital public goods
-
-Today, many components of the OpenHIE specification are now available as a digital public goods, as listed in @tbl-digital-public-goods.
-Typically, these open source components are intended to support deployments in small countries (population up to 10 million) or large NGOs out of the box, and should provide a stepping stone for customized deployments in medium-sized countries (population around 40 million).[^1] To further ease the development, configuration and deployment of health information exchanges, the concept of 'Instant OpenHIE' has been championed to (i) allow implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) have a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@InstantOpenHIEv2].
-Besides the core functional components of the OpenHIE architecture, the Instant OpenHIE toolkit allows packaging and integration of generic components such as Identity and Access Managment (IAM) and a reverse proxy gateway.
-In the following, we will evaluatie three of such configurations, with the aim to conceptualize and evaluate the proposed Data and Analytics Services domain of of the OpenHIE architecture.
-
-[^1]: Although the OpenHIE specification does not include details on dimensioning, these are typically the requirements that are used within the community.
-See [OpenHIE Community Wiki](https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements).
-
-| Category  | Component | Digital Public Good |
-|:----------|:----------|:--------------------|
-| **Interoperability layer (IOL)** | IOL | [OpenHIM](http://openhim.org) |
-| | | [mHero](https://www.mhero.org/technology) |
-| | | [OpenFN](https://docs.openfn.org/) |
-| **Registry Services** | Client Registry (CR) | [SanteMPI](https://help.santesuite.org/product-overview/santesuite-products/master-patient-index-santempi) |
-| | | [JeMPI](https://jembi.gitbook.io/jempi/) |
-| | | [OpenCR](https://www.openclientregistry.org/) |
-| | | [OpenCRVS](https://www.opencrvs.org/) |
-| | Facility Registry (FR) | [Global Open Facility Registry (GOFR)](https://github.com/intrahealth/gofr) |
-| | | [GeoPrism Registry (GPR)](https://geoprismregistry.com/) |
-| | | [Resource Map](http://instedd.org/technologies/resource-map/) |
-| | | [Healthsite.io](https://healthsites.io/) |
-| | | [GeoPrism Registry](https://geoprismregistry.com/) |
-| | | [DHIS2](https://github.com/dhis2) |
-| | Health Worker Registry (HWR) | [iHRIS](https://www.ihris.org/ihris-50) |
-| | Terminology Service (TS) | [OCL Terminology Service](https://openconceptlab.org/terminology-service/) |
-| | Product Catalog (PC) | [PCMT](https://productcatalog.io/) |
-| **Business Domain Services** | Shared Health Record (SHR) | [HAPI FHIR](https://hapifhir.io/) |
-| | | [Fhirbase](https://www.health-samurai.io/fhirbase) |
-| | | [FHIR Server for Azure](https://github.com/microsoft/fhir-server) |
-| | | [I-TECH-UW SHR](https://github.com/I-TECH-UW/shared-health-record) |
-| | Health Management Information System (HMIS)[^2] | [DHIS2](http://dhis2.org/) |
-| | Finance and Insurance Service (FIS) | [OpenIMIS](http://openimis.org/) |
-| | Logistics Management Information System (LMIS) | [OpenLMIS](http://openlmis.org/) |
-| | | [OpenBoxes](https://openboxes.com/features/) |
-| **Generic** | Identity and Access Management (IAM) | [Keycloak](https://www.keycloak.org/) |
-| | Gateway & proxy | [FHIR Information Gateway](https://github.com/google/fhir-gateway) |
-| | Admin dashboard for SHR | [OpenSRP FHIR Web](https://github.com/onaio/fhir-web) |
-| | Configuration and deployment | [Instant OpenHIE](https://jembi.gitbook.io/instant-v2/) |
-
-: Overview of current open source implementations of components that fit in the OpenHIE specification.
-The category Analytics Services is not a part of the original OpenHIE and is discussed in this paper.
-Point-of-Service systems are excluded for brevity.List compiled using [OpenHIE Reference Technologies](https://wiki.ohie.org/display/DR/OpenHIE+Reference+Technologies), [Global Goods for Digital Health](https://digitalsquare.org/digital-health-global-goods), [Digital Public Goods Alliance](https://digitalpublicgoods.net/) and search of open source code repositories.
-A systematic review of such digital public goods is beyond the scope of this document.
-{#tbl-digital-public-goods .striped .hover .responsive}
-
-[^2]: Note that there is often a confusion on the acronym HMIS.
-Strictly speaking, the OpenHIE specification uses HMIS to refer to a Health Management Information System that is part of the Business Domain Services.
-Sometimes HMIS is used to refer to a Hospital Management Systems in the Point-of-Service domain, synonymous with an Electronic Medical Record (EMR) system.
-
-## Evaluation
-
-### OpenHIM platform
-
-To evaluate the extended OpenHIE architecture described above, we first consider the OpenHIM Platform.
-The Open Health Information Mediator (OpenHIM, [http://openhim.org/](OpenHIMhttp://openhim.org/))) component is the reference implementation of the Interoperability Layer (IOL) as defined in the OpenHIE specification.
-The most current version (8.4.2 at the time of writing) provides all the core functions including central point of access for the services of the HIE; routing functions; central logging for auditing and debugging purposes; and orchestration/mediation mechanisms to co-ordinate requests.
-By extension, the OpenHIM Platform ([https://jembi.gitbook.io/openhim-platform](https://jembi.gitbook.io/openhim-platform)) is a reference implementation of a set of Instant OpenHIE configurations, refered to as 'recipes' in the documentation.
-In the following we will evaluate the recipe for "a central data repository with a data warehouse" that provides "A FHIR-based Shared Health record linked to a Master Patient Index (MPI) for linking and mathing patient demographics and a default reporting pipeline to transform and visualise FHIR data" ([https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing](https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing)).
-
-![Overview of the default data stack of the OpenHIM Platform.
-The default stack (top, red) consists of Kafka, Clickhouse and Superset.
-An alternative solution based on the ELK stack is also supported (bottom, orange), consisting of Elasticsearch, Logstash and Kibana.](images/openhim-platform.png){#fig-openhim-platform}
-
-@fig-openhim-platform shows a schematic overview of two data stacks that are supported in the OpenHIM platform.
-The Shared Health Record (SHR, implemented with HAPI FHIR server) and the Client Registry (CR, implemented with JeMPI server) are the sources that store clinical FHIR data and patient demographic data, respectively.
-The default data stack is based on streaming ingestion using Kafka into a Clickhouse database.
-As part of the ingestion, incoming FHIR bundles that contain multiple FHIR resources are unbundled in separate topics using a generic Kafka utility component.
-Subsequently, each FHIR resource topic is flatted with Kafka mappers that use FHIRPath.
-Superset is used as the tool for consuming the data to create dashboard visualizations.
-
-The OpenHIM platform also support data and analytics based on the ELK stack, where data is ingested in bulk using Logstash, stored in Elasticsearch and made available for consumption in Kibana.
-Also here, the incoming FHIR bundles are unbundled in Logstash into separate FHIR resources.
-However, given that Elasticsearch is a document-based search engine, the FHIR resources are stored as-is with no flattening.
-Exploring and analysing the data requires writing queries in Elasticsearch Query Language (ES|QL), either through the query interface of Elasticsearch or using Kibana.
-
-Evaluating these two data stacks, we see the following:
-
-- Pattern of flattening FHIR resources with FHIRPath expressions is very close to the idea of SQL-on-FHIR.
-Although it doesn't adhere to this new standard in the strict sense, the philosophy of generating tabular views is the same
-- When using the ELK stack, flattening is done at the end. Implementations of FHIRPath support Elasticsearch as an execution engine, also here 
-- Main limitations: both Clickhouse en Elasticsearch don't follow decomposition of storage, compute and UI. Therefore, downward scaleability is limited.
-
-#### Level of opennness
-
-TO DO: evaluate openness of OpenHIM platform
-
-#### Core maintenance functions
-
-TO DO: evaluate maintenance functions
-
-#### Downward scaleability
-
-TO DO: evaluate downward scalability
-
-### ONA Canopy
-
-TO DO: add section describing ONA Canopy.
-
-#### Level of opennness
-
-TO DO: evaluate openness of OpenHIM platform
-
-#### Core maintenance functions
-
-TO DO: evaluate maintenance functions
-
-#### Downward scaleability
-
-TO DO: evaluate downward scalability
-
-### Momcare programme
-
-MomCare was launched in Kenya [@huisman2022digital;@sanctis2022maintaining] and Tanzania [@shija2021access;@mrema2021application] in 2017 and 2019 respectively, with the objective to improve health outcomes for maternal and antenatal care.
-MomCare distinguishes two user groups: mothers are supported during their pregnancy through reminders and surveys, using SMS as the digital mode of engagement.
-Health workers are equipped with an Android-based application, in which visits, care activities and clinical observations are recorded.
-Reimbursements of the maternal clinic are based on the data captured with SMS and the app, thereby creating a conditional payment scheme, where providers are partially reimbursed up-front for a fixed bundle of activities, supplemented by bonus payments based on a predefined set of care activities.
-
-In its original form, the MomCare programme used closed digital platforms.
-In Kenya, M-TIBA is the primary digital platform, on top of which a relatively lightweight custom app has been built as the engagement layer for the health workers [@huisman2022digital].
-M-TIBA provides data access through its data warehouse platform for the MomCare programme, however, this is not a standardized, general purpose API.
-In the case of Tanzania, a stand-alone custom app is used which does not provide an interface of any kind for interacting with the platform [@mrema2021application].
-Given these constraints, the first iteration of the MomCare programme used a custom-built data warehouse environment as its main data platform, on which data extractions, transformations and analysis are performed to generate the operational reports.
-Feedback reports for the health workers, in the form of operational dashboards, are made accessible through the app.
-Similar reports are provided to the back-office for the periodic reimbursement to the clinics.
-
-Clearly, a more open and scaleable platform was required if MomCare was to be implemented in more regions.
-This need led to a redesign of the underlying technical infrastructure of the MomCare project.
-The objectives of this work were in fact to demonstrate a solution design that could support the first three types of data sharing listed in @tbl-types-data-sharing.
-First, to investigate the viability of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed to assess the complexity and effort required to implement the facade pattern to integrate the legacy system into the FHIR data standard.
-Using the longitudinal dataset from approximately 28 thousand patient records, FHIR transformations script were developed and deployed using the mediator function of the IOL.
-The data was transformed into 10 FHIR v4 resources and the conceptual data model of the existing MomCare app could readily be transformed into the FHIR standard using SQL and validated with a Python library [@islam2023fhir].
-The largest challenge during the transformation process pertained to the absence of unique business identifiers for patients and healthcare organizations.
-For patients, either the mobile phone number or the healthcare insurance number was taken, depending on availability.
-A combination of name, address and latitude/longitude coordinates were used to uniquely identify organizations and locations, as Tanzania does not have a system in place for this purpose.
-
-The second objective was to reproduce existing analytic reports, using the bulk FHIR data format as input.
-Here, the focus was to standardize the logic required for producing metrics and reports.
-The transformed and validated data is uploaded into the FHIR server on a daily basis using an automated cloud function.
-Analysis of bulk data was done by directly reading the standard newline delimited JSON into the Python pandas data analysis library.
-Cross checking the output with queries on the original data confirmed that the whole data pipeline produced consistent results.
-For example, the report of the antenatal coverage metric (number of pregnancies with four or more visits) could be reproduced per patient journey and aggregated (per year, per organization etc.) as required for the MomCare reports.
-
-TO DO: explain logic of patient-timeline table.
-Write standard transformation to go from FHIR resources to this standard table.
-On top of that the actual metrics and reporting.
-Explain serverless: we wanted to get rid of resource-heavy data visualization tools.
-This led to the idea of serverless: using duckdb-wasm and pipelines of cloud functions.
-
-The third objective was to run a technical feasibility test for federated analytics. Using the MPC platform of Roseman Labs, we managed to do aggregations in the blind ...
-TO DO: explain that we managed to reproduce the reports we generated in the clear, but then in the blind. Note, however, that in the remainder we will focus on first two types of data sharing.
-
-Based on these experiments, we arrived at the following design for the data & analytics services
-
-- Use 'serverless' file-based storage: bulk copy of data as-is in parquet
-  - Tension: how to manage change data capture
-  - Tension: how to manage access rights
-- Use SQL-on-FHIR-v2 to create tabular views.
-  - Example: patient timeline
-  - TO DO: rewrite patient timeline queries with SQL-on-FHIR-v2 and run it with Pathling
-- Use semantic modeling layer to define metrics
-  - There are many options: dbt, cube.dev
-  - Fulfills same function as ADX/mADX IHE profile in OpenHIE specification
-  - Tension: going from patient-timeline to reported metrics still isn't standardized.
-This is where Ibis/Substrait comes in.
-Substrait as IR for cross-language serialization for relational algebra.
-Can be executed on different backends.
-Write once, run on different engines.
-- Distribute and publish reports on resource-constrained devices 
-  - duckdb
-  - sveltekit
-
-TO DO: Add diagram
-
-#### Level of opennness
-
-TO DO: evaluate openness of OpenHIM platform
-
-#### Core maintenance functions
-
-TO DO: evaluate maintenance functions
-
-#### Downward scaleability
-
-TO DO: evaluate downward scalability
-
-## Discussion
-
-### Openness of data platforms
-
-We specifically address the notion of openness of HDPs in LMICs in terms of the design-related questions put forward by de Reuver at al.11:
-
-- Object of openness: what data-related resources should data platforms make available when opening up (e.g.
-data, data products, datadriven insights, analytics modules)?
-Which user groups derive value from accessing data-related resources from data platforms (e.g.
-data providers, data users, intermediaries, developers)?
-- Unit of analysis: what is platform-to-platform openness in the context of data platforms, given the expectation that different HDPs will emerge at various aggregation levels?
-How do we distinguish meta-platforms, forking, and platform interoperability?
-- Risk of openness: What are the novel (negative) implications of opening up data platforms? How can reflexivity in design help providers to resolve the negative implications of openness?
-- Answers/insights to above:
-  - Openness of standardized view on FHIR data and cross-language serialization of relational algebra makes it possible to fully standardize the workflow from start to finish
-  - Platform-to-platform: MPC
-  - Risk of openness: difficult to answer ...
-
-### Comparison with HMIS component
-
-- Workflow requirements: Report aggregate data (link): receiver is HMIS, mADX
-- Functional requirements: https://guides.ohie.org/arch-spec/openhie-component-specifications-1/openhie-health-management-information-system-hmis
-
-Requirements are similar, but implementation differs: Datamodel is non-FHIR, focused on DataValue, which conceptually equates to FHIR Measure
-
-### The need to a semantic layer?
-
-- FHIR and FAIR
-  - How does FHIR relate to approaches taken by the FAIR community, which tend to take more an approach of using knowledge graphs.
-For example, VODAN Africa [@gebreslassie2023fhir4fair;@purnamajati2022data].
-  - FAIR principles vs FHIR graph: is FHIR a FAIR Data Object
-
-- Since we use FHIR, we don't need a semantic layer because that is already provided
-- We do need different semantic layer, namely with metrics. Explain different types of semantics.
-  - The metrics layer same function as CQL. Discuss CQL vs generic metrics layer.
-
-### Attribute-based access control
-
-- TO DO: if you have generated flattened SQL tables, how are you going to manage security?
-- Cerbos, attribute based on lineage or anonymized tables
-- Catalogs solve this: Tabular.io, Google BigLake. What is open source option?
-
-### Federated learning and multiparty computation
-
-- data stations??!
-
-## Abbreviations
-
-|    |    |
-|:---|:---|
-| ACID | Atomicity, Consistency, Isolation, and Durability |
-| CLI | Command-line Interface |
-| CR | Client Registry |
-| ELK | Elasticsearch, Logstach and Kibana stack |
-| ELT | Extract, Load and Transform |
-| FAIR |Findable, Accessible, Interoperable and Reusable |
-| FHIR |Fast Healthcare Interoperability Resources |
-| FL | Federated learning |
-| HDP	|	Health data platform, explicitly differentiated from health digital platform |
-| HIE | Health Information Exchange |
-| IR | Intermediate Representation |
-| LMIC | Low- and middle income countries |
-| MPC | Multiparty Computation |
-| PET | Privacy-enhancing technologies |
-| SHR | Shared Health Record |

### 2024-08-13 index.tex
wip

@@
   pdftitle={Conceptualizing open health data platforms for low- and middle income countries},
-  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Paula van Brakel; Mark van der Graaf; Nicole Spieker; Antragama Abbas; Mark de Reuver},
+  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Paula van Brakel; Mark van der Graaf; Nicole Spieker},
   pdfkeywords={Analytics-on-FHIR, SQL-on-FHIR, HIE, data
@@
 \author[1]{Nicole Spieker%
-%
-}
-
-\author[8]{Antragama Abbas%
-%
-}
-
-\author[8]{Mark de Reuver%
 %
@@
 Centre},city={Brisbane},country={Australia},countrysep={,},postcodesep={}}
-\affiliation[8]{organization={Delft University of
-Technology},city={Delft},country={the
-Netherlands},countrysep={,},postcodesep={}}
 
@@
 health systems in LMICs
-\citep{mccool2022mobile, who2019recommendations, neumark2021digital}.
+\citep{mehl2023fullstac, mccool2022mobile, who2019recommendations, neumark2021digital}.
 
@@
 ecosystem perspective, data platforms have more moderate network effects
-and are more susceptible to fragmentation and heterogeneity.
-
-Particularly relevant in the context of health data platforms (HDPs), is
-the conceptualization of openness. The shift in perspective from digital
-platforms to data platforms coincides with the paradox of open
-\citep{keller2021paradox}. Originally, openness of digital platforms
-focused on open source, open standards and copyrights, which by has been
-superseded by ``\ldots{} conflicts about privacy, economic value
-extraction, the emergence of artificial intelligence, and the
+and are more susceptible to fragmentation and heterogeneity
+\citep{dereuver2022openness}.
+
+Particularly relevant in the context of open health data platforms
+(OHDPs), is the conceptualization of openness. The shift in perspective
+from digital platforms to data platforms coincides with the paradox of
+open \citep{keller2021paradox}. Originally, openness of digital
+platforms focused on open source, open standards and copyrights, which
+by has been superseded by ``\ldots{} conflicts about privacy, economic
+value extraction, the emergence of artificial intelligence, and the
 destabilizing effects of dominant platforms on (democratic) societies.
@@
 data solidarity
-\citep{kickbusch2021lancet, prainsack2022data, prainsack2023beyond}.
+\citep{kickbusch2021lancet, prainsack2022data, prainsack2023beyond, purtova2023data}.
 Openness is particularly relevant if we are to realize a
@@
 This paper is motivated by the conflation of a number of developments
-relevant to the design and implementation of open, solidarity-based HDPs
-in LMICs. First, the OpenHIE framework \citep{openhie} has been adopted
-by many sub-Saharan African countries \citep{mamuye2022health} as the
-architectural blueprint for implementing nation-wide health information
-exchanges (HIE), including Nigeria \citep{dalhatu2023paper}, Kenya
-\citep{thaiya2021adoption} and Tanzania \citep{nsaghurwe2021one}. These
-countries have, as a matter of course, extended the framework to include
-``data \& analytics services'' as an additional domain. The rationale
-for this addition is to facilitate secondary reuse of health data for
-academic research, real-world evidence studies etc. which can be framed
-within the context of ongoing efforts towards Findable, Accessible,
-Interoperable and Reusable (FAIR) sharing of health data
+relevant to the design and implementation of open, solidarity-based
+OHDPs in LMICs. First, the OpenHIE framework \citep{openhie} has been
+adopted by many sub-Saharan African countries \citep{mamuye2022health}
+as the architectural blueprint for implementing nation-wide health
+information exchanges (HIE), including Nigeria \citep{dalhatu2023paper},
+Kenya \citep{thaiya2021adoption} and Tanzania \citep{nsaghurwe2021one}.
+These countries have, as a matter of course, extended the framework to
+include ``data \& analytics services'' as an additional domain. The
+rationale for this addition is to facilitate secondary reuse of health
+data for academic research, real-world evidence studies etc. which can
+be framed within the context of ongoing efforts towards Findable,
+Accessible, Interoperable and Reusable (FAIR) sharing of health data
 \citep{guillot2023fair}. In doing so, however, we have implicitly moved
@@
 consensus of this addition to OpenHIE currently stands in the way of
-development projects that aim to establish HDPs in LMICs. The purpose of
-this paper, therefore, is to specificy how new standards and
+development projects that aim to establish OHDPs in LMICs. The purpose
+of this paper, therefore, is to specificy how new standards and
 technologies can be integrated into the OpenHIE architecture such that
-an open HDP can be realized that supports the following different types
+an open OHDP can be realized that supports the following different types
 of data sharing.
@@
 In this paper, we present a design that extends the OpenHIE
-specification to include the four types of data sharing mentioned above.
-Using the full-STAC approach \citep{mehl2023fullstac} we combine open
-standards, open technologies and open architectures into a coherent
-modular HDP platform that can be configured and reused across a variety
-of use-cases. We employ a formative, naturalistic evaluation to assess
-the technical risk and efficacy of the design \citep{venable2016feds}.
-Given that it is prohibitively expensive to evaluate the design a
-real-world setting, we aim to minimize technological risks and maximize
-the efficacy of the design by considering three examples of health data
+specification to include the three types of data sharing mentioned
+above. Using the full-STAC approach \citep{mehl2023fullstac} we combine
+open standards, open technologies and open architectures into a coherent
+modular OHDP that can be configured and reused across a variety of
+use-cases. We employ a formative, naturalistic evaluation to assess the
+technical risk and efficacy of the design \citep{venable2016feds}. Given
+that it is prohibitively expensive to evaluate the design a real-world
+setting, we aim to minimize technological risks and maximize the
+efficacy of the design by considering three examples of health data
 platforms in LMICs, namely:
@@
   What is the level of openness of the implementation, specifically in
-  terms of the four types of data sharing and platform-to-platform
+  terms of the three types of data sharing and platform-to-platform
   opennenss?
 \item
-  How are the core maintenance functions of the HDP implemented, where
+  How are the core maintenance functions of the OHDP implemented, where
   we particularly focus on functions related to metadata management,
@@
 
+\subsection{Open content}\label{open-content}
+
+\begin{itemize}
+\tightlist
+\item
+\end{itemize}
+
 \section{Evaluation}\label{evaluation}
 
-\subsection{OpenHIM platform}\label{openhim-platform}
+\subsection{Jembi OpenHIM platform}\label{jembi-openhim-platform}
 
@@
 
-\subsubsection{Level of opennness}\label{level-of-opennness}
-
-TO DO: evaluate openness of OpenHIM platform
-
-\subsubsection{Core maintenance
-functions}\label{core-maintenance-functions}
-
-TO DO: evaluate maintenance functions
-
-\subsubsection{Downward scaleability}\label{downward-scaleability}
-
-TO DO: evaluate downward scalability
-
-\subsection{ONA Canopy}\label{ona-canopy}
-
-TO DO: add section describing ONA Canopy.
-
-\subsubsection{Level of opennness}\label{level-of-opennness-1}
-
-TO DO: evaluate openness of OpenHIM platform
-
-\subsubsection{Core maintenance
-functions}\label{core-maintenance-functions-1}
-
-TO DO: evaluate maintenance functions
-
-\subsubsection{Downward scaleability}\label{downward-scaleability-1}
-
-TO DO: evaluate downward scalability
-
-\subsection{Momcare programme}\label{momcare-programme}
+\subsection{ONA OpenSRP 2}\label{ona-opensrp-2}
+
+Continuing our evaluation of the extended OpenHIE architecture, we can
+see a different flavor in the implementation driven by Ona. Ona is a
+social enterprise that has pioneered the adoption of FHIR data standard
+via the development of OpenSRP2, a FHIR-based data collection app built
+using Google's FHIR SDK and focused on enabling offline-first workflows
+for community-based care. OpenSRP 2 is a complete rewrite of the
+original OpenSRP application, a global public good maintained by Ona and
+deployed in XX countries worldwide.
+
+OpenSRP2 applications are currently implemented in the field in three
+countries (Uganda, Liberia, and Madagascar) in collaboration with local
+Ministries of Health and with international donors such as UNICEF,
+supporting a variety of different workflows including antenatal care
+(ANC), postnatal Care (PNC), immunization, and last-mile logistics.
+Besides the OpenSRP Android application and HAPI-FHIR backend, in each
+of its projects Ona also implements a companion set of tools that
+support analytics and various reporting needs.
+
+\subsubsection{Requirements for data
+sharing}\label{requirements-for-data-sharing}
+
+Based on years of work in global health, Ona has learned that the data
+stack implemented to support a national-scale implementation of its
+FHIR-based application responds to the following requirements.
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2857}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7143}}@{}}
+\caption{Requirements and rationale for open health data platform
+developed and used by ONA, based on OpenSRP 2
+(\url{https://opensrp.io/}).}\label{tbl-ona-requirements}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Requirement
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Rationale
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Requirement
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Rationale
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Ingest data from multiple sources, both FHIR and non-FHIR based. & While
+most health record data can be collected and aggregated in FHIR,
+Ministries of Health rely on other data sources to govern their
+operations. For example, operationalizing an immunization campaign
+usually includes tracking against specific targets for locations to be
+visited on specific days and number of children to immunize per day.
+Such targets are often stored in spreadsheets or other applications
+where the data is not FHIR. \\
+Ingest data in batches. & Most data ingestion can happen in batches,
+since Ona's applications are deployed in hard to reach areas where
+connectivity is an issue. Data ingestion closer to real-time can be
+relevant for disaster-response and other time-sensitive applications,
+but this is not a priority. \\
+Support national-scale data volumes. & A data store that can grow from
+dozens to thousands of devices and where data can be aggregated up to
+the national level, matching the scale of implementation of data
+collection applications in the field. \\
+Pre-compute complex business metrics. & Reporting on health systems
+requires pre-computing complex metrics and often performing cohort
+analyses to map trends in service provision. For example, understanding
+quality of care for children requires computing metrics such as the
+percentage of children fully immunized on schedule (i.e.~children 6-59
+months that have received the set of vaccines required by the Ministry
+of Health, and have received each of those vaccines within the expected
+age-window). For Business Intelligence applications, calculating such a
+vital metric cannot be performed at run time, to avoid long and
+expensive queries. \\
+Outbound integrations. & While aggregated data and reports should be
+accessible by other applications such as BI platforms via pulls, there
+should be an easy integration framework to push data to other
+applications used by the Ministry of Health for other purposes, such as
+DHIS2 for health systems management or RapidPro for communications with
+program beneficiaries. \\
+Open source and easily deployable in-country. & Given the extremely
+sensitive nature of health data, it is paramount for governments to have
+the flexibility to deploy the stack in various different environments,
+both on premise and in private clouds. \\
+\end{longtable}
+
+The architecture Learning from experience in the field and internal
+research and development, Ona has developed preferences for a specific
+data stack responding to the aforementioned requirements.
+
+{[}{[}graphic{]}{]}
+
+Core toolings in the stack include: Data ingestion with Airbyte. Ona
+uses Airbyte as the primary data ingestion tool, leveraging the wide
+array of connectors that come standard with the application as well as a
+dedicated suite of connectors developed internally by Ona, including
+HAPI FHIR, RapidPro, Ona Data, Kobo Toolbox and others.\\
+Data storage with Clickhouse. While different health projects have
+varying requirements, Ona has found success in using Clickhouse as the
+main analytics data store in its most recent implementations. Clickhouse
+supports the scale required for analytics at a national level, as well
+as the speed that enables cross-application integrations and more
+real-time analytics. For example, in Madagascar Ona uses its reporting
+suite to identify facilities with stock in need of maintenance and can
+trigger the scheduling of a maintenance visit ad hoc. Data
+transformation with dbt. Following global best practice, Ona leverages
+dbt to segregate the data warehouse in different levels (staging, marts,
+metrics), as well as pre-computing complex indicators for ease of
+reporting and for transmissions into other systems. For example, in
+Liberia Ona implements OpenSRP at community health worker level, but can
+aggregate immunization data at facility level in the data warehouse and
+then push quarterly summary metrics to DHIS2. No recommendation on
+reporting / BI tooling. Ona recognizes that business users have their
+own strong preferences for BI tooling, and some already have licenses
+for specific software, so the architecture is flexible to provide easy
+connections to different BI tools.
+
+Evaluation
+
+Evaluating the data stack, we see the following: Use of generic
+best-of-breed tooling. Ona focused on utilizing Open HIE tools that are
+widely adopted outside of the global health and development sectors.
+This approach aims to provide assurance on two main fronts, the ability
+to handle performance at scale and the long term dependability of the
+tools, rather than relying on smaller projects with uncertain long term
+funding or unproven implementations. Columnar data warehouse for
+analytics. The scale of Ona's project requires the implementation of a
+dedicated database for analytics. While original data can still be
+stored as parquet or other file system, being able to ingest it into a
+relational data store allows to create well defined indicators. Using
+clickhouse as a tool helps and combine the need accuracy with the speed
+of reporting as new data is ingested.\\
+Strong emphasis on SQL. While Ona has tested and experimented with
+FHIR-specific tooling, such as the definition of data projections using
+sql-on-fhir, Ona found that relying on sql for coding business logic
+remained the faster and most scalable approach.
+
+In summary, for Ona building analytics with FHIR data looks similar to
+building analytics with any other type of data. While FHIR provides a
+clear and standard data model, managing information for most health
+systems requires custom integration of data between different sources,
+as well as computing indicators using business logic specific to the
+needs of the local users. Building upon well established best-of-breed
+tools allows Ona to implement FHIR applications at scale and provide
+trusted analytics on top.
+
+\subsection{PharmAccess demonstrator Momcare
+programme}\label{pharmaccess-demonstrator-momcare-programme}
 
@@
 objectives of this work were in fact to demonstrate a solution design
-that could support the first three types of data sharing listed in
-\textbf{?@tbl-types-data-sharing}. First, to investigate the viability
-of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed
-to assess the complexity and effort required to implement the facade
-pattern to integrate the legacy system into the FHIR data standard.
-Using the longitudinal dataset from approximately 28 thousand patient
-records, FHIR transformations script were developed and deployed using
-the mediator function of the IOL. The data was transformed into 10 FHIR
-v4 resources and the conceptual data model of the existing MomCare app
-could readily be transformed into the FHIR standard using SQL and
-validated with a Python library \citep{islam2023fhir}. The largest
-challenge during the transformation process pertained to the absence of
-unique business identifiers for patients and healthcare organizations.
-For patients, either the mobile phone number or the healthcare insurance
-number was taken, depending on availability. A combination of name,
-address and latitude/longitude coordinates were used to uniquely
-identify organizations and locations, as Tanzania does not have a system
-in place for this purpose.
+that could support the first three types of data sharing. First, to
+investigate the viability of using FHIR for bulk data sharing, MomCare
+Tanzania was used a testbed to assess the complexity and effort required
+to implement the facade pattern to integrate the legacy system into the
+FHIR data standard. Using the longitudinal dataset from approximately 28
+thousand patient records, FHIR transformations script were developed and
+deployed using the mediator function of the IOL. The data was
+transformed into 10 FHIR v4 resources and the conceptual data model of
+the existing MomCare app could readily be transformed into the FHIR
+standard using SQL and validated with a Python library
+\citep{islam2023fhir}. The largest challenge during the transformation
+process pertained to the absence of unique business identifiers for
+patients and healthcare organizations. For patients, either the mobile
+phone number or the healthcare insurance number was taken, depending on
+availability. A combination of name, address and latitude/longitude
+coordinates were used to uniquely identify organizations and locations,
+as Tanzania does not have a system in place for this purpose.
 
@@
 
-\subsubsection{Level of opennness}\label{level-of-opennness-2}
+\subsubsection{Level of opennness}\label{level-of-opennness}
 
@@
 \subsubsection{Core maintenance
-functions}\label{core-maintenance-functions-2}
+functions}\label{core-maintenance-functions}
 
@@
 
-\subsubsection{Downward scaleability}\label{downward-scaleability-2}
+\subsubsection{Downward scaleability}\label{downward-scaleability}
 
@@
 
-We specifically address the notion of openness of HDPs in LMICs in terms
-of the design-related questions put forward by de Reuver at al.11:
+We specifically address the notion of openness of OHDPs in LMICs in
+terms of the design-related questions put forward by de Reuver at al.11:
 
@@
   Unit of analysis: what is platform-to-platform openness in the context
-  of data platforms, given the expectation that different HDPs will
+  of data platforms, given the expectation that different OHDPs will
   emerge at various aggregation levels? How do we distinguish
@@
 FL & Federated learning \\
-HDP & Health data platform, explicitly differentiated from health
+OHDP & Health data platform, explicitly differentiated from health
 digital platform \\

### 2024-08-13 rewrite.qmd
wip

@@
     code-fold: true
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: JIMR
-      cite-style: authoryear
+  pdf: default
+    # keep-tex: true
+    # journal:
+    #   name: JIMR
+    #   cite-style: authoryear
 author:
@@
 
+### The global digital health ecosystem is still project-centric
+
 Digital transformation have the potential to make inclusive health systems in low- and middle income countries (LMICs) a reality by connecting patients and professionals, using data as a currency of trust.
@@
 
-In this context it is essential to have a good understanding of digital platforms and their associated ecosystem as an essential enabling socio-technical infrastructure upon  which digital health interventions can be realized in a more standardized, scaleable and sustainable fashion.
+To achieve scaleability of digital health interventions, we need to conceptualize and implement digital platforms and their associated ecosystems that constitute an important enabling socio-technical infrastructure upon which such interventions can be realized in a more standardized and sustainable fashion.
 Digital platforms are a challenging research object because of their distributed nature and intertwinement with institutions, markets and technologies [@dereuver2018digital].
 Digital platform ecosystems vary according to three core building blocks: (1) platform ownership, (2) value-creating mechanisms, and (3) complementor autonomy [@hein2020digital].
-Digital platforms have disrupted many sectors but have just started to make inroads into highly regulated industries such as healthcare [@ozalp2022digital].
+Various studies have investigated why the healthcare domain has been slow in adopting platforms and their ecosystems [@hermes2020digital;@ozalp2022digital].
+We highlight three paradigmatic issues that are relevant to the work presented in this paper.
+
+### From centralized to decentralized platforms
+
+Digital platforms have predominantly flourished in a business setting, as is exemplified by the well known success stories of cloud platforms providers (AWS, Microsoft, Google, Alibaba), two-sided market platforms (Airbnb, Spotify, Uber) and social media platforms (Meta, WeChat).
+While these examples allude to the potentional to generate social and economic value by using platforms as a key enabler of digital transformation of healthcare in LMICs, it is less obvious in what ways a digital platform can trigger specific positive effects *for* development [@bonina2021digital].
+Part of the problem in grasping the implications of platforms for development in the field of information systems has been the lack of clarity regarding the understanding of what digital platforms are, what their main features are and how they generate value [@dereuver2018digital].
+Notably, even valuable landscape papers in information systems largely ignore the development implications of digital platforms [@dereuver2018digital] or may summarize them in rather naïve ways [@bonina2021digital].
+Put more strongly, issues pertaining to the winner-takes-all nature of digital platforms have resulted in so-called transgression of spheres where commercial platform operators acquire disproprotionately large market power which is at odds with realizing public value [@sharon2018when].
+
+It almost seems that we have forgotten that digital platforms need not always be centralized, but that we have a choice to implement decentralized or distributed platforms.
+In his seminal paper in 1963, Baran distinguished three types of network topologies ([@fig-network-topologies]).
+In a centralized network, all communication is facilitated through a single, central hub, which is obviously vulnerable as "...destruction of single node destroys communication between the end stations" [@baran1964distributed].
+At the opposite end of the spectrum, in a fully distributed network stations are connected to all adjacent stations.
+In practice, a mixture of centralized and distributed components is used to form a so-called decentralized network.
+Complete reliance upon a single point is no longer required, while at the same time where is a hierarchy where smaller centralized networks are connected to higher-lying networks through interconnecting hubs (encircled in blue).
+Note that the internet has both the characteristics of a distributed network (for example, traffic and data processing are shared across multiple paths and servers) as well as that of a decentralized network (there is no single entity that controls the entire internet.
+Instead, numerous internet service providers (ISPs), organizations, and governments manage different parts of the network).
+
+![Three network topologies, taken from Baran (1964).
+The nodes encircled in blue in the decentralized network are hubs that provide connections between sub-networks.](images/network-topologies.png){#fig-network-topologies}
+
+We are inspired by various fields of study that have shown that decentralized and/or distributed systems are often 'resilient by design' [@fiksel2003designing;@liu2010architectural;@mayar2022resilience], for example, allowing efficient deployments at different geographic scales.
+Given the LMICs are resource-constrained, we consider downward scaleability as a key design criterium [@fayad2005scalable;@duboc2006framework].
+
+, one of the main drivers of the work presented here is to conceptualize and implement a platform that is conducive to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].
+With this approach, we purposefully challenge the dominant paradigm of designing and implementing centralized platforms to support the digital transformation of healthcare in LMICs
+
+### From ... to ... (openness)
+
+### From digital platforms to data platforms
+
+This paper is motivated by the conflation of a number of developments relevant to the design and implementation of open, solidarity-based OHDPs in LMICs.
+First, the OpenHIE framework [@openhie] has been adopted by many sub-Saharan African countries [@mamuye2022health] as the architectural blueprint for implementing nation-wide health information exchanges (HIE), including Nigeria [@dalhatu2023paper], Kenya [@thaiya2021adoption] and Tanzania [@nsaghurwe2021one].
+These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain.
+The rationale for this addition is to facilitate secondary reuse of health data for academic research, real-world evidence studies etc.
+which can be framed within the context of ongoing efforts towards Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair].
+In doing so, however, we have implicitly moved from conceptualizing digital health platforms for primary data sharing (the original OpenHIE specification) to health data platforms for secondary data sharing.
+This is problematic because the notion of openness, which is assumed to be essential in establishing solidarity-based approaches to data sharing, is inherently different for a data platform compared to a digital platform.
+
+Conceptually, the OpenHIE framework constitutes a framework for an open digital platform.
+Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods [@digitalpublicgoods].
+If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for secondary data sharing and reuse.
+The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish OHDPs in LMICs.
+The purpose of this paper, therefore, is to specificy how new standards and technologies can be integrated into the OpenHIE architecture such that an open OHDP can be realized that supports the following different types of data sharing.
+
+#### 1. Sharing of data sets at the most granular (patient) level, persisted as longitudinal records
+The Shared Health Record (SHR) in OpenHIE provides an operational, real-time transactional data source which is intended for primary data sharing.
+The specification explicitly states that the SHR is distinct from a datawarehouse.
+At the same time, the Health Management Information System (HMIS) component is specified to fulfil functional requirements that are typically provided by datawarehouses.
+
+- **Q1:** How can proven solution designs and modern technologies from data warehousing and engineering be integrated in the OpenHIE specification and how does this relate to the curren specifications of the SHR and HMIS.
+
+#### 2. Sharing of data products derived from the original data set
+OpenHIE specifies that the Health Management Information System (HMIS) component should support a workflow to validate and save aggregate data based on the emerging [IHE Aggregate Data Exchange (ADX) standard](https://www.ihe.net/uploadedFiles/Documents/QRPH/IHE_QRPH_Suppl_ADX.pdf).
+This is just one specific workflow within a larger data analytics value chain that covers i) collection, ii) standardization, iii) cleaning, iv) storage, v) analysis and vi) distribution [@curry2016big].
+Sharing of benchmarking, and decision-support tools based on various statistical (machine learning) models that have been trained on data sets and/or data product, can be shared for standalone use.
+
+- **Q2:** How can proven solutions designs from the data warehousing and engineering community be included in OpenHIE to support sharing of wider variety of data products.
+
+#### 3. Sharing of secure computational environments to access and work with the data
+
+Current solution design of centralized data storage is increasingly being challenged.
+Recent work points to federated learning (FL) [@rieke2020future] and privacy- enhancing technologies (PETs) [@scheibner2021revolutionizing;@jordan2022selecting] as a better long-term solution to secure and equitable sharing of data.
+Given the resource constrained environments of LMICs, is is particularly challenging to devise a solution that is economically feasible and ...
+
+### Design principles
+
+The work presented there:
+
+- inclusive-by-design, based on the notions of datasolidarity and maximising autonomy of participants in the ecosystem
+- scaleable by design, particularly focusing on downward scaleability to support a decentralized platform topology that supports a hierarchy of platforms of local care networks --> county-level networks --> national networks
+-  fits a socio-political context where counties are enabled to implement regional datwhich mirrors the administrative organizations of
+
+## Points to ponder
+
+- Most succesful examples of digital platforms have a private keystone actor
+-
 

### 2024-08-14 index.qmd
Publish full rewrite

@@
 ---
-title: "Conceptualizing open health data platforms for low- and middle income countries"
+title: "Conceptualizing and implementing open, decentralized health data sharing platforms for low- and middle income countries"
 format:
@@
 
-## Introduction {#sec-intro}
+## Introduction {#sec-introduction}
 
-### The paradox of open for digital vs. data platforms in healthcare
+### The global digital health ecosystem is still project-centric
 
-It is a widely held belief that digital technologies have an important role to play in strengthening health systems in low- and middle income countries (LMICs), as exemplified by the WHO global strategy on digital health [@who2021global].
-The adoption rate of mobile phones in LMICs has been an important driver in implementing digital health solutions [@mccool2022mobile].
-Yet, there are many shortcomings and challenges, including the current fragmentation of digital platforms and the lack of clear-cut pathways of scaling up digital health programmes, such that they can support sustainable and equitable change of national health systems in LMICs [@mehl2023fullstac;@mccool2022mobile;@who2019recommendations;@neumark2021digital].
+Digital transformation have the potential to make inclusive health systems in low- and middle income countries (LMICs) a reality by connecting patients and professionals, using data as a currency of trust.
+Adoption of digital technologies in healthcare means more than simply increasing efficiency or cutting costs.
+Instead, digital technologies have the potential to i) increase the availability, accessibility, acceptability, and quality of health services as we know them; ii) change nature and direction of health care, and the possibility of making it more preventive, personalised, and mobile through the use of such technologies; and iii) enfranchise patients and communities (and particularly vulnerable groups including children and young people) and evolve their relationship with health professionals and providers, thus helping shape the health system according to the needs of the patients and communities [@kickbusch2021lancet].
 
-A commonly used perspective to scrutinize digital health is to consider it as a digital platform [@dereuver2018digital].
-Digital platforms have disrupted many sectors but have just started to make inroads into highly regulated industries such as healthcare [@ozalp2022digital].
-In this light, the challenges faced by LMICs in establishing national digital health platforms have a lot in common with those faced by high income countries.
-From a technological perspective, interoperability issues, weak integrations, siloed data repositories and overall lack of openness are often reported as key impediments [@malm-nicolais2023exploring;@mehl2023fullstac].
-From a societal perspective, issues pertaining to the winner-takes-all nature of digital platforms are hotly debated as many jurisdictions make work to ensure these new digital health platforms indeed serve the common good of achieving universal health coverage [@sharon2018when].
+Yet, to date the global digital health ecosystem is still project-centric, resulting in data fragmentation and technology lock-in, compromising health care delivery [@mehl2023fullstac].
+There is no dearth of the number of digital health interventions: sub-Saharan Africa has seen over 700 such projects in the past 10 years.
+There is, however, a fundamental lack in coordination, integration, scalability, sustainability, and equitable distribution of investments in digital health.
+Health policymakers and the global health community at large need to urgently institute coordination mechanisms to terminate unending duplication and disjointed vertical implementations and manage solutions for scale [@karamagi2022ehealth].
+
+To achieve scaleability of digital health interventions, we need to conceptualize and implement digital platforms and their associated ecosystems that constitute an important enabling socio-technical infrastructure upon which such interventions can be realized in a more standardized and sustainable fashion.
+Digital platforms are a challenging research object because of their distributed nature and intertwinement with institutions, markets and technologies [@dereuver2018digital].
+Digital platform ecosystems vary according to three core building blocks: (1) platform ownership, (2) value-creating mechanisms, and (3) complementor autonomy [@hein2020digital].
+Various studies have investigated why the healthcare domain has been slow in adopting platforms and their ecosystems [@hermes2020digital;@ozalp2022digital].
+Before presenting our main contributions, we highlight two paradigmatic themes that are relevant to our paper.
+
+### From centralized to decentralized platforms
+
+Digital platforms have predominantly flourished in a business setting, as is exemplified by the well known success stories of cloud platforms providers (AWS, Microsoft, Google, Alibaba), two-sided market platforms (Airbnb, Spotify, Uber) and social media platforms (Meta, WeChat).
+While these examples allude to the potentional to generate social and economic value by using platforms as a key enabler of digital transformation of healthcare in LMICs, it is less obvious in what way a digital platform can trigger specific positive effects *for* development [@bonina2021digital].
+Part of the problem in grasping the implications of platforms for development in the field of information systems has been the lack of clarity regarding the understanding of what digital platforms are, what their main features are and how they generate value [@dereuver2018digital].
+Notably, even valuable landscape papers in information systems largely ignore the development implications of digital platforms [@dereuver2018digital] or may summarize them in rather naïve ways [@bonina2021digital].
+Put more strongly, issues pertaining to the winner-takes-all nature of digital platforms have resulted in so-called 'transgression of spheres' where commercial platform providers acquire disproprotionately large market power which is at odds with realizing public value for the common good [@sharon2018when].
+
+It almost seems we have forgotten that digital platforms need not always be centralized, but that we have a choice to implement decentralized or distributed platforms.
+In his seminal paper in 1963, Baran distinguished three types of network topologies ([@fig-network-topologies]).
+In a centralized network, all communication is facilitated through a single, central hub, which is obviously vulnerable as "...destruction of single node destroys communication between the end stations" [@baran1964distributed].
+At the opposite end of the spectrum, in a fully distributed network, stations are connected to all adjacent stations.
+In practice, a mixture of centralized and distributed components is the most commonly used topology, aptly called decentralized networks as they remove complete reliance upon a single point, while at the same time where is a hierarchy where smaller centralized networks are connected to higher-lying networks through interconnecting hubs (encircled in blue).
+Note that the internet has both the characteristics of a distributed network (for example, traffic and data processing are shared across multiple paths and servers) as well as that of a decentralized network (there is no single entity that controls the entire internet.
+Instead, numerous internet service providers (ISPs), organizations, and governments manage different parts of the network).
+
+![Three network topologies, taken from Baran (1964).
+The nodes encircled in blue in the decentralized network are hubs that provide connections between sub-networks.](images/network-topologies.png){#fig-network-topologies}
+
+We are inspired by various fields of study that have shown that decentralized and/or distributed systems are often 'resilient by design' [@fiksel2003designing;@liu2010architectural;@mayar2022resilience], for example, allowing efficient deployments at different geographical scales.
+Given that LMICs are resource-constrained, we consider downward scaleability as a key design criterium [@fayad2005scalable;@duboc2006framework].
+We consider the insights from the 'hourglass model', which provides clear design principles how the use of minimal standards in a layered, distributed systems allows implementation information systetms using a great diversity of services [@beck2019hourglass].
+
+We also posit that a decentralized platform is more conducive to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].
+With this approach, we purposefully challenge the dominant paradigm of designing and implementing centralized platforms to support the digital transformation of healthcare in LMICs [@ogundaini2022systematic] with the aim to make digital platforms work *for* development [@hermes2020digital].
+
+### From open source to open data sharing
 
@@
 to consider data platforms as a phenomenon distinct from digital platforms [@dereuver2022openness].
-Generally, data platforms inherit the characteristics of digital platforms.
-From an economic perspective, for example, both types exhibit multi-sided markets.
-At the same time, data platforms differ as their main offerings revolve around data.
-From an ecosystem perspective, data platforms have more moderate network effects and are more susceptible to fragmentation and heterogeneity [@dereuver2022openness].
+@tbl-comparing-platforms compares the key characteristics of digital vs.
+data platforms.
+The concept of health data spaces, currently under development in Europe, is an archetypical example of a data platform [@otto2022designing;@shabani2022will].
 
-Particularly relevant in the context of open health data platforms (OHDPs), is the conceptualization of openness.
+| | Digital platforms (e.g. software, gaming) | data platforms (e.g. data marketplaces) |
+|:---|:-----|:-----|
+| **User groups** | Developers, Consumers | Data consumers, Data owners, Solution providers |
+| **Object of openness** | Software/hardware modules | Data, aggregated data, trained models; Data analytics modules |
+| **Ways to realize openness** | Open source, open boundary resources (APIs) | Data-driven mechanisms (secure computational environments) |
+| **Market consolidation** | Strong network effects; Dominant cross-industry platforms | Moderate network effects; Fragmentation and heterogeneity |
+
+: Comparing digital and health data platforms, taken from de Reuver et al. (2022) {#tbl-comparing-platforms .striped .hover .responsive}
+
+To illustrate the relevance of conceptualising health platforms as data platforms, consider the OpenHIE framework [@openhie] which has been adopted by many sub-Saharan African countries as the architectural blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one].
+The OpenHIE standard is best described as a digital platform as it is designed to facilitate information exchange, using specific workflows and information standards (open boundary resources in @tbl-comparing-platforms), with a primary focus is on creating a cohesive infrastructure that supports the interaction and integration of various health information systems.
+Since its inception in 2013, the ICT for Development community (ICT4D) has pushed for the development of open source implementations of OpenHIE components, which are now available as digital public goods [@digitalpublicgoods].
+@tbl-digital-public-goods list the currently available open source components that adhere to the OpenHIE specification.
+
+| Category  | Component | Digital Public Good |
+|:----------|:----------|:--------------------|
+| **Interoperability layer (IOL)** | IOL | [OpenHIM](http://openhim.org) |
+| | | [mHero](https://www.mhero.org/technology) |
+| | | [OpenFN](https://docs.openfn.org/) |
+| **Registry Services** | Client Registry (CR) | [SanteMPI](https://help.santesuite.org/product-overview/santesuite-products/master-patient-index-santempi) |
+| | | [JeMPI](https://jembi.gitbook.io/jempi/) |
+| | | [OpenCR](https://www.openclientregistry.org/) |
+| | | [OpenCRVS](https://www.opencrvs.org/) |
+| | Facility Registry (FR) | [Global Open Facility Registry (GOFR)](https://github.com/intrahealth/gofr) |
+| | | [GeoPrism Registry (GPR)](https://geoprismregistry.com/) |
+| | | [Resource Map](http://instedd.org/technologies/resource-map/) |
+| | | [Healthsite.io](https://healthsites.io/) |
+| | | [GeoPrism Registry](https://geoprismregistry.com/) |
+| | | [DHIS2](https://github.com/dhis2) |
+| | Health Worker Registry (HWR) | [iHRIS](https://www.ihris.org/ihris-50) |
+| | Terminology Service (TS) | [OCL Terminology Service](https://openconceptlab.org/terminology-service/) |
+| | Product Catalog (PC) | [PCMT](https://productcatalog.io/) |
+| **Business Domain Services** | Shared Health Record (SHR) | [HAPI FHIR](https://hapifhir.io/) |
+| | | [Fhirbase](https://www.health-samurai.io/fhirbase) |
+| | | [FHIR Server for Azure](https://github.com/microsoft/fhir-server) |
+| | | [I-TECH-UW SHR](https://github.com/I-TECH-UW/shared-health-record) |
+| | Health Management Information System (HMIS)[^1] | [DHIS2](http://dhis2.org/) |
+| | Finance and Insurance Service (FIS) | [OpenIMIS](http://openimis.org/) |
+| | Logistics Management Information System (LMIS) | [OpenLMIS](http://openlmis.org/) |
+| | | [OpenBoxes](https://openboxes.com/features/) |
+| **Generic** | Identity and Access Management (IAM) | [Keycloak](https://www.keycloak.org/) |
+| | Gateway & proxy | [FHIR Information Gateway](https://github.com/google/fhir-gateway) |
+| | Admin dashboard for SHR | [OpenSRP FHIR Web](https://github.com/onaio/fhir-web) |
+| | Configuration and deployment | [Instant OpenHIE](https://jembi.gitbook.io/instant-v2/) |
+
+: Overview of current open source implementations of components that fit in the OpenHIE specification.
+Point-of-Service systems are excluded for brevity.
+List compiled using [OpenHIE Reference Technologies](https://wiki.ohie.org/display/DR/OpenHIE+Reference+Technologies), [Global Goods for Digital Health](https://digitalsquare.org/digital-health-global-goods), [Digital Public Goods Alliance](https://digitalpublicgoods.net/) and search of open source code repositories.
+A systematic review of such digital public goods is beyond the scope of this paper.
+{#tbl-digital-public-goods .striped .hover .responsive}
+
+[^1]: Note that there is often a confusion on the acronym HMIS.
+Strictly speaking, the OpenHIE specification uses HMIS to refer to a Health Management Information System that is part of the Business Domain Services.
+Sometimes HMIS is used to refer to a Hospital Management Systems in the Point-of-Service domain, synonymous with an Electronic Medical Record (EMR) system.
+
 The shift in perspective from digital platforms to data platforms coincides with the paradox of open [@keller2021paradox].
-Originally, openness of digital platforms focused on open source, open standards and copyrights, which by has been superseded by “...
+Originally, openness of digital platforms focused on open source and open standards (as shown above for OpenHIE) which by has been superseded by “...
 conflicts about privacy, economic value extraction, the emergence of artificial intelligence, and the destabilizing effects of dominant platforms on (democratic) societies.
 Instead of access to information, the control of personal data has emerged in the age of platforms as the critical contention.” [@keller2021paradox].
-These conflicts are particularly salient in the healthcare domain, where people are generally willing to share their health data to receive the best care (primary use), while the attitude towards secondary use of health data varies greatly depending on the type and context [@cascini2024health].
+These conflicts are particularly salient in the healthcare domain, where people are generally willing to share their health data to receive the best care (primary use, which is aligned with the concept of digital platforms), while the attitude towards secondary use of health data (conceptually aligned with a data platform) varies greatly depending on the type and context [@cascini2024health].
 The shift in perspective from digital platforms supporting primary data sharing toward data platforms supporting secondary data sharing is one of the key issues surrounding the polemic of data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond;@purtova2023data].
-Openness is particularly relevant if we are to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].
 
-### From health information exchanges to health data platforms
+This shift in perspective also has real implications for the design and implementation of OpenHIE.
+We observe that some African countries have, as a matter of course, extended the OpenHIE framework to include “data & analytics services” as an additional domain in their policies [@mbugua2021adoption;@dalhatu2023paper].
+The rationale for this addition is indeed to facilitate secondary reuse of health data for academic research, real-world evidence studies etc.
+using the principles of Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair].
+In doing so, however, we have implicitly moved from conceptualizing digital health platforms for primary data sharing (the original OpenHIE specification) to health data platforms for secondary data sharing.
+If we are to use the OpenHIE framework as an open data platform as well, we need to extend the standards, technologies and architecture to include functionality for secondary data sharing and reuse.
+The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish more comprehensive platforms to support primary and secondary health data sharing in LMICs.
+One of the motivations for this paper is integrate recent standards and open source implementations from the data engineering community into the OpenHIE framework.
 
-This paper is motivated by the conflation of a number of developments relevant to the design and implementation of open, solidarity-based OHDPs in LMICs.
-First, the OpenHIE framework [@openhie] has been adopted by many sub-Saharan African countries [@mamuye2022health] as the architectural blueprint for implementing nation-wide health information exchanges (HIE), including Nigeria [@dalhatu2023paper], Kenya [@thaiya2021adoption] and Tanzania [@nsaghurwe2021one].
-These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain.
-The rationale for this addition is to facilitate secondary reuse of health data for academic research, real-world evidence studies etc.
-which can be framed within the context of ongoing efforts towards Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair].
-In doing so, however, we have implicitly moved from conceptualizing digital health platforms for primary data sharing (the original OpenHIE specification) to health data platforms for secondary data sharing.
-This is problematic because the notion of openness, which is assumed to be essential in establishing solidarity-based approaches to data sharing, is inherently different for a data platform compared to a digital platform.
+### Conceptualizing and implementing open health data sharing platforms
 
-Conceptually, the OpenHIE framework constitutes a framework for an open digital platform.
-Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods [@digitalpublicgoods].
-If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for secondary data sharing and reuse.
-The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish OHDPs in LMICs.
-The purpose of this paper, therefore, is to specificy how new standards and technologies can be integrated into the OpenHIE architecture such that an open OHDP can be realized that supports the following different types of data sharing.
+In the light of these nascent paradigm shifts, the objective of this paper is the following.
+Given high-level design principles that we believe are supportive of realising solidarity-based, namely [TO DO: this is just first shot at formulating our design principles; needs refinement]:
+
+- inclusive-by-design, based on the notions of datasolidarity and maximising autonomy of all future participants in the ecosystem;
+- scaleable-by-design, particularly focusing on downward scaleability to support a decentralized platform topology to allows for bottom-up deployment scenarios (from local care networks --> county-level networks --> national networks) instead of top-down national roll-out;
+- open-by-design, whereby a balance is found to resort to minimal standards and allow for a large diversity of partners and technologies to be used;
+
+we seek to specifiy and demonstrate how this can be implemented using the full-STAC approach [@mehl2023fullstac], that consists of:
+
+1. FHIR as the main open standard
+2. the Instant HIE framework for integrating open technologies
+3. an extended version of OpenHIE as the open architecture that includes a more detailed architecture for the data platform
+4. open content with various demonstrators in the domain of antenatal and maternal care (ANC) as a first version of hands-on digital adaptation kits.
 
@@
 Given the resource constrained environments of LMICs, is is particularly challenging to devise a solution that is economically feasible and ...
-[TO DO: add feedback from Mark vdG]
-
-Federated solutions:
-
-- KETOS OMOP-FHIR [@gruendner2019ketos]
-- Personal Health Train on FHIR [@choudhury2020personal]
-- OHDSI analytics [@khalid2021standardized]
-- CODA project [@mullie2023coda]
-- GenoMed4All, in evalution phase for -omis [@cremonesi2023need]
-
-Similar approaches taken in the context of data sharing for pandemic response, for example:
-
-- VODAN Africa, based on electronic data capture with templates [@purnamajati2022data]
-- Global Data Sharing Initiative (GDSI) [@pirmani2023journey]
-
-- **Q3:** How can a solution be designed that is economically viable, can be deployed in a bottom-up fashion, support data governance principles ...
 
@@
 
-## Design
+## Results
 
-### Open standards: using FHIR as the common data model
+### Design
+
+#### Open standards: using FHIR as the common data model
 
@@
 
-TO DO: re-use these elements from table 1 into this section:
+TO DO:
+- tie-in paper [@tsafnat2024converge] on why we use FHIR
+- re-use these elements from table 1 into this section:
 
@@
 
-### Open architecture: extending OpenHIE with a composable data stack
-
-#### Evolving the data lakehouse to a composable data stack
+#### Open architecture: extending OpenHIE with a composable data stack
 
@@
 Today, many components of the OpenHIE specification are now available as a digital public goods, as listed in @tbl-digital-public-goods.
-Typically, these open source components are intended to support deployments in small countries (population up to 10 million) or large NGOs out of the box, and should provide a stepping stone for customized deployments in medium-sized countries (population around 40 million).[^1] To further ease the development, configuration and deployment of health information exchanges, the concept of 'Instant OpenHIE' has been championed to (i) allow implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) have a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@InstantOpenHIEv2].
+Typically, these open source components are intended to support deployments in small countries (population up to 10 million) or large NGOs out of the box, and should provide a stepping stone for customized deployments in medium-sized countries (population around 40 million).[^2] To further ease the development, configuration and deployment of health information exchanges, the concept of 'Instant OpenHIE' has been championed to (i) allow implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) have a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@InstantOpenHIEv2].
 Besides the core functional components of the OpenHIE architecture, the Instant OpenHIE toolkit allows packaging and integration of generic components such as Identity and Access Managment (IAM) and a reverse proxy gateway.
@@
 
-[^1]: Although the OpenHIE specification does not include details on dimensioning, these are typically the requirements that are used within the community.
+[^2]: Although the OpenHIE specification does not include details on dimensioning, these are typically the requirements that are used within the community.
 See [OpenHIE Community Wiki](https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements).
-
-| Category  | Component | Digital Public Good |
-|:----------|:----------|:--------------------|
-| **Interoperability layer (IOL)** | IOL | [OpenHIM](http://openhim.org) |
-| | | [mHero](https://www.mhero.org/technology) |
-| | | [OpenFN](https://docs.openfn.org/) |
-| **Registry Services** | Client Registry (CR) | [SanteMPI](https://help.santesuite.org/product-overview/santesuite-products/master-patient-index-santempi) |
-| | | [JeMPI](https://jembi.gitbook.io/jempi/) |
-| | | [OpenCR](https://www.openclientregistry.org/) |
-| | | [OpenCRVS](https://www.opencrvs.org/) |
-| | Facility Registry (FR) | [Global Open Facility Registry (GOFR)](https://github.com/intrahealth/gofr) |
-| | | [GeoPrism Registry (GPR)](https://geoprismregistry.com/) |
-| | | [Resource Map](http://instedd.org/technologies/resource-map/) |
-| | | [Healthsite.io](https://healthsites.io/) |
-| | | [GeoPrism Registry](https://geoprismregistry.com/) |
-| | | [DHIS2](https://github.com/dhis2) |
-| | Health Worker Registry (HWR) | [iHRIS](https://www.ihris.org/ihris-50) |
-| | Terminology Service (TS) | [OCL Terminology Service](https://openconceptlab.org/terminology-service/) |
-| | Product Catalog (PC) | [PCMT](https://productcatalog.io/) |
-| **Business Domain Services** | Shared Health Record (SHR) | [HAPI FHIR](https://hapifhir.io/) |
-| | | [Fhirbase](https://www.health-samurai.io/fhirbase) |
-| | | [FHIR Server for Azure](https://github.com/microsoft/fhir-server) |
-| | | [I-TECH-UW SHR](https://github.com/I-TECH-UW/shared-health-record) |
-| | Health Management Information System (HMIS)[^2] | [DHIS2](http://dhis2.org/) |
-| | Finance and Insurance Service (FIS) | [OpenIMIS](http://openimis.org/) |
-| | Logistics Management Information System (LMIS) | [OpenLMIS](http://openlmis.org/) |
-| | | [OpenBoxes](https://openboxes.com/features/) |
-| **Generic** | Identity and Access Management (IAM) | [Keycloak](https://www.keycloak.org/) |
-| | Gateway & proxy | [FHIR Information Gateway](https://github.com/google/fhir-gateway) |
-| | Admin dashboard for SHR | [OpenSRP FHIR Web](https://github.com/onaio/fhir-web) |
-| | Configuration and deployment | [Instant OpenHIE](https://jembi.gitbook.io/instant-v2/) |
-
-: Overview of current open source implementations of components that fit in the OpenHIE specification.
-The category Analytics Services is not a part of the original OpenHIE and is discussed in this paper.
-Point-of-Service systems are excluded for brevity.List compiled using [OpenHIE Reference Technologies](https://wiki.ohie.org/display/DR/OpenHIE+Reference+Technologies), [Global Goods for Digital Health](https://digitalsquare.org/digital-health-global-goods), [Digital Public Goods Alliance](https://digitalpublicgoods.net/) and search of open source code repositories.
-A systematic review of such digital public goods is beyond the scope of this document.
-{#tbl-digital-public-goods .striped .hover .responsive}
-
-[^2]: Note that there is often a confusion on the acronym HMIS.
-Strictly speaking, the OpenHIE specification uses HMIS to refer to a Health Management Information System that is part of the Business Domain Services.
-Sometimes HMIS is used to refer to a Hospital Management Systems in the Point-of-Service domain, synonymous with an Electronic Medical Record (EMR) system.
 
@@
 
--
+...
 
-## Evaluation
+### Evaluation of implementations
 
-### Jembi OpenHIM platform
+#### Jembi OpenHIM platform
 
@@
 
-### ONA OpenSRP 2
+#### ONA OpenSRP 2
 
@@
 
-#### Requirements for data sharing
+##### Requirements for data sharing
 
@@
 
-### Openness of data platforms
+### Openness of health data sharing platforms
 
@@
   - Risk of openness: difficult to answer ...
+- Paradox of open in disucssion: we started with hypotheses that a decentralized approach will lead to distribution of power, hence ...
+But is this really the case?
+Will open source not backfire and strengthen their position?
 
@@
 
-- Workflow requirements: Report aggregate data (link): receiver is HMIS, mADX
+- Workflow requirements: Report aggregate data (link): receiver is HMIS, mADX; this is not analytics!
 - Functional requirements: https://guides.ohie.org/arch-spec/openhie-component-specifications-1/openhie-health-management-information-system-hmis
@@
 
+## Acknowledgements
+
+## Conflicts of interest
+
+DK received funding from PharmAccess to conduct this work as a contractor.
+SW/IntelliSOFT received funding from PharmAccess as an implementation partner in various projects.
+AP/ONA received funding from PharmAccess to develop a new version of components of the open source OpenSRP 2 framework.
+
 ## Abbreviations
@@
 | CR | Client Registry |
+| DHI | Digital Health Intervention |
 | ELK | Elasticsearch, Logstach and Kibana stack |
@@
 | FL | Federated learning |
-| OHDP	|	Health data platform, explicitly differentiated from health digital platform |
+| HDSP	|	Health data sharing platform |
 | HIE | Health Information Exchange |
+| ICT4D | ICT for Development |
 | IR | Intermediate Representation |

### 2024-08-26 previous-versions/20240822.qmd
wip

--- 2024-08-14 index.qmd
@@
 Given the resource constrained environments of LMICs, is is particularly challenging to devise a solution that is economically feasible and ...
+
+| Level | Description | Number of facilities | 
+|:---:|----------------------|--------:|
+| 2 | Dispensaries and private clincs, typically located in a school, industrial plant or other organization that dispenses medication and sometimes basic medical and dental treatment | 8,806 |
+| 3 | Health centres, medium-sized units which cater for a population of about 80,000 people | 2,559 |
+| 4 | Sub-county hospital, similar to health centres with additional facilities for more complex procedures |   971 |
+| 5 | County referral hospital, regional centres which provide specialised care |    34 |
+| 6 | National referral hospital | 5 |
+
+: Number of healthcare facilities in Kenya. Source: Kenya Health Facility Census, Ministry of Health, September 2023. {#tbl-kenya-facilities}
 

### 2024-08-26 index.qmd
wip

@@
 ---
-title: "Conceptualizing and implementing open, decentralized health data sharing platforms for low- and middle income countries"
+title: "Design and implemention of open health data systems with standardized data and analytics capabilities for low- and middle income countries"
 format:
@@
         country: the Netherlands
+  - name: Irni Gemzon
+    affiliations:
+      - name: PharmAccess Foundation
+        city: Amsterdam
+        country: the Netherlands
   - name: Chris Ihure
@@
         country: Australia
-  - name: Paula van Brakel
-    affiliations:
-      - name: PharmAccess Foundation
-        city: Amsterdam
-        country: the Netherlands
-
-  - name: Mark van der Graaf
-    affiliations:
-      - name: PharmAccess Foundation
-        city: Amsterdam
-        country: the Netherlands
-  - name: Nicole Spieker
-    affiliations:
-      - name: PharmAccess Foundation
-        city: Amsterdam
-        country: the Netherlands
+  # - name: Paula van Brakel
+  #   affiliations:
+  #     - name: PharmAccess Foundation
+  #       city: Amsterdam
+  #       country: the Netherlands
+  # - name: Mark van der Graaf
+  #   affiliations:
+  #     - name: PharmAccess Foundation
+  #       city: Amsterdam
+  #       country: the Netherlands
+  # - name: Nicole Spieker
+  #   affiliations:
+  #     - name: PharmAccess Foundation
+  #       city: Amsterdam
+  #       country: the Netherlands
 
@@
 Digital transformation have the potential to make inclusive health systems in low- and middle income countries (LMICs) a reality by connecting patients and professionals, using data as a currency of trust.
-Adoption of digital technologies in healthcare means more than simply increasing efficiency or cutting costs.
-Instead, digital technologies have the potential to i) increase the availability, accessibility, acceptability, and quality of health services as we know them; ii) change nature and direction of health care, and the possibility of making it more preventive, personalised, and mobile through the use of such technologies; and iii) enfranchise patients and communities (and particularly vulnerable groups including children and young people) and evolve their relationship with health professionals and providers, thus helping shape the health system according to the needs of the patients and communities [@kickbusch2021lancet].
+Digital technologies have the potential to increase the availability, accessibility, acceptability, and quality of health services as we know them; change nature and direction of health care, and the possibility of making it more preventive, personalised, and mobile through the use of such technologies; and enfranchise patients and communities (and particularly vulnerable groups including children and young people) and evolve their relationship with health professionals and providers, thus helping shape the health system according to the needs of the patients and communities [@kickbusch2021lancet].
 
@@
 
-To achieve scaleability of digital health interventions, we need to conceptualize and implement digital platforms and their associated ecosystems that constitute an important enabling socio-technical infrastructure upon which such interventions can be realized in a more standardized and sustainable fashion.
-Digital platforms are a challenging research object because of their distributed nature and intertwinement with institutions, markets and technologies [@dereuver2018digital].
-Digital platform ecosystems vary according to three core building blocks: (1) platform ownership, (2) value-creating mechanisms, and (3) complementor autonomy [@hein2020digital].
-Various studies have investigated why the healthcare domain has been slow in adopting platforms and their ecosystems [@hermes2020digital;@ozalp2022digital].
-Before presenting our main contributions, we highlight two paradigmatic themes that are relevant to our paper.
+To achieve scaleability of digital health interventions, we need to design and implement standardized open health data systems (OHDS) and their associated ecosystems that can support the wider health sector in improving along five dimensions, namely [@kelley2020digital]:
 
-### From centralized to decentralized platforms
+1. overall quality and continuity of care;
+2. adherence to clinical guidelines and best practices;
+3. efficiency and affordability of services and health commodities, by reducing duplication of effort and ensuring effective use of time and resources;
+4. health-financing models and processes, regulation, oversight, and patient safety resulting from increased availability of performance data and reductions in errors; and
+5. health policy-making and resource allocation based on better quality data .
 
-Digital platforms have predominantly flourished in a business setting, as is exemplified by the well known success stories of cloud platforms providers (AWS, Microsoft, Google, Alibaba), two-sided market platforms (Airbnb, Spotify, Uber) and social media platforms (Meta, WeChat).
-While these examples allude to the potentional to generate social and economic value by using platforms as a key enabler of digital transformation of healthcare in LMICs, it is less obvious in what way a digital platform can trigger specific positive effects *for* development [@bonina2021digital].
-Part of the problem in grasping the implications of platforms for development in the field of information systems has been the lack of clarity regarding the understanding of what digital platforms are, what their main features are and how they generate value [@dereuver2018digital].
-Notably, even valuable landscape papers in information systems largely ignore the development implications of digital platforms [@dereuver2018digital] or may summarize them in rather naïve ways [@bonina2021digital].
-Put more strongly, issues pertaining to the winner-takes-all nature of digital platforms have resulted in so-called 'transgression of spheres' where commercial platform providers acquire disproprotionately large market power which is at odds with realizing public value for the common good [@sharon2018when].
+### Data & analytics functions are essential in open health data systems
 
-It almost seems we have forgotten that digital platforms need not always be centralized, but that we have a choice to implement decentralized or distributed platforms.
-In his seminal paper in 1963, Baran distinguished three types of network topologies ([@fig-network-topologies]).
-In a centralized network, all communication is facilitated through a single, central hub, which is obviously vulnerable as "...destruction of single node destroys communication between the end stations" [@baran1964distributed].
-At the opposite end of the spectrum, in a fully distributed network, stations are connected to all adjacent stations.
-In practice, a mixture of centralized and distributed components is the most commonly used topology, aptly called decentralized networks as they remove complete reliance upon a single point, while at the same time where is a hierarchy where smaller centralized networks are connected to higher-lying networks through interconnecting hubs (encircled in blue).
-Note that the internet has both the characteristics of a distributed network (for example, traffic and data processing are shared across multiple paths and servers) as well as that of a decentralized network (there is no single entity that controls the entire internet.
-Instead, numerous internet service providers (ISPs), organizations, and governments manage different parts of the network).
+Learnings from digital health interventions have shown that improvements along the five aforementioned dimensions is indeed within reach.
+As an example, in our experience with the MomCare programme, we have demonstrated that routinely collected health data, combined with financial data, can be effectively used to gain insight into the continuity of care, improve clinical adherence whilst maintaining efficiency and affordability of health services in LMICs.
+By actively coaxing pregnant mothers to undergo 4 antenatal check-ups, outcomes were improved whilst maintaining the average cost of MCH services [@huisman2022digital;@sanctis2022maintaining;@izudi2023experiences].
 
-![Three network topologies, taken from Baran (1964).
-The nodes encircled in blue in the decentralized network are hubs that provide connections between sub-networks.](images/network-topologies.png){#fig-network-topologies}
+MomCare, like most other digital health interventions, is critically dependent on the availability of data and analytics functions within the OHDS.
+For example, the widely used DHIS2 reports that are most commonly used in LMICs contained insufficient information to run the programme.
+More detailed analytics at the patient level were required.
+It is, in fact, a widely held belief that data & analytics are an essential part of OHDSs.
+This is exemplified how the OpenHIE reference architecture [@openhie] has been adopted by many sub-Saharan African countries as the blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one].
+These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain [@mbugua2021adoption;@dalhatu2023paper].
+The rationale for this addition is to facilitate secondary reuse of health data for academic research, real-world evidence studies etc.
+which can be framed within the context of ongoing efforts towards Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair].
 
-We are inspired by various fields of study that have shown that decentralized and/or distributed systems are often 'resilient by design' [@fiksel2003designing;@liu2010architectural;@mayar2022resilience], for example, allowing efficient deployments at different geographical scales.
-Given that LMICs are resource-constrained, we consider downward scaleability as a key design criterium [@fayad2005scalable;@duboc2006framework].
-We consider the insights from the 'hourglass model', which provides clear design principles how the use of minimal standards in a layered, distributed systems allows implementation information systetms using a great diversity of services [@beck2019hourglass].
+## Extending OpenHIE to include modern data and analytics standards
 
-We also posit that a decentralized platform is more conducive to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].
-With this approach, we purposefully challenge the dominant paradigm of designing and implementing centralized platforms to support the digital transformation of healthcare in LMICs [@ogundaini2022systematic] with the aim to make digital platforms work *for* development [@hermes2020digital].
+If we intend to use the OpenHIE standard not only for primary health data sharing, but also to support data and analytics for secondary health data sharing, we need to extend its specification.
+The main contribution of this paper is to propose how recent standards and open source implementations from the data engineering community could be integrated into the OpenHIE framework, as illustrated in @fig-ohie.
 
-### From open source to open data sharing
+![Proposed extension of the OpenHIE architecture that includes "Data and Analytics Services" as an additional service domain.](./images/openhie-extended-architecture.png){#fig-ohie}
 
-Case studies on digital platforms in healthcare point to an emerging pattern where the focus shifts from the digital platform with its defining software and hardware components, to the data as the primary object of interest in and of itself [@ozalp2022digital;@alaimo2022organizations].
-This observation ties into with the proposed research agenda by de Reuver et al.
-to consider data platforms as a phenomenon distinct from digital platforms [@dereuver2022openness].
-@tbl-comparing-platforms compares the key characteristics of digital vs.
-data platforms.
-The concept of health data spaces, currently under development in Europe, is an archetypical example of a data platform [@otto2022designing;@shabani2022will].
-
-| | Digital platforms (e.g. software, gaming) | data platforms (e.g. data marketplaces) |
-|:---|:-----|:-----|
-| **User groups** | Developers, Consumers | Data consumers, Data owners, Solution providers |
-| **Object of openness** | Software/hardware modules | Data, aggregated data, trained models; Data analytics modules |
-| **Ways to realize openness** | Open source, open boundary resources (APIs) | Data-driven mechanisms (secure computational environments) |
-| **Market consolidation** | Strong network effects; Dominant cross-industry platforms | Moderate network effects; Fragmentation and heterogeneity |
-
-: Comparing digital and health data platforms, taken from de Reuver et al. (2022) {#tbl-comparing-platforms .striped .hover .responsive}
-
-To illustrate the relevance of conceptualising health platforms as data platforms, consider the OpenHIE framework [@openhie] which has been adopted by many sub-Saharan African countries as the architectural blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one].
-The OpenHIE standard is best described as a digital platform as it is designed to facilitate information exchange, using specific workflows and information standards (open boundary resources in @tbl-comparing-platforms), with a primary focus is on creating a cohesive infrastructure that supports the interaction and integration of various health information systems.
-Since its inception in 2013, the ICT for Development community (ICT4D) has pushed for the development of open source implementations of OpenHIE components, which are now available as digital public goods [@digitalpublicgoods].
-@tbl-digital-public-goods list the currently available open source components that adhere to the OpenHIE specification.
-
-| Category  | Component | Digital Public Good |
-|:----------|:----------|:--------------------|
-| **Interoperability layer (IOL)** | IOL | [OpenHIM](http://openhim.org) |
-| | | [mHero](https://www.mhero.org/technology) |
-| | | [OpenFN](https://docs.openfn.org/) |
-| **Registry Services** | Client Registry (CR) | [SanteMPI](https://help.santesuite.org/product-overview/santesuite-products/master-patient-index-santempi) |
-| | | [JeMPI](https://jembi.gitbook.io/jempi/) |
-| | | [OpenCR](https://www.openclientregistry.org/) |
-| | | [OpenCRVS](https://www.opencrvs.org/) |
-| | Facility Registry (FR) | [Global Open Facility Registry (GOFR)](https://github.com/intrahealth/gofr) |
-| | | [GeoPrism Registry (GPR)](https://geoprismregistry.com/) |
-| | | [Resource Map](http://instedd.org/technologies/resource-map/) |
-| | | [Healthsite.io](https://healthsites.io/) |
-| | | [GeoPrism Registry](https://geoprismregistry.com/) |
-| | | [DHIS2](https://github.com/dhis2) |
-| | Health Worker Registry (HWR) | [iHRIS](https://www.ihris.org/ihris-50) |
-| | Terminology Service (TS) | [OCL Terminology Service](https://openconceptlab.org/terminology-service/) |
-| | Product Catalog (PC) | [PCMT](https://productcatalog.io/) |
-| **Business Domain Services** | Shared Health Record (SHR) | [HAPI FHIR](https://hapifhir.io/) |
-| | | [Fhirbase](https://www.health-samurai.io/fhirbase) |
-| | | [FHIR Server for Azure](https://github.com/microsoft/fhir-server) |
-| | | [I-TECH-UW SHR](https://github.com/I-TECH-UW/shared-health-record) |
-| | Health Management Information System (HMIS)[^1] | [DHIS2](http://dhis2.org/) |
-| | Finance and Insurance Service (FIS) | [OpenIMIS](http://openimis.org/) |
-| | Logistics Management Information System (LMIS) | [OpenLMIS](http://openlmis.org/) |
-| | | [OpenBoxes](https://openboxes.com/features/) |
-| **Generic** | Identity and Access Management (IAM) | [Keycloak](https://www.keycloak.org/) |
-| | Gateway & proxy | [FHIR Information Gateway](https://github.com/google/fhir-gateway) |
-| | Admin dashboard for SHR | [OpenSRP FHIR Web](https://github.com/onaio/fhir-web) |
-| | Configuration and deployment | [Instant OpenHIE](https://jembi.gitbook.io/instant-v2/) |
-
-: Overview of current open source implementations of components that fit in the OpenHIE specification.
-Point-of-Service systems are excluded for brevity.
-List compiled using [OpenHIE Reference Technologies](https://wiki.ohie.org/display/DR/OpenHIE+Reference+Technologies), [Global Goods for Digital Health](https://digitalsquare.org/digital-health-global-goods), [Digital Public Goods Alliance](https://digitalpublicgoods.net/) and search of open source code repositories.
-A systematic review of such digital public goods is beyond the scope of this paper.
-{#tbl-digital-public-goods .striped .hover .responsive}
-
-[^1]: Note that there is often a confusion on the acronym HMIS.
-Strictly speaking, the OpenHIE specification uses HMIS to refer to a Health Management Information System that is part of the Business Domain Services.
-Sometimes HMIS is used to refer to a Hospital Management Systems in the Point-of-Service domain, synonymous with an Electronic Medical Record (EMR) system.
-
-The shift in perspective from digital platforms to data platforms coincides with the paradox of open [@keller2021paradox].
-Originally, openness of digital platforms focused on open source and open standards (as shown above for OpenHIE) which by has been superseded by “...
-conflicts about privacy, economic value extraction, the emergence of artificial intelligence, and the destabilizing effects of dominant platforms on (democratic) societies.
-Instead of access to information, the control of personal data has emerged in the age of platforms as the critical contention.” [@keller2021paradox].
-These conflicts are particularly salient in the healthcare domain, where people are generally willing to share their health data to receive the best care (primary use, which is aligned with the concept of digital platforms), while the attitude towards secondary use of health data (conceptually aligned with a data platform) varies greatly depending on the type and context [@cascini2024health].
-The shift in perspective from digital platforms supporting primary data sharing toward data platforms supporting secondary data sharing is one of the key issues surrounding the polemic of data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond;@purtova2023data].
-
-This shift in perspective also has real implications for the design and implementation of OpenHIE.
-We observe that some African countries have, as a matter of course, extended the OpenHIE framework to include “data & analytics services” as an additional domain in their policies [@mbugua2021adoption;@dalhatu2023paper].
-The rationale for this addition is indeed to facilitate secondary reuse of health data for academic research, real-world evidence studies etc.
-using the principles of Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair].
-In doing so, however, we have implicitly moved from conceptualizing digital health platforms for primary data sharing (the original OpenHIE specification) to health data platforms for secondary data sharing.
-If we are to use the OpenHIE framework as an open data platform as well, we need to extend the standards, technologies and architecture to include functionality for secondary data sharing and reuse.
-The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish more comprehensive platforms to support primary and secondary health data sharing in LMICs.
-One of the motivations for this paper is integrate recent standards and open source implementations from the data engineering community into the OpenHIE framework.
-
-### Conceptualizing and implementing open health data sharing platforms
-
-In the light of these nascent paradigm shifts, the objective of this paper is the following.
-Given high-level design principles that we believe are supportive of realising solidarity-based, namely [TO DO: this is just first shot at formulating our design principles; needs refinement]:
-
-- inclusive-by-design, based on the notions of datasolidarity and maximising autonomy of all future participants in the ecosystem;
-- scaleable-by-design, particularly focusing on downward scaleability to support a decentralized platform topology to allows for bottom-up deployment scenarios (from local care networks --> county-level networks --> national networks) instead of top-down national roll-out;
-- open-by-design, whereby a balance is found to resort to minimal standards and allow for a large diversity of partners and technologies to be used;
-
-we seek to specifiy and demonstrate how this can be implemented using the full-STAC approach [@mehl2023fullstac], that consists of:
-
-1. FHIR as the main open standard
-2. the Instant HIE framework for integrating open technologies
-3. an extended version of OpenHIE as the open architecture that includes a more detailed architecture for the data platform
-4. open content with various demonstrators in the domain of antenatal and maternal care (ANC) as a first version of hands-on digital adaptation kits.
-
-#### 1. Sharing of data sets at the most granular (patient) level, persisted as longitudinal records
-The Shared Health Record (SHR) in OpenHIE provides an operational, real-time transactional data source which is intended for primary data sharing.
-The specification explicitly states that the SHR is distinct from a datawarehouse.
-At the same time, the Health Management Information System (HMIS) component is specified to fulfil functional requirements that are typically provided by datawarehouses.
-
-- **Q1:** How can proven solution designs and modern technologies from data warehousing and engineering be integrated in the OpenHIE specification and how does this relate to the curren specifications of the SHR and HMIS.
-
-#### 2. Sharing of data products derived from the original data set
-OpenHIE specifies that the Health Management Information System (HMIS) component should support a workflow to validate and save aggregate data based on the emerging [IHE Aggregate Data Exchange (ADX) standard](https://www.ihe.net/uploadedFiles/Documents/QRPH/IHE_QRPH_Suppl_ADX.pdf).
-This is just one specific workflow within a larger data analytics value chain that covers i) collection, ii) standardization, iii) cleaning, iv) storage, v) analysis and vi) distribution [@curry2016big].
-Sharing of benchmarking, and decision-support tools based on various statistical (machine learning) models that have been trained on data sets and/or data product, can be shared for standalone use.
-
-- **Q2:** How can proven solutions designs from the data warehousing and engineering community be included in OpenHIE to support sharing of wider variety of data products.
-
-#### 3. Sharing of secure computational environments to access and work with the data
-
-Current solution design of centralized data storage is increasingly being challenged.
-Recent work points to federated learning (FL) [@rieke2020future] and privacy- enhancing technologies (PETs) [@scheibner2021revolutionizing;@jordan2022selecting] as a better long-term solution to secure and equitable sharing of data.
-Given the resource constrained environments of LMICs, is is particularly challenging to devise a solution that is economically feasible and ...
-
-## Methods
-
-In this paper, we present a design that extends the OpenHIE specification to include the three types of data sharing mentioned above.
-Using the full-STAC approach [@mehl2023fullstac] we combine open standards, open technologies and open architectures into a coherent modular OHDP that can be configured and reused across a variety of use-cases.
-We employ a formative, naturalistic evaluation to assess the technical risk and efficacy of the design [@venable2016feds].
-Given that it is prohibitively expensive to evaluate the design a real-world setting, we aim to minimize technological risks and maximize the efficacy of the design by considering three examples of health data platforms in LMICs, namely:
-
- 1. the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/))
- 2. the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/)
- 3. the work conducted at PharmAccess Foundation as part of the MomCare programme ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org)).
-
-These three real-world implementations are evaluated along the following dimensions:
-
-- What is the level of openness of the implementation, specifically in terms of the three types of data sharing and platform-to-platform opennenss?
-- How are the core maintenance functions of the OHDP implemented, where we particularly focus on functions related to metadata management, data lineage and access control mechanisms?
-- Is the solution suitable for downward scalability, where we focus on the computer and storage requirements of the solution and assess the minimal requirements for on-site resources (computers, edge devices)
-
-As part of our design research, we have taken a narrative approach in surveying existing scientific studies on health data platforms, focusing on the seminal reports and subsequently searching forward citations.
-In addition, we have searched the open source repositories (most notably GitHub) and the online communities (OpenHIE community, FHIR community) to search for relevant open standards, technologies and architectures.
-This paper should not be considered as a proper systematic review.
-
-The main contributions of this paper are i) description of a framework for the components of the Data & Analysis Services that builds on current best practices from the data engineering community into the OpenHIE framework; ii) evaluation of different implementations and design options for type 1 and 2 data sharing within an extended OpenHIE architecture; and iii) provide open content for the MomCare implementation to facilitate adaptation and deployment.
-As such, it aims to inform future developments and implementation of open digital health platforms in LMICs.
-
-## Results
-
-### Design
-
-#### Open standards: using FHIR as the common data model
-
-The recent convergence to FHIR as the de facto standard for information exchange has fuelled the development of OpenHIE.
-FHIR is currently used both for routine healthcare settings [@amar2024electronic;@ayaz2021fast] and clinical research settings [@duda2022hl7;@vorisek2022fast] and is increasingly being used in LMICs as well.
-The guidelines and standards of the African Union explicitly state FHIR is to be used as the messaging standard [@2023african].
-The FHIR-native OpenSRP platform [@mehl2020open] has been deployed in 14 countries targeting various patient populations, amongst which a reference implementation of the WHO antenatal and neonatal care guidelines for midwives in Lombok, Indonesia [@summitinstitutefordevelopment2023bunda;@kurniawan2019midwife].
-In India, FHIR is used as the underlying technology for the open Health Claims Exchange protocol specification, which has been adopted by the Indian government as the standard for e-claims handling [@hcx].
-This range of utilizations showcase the standards’ widespread applicability.
-The proceedings of the OpenHIE conference 2023 attest to the fact that FHIR and open source technologies are embraced as critical enablers in implementing health information exchanges in LMICs [@ohie2023unconference].
-
-Various studies have investigated the merits of FHIR and its performance vis-a-vis other healthcare standards.
-Comparisons between OpenEHR, ISO 13606, OMOP and FHIR have been made [@ayaz2023transforming;@mullie2023coda;@rinaldi2021openehr;@cremonesi2023need;@sinaci2023data;@tsafnat2024converge].
-A study involving 10 experts comparing OpenEHR, ISO 13606 and FHIR concluded that i) these three standards are functionally and technically compatible, and therefore can be used side by side; and that ii) each of these standards have their strengths and limitations that correlate with their intended use as summarized in the @tbl-comparison.
-
-![Comparison of OpenEHR, ISO 13606 and FHIR standards](images/comparison-ehr-standards.png){width=auto #tbl-comparison}
-
-For an infectious diseases dataset with a limited scope, OpenEHR, OMOP and FHIR have been compared and found all to be equally suitable [@rinaldi2021openehr].
-Comparing OMOP and FHIR, the latter has been found to support more granular mappings required for analytics and was therefore chosen as the standard for the CODA project [@mullie2023coda].
-
-Although FHIR was originally designed only for exchange between systems, we propose to use it as the common data model for the design presented here for the following reasons:
-
-- Industry adoption has significantly increased, as exemplified by FHIR-based offering by major cloud providers such as Google, Azure and AWS.
-Also, Africa CDC has explicitly chosen FHIR as the preferred standard;
-- The widespread availibility of the Bulk FHIR API [@mandl2020push;@jones2021landscape] enables bulk, file-based batchwise processing for analytics using the lakehouse architecture as detailed in the next section;
-- The concept of FHIR Profiles allow localisation to tailor the standard to a specific use case.
-A profile defines rules, extensions, and constraints for a resource.
-We posit that the possible penalty of this flexibility, namely having to manage different FHIR versions and/or profiles, is less of an issue in the context of LMICs where first priority is to exchange datasets such as the International Patient Summary (IPS) that are less complex compared to the requirements for high income countries;
-- Being based on webstandards, the FHIR standard lends itself best for further separation of concerns as envisioned by the composable data stack.
-This is an important enabler for the downward scalability of the solution;
-- With its inherent, graph-like nature, FHIR can be readily incorporated into the principles of FAIR data sharing, where FHIR-based data repositories can be integrated in an overarching netwerk of FAIR data stations [@sinaci2023data;@pedrera-jimenez2023can].
-
-TO DO:
-- tie-in paper [@tsafnat2024converge] on why we use FHIR
-- re-use these elements from table 1 into this section:
-
-  - Given that the Bulk FHIR API has by now been incorporated in all major FHIR implementations [@mandl2020push;@jones2021landscape], the FHIR standard can be readily used to bridge the gap between primary (transactional) and secondary (analytics) data sharing.
-  - __SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows.
-Builds on FHIRPath [@hl72020fhirpath] expressions in a logical structure to specify things like column names and unnested items.
-Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].
-Modern data platforms take this Within the HL7 ecosystem much progress has been made to support "analytics-on-FHIR" with standards including [FHIRPath](https://hl7.org/fhirpath/) and the new SQL-on-FHIR specification [@sql-on-fhir].
-Given the use of FHIR as the common data model (the rationale of which will be described later), we formulate the second design question:
-
-Possible risks pertaining to the use of FHIR as the common data model, most notably the possible incompatibilities and/or high costs of maintenance in supporting different versions, will be addressed in the Discussion.
-
-#### Open architecture: extending OpenHIE with a composable data stack
-
-Data management and analytics platforms have undergone significant changes since the first generation of data warehouses were introduced.
-Recent studies have shown that the current practice has converged towards the lakehouse as one of the most commonly used solution designs [@armbrust2021lakehouse;@hai2023data;@harby2022data].
+The original OpenHIE specification discerns four domains, namely Point-of-Service systems, the Interoperability Layer, Common Services and Business Services.
+We propose to extend the OpenHIE architecture with a "Data and Analytics Services" domain with different zones taken from the data lakehouse architecture, which currently is the most commonly used design pattern in this domain [@armbrust2021lakehouse;@hai2023data;@harby2022data;@harby2024data].
 Lakehouses typically have a zonal architecture that follow the Extract-Load-Transform pattern (ELT) where data is ingested from the source systems in bulk (E), delivered to storage with aligned schemas (L) and transformed into a format ready for analysis (T) [@hai2023data].
@@
 
-With respect to current implementations of lakehouse data platforms, we observe a proliferation of tools with as yet limited standards to improve technical interoperability.
-In the analysis of Pedreira et al.
-[@pedreira2023composable] the requirement for specialization in data management systems has evolved faster than our software development practices.
-This situation has created a siloed landscape composed of hundreds of products developed and maintained as monoliths, with limited reuse between systems.
-It has also affected the end users, who are often required to learn the idiosyncrasies of dozens of incompatible SQL and non-SQL API dialects, and settle for systems with incomplete functionality and inconsistent semantics.
-To remedy this, Pedreira et al.
-call to (re-)design and implement modern data platforms in terms of a 'composable data stack’ as a means to decrease development and maintenance cost and pick-up the speed of innovation.
+Following the terminology proposed by Hai et al.
+[@hai2023data], @tbl-data-and-analysis-services summarizes how the different zones of the lakehouse architecture can be adapted for healthcare and integrated into the OpeHIE specification.
+Taking Fast Healthcare Interoperability Resources (FHIR) as the open data standard, we envisage the extended OpenHIE architecture to include a 'FHIR lakehouse'[^1].
 
-While the lakehouse architecture separates the concerns of compute and storage, the composable data stack takes the separation of concerns is taken one step further.
-A composable data system (@fig-composable-data-stack), not only separates the storage (layer 3) and execution (layer 2), but also separates the user interface (layer 1) from the execution engine by introducing standards including Substrait for Intermediate Representation (standard A, IR) and Apache Arrow for data connectivity and data memory layout (standards B and C, respectively).
-The first generation of open source components are already available.
-For example, the Ibis user interface is currently sufficiently mature to offer a standardized dataframe interface to 19 different execution engines [@ibisproject].
-
-::: {#fig-composable-data-stack layout-nrow=2 fig-cap="Schematic overview of the composable data stack showing a) the overall architecture and b) examples of implementations.
-Images taken from [https://voltrondata.com/codex](https://voltrondata.com/codex)."}
-![a)](images/composable-data-stack.png)
-
-![b)](images/composable-data-stack-implementation-2.png)
-:::
-
-#### SQL-on-FHIR v2 as an intermediate representation for FHIR data in tabular format
-
-The premise of separating the user interface from the execution engine is directly related to the key objective of the SQL-on-FHIR project ([https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/](https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/)), namely to make large-scale analysis of FHIR data accessible to a larger audience, portable between systems and to make FHIR data work well with the best available analytic tools, regardless of the technology stack.
-However, to use FHIR effectively analysts require a thorough understanding of the specification as FHIR is represented as a graph of resources, with detailed semantics defined for references between resources, data types, terminology, extensions, and many other aspects of the specification.
-Most analytic and machine learning use cases require the preparation of FHIR data using transformations and tabular projections from its original form.
-The task of authoring these transformations and projections is not trivial and there is currently no standard mechanisms to support reuse.
-
-The solution of the SQL-on-FHIR project is to provide a specification for defining tabular, use case-specific views of FHIR data.
-The view definition and the execution of the view are separated, in such a way that the definition is portable across systems while the execution engine (called runners) are system-specific tools or libraries that apply view definitions to the underlying data layer, optionally making use of annotations to optimize performance.
-
-#### Extending OpenHIE with a FHIR-based composable data stack
-
-We propose to extend the OpenHIE architecture with a "Data and Analytics Services" domain with different service layers by synthesizing the current best practices of the a lakehouse architecture of [@hai2023data;@harby2022data;@harby2024data] and the composable data stack [@pedreira2023composable] (@fig-ohie, @tbl-data-and-analysis-services).
-These 5 services are considered the core of the FHIR data platform, and although in practice often a downstream dashboarding or visualization component is used, that component is not the main focus of our analysis.
-Rather, we aim to elucidate and conceptualize the core "FHIR Data Lakehouse".
-
-![Proposed extension of the OpenHIE architecture that includes "Data and Analytics Services" as an additional service domain.](./images/openhie-extended-architecture.png){#fig-ohie}
+[^1]: We have argued the choice of FHIR as the common data model elsewhere, working paper to be submitted ([link](https://plugin-healthcare.github.io/viewpoint-jimr/)).
 
@@
 +-------------------+---------------------------------------------------------------------------------------+
+: Adaptation of lakehouse zones for healthcare specific standardization of Data and Analysis Services {#tbl-data-and-analysis-services}
 
-: Definition of Data and Analysis Services {#tbl-data-and-analysis-services}
+[TO DO: need to explain table (is too condensed now) and especially FHIR specific components.]
 
-##### Ingestion
+In the remainder of this paper, we first describe in more detail how the FHIR lakehouse pattern can be implemented using existing open source technologies.
+We do so by presenting work done at PharmAccess.
+Code and digital artifacts are made available [TO DO: include links to support GitHub repositories].
 
-Default workflow is extraction of data from SHR using Bulk FHIR API.
-Data contains metadata (incl.
-FHIR versions) and fully qualified semantics, for example, coding systems.
-Despite this, metadata extraction and metadata modeling is still required to meet the FAIR requirements.
-Issues that need to be solved by these services:
+Subquently, we compare this solution design with two widely used OpenHIE-compliant frameworks, namely the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)) and the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/).
+We take a narrative approach in presenting our design and the case studies, surveying existing scientific studies on OHDSs, focusing on the seminal reports and subsequently searching forward citations.
+In addition, we have searched the open source repositories (most notably GitHub) and the online communities (OpenHIE community, FHIR community) to search for relevant open standards, technologies and architectures.
+This paper should not be considered as a proper systematic review.
 
-- To prepare for future updates of FHIR versions
-- Implement late-binding principle of having increasingly more specific FHIR profiles as bulk FHIR data propagates through lakehouse
+Finally, we discuss our findings and propose routes for future development.
 
-##### Storage
+The main contributions of this paper are i) description of a framework for the components of the Data & Analysis Services that builds on current best practices from the data engineering community into the OpenHIE framework; ii) evaluation of different implementations and design options for type 1 and 2 data sharing within an extended OpenHIE architecture; and iii) provide open content for the MomCare implementation to facilitate adaptation and deployment.
+As such, it aims to inform future developments and implementation of open digital health platforms in LMICs.
 
-- File-based:
-  - from ndjson to parquet
-  - possibly used delta lake for time versioning
-  - separation of storage from compute not only for benefits of lower TCO, but also be ready for federated learning and MPC in future
+## Case study of implementations
 
-- OLAP DBMS
-  - Often columnar, like Clickhouse and BigQuery
-  -
+### PharmAccess demonstrator Momcare programme
 
-##### Query & Processing
+MomCare was launched in Kenya [@huisman2022digital;@sanctis2022maintaining] and Tanzania [@shija2021access;@mrema2021application] in 2017 and 2019 respectively, with the objective to improve health outcomes for maternal and antenatal care.
+MomCare distinguishes two user groups: mothers are supported during their pregnancy through reminders and surveys, using SMS as the digital mode of engagement.
+Health workers are equipped with an Android-based application, in which visits, care activities and clinical observations are recorded.
+Reimbursements of the maternal clinic are based on the data captured with SMS and the app, thereby creating a conditional payment scheme, where providers are partially reimbursed up-front for a fixed bundle of activities, supplemented by bonus payments based on a predefined set of care activities.
 
-- fit in structure of OpenHIE specification
-- check which workflows are related to analytics
-- Hai calls this 'Maintenance'
+In its original form, the MomCare programme used closed digital platforms.
+In Kenya, M-TIBA is the primary digital platform, on top of which a relatively lightweight custom app has been built as the engagement layer for the health workers [@huisman2022digital].
+M-TIBA provides data access through its data warehouse platform for the MomCare programme, however, this is not a standardized, general purpose API.
+In the case of Tanzania, a stand-alone custom app is used which does not provide an interface of any kind for interacting with the platform [@mrema2021application].
+Given these constraints, the first iteration of the MomCare programme used a custom-built data warehouse environment as its main data platform, on which data extractions, transformations and analysis are performed to generate the operational reports.
+Feedback reports for the health workers, in the form of operational dashboards, are made accessible through the app.
+Similar reports are provided to the back-office for the periodic reimbursement to the clinics.
 
-##### Maintenance
+Clearly, a more open and scaleable platform was required if MomCare was to be implemented in more regions.
+This need led to a redesign of the underlying technical infrastructure of the MomCare project.
+The objectives of this work were in fact to demonstrate a solution design that could support the first three types of data sharing.
+First, to investigate the viability of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed to assess the complexity and effort required to implement the facade pattern to integrate the legacy system into the FHIR data standard.
+Using the longitudinal dataset from approximately 28 thousand patient records, FHIR transformations script were developed and deployed using the mediator function of the IOL.
+The data was transformed into 10 FHIR v4 resources and the conceptual data model of the existing MomCare app could readily be transformed into the FHIR standard using SQL and validated with a Python library [@islam2023fhir].
+The largest challenge during the transformation process pertained to the absence of unique business identifiers for patients and healthcare organizations.
+For patients, either the mobile phone number or the healthcare insurance number was taken, depending on availability.
+A combination of name, address and latitude/longitude coordinates were used to uniquely identify organizations and locations, as Tanzania does not have a system in place for this purpose.
 
-- SQL-on-FHIR Views provide new standard to support mADX aggregate reporting !! We need to stress this, because this is an existing OpenHIE workflow
-- Maintenance-related functions remain the same
-- NB: orchestration falls under data provenance
+The second objective was to reproduce existing analytic reports, using the bulk FHIR data format as input.
+Here, the focus was to standardize the logic required for producing metrics and reports.
+The transformed and validated data is uploaded into the FHIR server on a daily basis using an automated cloud function.
+Analysis of bulk data was done by directly reading the standard newline delimited JSON into the Python pandas data analysis library.
+Cross checking the output with queries on the original data confirmed that the whole data pipeline produced consistent results.
+For example, the report of the antenatal coverage metric (number of pregnancies with four or more visits) could be reproduced per patient journey and aggregated (per year, per organization etc.) as required for the MomCare reports.
 
-##### Data consumers
+TO DO: explain logic of patient-timeline table.
+Write standard transformation to go from FHIR resources to this standard table.
+On top of that the actual metrics and reporting.
+Explain serverless: we wanted to get rid of resource-heavy data visualization tools.
+This led to the idea of serverless: using duckdb-wasm and pipelines of cloud functions.
 
-- Many tools, often focus on creating information dashboards and visualizations
-- Compatibility with processing & API: which query languages and interfaces are supported. Some dialect of SQL, some dialect of NoSQL, dataframe API, all of the above?
+The third objective was to run a technical feasibility test for federated analytics. Using the MPC platform of Roseman Labs, we managed to do aggregations in the blind ...
+TO DO: explain that we managed to reproduce the reports we generated in the clear, but then in the blind. Note, however, that in the remainder we will focus on first two types of data sharing.
 
-### Open technologies: deploying Instant OpenHIE with digital public goods
+Based on these experiments, we arrived at the following design for the data & analytics services
 
-Today, many components of the OpenHIE specification are now available as a digital public goods, as listed in @tbl-digital-public-goods.
-Typically, these open source components are intended to support deployments in small countries (population up to 10 million) or large NGOs out of the box, and should provide a stepping stone for customized deployments in medium-sized countries (population around 40 million).[^2] To further ease the development, configuration and deployment of health information exchanges, the concept of 'Instant OpenHIE' has been championed to (i) allow implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) have a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@InstantOpenHIEv2].
-Besides the core functional components of the OpenHIE architecture, the Instant OpenHIE toolkit allows packaging and integration of generic components such as Identity and Access Managment (IAM) and a reverse proxy gateway.
-In the following, we will evaluatie three of such configurations, with the aim to conceptualize and evaluate the proposed Data and Analytics Services domain of of the OpenHIE architecture.
+- Use 'serverless' file-based storage: bulk copy of data as-is in parquet
+  - Tension: how to manage change data capture
+  - Tension: how to manage access rights
+- Use SQL-on-FHIR-v2 to create tabular views.
+  - Example: patient timeline
+  - TO DO: rewrite patient timeline queries with SQL-on-FHIR-v2 and run it with Pathling
+- Use semantic modeling layer to define metrics
+  - There are many options: dbt, cube.dev
+  - Fulfills same function as ADX/mADX IHE profile in OpenHIE specification
+  - Tension: going from patient-timeline to reported metrics still isn't standardized.
+This is where Ibis/Substrait comes in.
+Substrait as IR for cross-language serialization for relational algebra.
+Can be executed on different backends.
+Write once, run on different engines.
+- Distribute and publish reports on resource-constrained devices 
+  - duckdb
+  - sveltekit
 
-[^2]: Although the OpenHIE specification does not include details on dimensioning, these are typically the requirements that are used within the community.
-See [OpenHIE Community Wiki](https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements).
+TO DO: Add diagram
 
-### Open content
-
-...
-
-### Evaluation of implementations
-
-#### Jembi OpenHIM platform
+### Jembi OpenHIM platform
 
@@
 
-#### ONA OpenSRP 2
+### ONA OpenSRP 2
 
@@
 
-### PharmAccess demonstrator Momcare programme
+### Open technologies: deploying Instant OpenHIE with digital public goods
 
-MomCare was launched in Kenya [@huisman2022digital;@sanctis2022maintaining] and Tanzania [@shija2021access;@mrema2021application] in 2017 and 2019 respectively, with the objective to improve health outcomes for maternal and antenatal care.
-MomCare distinguishes two user groups: mothers are supported during their pregnancy through reminders and surveys, using SMS as the digital mode of engagement.
-Health workers are equipped with an Android-based application, in which visits, care activities and clinical observations are recorded.
-Reimbursements of the maternal clinic are based on the data captured with SMS and the app, thereby creating a conditional payment scheme, where providers are partially reimbursed up-front for a fixed bundle of activities, supplemented by bonus payments based on a predefined set of care activities.
+Today, many components of the OpenHIE specification are now available as a digital public goods, as listed in @tbl-digital-public-goods.
+Typically, these open source components are intended to support deployments in small countries (population up to 10 million) or large NGOs out of the box, and should provide a stepping stone for customized deployments in medium-sized countries (population around 40 million).[^2] To further ease the development, configuration and deployment of health information exchanges, the concept of 'Instant OpenHIE' has been championed to (i) allow implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) have a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@InstantOpenHIEv2].
+Besides the core functional components of the OpenHIE architecture, the Instant OpenHIE toolkit allows packaging and integration of generic components such as Identity and Access Managment (IAM) and a reverse proxy gateway.
+In the following, we will evaluatie three of such configurations, with the aim to conceptualize and evaluate the proposed Data and Analytics Services domain of of the OpenHIE architecture.
 
-In its original form, the MomCare programme used closed digital platforms.
-In Kenya, M-TIBA is the primary digital platform, on top of which a relatively lightweight custom app has been built as the engagement layer for the health workers [@huisman2022digital].
-M-TIBA provides data access through its data warehouse platform for the MomCare programme, however, this is not a standardized, general purpose API.
-In the case of Tanzania, a stand-alone custom app is used which does not provide an interface of any kind for interacting with the platform [@mrema2021application].
-Given these constraints, the first iteration of the MomCare programme used a custom-built data warehouse environment as its main data platform, on which data extractions, transformations and analysis are performed to generate the operational reports.
-Feedback reports for the health workers, in the form of operational dashboards, are made accessible through the app.
-Similar reports are provided to the back-office for the periodic reimbursement to the clinics.
+[^2]: Although the OpenHIE specification does not include details on dimensioning, these are typically the requirements that are used within the community.
+See [OpenHIE Community Wiki](https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements).
 
-Clearly, a more open and scaleable platform was required if MomCare was to be implemented in more regions.
-This need led to a redesign of the underlying technical infrastructure of the MomCare project.
-The objectives of this work were in fact to demonstrate a solution design that could support the first three types of data sharing.
-First, to investigate the viability of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed to assess the complexity and effort required to implement the facade pattern to integrate the legacy system into the FHIR data standard.
-Using the longitudinal dataset from approximately 28 thousand patient records, FHIR transformations script were developed and deployed using the mediator function of the IOL.
-The data was transformed into 10 FHIR v4 resources and the conceptual data model of the existing MomCare app could readily be transformed into the FHIR standard using SQL and validated with a Python library [@islam2023fhir].
-The largest challenge during the transformation process pertained to the absence of unique business identifiers for patients and healthcare organizations.
-For patients, either the mobile phone number or the healthcare insurance number was taken, depending on availability.
-A combination of name, address and latitude/longitude coordinates were used to uniquely identify organizations and locations, as Tanzania does not have a system in place for this purpose.
+We also posit that a decentralized platform is more conducive to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].
+With this approach, we purposefully challenge the dominant paradigm of designing and implementing centralized platforms to support the digital transformation of healthcare in LMICs [@ogundaini2022systematic] with the aim to make digital platforms work *for* development [@hermes2020digital].
 
-The second objective was to reproduce existing analytic reports, using the bulk FHIR data format as input.
-Here, the focus was to standardize the logic required for producing metrics and reports.
-The transformed and validated data is uploaded into the FHIR server on a daily basis using an automated cloud function.
-Analysis of bulk data was done by directly reading the standard newline delimited JSON into the Python pandas data analysis library.
-Cross checking the output with queries on the original data confirmed that the whole data pipeline produced consistent results.
-For example, the report of the antenatal coverage metric (number of pregnancies with four or more visits) could be reproduced per patient journey and aggregated (per year, per organization etc.) as required for the MomCare reports.
+| Level | Description | Number of facilities | 
+|:---:|----------------------|--------:|
+| 2 | Dispensaries and private clincs, typically located in a school, industrial plant or other organization that dispenses medication and sometimes basic medical and dental treatment | 8,806 |
+| 3 | Health centres, medium-sized units which cater for a population of about 80,000 people | 2,559 |
+| 4 | Sub-county hospital, similar to health centres with additional facilities for more complex procedures |   971 |
+| 5 | County referral hospital, regional centres which provide specialised care |    34 |
+| 6 | National referral hospital | 5 |
 
-TO DO: explain logic of patient-timeline table.
-Write standard transformation to go from FHIR resources to this standard table.
-On top of that the actual metrics and reporting.
-Explain serverless: we wanted to get rid of resource-heavy data visualization tools.
-This led to the idea of serverless: using duckdb-wasm and pipelines of cloud functions.
+: Number of healthcare facilities in Kenya. Source: Kenya Health Facility Census, Ministry of Health, September 2023. {#tbl-kenya-facilities}
 
-The third objective was to run a technical feasibility test for federated analytics. Using the MPC platform of Roseman Labs, we managed to do aggregations in the blind ...
-TO DO: explain that we managed to reproduce the reports we generated in the clear, but then in the blind. Note, however, that in the remainder we will focus on first two types of data sharing.
+#### SQL-on-FHIR v2 as an intermediate representation for FHIR data in tabular format
 
-Based on these experiments, we arrived at the following design for the data & analytics services
+The premise of separating the user interface from the execution engine is directly related to the key objective of the SQL-on-FHIR project ([https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/](https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/)), namely to make large-scale analysis of FHIR data accessible to a larger audience, portable between systems and to make FHIR data work well with the best available analytic tools, regardless of the technology stack.
+However, to use FHIR effectively analysts require a thorough understanding of the specification as FHIR is represented as a graph of resources, with detailed semantics defined for references between resources, data types, terminology, extensions, and many other aspects of the specification.
+Most analytic and machine learning use cases require the preparation of FHIR data using transformations and tabular projections from its original form.
+The task of authoring these transformations and projections is not trivial and there is currently no standard mechanisms to support reuse.
 
-- Use 'serverless' file-based storage: bulk copy of data as-is in parquet
-  - Tension: how to manage change data capture
-  - Tension: how to manage access rights
-- Use SQL-on-FHIR-v2 to create tabular views.
-  - Example: patient timeline
-  - TO DO: rewrite patient timeline queries with SQL-on-FHIR-v2 and run it with Pathling
-- Use semantic modeling layer to define metrics
-  - There are many options: dbt, cube.dev
-  - Fulfills same function as ADX/mADX IHE profile in OpenHIE specification
-  - Tension: going from patient-timeline to reported metrics still isn't standardized.
-This is where Ibis/Substrait comes in.
-Substrait as IR for cross-language serialization for relational algebra.
-Can be executed on different backends.
-Write once, run on different engines.
-- Distribute and publish reports on resource-constrained devices 
-  - duckdb
-  - sveltekit
+The solution of the SQL-on-FHIR project is to provide a specification for defining tabular, use case-specific views of FHIR data.
+The view definition and the execution of the view are separated, in such a way that the definition is portable across systems while the execution engine (called runners) are system-specific tools or libraries that apply view definitions to the underlying data layer, optionally making use of annotations to optimize performance.
 
-TO DO: Add diagram
+##### Ingestion
 
-#### Level of opennness
+Default workflow is extraction of data from SHR using Bulk FHIR API.
+Data contains metadata (incl.
+FHIR versions) and fully qualified semantics, for example, coding systems.
+Despite this, metadata extraction and metadata modeling is still required to meet the FAIR requirements.
+Issues that need to be solved by these services:
 
-TO DO: evaluate openness of OpenHIM platform
+- To prepare for future updates of FHIR versions
+- Implement late-binding principle of having increasingly more specific FHIR profiles as bulk FHIR data propagates through lakehouse
 
-#### Core maintenance functions
+##### Storage
 
-TO DO: evaluate maintenance functions
+- File-based:
+  - from ndjson to parquet
+  - possibly used delta lake for time versioning
+  - separation of storage from compute not only for benefits of lower TCO, but also be ready for federated learning and MPC in future
 
-#### Downward scaleability
+- OLAP DBMS
+  - Often columnar, like Clickhouse and BigQuery
+  -
 
-TO DO: evaluate downward scalability
+##### Query & Processing
+
+- fit in structure of OpenHIE specification
+- check which workflows are related to analytics
+- Hai calls this 'Maintenance'
+
+##### Maintenance
+
+- SQL-on-FHIR Views provide new standard to support mADX aggregate reporting !! We need to stress this, because this is an existing OpenHIE workflow
+- Maintenance-related functions remain the same
+- NB: orchestration falls under data provenance
+
+##### Data consumers
+
+- Many tools, often focus on creating information dashboards and visualizations
+- Compatibility with processing & API: which query languages and interfaces are supported. Some dialect of SQL, some dialect of NoSQL, dataframe API, all of the above?
 
@@
 ### Openness of health data sharing platforms
+
+The shift in perspective from digital platforms to data platforms coincides with the paradox of open [@keller2021paradox].
+Originally, openness of digital platforms focused on open source and open standards (as shown above for OpenHIE) which by has been superseded by “...
+conflicts about privacy, economic value extraction, the emergence of artificial intelligence, and the destabilizing effects of dominant platforms on (democratic) societies.
+Instead of access to information, the control of personal data has emerged in the age of platforms as the critical contention.” [@keller2021paradox].
+These conflicts are particularly salient in the healthcare domain, where people are generally willing to share their health data to receive the best care (primary use, which is aligned with the concept of digital platforms), while the attitude towards secondary use of health data (conceptually aligned with a data platform) varies greatly depending on the type and context [@cascini2024health].
+The shift in perspective from digital platforms supporting primary data sharing toward data platforms supporting secondary data sharing is one of the key issues surrounding the polemic of data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond;@purtova2023data].
+
+### Conceptualizing and implementing open health data sharing platforms
+
+In the light of these nascent paradigm shifts, the objective of this paper is the following.
+Given high-level design principles that we believe are supportive of realising solidarity-based, namely [TO DO: this is just first shot at formulating our design principles; needs refinement]:
+
+- inclusive-by-design, based on the notions of datasolidarity and maximising autonomy of all future participants in the ecosystem;
+- scaleable-by-design, particularly focusing on downward scaleability to support a decentralized platform topology to allows for bottom-up deployment scenarios (from local care networks --> county-level networks --> national networks) instead of top-down national roll-out;
+- open-by-design, whereby a balance is found to resort to minimal standards and allow for a large diversity of partners and technologies to be used;
 
@@
 SW/IntelliSOFT received funding from PharmAccess as an implementation partner in various projects.
-AP/ONA received funding from PharmAccess to develop a new version of components of the open source OpenSRP 2 framework.
+AP/ONA received funding from PharmAccess to develop an improved version of one of the components of the open source OpenSRP 2 framework.
 
@@
 | FL | Federated learning |
-| HDSP	|	Health data sharing platform |
 | HIE | Health Information Exchange |
@@
 | PET | Privacy-enhancing technologies |
+| OHDS | Open health data system |
 | SHR | Shared Health Record |

### 2024-08-26 index.qmd
Restructured whole paper

@@
 
-### The global digital health ecosystem is still project-centric
+### The need for standardized open health data systems
 
-Digital transformation have the potential to make inclusive health systems in low- and middle income countries (LMICs) a reality by connecting patients and professionals, using data as a currency of trust.
-Digital technologies have the potential to increase the availability, accessibility, acceptability, and quality of health services as we know them; change nature and direction of health care, and the possibility of making it more preventive, personalised, and mobile through the use of such technologies; and enfranchise patients and communities (and particularly vulnerable groups including children and young people) and evolve their relationship with health professionals and providers, thus helping shape the health system according to the needs of the patients and communities [@kickbusch2021lancet].
+Digital transformation should be considered as a key determinant of health, but also requires a radical rethink how digital technologies should be used [@kickbusch2021lancet].
+Digital technologies have the potential to increase the availability, accessibility, acceptability, and quality of health services as we know them; change nature and direction of health care, and the possibility of making it more preventive, personalised, and mobile; and enfranchise patients and communities (particularly vulnerable groups), thus helping shape the health system according to the needs of the patients and communities.
 
@@
 
-To achieve scaleability of digital health interventions, we need to design and implement standardized open health data systems (OHDS) and their associated ecosystems that can support the wider health sector in improving along five dimensions, namely [@kelley2020digital]:
+To achieve scaleability of digital health interventions, we need to design and implement standardized open health data systems (OHDS) and their associated ecosystems that can support improvements in the wider health sector along five dimensions, namely [@kelley2020digital]:
 
@@
 4. health-financing models and processes, regulation, oversight, and patient safety resulting from increased availability of performance data and reductions in errors; and
-5. health policy-making and resource allocation based on better quality data .
+5. health policy-making and resource allocation based on better quality data.
 
@@
 
-Learnings from digital health interventions have shown that improvements along the five aforementioned dimensions is indeed within reach.
+Learnings from digital health interventions have shown that improvements along the five aforementioned dimensions is within reach.
 As an example, in our experience with the MomCare programme, we have demonstrated that routinely collected health data, combined with financial data, can be effectively used to gain insight into the continuity of care, improve clinical adherence whilst maintaining efficiency and affordability of health services in LMICs.
-By actively coaxing pregnant mothers to undergo 4 antenatal check-ups, outcomes were improved whilst maintaining the average cost of MCH services [@huisman2022digital;@sanctis2022maintaining;@izudi2023experiences].
+By actively coaxing pregnant mothers to undergo 4 antenatal check-ups, outcomes were improved whilst maintaining the average cost of maternal, newborn and child health (MNCH) services [@huisman2022digital;@sanctis2022maintaining;@izudi2023experiences].
 
-MomCare, like most other digital health interventions, is critically dependent on the availability of data and analytics functions within the OHDS.
-For example, the widely used DHIS2 reports that are most commonly used in LMICs contained insufficient information to run the programme.
-More detailed analytics at the patient level were required.
-It is, in fact, a widely held belief that data & analytics are an essential part of OHDSs.
+MomCare, like most other digital health interventions, is critically dependent on the availability of data and analytics functions within its underlying supportive OHDS.
+For example, the widely used DHIS2 reports that are most commonly used in LMICs contained insufficient information and analytical functions to support such interventions.
+To implement the value-based healthcare business logic, more detailed analytics of patient journeys is required.
+
+It is in fact a widely held belief in the global digital health community that data & analytics services are an essential capibility of OHDSs.
 This is exemplified how the OpenHIE reference architecture [@openhie] has been adopted by many sub-Saharan African countries as the blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one].
 These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain [@mbugua2021adoption;@dalhatu2023paper].
-The rationale for this addition is to facilitate secondary reuse of health data for academic research, real-world evidence studies etc.
-which can be framed within the context of ongoing efforts towards Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair].
+In terms of the often-used distinction between primary and secondary health data use [@cascini2024health], these countries aim to extend OpenHIE beyond it original scope of primary data sharing to also include secondary use of health data for academic research, real-world evidence studies etc.
+If we are to use the OpenHIE framework for data & analytics as well, we need to extend the standards, technologies and architecture to include functionality to do so.
+The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish more comprehensive platforms to support primary and secondary health data sharing in LMICs.
+
+The main contribution of this paper is to propose how recent standards and open source implementations from the data engineering community can be integrated into the OpenHIE framework.
+In the following, we first describe how the lakehouse design pattern, being the most widely used data & analytics solution architecture, can be integrated in OpenHIE.
+To demonstrate the feasibility of this design, we present a proof-of-concept impelementation using open source technologies within the context of the MomCare programme.
+Code and digital artifacts of this demonstrator is available as supplementary material [TO DO: include links to support GitHub repositories].
+Subquently, we compare this solution design with two widely used and operational OpenHIE-compliant open source frameworks, namely the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)) and the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/).
+Finally, we discuss our findings and propose routes for future development.
+
+We take a narrative approach in presenting our design and the case studies, surveying existing scientific studies on OHDSs, focusing on the seminal reports and subsequently searching forward citations.
+In addition, we have searched the open source repositories (most notably GitHub) and the online communities (OpenHIE community, FHIR community) to search for relevant open standards, technologies and architectures.
+This paper should not be considered as a proper systematic review.
+The main contributions of this paper are i) description of a framework for the components of the Data & Analysis Services that builds on current best practices from the data engineering community into the OpenHIE framework; and ii) evaluation of different implementations and design options for various data sharing scenarios within an extended OpenHIE architecture.
 
@@
 
-If we intend to use the OpenHIE standard not only for primary health data sharing, but also to support data and analytics for secondary health data sharing, we need to extend its specification.
-The main contribution of this paper is to propose how recent standards and open source implementations from the data engineering community could be integrated into the OpenHIE framework, as illustrated in @fig-ohie.
+### High-level solution design
 
-![Proposed extension of the OpenHIE architecture that includes "Data and Analytics Services" as an additional service domain.](./images/openhie-extended-architecture.png){#fig-ohie}
-
-The original OpenHIE specification discerns four domains, namely Point-of-Service systems, the Interoperability Layer, Common Services and Business Services.
+The original OpenHIE specification discerns four domains, namely Point-of-Service systems, the Interoperability Layer, Common Services and Business Services (@fig-ohie).
 We propose to extend the OpenHIE architecture with a "Data and Analytics Services" domain with different zones taken from the data lakehouse architecture, which currently is the most commonly used design pattern in this domain [@armbrust2021lakehouse;@hai2023data;@harby2022data;@harby2024data].
@@
 
+![Proposed extension of the OpenHIE architecture that includes "Data and Analytics Services" as an additional service domain.](./images/openhie-extended-architecture.png){#fig-ohie}
+
 Following the terminology proposed by Hai et al.
-[@hai2023data], @tbl-data-and-analysis-services summarizes how the different zones of the lakehouse architecture can be adapted for healthcare and integrated into the OpeHIE specification.
-Taking Fast Healthcare Interoperability Resources (FHIR) as the open data standard, we envisage the extended OpenHIE architecture to include a 'FHIR lakehouse'[^1].
+[@hai2023data], summarizes how the different zones of the lakehouse architecture can be adapted for healthcare and integrated into the OpeHIE specification.
+Taking Fast Healthcare Interoperability Resources (FHIR) as the open data standard[^1], we envisage the extended OpenHIE architecture to include a 'OpenHIE Lakehouse' with the following healthcare specific adaptations of the various data & analytics services:
+
+- **Ingestion Services:** use of FHIR standard to harmonize all incoming healthcare data to a common data model, including metadata extraction and metadata modeling.
+Should support both single-records streaming ingest as well as bulk data ingestion in batches using the Bulk FHIR API as interface [@mandl2020push].
+- **Storage Service:** should support columnar storage engines optimized for analytical workloads.
+In case of file-based storage, use columnar file formats such as Apache Parquet.
+In case of databases, prefer open source engines such as Clickhouse or PostgreSQL that also support external, file-based tabels [@pedreira2023composable].
+- **Maintenance:** use one of the open table formats such as Apache Iceberg, Apache Hudi and/or Delta Lake [@jain2023analyzing] to realize dataset organization, data integration, schema evolution and data provenance.
+Use of SQL-on-FHIR v2 View definitions to facilitate access to FHIR resources in flattened, tabular format [@sql-on-fhir]
+- **Exploration:** use on-demand, read-only analytical processing engines to provide a unified querying interface to access the heterogeneously structured data using new dataprocessing technologies such as DuckDB, polars etc.
+Should support SQL-on-FHIR Runners, such as Pathling, to generate the standardized views on demand.
+
+Strictly speaking, data consumer services are not part of the lakehouse solution design.
+In practice, these services are implemented using a combination of business intelligence (BI) reporting tools, and interactive development environment (IDE) to perform SQL queries and/or an interactive notebook computing environment [@granger2021jupyter].
+In the discussion we will address future support of federated learning and/or secure multiparty computation network.
 
@@
 
-+-------------------+---------------------------------------------------------------------------------------+
-| Service           | Functional requirements                                                               |
-+===================+=======================================================================================+
-| Ingestion         | - Bulk                                                                                |
-|                   | - Streaming                                                                           |
-+-------------------+---------------------------------------------------------------------------------------+
-| Storage           | - File-based blob storage                                                             |
-|                   | - Database optimized for online analytical processing (OLAP)
-+-------------------+---------------------------------------------------------------------------------------+
-| Maintenance       | - SQL-on-FHIR View defintions                                                         |
-|                   | - Catalog and other maintenance-related functions as defined by Hai et al.            |
-+-------------------+---------------------------------------------------------------------------------------+
-| Processing        | - SQL-on-FHIR Runner                                                                  |
-| & API             | - Execution engine on tabular data as defined in composable data stack                |
-|                   | - Capability to participate as a node in federated learning / MPC network             |
-|                   | - Read-only access to storage                                                         |  
-+-------------------+---------------------------------------------------------------------------------------+
-| Data              | - SQL interactive development environment (IDE)                                       |
-| Consumer          | - Interactive notebook computing environment [@granger2021jupyter]                    |
-|                   | - BI reporting, dashboarding and data visualization                                   |  
-+-------------------+---------------------------------------------------------------------------------------+
-: Adaptation of lakehouse zones for healthcare specific standardization of Data and Analysis Services {#tbl-data-and-analysis-services}
+### Parking lot: following snippets need reviewing (integrate or discard)
 
-[TO DO: need to explain table (is too condensed now) and especially FHIR specific components.]
+#### SQL-on-FHIR v2 as an intermediate representation for FHIR data in tabular format
 
-In the remainder of this paper, we first describe in more detail how the FHIR lakehouse pattern can be implemented using existing open source technologies.
-We do so by presenting work done at PharmAccess.
-Code and digital artifacts are made available [TO DO: include links to support GitHub repositories].
+The premise of separating the user interface from the execution engine is directly related to the key objective of the SQL-on-FHIR project ([https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/](https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/)), namely to make large-scale analysis of FHIR data accessible to a larger audience, portable between systems and to make FHIR data work well with the best available analytic tools, regardless of the technology stack.
+However, to use FHIR effectively analysts require a thorough understanding of the specification as FHIR is represented as a graph of resources, with detailed semantics defined for references between resources, data types, terminology, extensions, and many other aspects of the specification.
+Most analytic and machine learning use cases require the preparation of FHIR data using transformations and tabular projections from its original form.
+The task of authoring these transformations and projections is not trivial and there is currently no standard mechanisms to support reuse.
 
-Subquently, we compare this solution design with two widely used OpenHIE-compliant frameworks, namely the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)) and the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/).
-We take a narrative approach in presenting our design and the case studies, surveying existing scientific studies on OHDSs, focusing on the seminal reports and subsequently searching forward citations.
-In addition, we have searched the open source repositories (most notably GitHub) and the online communities (OpenHIE community, FHIR community) to search for relevant open standards, technologies and architectures.
-This paper should not be considered as a proper systematic review.
+The solution of the SQL-on-FHIR project is to provide a specification for defining tabular, use case-specific views of FHIR data.
+The view definition and the execution of the view are separated, in such a way that the definition is portable across systems while the execution engine (called runners) are system-specific tools or libraries that apply view definitions to the underlying data layer, optionally making use of annotations to optimize performance.
 
-Finally, we discuss our findings and propose routes for future development.
+##### Ingestion
 
-The main contributions of this paper are i) description of a framework for the components of the Data & Analysis Services that builds on current best practices from the data engineering community into the OpenHIE framework; ii) evaluation of different implementations and design options for type 1 and 2 data sharing within an extended OpenHIE architecture; and iii) provide open content for the MomCare implementation to facilitate adaptation and deployment.
-As such, it aims to inform future developments and implementation of open digital health platforms in LMICs.
+- Default workflow is extraction of data from SHR using Bulk FHIR API.
+Data contains metadata (incl.
+FHIR versions) and fully qualified semantics, for example, coding systems.
+Despite this, metadata extraction and metadata modeling is still required to meet the FAIR requirements.
+Issues that need to be solved by these services:
+- To prepare for future updates of FHIR versions
+- Implement late-binding principle of having increasingly more specific FHIR profiles as bulk FHIR data propagates through lakehouse
+- FHIR vs. FAIR
+  - How does FHIR relate to approaches taken by the FAIR community, which tend to take more an approach of using knowledge graphs.
+For example, VODAN Africa [@gebreslassie2023fhir4fair;@purnamajati2022data].
+  - FAIR principles vs FHIR graph: is FHIR a FAIR Data Object
+- Since we use FHIR, we don't need a semantic layer because that is already provided
+- We do need different semantic layer, namely with metrics. Explain different types of semantics.
+  - The metrics layer same function as CQL. Discuss CQL vs generic metrics layer.
+
+##### Storage
+
+- File-based:
+  - from ndjson to parquet
+  - possibly used delta lake for time versioning
+  - separation of storage from compute not only for benefits of lower TCO, but also be ready for federated learning and MPC in future
+
+##### Maintenance
+
+- SQL-on-FHIR Views provide new standard to support mADX aggregate reporting !! We need to stress this, because this is an existing OpenHIE workflow
+- Maintenance-related functions remain the same
+- NB: orchestration falls under data provenance
+- NB: make comparison with HMIS component
+  - workflow requirements: Report aggregate data (link): receiver is HMIS, mADX; this is not analytics!
+  - Functional requirements: https://guides.ohie.org/arch-spec/openhie-component-specifications-1/openhie-health-management-information-system-hmis
+  - Requirements are similar, but implementation differs: Datamodel is non-FHIR, focused on DataValue, which conceptually equates to FHIR Measure
 
@@
 
-### Open technologies: deploying Instant OpenHIE with digital public goods
+## Discussion
 
-Today, many components of the OpenHIE specification are now available as a digital public goods, as listed in @tbl-digital-public-goods.
+Given the solution design, and evaluation of the three case studies, we discuss the pros and cons from different perspectives that we believe are essential design principles to realise solidarity-based OHDSs, namely [TO DO: this is just first shot at formulating our design principles; needs refinement]:
+
+- inclusive-by-design, based on the notions of datasolidarity and maximising autonomy of all future participants in the ecosystem;
+- scaleable-by-design, particularly focusing on downward scaleability to support a decentralized platform topology to allows for bottom-up deployment scenarios (from local care networks --> county-level networks --> national networks) instead of top-down national roll-out;
+- open-by-design, whereby a balance is found to resort to minimal standards and allow for a large diversity of partners and technologies to be used;
+
+### Inclusive-by-design: datasolidarity, FAIRness and autonomy
+
+- which can be framed within the context of ongoing efforts towards Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair].
+- equitable data sharing requires more than just FAIRification [@evertz2023what]
+
+### Scaleable-by-design
+
+Today, many components of the OpenHIE specification are now available as a digital public goods.
 Typically, these open source components are intended to support deployments in small countries (population up to 10 million) or large NGOs out of the box, and should provide a stepping stone for customized deployments in medium-sized countries (population around 40 million).[^2] To further ease the development, configuration and deployment of health information exchanges, the concept of 'Instant OpenHIE' has been championed to (i) allow implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) have a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@InstantOpenHIEv2].
@@
 With this approach, we purposefully challenge the dominant paradigm of designing and implementing centralized platforms to support the digital transformation of healthcare in LMICs [@ogundaini2022systematic] with the aim to make digital platforms work *for* development [@hermes2020digital].
+
+[TO DO: elaborate on how we see this solution design can be implemented from the bottom-up, typically in a primary care network serving a population of around 80,000 people with level 3 facilities that have limited resources]
 
@@
 
-#### SQL-on-FHIR v2 as an intermediate representation for FHIR data in tabular format
-
-The premise of separating the user interface from the execution engine is directly related to the key objective of the SQL-on-FHIR project ([https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/](https://build.fhir.org/ig/FHIR/sql-on-fhir-v2/)), namely to make large-scale analysis of FHIR data accessible to a larger audience, portable between systems and to make FHIR data work well with the best available analytic tools, regardless of the technology stack.
-However, to use FHIR effectively analysts require a thorough understanding of the specification as FHIR is represented as a graph of resources, with detailed semantics defined for references between resources, data types, terminology, extensions, and many other aspects of the specification.
-Most analytic and machine learning use cases require the preparation of FHIR data using transformations and tabular projections from its original form.
-The task of authoring these transformations and projections is not trivial and there is currently no standard mechanisms to support reuse.
-
-The solution of the SQL-on-FHIR project is to provide a specification for defining tabular, use case-specific views of FHIR data.
-The view definition and the execution of the view are separated, in such a way that the definition is portable across systems while the execution engine (called runners) are system-specific tools or libraries that apply view definitions to the underlying data layer, optionally making use of annotations to optimize performance.
-
-##### Ingestion
-
-Default workflow is extraction of data from SHR using Bulk FHIR API.
-Data contains metadata (incl.
-FHIR versions) and fully qualified semantics, for example, coding systems.
-Despite this, metadata extraction and metadata modeling is still required to meet the FAIR requirements.
-Issues that need to be solved by these services:
-
-- To prepare for future updates of FHIR versions
-- Implement late-binding principle of having increasingly more specific FHIR profiles as bulk FHIR data propagates through lakehouse
-
-##### Storage
-
-- File-based:
-  - from ndjson to parquet
-  - possibly used delta lake for time versioning
-  - separation of storage from compute not only for benefits of lower TCO, but also be ready for federated learning and MPC in future
-
-- OLAP DBMS
-  - Often columnar, like Clickhouse and BigQuery
-  -
-
-##### Query & Processing
-
-- fit in structure of OpenHIE specification
-- check which workflows are related to analytics
-- Hai calls this 'Maintenance'
-
-##### Maintenance
-
-- SQL-on-FHIR Views provide new standard to support mADX aggregate reporting !! We need to stress this, because this is an existing OpenHIE workflow
-- Maintenance-related functions remain the same
-- NB: orchestration falls under data provenance
-
-##### Data consumers
-
-- Many tools, often focus on creating information dashboards and visualizations
-- Compatibility with processing & API: which query languages and interfaces are supported. Some dialect of SQL, some dialect of NoSQL, dataframe API, all of the above?
-
-## Discussion
-
-### Openness of health data sharing platforms
+### Open-by-design: mitigating risk and rebalancing asymmetries
 
@@
 
-### Conceptualizing and implementing open health data sharing platforms
-
-In the light of these nascent paradigm shifts, the objective of this paper is the following.
-Given high-level design principles that we believe are supportive of realising solidarity-based, namely [TO DO: this is just first shot at formulating our design principles; needs refinement]:
-
-- inclusive-by-design, based on the notions of datasolidarity and maximising autonomy of all future participants in the ecosystem;
-- scaleable-by-design, particularly focusing on downward scaleability to support a decentralized platform topology to allows for bottom-up deployment scenarios (from local care networks --> county-level networks --> national networks) instead of top-down national roll-out;
-- open-by-design, whereby a balance is found to resort to minimal standards and allow for a large diversity of partners and technologies to be used;
-
-We specifically address the notion of openness of OHDPs in LMICs in terms of the design-related questions put forward by de Reuver at al.11:
-
-- Object of openness: what data-related resources should data platforms make available when opening up (e.g.
-data, data products, datadriven insights, analytics modules)?
-Which user groups derive value from accessing data-related resources from data platforms (e.g.
-data providers, data users, intermediaries, developers)?
-- Unit of analysis: what is platform-to-platform openness in the context of data platforms, given the expectation that different OHDPs will emerge at various aggregation levels?
-How do we distinguish meta-platforms, forking, and platform interoperability?
 - Risk of openness: What are the novel (negative) implications of opening up data platforms? How can reflexivity in design help providers to resolve the negative implications of openness?
@@
 
-### Comparison with HMIS component
+### Limitation and future work
 
-- Workflow requirements: Report aggregate data (link): receiver is HMIS, mADX; this is not analytics!
-- Functional requirements: https://guides.ohie.org/arch-spec/openhie-component-specifications-1/openhie-health-management-information-system-hmis
-
-Requirements are similar, but implementation differs: Datamodel is non-FHIR, focused on DataValue, which conceptually equates to FHIR Measure
-
-### The need to a semantic layer?
-
-- FHIR and FAIR
-  - How does FHIR relate to approaches taken by the FAIR community, which tend to take more an approach of using knowledge graphs.
-For example, VODAN Africa [@gebreslassie2023fhir4fair;@purnamajati2022data].
-  - FAIR principles vs FHIR graph: is FHIR a FAIR Data Object
-
-- Since we use FHIR, we don't need a semantic layer because that is already provided
-- We do need different semantic layer, namely with metrics. Explain different types of semantics.
-  - The metrics layer same function as CQL. Discuss CQL vs generic metrics layer.
-
-### Attribute-based access control
-
+- Access control is still a pain-point, can we move to Attribute-based access control?
   - TO DO: if you have generated flattened SQL tables, how are you going to manage security?
@@
 
-### Federated learning and multiparty computation
+- Federated learning and multiparty computation
+  - Lakehouse serves as datastations
+  - Explain first results Roseman Labs
 
-- data stations??!
+- ... [more future work items here]
+
+## Conclusion
 
@@
 | HIE | Health Information Exchange |
-| ICT4D | ICT for Development |
-| IR | Intermediate Representation |
 | LMIC | Low- and middle income countries |

### 2024-08-27 index.qmd
Add table 1 with types of data sharing

@@
 
-Digital transformation should be considered as a key determinant of health, but also requires a radical rethink how digital technologies should be used [@kickbusch2021lancet].
-Digital technologies have the potential to increase the availability, accessibility, acceptability, and quality of health services as we know them; change nature and direction of health care, and the possibility of making it more preventive, personalised, and mobile; and enfranchise patients and communities (particularly vulnerable groups), thus helping shape the health system according to the needs of the patients and communities.
+It is a widely held belief in the global health community that digital technologies have an important role to play in strengthening healthcare in low- and middle income countries (LMICs).
+Digital technologies have the potential to increase the availability, accessibility, acceptability, and quality of health services; make healthcare more preventive, personalised, and mobile; and enfranchise patients and communities, particularly those who are most vulnerable [@kickbusch2021lancet].
 
@@
 
-Learnings from digital health interventions have shown that improvements along the five aforementioned dimensions is within reach.
+Learnings from digital health interventions have shown that improvements along the five aforementioned dimensions are within reach.
 As an example, in our experience with the MomCare programme, we have demonstrated that routinely collected health data, combined with financial data, can be effectively used to gain insight into the continuity of care, improve clinical adherence whilst maintaining efficiency and affordability of health services in LMICs.
@@
 
-MomCare, like most other digital health interventions, is critically dependent on the availability of data and analytics functions within its underlying supportive OHDS.
-For example, the widely used DHIS2 reports that are most commonly used in LMICs contained insufficient information and analytical functions to support such interventions.
-To implement the value-based healthcare business logic, more detailed analytics of patient journeys is required.
+MomCare is critically dependent on the availability of data and analytics functions within its underlying supportive OHDS.
+For example, to implement the value-based healthcare business logic of MomCare, detailed analysis of patient journeys is required.
+This functionality is currently not available as a standard; standardized reports that are available, such as DHIS2, contained insufficient information and analytical functions to support intervention that aim to improve continuity of care and adherence to clinical guidelines.
+As such, a large part of the day-to-day operation of MomCare revolved performaing data and analytics: data acquisition, data integration, analysis etc.
 
-It is in fact a widely held belief in the global digital health community that data & analytics services are an essential capibility of OHDSs.
+This situation is not unique to MomCare.
+In fact, many in the global digital health community think that data & analytics services should be an essential capibility of OHDSs going forward.
 This is exemplified how the OpenHIE reference architecture [@openhie] has been adopted by many sub-Saharan African countries as the blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one].
@@
 In terms of the often-used distinction between primary and secondary health data use [@cascini2024health], these countries aim to extend OpenHIE beyond it original scope of primary data sharing to also include secondary use of health data for academic research, real-world evidence studies etc.
-If we are to use the OpenHIE framework for data & analytics as well, we need to extend the standards, technologies and architecture to include functionality to do so.
+If we are to use the OpenHIE framework for secondary use, supported by data & analytics as well, we need to extend the standards, technologies and architecture to include functionality to do so.
 The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish more comprehensive platforms to support primary and secondary health data sharing in LMICs.
+
+### Objects of openness in secondary data sharing
+
+To set the scene of for the main contributions of this viewpoint paper, consider four types of secondary data sharing as shown in @tbl-types-data-sharing, where we follow the research agenda proposed by de Reuver et al.
+to scrutinize how openness of data platforms can be achieved [@dereuver2022openness].
+
+|   | __Type of data sharing__ | __Relevance to extension of OpenHIE specification__ |
+|:-:|:-------|:-------------|
+| __1__ | Data at the most granular level with which the patient journey (timeline) can be reconstructed and used for various analytic tasks.
+| The Shared Health Record (SHR) is specified as an operational, real-time transactional data source, distinct from a data warehouse.
+A seperate specification of data and analytics functions, typically provided by a datawarehouse, is required.
+|
+| __2__ | Aggregated data, typically used for routine reports and benchmarking.
+| The Health Management Information System (HMIS) should support Aggregate Data Exchange (ADX) workflow standard.
+More flexible and extensive workflows are emerging based on FHIR.
+| __SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows.
+Builds on FHIRPath [@hl72020fhirpath] expressions in a logical structure to specify things like column names and unnested items.
+Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].|
+| __3__ | Data analytics modules, that provide secure and privacy-preserving computational environments to work with the data.
+| Federated learning (FL) [@rieke2020future] and privacy- enhancing technologies (PETs) [@scheibner2021revolutionizing;@jordan2022selecting] provide new paradigms that address the problem of data governance and privacy by training algorithms collaboratively without exchanging the data itself.
+Requires use of a common data model, such as FHIR, to analyze the data in a collaborative, decentralized fashion.
+|
+| __4__ | Trained models that have been derived from the data and can be used stand-alone for decision support.
+| Increasing need to open source trained AI models [@osi2024open], enabled by technologies such as ONNX [@onnx].
+|
+
+: Types of secondary data sharing, and there relevance to the proposed extension of the OpenHIE specification. {#tbl-types-data-sharing .striped .hover .responsive}
+
+[TO DO: add more text here to clarify our position and intention with this paper.]
+
+### Outline
 
@@
 - **Ingestion Services:** use of FHIR standard to harmonize all incoming healthcare data to a common data model, including metadata extraction and metadata modeling.
-Should support both single-records streaming ingest as well as bulk data ingestion in batches using the Bulk FHIR API as interface [@mandl2020push].
+Should support both single-records streaming ingest as well as bulk data ingestion in batches using the Bulk FHIR API as interface [@mandl2020push;@jones2021landscape].
 - **Storage Service:** should support columnar storage engines optimized for analytical workloads.

### 2024-08-29 index.qmd
Editorial suggestions

@@
 
-To achieve scaleability of digital health interventions, we need to design and implement standardized open health data systems (OHDS) and their associated ecosystems that can support improvements in the wider health sector along five dimensions, namely [@kelley2020digital]:
+To achieve scalability of digital health interventions, we need to design and implement standardized open health data systems (OHDS) and their associated ecosystems that can support improvements in the wider health sector along five dimensions, namely [@kelley2020digital]:
 
@@
 As an example, in our experience with the MomCare programme, we have demonstrated that routinely collected health data, combined with financial data, can be effectively used to gain insight into the continuity of care, improve clinical adherence whilst maintaining efficiency and affordability of health services in LMICs.
-By actively coaxing pregnant mothers to undergo 4 antenatal check-ups, outcomes were improved whilst maintaining the average cost of maternal, newborn and child health (MNCH) services [@huisman2022digital;@sanctis2022maintaining;@izudi2023experiences].
+By actively encouraging pregnant mothers to undergo 4 antenatal check-ups, outcomes were improved whilst maintaining the average cost of maternal, newborn and child health (MNCH) services [@huisman2022digital;@sanctis2022maintaining;@izudi2023experiences].
 
@@
 For example, to implement the value-based healthcare business logic of MomCare, detailed analysis of patient journeys is required.
-This functionality is currently not available as a standard; standardized reports that are available, such as DHIS2, contained insufficient information and analytical functions to support intervention that aim to improve continuity of care and adherence to clinical guidelines.
-As such, a large part of the day-to-day operation of MomCare revolved performaing data and analytics: data acquisition, data integration, analysis etc.
+This functionality is currently not available in a standard form; standardized reports that are available, such as DHIS2, contained insufficient information and analytical functions to support intervention that aim to improve continuity of care and adherence to clinical guidelines.
+As such, a large part of the day-to-day operation of MomCare involved performing data and analytics tasks: data acquisition, data integration, analysis etc.
 
 This situation is not unique to MomCare.
-In fact, many in the global digital health community think that data & analytics services should be an essential capibility of OHDSs going forward.
-This is exemplified how the OpenHIE reference architecture [@openhie] has been adopted by many sub-Saharan African countries as the blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one].
+In fact, many in the global digital health community think that data & analytics services should be an essential capability of OHDSs going forward.
+This is exemplified by the adoption of the OpenHIE reference architecture [@openhie] by many sub-Saharan African countries as the blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one].
 These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain [@mbugua2021adoption;@dalhatu2023paper].
-In terms of the often-used distinction between primary and secondary health data use [@cascini2024health], these countries aim to extend OpenHIE beyond it original scope of primary data sharing to also include secondary use of health data for academic research, real-world evidence studies etc.
-If we are to use the OpenHIE framework for secondary use, supported by data & analytics as well, we need to extend the standards, technologies and architecture to include functionality to do so.
-The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish more comprehensive platforms to support primary and secondary health data sharing in LMICs.
+In terms of the often-used distinction between primary and secondary health data use [@cascini2024health], these countries aim to extend OpenHIE beyond its original scope of primary data sharing to also include secondary use of health data for academic research, real-world evidence studies, etc.
+If we are to use the OpenHIE framework for secondary use, supported by data and analytics as well, we need to extend the standards, technologies and architecture to include functionality to do so.
+The lack of detailed specifications and consensus for this addition to OpenHIE currently stands in the way of development projects that aim to establish more comprehensive platforms to support primary and secondary health data sharing in LMICs.
 
@@
 In the following, we first describe how the lakehouse design pattern, being the most widely used data & analytics solution architecture, can be integrated in OpenHIE.
-To demonstrate the feasibility of this design, we present a proof-of-concept impelementation using open source technologies within the context of the MomCare programme.
+To demonstrate the feasibility of this design, we present a proof-of-concept implementation using open source technologies within the context of the MomCare programme.
 Code and digital artifacts of this demonstrator is available as supplementary material [TO DO: include links to support GitHub repositories].
-Subquently, we compare this solution design with two widely used and operational OpenHIE-compliant open source frameworks, namely the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)) and the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/).
+Subsequently, we compare this solution design with two widely used and operational OpenHIE-compliant open source frameworks, namely the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/)) and the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/).
 Finally, we discuss our findings and propose routes for future development.
@@
 Most analytic and machine learning use cases require the preparation of FHIR data using transformations and tabular projections from its original form.
-The task of authoring these transformations and projections is not trivial and there is currently no standard mechanisms to support reuse.
+The task of authoring these transformations and projections is not trivial and there are currently no standard mechanisms to support reuse.
 
@@
 
-Clearly, a more open and scaleable platform was required if MomCare was to be implemented in more regions.
+Clearly, a more open and scalable platform was required if MomCare was to be implemented in more regions.
 This need led to a redesign of the underlying technical infrastructure of the MomCare project.
@@
 Analysis of bulk data was done by directly reading the standard newline delimited JSON into the Python pandas data analysis library.
-Cross checking the output with queries on the original data confirmed that the whole data pipeline produced consistent results.
+Cross-checking the output with queries on the original data confirmed that the whole data pipeline produced consistent results.
 For example, the report of the antenatal coverage metric (number of pregnancies with four or more visits) could be reproduced per patient journey and aggregated (per year, per organization etc.) as required for the MomCare reports.
@@
 As part of the ingestion, incoming FHIR bundles that contain multiple FHIR resources are unbundled in separate topics using a generic Kafka utility component.
-Subsequently, each FHIR resource topic is flatted with Kafka mappers that use FHIRPath.
+Subsequently, each FHIR resource topic is flattened with Kafka mappers that use FHIRPath.
 Superset is used as the tool for consuming the data to create dashboard visualizations.
@@
 - When using the ELK stack, flattening is done at the end. Implementations of FHIRPath support Elasticsearch as an execution engine, also here 
-- Main limitations: both Clickhouse en Elasticsearch don't follow decomposition of storage, compute and UI. Therefore, downward scaleability is limited.
+- Main limitations: both Clickhouse en Elasticsearch don't follow decomposition of storage, compute and UI. Therefore, downward scalability is limited.
 
